---
title: "Join samples with survey data"
author: "Cob Staines"
output: html_document
---
# Description

This script aligns sequence filed with sample metadata in the RIBBiTR database, for easy lookups and referencing of eDNA collection downstream. To establish a connection with the RIBBiTR database (for the first time) check out this [Connection Tutorial](https://ribbitr-bii.github.io/ribbitr-data-access/tutorial_series/01_connection_setup.html)

This script also assigns each sample a `study_system` (see `study_system_lookup` below), derived from its RIBBiTR `region`. This runs early (script 03) specifically so that downstream scripts -- notably `05_query_taxonomy_geography.Rmd`, which needs to know which study system(s) are present in a run before it can query GBIF per system -- can rely on it.

# Config
```{r}
librarian::shelf(tidyverse, here, yaml, dbplyr, RPostgres, DBI, ribbitrrr, janitor, uuid)

# # Point to RIBBiTR eDNA Processing 22-25 lookup table for extraction_ids & study systems
# sample_lookup_file = here(Sys.getenv("data_dir"), "edna_results", "RIBBiTR eDNA Processing 22-25_2026-08-23.xlsx")

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R); this is read
# directly so the same values apply whether this script is run manually or via the driver
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

# import run config file
run_dir = here(pipeline_config$run_dir)

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# sample map out file
map_out = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))

# establish database connection
dbcon = hopToDB("ribbitr")

```

# Load data
```{r}

# db table pointers
db_edna_sequencing = tbl(dbcon, Id("survey_data", "edna_sequencing"))
db_edna_extraction = tbl(dbcon, Id("survey_data", "edna_extraction"))
db_edna_sample = tbl(dbcon, Id("survey_data", "edna_sample"))
db_env = tbl(dbcon, Id("survey_data", "environmental"))
db_survey = tbl(dbcon, Id("survey_data", "survey"))
db_visit = tbl(dbcon, Id("survey_data", "visit"))
db_site = tbl(dbcon, Id("survey_data", "site"))
db_region = tbl(dbcon, Id("survey_data", "region"))
db_country = tbl(dbcon, Id("survey_data", "country"))

all_edna_data = db_edna_sequencing %>%
  left_join(db_edna_extraction, by =  "edna_extraction_id") %>%
  left_join(db_edna_sample, by = "edna_sample_id") %>%
  left_join(db_survey, by = "survey_id") %>%
  left_join(db_visit, by = "visit_id") %>%
  left_join(db_site, by = "site_id") %>%
  left_join(db_region, by = "region_id") %>%
  left_join(db_country, by = "country_id") %>%
  select(all_of(colnames(db_edna_sequencing)),
         all_of(colnames(db_edna_extraction)),
         all_of(colnames(db_edna_sample)),
         visit_id,
         time_of_day,
         date,
         transect,
         site,
         site_latitude,
         site_longitude,
         region,
         country) %>%
  filter(read_direction == "forward") %>%
  collect() %>%
  mutate(sequencing_base = basename(sequencing_s3_filepath)) %>%
  group_by(edna_sequencing_name, read_direction, sequencing_date) %>%
  slice_head(n = 1) %>%
  ungroup()

# import feature table
feature_table =
  read_tsv(here(run_dir, "/analysis/s06_denoised_16S_eDNA/sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

norm = function(x) tolower(gsub("[-_]", "", x))

seq_bases = sort(all_edna_data$sequencing_base)
seq_features = sort(colnames(feature_table %>% select(-asv_id)))
seq_bases_norm = norm(seq_bases)

sample_map = tibble(feature_id = seq_features) %>%
  mutate(
    edna_sequencing_alignment = purrr::map_chr(feature_id, function(b) {
      bn = norm(b)
      hits = seq_bases[startsWith(seq_bases_norm, bn)]
      if (length(hits) == 0) return(NA_character_)
      hits[which.max(nchar(hits))]
    })
  ) %>%
  left_join(all_edna_data, by = c("edna_sequencing_alignment" = "sequencing_base"))


```

# environmental data
```{r}
# survey data
env_obs = db_env %>%
  left_join(db_survey, by = "survey_id") %>%
  left_join(db_visit, by = "visit_id") %>%
  filter(visit_id %in% sample_map$visit_id) %>%
  select(visit_id,
         air_time,
         air_temp_c,
         wind_speed_m_s,
         water_time,
         water_temp_c,
         p_h,
         tds_ppm,
         conductivity_us_cm) %>%
  collect() %>%
  arrange(visit_id,
          water_time,
          air_time) %>%
  group_by(visit_id) %>%
  slice_head(n = 1) %>%
  ungroup()

sample_map_env = sample_map %>%
  left_join(env_obs, by = "visit_id") %>%
  mutate(study_system = case_when(
    country == "panama" ~ "panama",
    country == "brazil" ~ "brazil",
    region == "pennsylvania" ~ "pennsylvania",
    region == "california" ~ "sierra_nevada",
    pcr_control | extraction_negative_control ~ "lab_control",
    edna_sequencing_name == "Tissue_63" ~ "pennsylvania",
    edna_sequencing_name %in% c("RP_330", "RP_343", "RP_345", "RP_346") ~ "brazil",
    .default = NA
  ))

```

# Study system

Assign each sample a canonical `study_system` label, derived from its RIBBiTR `region`. This is what downstream scripts (starting with `05_query_taxonomy_geography.Rmd`) use to know which study system(s) are present in a run -- so a run can freely mix samples from multiple systems.

`study_system_lookup` below maps a normalized `region` value (lowercased, non-alphanumeric characters collapsed to `_`) to a canonical study_system label. If any regions in your run don't map (see the warning below), add them here.
```{r}
unmapped_regions = sample_map_env %>%
  filter(is.na(study_system)) %>%
  distinct(region)

if (nrow(unmapped_regions) > 0) {
  cat("WARNING: the following region values did not map to a known study_system.\n",
      "Add them to study_system_lookup above:\n")
  print(unmapped_regions)
}

cat("\nStudy systems present in this run:\n")
print(sample_map_env %>% filter(!is.na(study_system)) %>% distinct(study_system))
```

# export
```{r}
write.csv(sample_map_env, map_out, row.names = FALSE)
```