---
title: "Query taxonomy geography"
author: "Cob Staines"
output: html_document
---

# Description

This script queries the geographic range of taxonomies corresponding to vsearch hits, to support a more accurate, locally informed classification.

This script requires API keys for GBIF in .Renviron (see this [Tutorial](https://docs.ropensci.org/rgbif/articles/gbif_credentials.html)).

If you want to skip this step you can, just set the parameter `gbif_query` to `FALSE` in the subsequent script (`06_classify_asv.Rmd`).

## Multi-system runs & resumability

A run may contain samples from several study systems (see `study_system` in `03_sample_map.Rmd`, which must be run first). GBIF occurrence counts are always queried separately per system -- combining systems into one query would defeat the point of a locally-informed check, and GBIF's own rate limits make a combined query impractical anyway.

The query is split into two stages that behave differently under repeated runs:

- **Taxon resolution ("backbone")** -- resolving each hit taxon to a GBIF `usageKey` -- is the same regardless of study system, so it only runs once for the whole run, not once per system. Each sub-query (by scientific name, by epithet, by genus, by family) is checkpointed to its own file; if you re-run this script, any sub-query whose checkpoint file already exists is skipped rather than re-queried. This is what used to require manually re-running individual code chunks by hand when GBIF's servers hiccupped -- now you can just re-run the whole script and only the parts that didn't finish will re-query.
- **Occurrence counting** is per system, and is checkpointed the same way, one file per system (`..._05_gbif_occ_data_<system>.csv`). Re-running the script picks up any systems that haven't finished yet and skips ones that have. You can run this script as many times as you need across a session (or multiple sessions) until every system has a checkpoint file.

Both stages also retry transient GBIF failures automatically (see `with_retry` below) before giving up.

The final "Process GBIF data" step builds a composite table from whichever per-system occurrence files exist *at the time it's run* -- so it's safe to run before every system has finished; it'll just reflect however many systems are done so far, and you can re-run it again later once the rest complete.

# Config
```{r}
## Load in Libraries (install as needed)
librarian::shelf(tidyverse, yaml, here, rgbif)

# requires GBIF token in .Renviron

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R); this is read
# directly so the same values apply whether this script is run manually or via the driver
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

# import run config file
run_dir = here(pipeline_config$run_dir)

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# study systems present in this run, derived from the sample map (03_sample_map.Rmd, which
# must be run first) rather than configured by hand -- a run may mix multiple systems
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))

if (!file.exists(sample_map_file)) {
  stop("Sample map not found at: ", sample_map_file, "\nRun 03_sample_map.Rmd first.")
}

study_systems = read_csv(sample_map_file, show_col_types = FALSE) %>%
  filter(!is.na(study_system)) %>%
  distinct(study_system) %>%
  arrange(study_system) %>%
  pull(study_system)

cat("Study systems found in run:", paste(study_systems, collapse = ", "), "\n")

# checkpoint file paths (backbone sub-queries, run once for the whole run)
backbone_file = function(label) here(run_dir, "output", paste0(config$run$name, "_05_gbif_backbone_", label, ".csv"))

# checkpoint file path (occurrence counts, one per study system)
occ_file = function(system) here(run_dir, "output", paste0(config$run$name, "_05_gbif_occ_data_", system, ".csv"))

# composite output file (built from whichever per-system occurrence files exist so far)
taxa_out_file = here(run_dir, "output", paste0(config$run$name, "_05_taxonomy_table_gbif_occ.csv"))

# retry a GBIF call up to `attempts` times, waiting `wait_sec` between failures -- GBIF's
# servers intermittently error/timeout regardless of query validity, so this replaces manually
# re-running a chunk until it happens to succeed
with_retry = function(expr_fn, label, attempts = 5, wait_sec = 30) {
  for (attempt in seq_len(attempts)) {
    result = tryCatch(list(ok = TRUE, value = expr_fn()),
                       error = function(e) list(ok = FALSE, error = e))
    if (result$ok) return(result$value)
    cat(sprintf("  [%s] attempt %d/%d failed: %s\n", label, attempt, attempts, conditionMessage(result$error)))
    if (attempt < attempts) Sys.sleep(wait_sec)
  }
  stop(paste0(label, ": all ", attempts, " attempts failed"))
}
```

# Import taxa
```{r}
# load taxonomy table
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_02_taxonomy_table_filtered.csv")))) {
  taxa_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_02_taxonomy_table_filtered.csv")))
} else {
  cat("No taxonomy table found, skipping.\n")
  taxa_raw = tibble(c())
}

# load web blast table
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv")))) {
  taxa_web_blast_metazoa = read_csv(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv"))) %>%
    mutate(species = sub("^\\S+\\s+", "", species)) %>%
    select("accession",
           "domain",
           "kingdom",
           "phylum",
           "class",
           "order",
           "family",
           "genus",
           "species") %>%
    filter(kingdom == "Metazoa")
} else {
  cat("No web blast results found, skipping.\n")
  taxa_web_blast_metazoa = tibble(c())
}

# clean up taxa
taxa_clean = bind_rows(taxa_raw,
                       taxa_web_blast_metazoa)%>%
  mutate(
    scientificName = case_when(
      !is.na(genus) & !is.na(species) ~ paste0(genus, " ", species),
      !is.na(genus) & is.na(species) ~ genus,
      .default = NA),
    species_simple = species,
    species_simple = if_else(grepl("^sp\\.", species_simple), NA, species_simple),
    species_simple = gsub(" x .*", "", species_simple),
    species_simple = gsub("^cf\\. *", "", species_simple),
    species_simple = gsub("^aff\\. *", "", species_simple),
    specificEpithet = gsub(" .*", "", species_simple),
    intraspecificEpithet = gsub("^[^ ]+ ?", "", species_simple),
    intraspecificEpithet = if_else(intraspecificEpithet == "", NA, intraspecificEpithet))

# copy for output
taxa_out = taxa_clean

# get unique for querying
taxa_unique = taxa_clean %>%
  select(-accession) %>%
  distinct() %>%
  arrange(domain, kingdom, phylum, class, order, family, genus, species)
```

# Resolve taxon backbone (run once, checkpointed per sub-query)
```{r}
cat("\nResolving taxa to GBIF backbone (system-independent, runs once)...\n")

## species, by full scientific name
if (file.exists(backbone_file("species_sn"))) {
  cat("\tSpecies IDs (by scientificName) already resolved, reading checkpoint\n")
  gbif_backbone_key_species_sn = read_csv(backbone_file("species_sn"), show_col_types = FALSE)
} else {
  cat("\tLooking up GBIF species IDs (by scientificName)\n")

  gbif_query_species_sn = taxa_unique %>%
    filter(!is.na(species)) %>%
    select(class,
           order,
           family,
           scientificName) %>%
    distinct()

  gbif_backbone_species_sn = with_retry(
    function() name_backbone_checklist(gbif_query_species_sn, bucket_size = 3500, sleep = 10),
    label = "backbone species (scientificName)"
  )

  rownames(gbif_query_species_sn) = NULL
  rownames(gbif_backbone_species_sn) = NULL
  gbif_backbone_key_species_sn = bind_cols(gbif_query_species_sn %>%
                                             rename(class_q = class,
                                                    order_q = order,
                                                    family_q = family,
                                                    scientific_name_q = scientificName),
                                           gbif_backbone_species_sn) %>%
    mutate(local_occ_count = NA) %>%
    filter(rank %in% c("SPECIES", "SUBSPECIES"))

  write_csv(gbif_backbone_key_species_sn, backbone_file("species_sn"))
}

## species, by specific epithet, for those which returned NA above
if (file.exists(backbone_file("species_e"))) {
  cat("\tSpecies IDs (by specificEpithet) already resolved, reading checkpoint\n")
  gbif_backbone_key_species_e = read_csv(backbone_file("species_e"), show_col_types = FALSE)
} else {
  cat("\tLooking up GBIF species IDs (by specificEpithet)\n")

  gbif_query_species_e = taxa_unique %>%
    filter(!is.na(specificEpithet)) %>%
    select(class,
           order,
           family,
           genus,
           specificEpithet,
           scientificName) %>%
    anti_join(gbif_backbone_key_species_sn, by = c("class" = "class_q",
                                                   "order" = "order_q",
                                                   "family" = "family_q",
                                                   "scientificName" = "scientific_name_q")) %>%
    select(-scientificName) %>%
    distinct()

  gbif_backbone_species_e = with_retry(
    function() name_backbone_checklist(gbif_query_species_e, bucket_size = 100, sleep = 10),
    label = "backbone species (specificEpithet)"
  )

  rownames(gbif_query_species_e) = NULL
  rownames(gbif_backbone_species_e) = NULL
  gbif_backbone_key_species_e = bind_cols(gbif_query_species_e %>%
                                            rename(class_q = class,
                                                   order_q = order,
                                                   family_q = family,
                                                   genus_q = genus,
                                                   specific_epithet_q = specificEpithet),
                                          gbif_backbone_species_e) %>%
    mutate(local_occ_count = NA) %>%
    filter(rank %in% c("SPECIES", "SUBSPECIES"))

  write_csv(gbif_backbone_key_species_e, backbone_file("species_e"))
}

# combine specific results
gbif_backbone_key_species = bind_rows(gbif_backbone_key_species_sn,
                                      gbif_backbone_key_species_e)

## genus
if (file.exists(backbone_file("genus"))) {
  cat("\tGenus IDs already resolved, reading checkpoint\n")
  gbif_backbone_key_genus = read_csv(backbone_file("genus"), show_col_types = FALSE)
} else {
  cat("\tLooking up genus IDs\n")

  gbif_query_genus = taxa_unique %>%
    select(class,
           order,
           family,
           genus) %>%
    filter(!is.na(genus)) %>%
    distinct()

  gbif_backbone_genus = with_retry(
    function() name_backbone_checklist(gbif_query_genus, bucket_size = 1500, sleep = 10),
    label = "backbone genus"
  )

  rownames(gbif_query_genus) = NULL
  rownames(gbif_backbone_genus) = NULL

  gbif_backbone_key_genus = bind_cols(gbif_query_genus %>%
                                        rename(class_q = class,
                                               order_q = order,
                                               family_q = family,
                                               genus_q = genus),
                                      gbif_backbone_genus) %>%
    mutate(local_occ_count = NA) %>%
    filter(rank == "GENUS")

  write_csv(gbif_backbone_key_genus, backbone_file("genus"))
}

## family
if (file.exists(backbone_file("family"))) {
  cat("\tFamily IDs already resolved, reading checkpoint\n")
  gbif_backbone_key_family = read_csv(backbone_file("family"), show_col_types = FALSE)
} else {
  cat("\tLooking up family IDs\n")

  gbif_query_family = taxa_unique %>%
    select(class,
           order,
           family) %>%
    filter(!is.na(family)) %>%
    distinct()

  gbif_backbone_family = with_retry(
    function() name_backbone_checklist(gbif_query_family, bucket_size = 500, sleep = 10),
    label = "backbone family"
  )

  rownames(gbif_query_family) = NULL
  rownames(gbif_backbone_family) = NULL

  gbif_backbone_key_family = bind_cols(gbif_query_family %>%
                                         rename(class_q = class,
                                                order_q = order,
                                                family_q = family),
                                       gbif_backbone_family) %>%
    mutate(local_occ_count = NA) %>%
    filter(rank == "FAMILY")

  write_csv(gbif_backbone_key_family, backbone_file("family"))
}

# row bind all queries into the full backbone, shared by every system's occurrence query below
gbif_backbone_key = bind_rows(gbif_backbone_key_species,
                              gbif_backbone_key_genus,
                              gbif_backbone_key_family)

valid_keys = unique(na.omit(gbif_backbone_key$usageKey))

cat("Backbone resolution done:", length(valid_keys), "valid GBIF keys.\n")
```

# Query local occurrences (per system, checkpointed)
```{r}
# lookup table
# gbif_wkt_polygon (optional): a WKT POLYGON string defining the study system's geographic
# extent. When present for a study system, occurrences are queried within this polygon
# instead of by gbif_country_code / gbif_gid1_code (see predicate construction below).
study_system_lookup = data.frame(
  study_system = c("brazil", "panama", "sierra_nevada", "pennsylvania"),
  gbif_country_code = c("BR", "PA", "US", "US"),
  gbif_gid1_code = c("BRA.25_1", NA, "USA.5_1", "USA.39_1"),  # state/provincial codes
  gbif_wkt_polygon = c(NA,
                       NA,
                       "POLYGON ((-121.27333176175197593 40.27541263975778207, -119.98747131851729364 37.76972609026517347, -118.74167109545859944 36.2741249368575609, -118.01260970478564616 36.38001696734033175, -118.77690357142141409 37.80071271972609281, -119.92310158236205098 38.82366483213162667, -120.91068744657029299 40.3077386192145255, -121.27333176175197593 40.27541263975778207))",
                       NA),
  stringsAsFactors = FALSE
)

for (system in study_systems) {

  if (file.exists(occ_file(system))) {
    cat(paste0("\nOccurrences for '", system, "' already downloaded, skipping\n"))
    next
  }

  this_row = study_system_lookup[study_system_lookup$study_system == system, ]

  if (nrow(this_row) == 0) {
    cat(paste0("\nNo GBIF country/region lookup defined for study_system '", system,
               "' -- add it to study_system_lookup above. Skipping for now.\n"))
    next
  }

  cat(paste0("\nQuerying local occurrences for '", system, "'...\n"))

  gbif_country_code = this_row$gbif_country_code
  gbif_gid1_code = this_row$gbif_gid1_code
  gbif_wkt_polygon = this_row$gbif_wkt_polygon

  # default to a polygon query when one is provided for this study system, otherwise fall
  # back to country (+ optional state/province) codes
  if (!is.na(gbif_wkt_polygon)) {
    preds = list(
      pred_within(gbif_wkt_polygon),
      pred_in("taxonKey", valid_keys)
    )
  } else {
    preds = list(
      pred("country", gbif_country_code),
      pred_in("taxonKey", valid_keys)
    )
    if (!is.na(gbif_gid1_code)) {
      preds = c(preds, list(pred("gadm", gbif_gid1_code)))
    }
  }

  # submit + wait + download + import as one retry unit -- if any part fails, resubmit fresh
  occ_data_raw = with_retry(function() {
    download_submission = occ_download(
      pred_and(.list = preds),
      format = "SIMPLE_CSV"
    )

    occ_download_wait(download_submission)

    occ_download_file = occ_download_get(download_submission, path = here(run_dir, "output"), overwrite = TRUE)
    data = occ_download_import(occ_download_file)

    if (file.exists(as.character(occ_download_file))) {
      file.remove(as.character(occ_download_file))
    }

    data
  }, label = paste0("occurrence download (", system, ")"))

  # clean, exact per-key counts — no facet, no hierarchy contamination
  occ_data_summary = occ_data_raw %>%
    count(taxonKey, name = "occ_count") %>%
    rename(usageKey = taxonKey)

  write_csv(occ_data_summary, occ_file(system))

  cat(paste0("Done with '", system, "'.\n"))
}
```

# Process GBIF data

Builds the composite output from whichever per-system occurrence files exist right now. Safe to run before every system has finished -- re-run later to pick up the rest.
```{r}
completed_systems = study_systems[file.exists(occ_file(study_systems))]
pending_systems = setdiff(study_systems, completed_systems)

if (length(pending_systems) > 0) {
  cat("NOTE: still waiting on GBIF occurrence queries for:", paste(pending_systems, collapse = ", "), "\n",
      "Re-run this script to pick them up; the composite output below only reflects:", paste(completed_systems, collapse = ", "), "\n\n")
}

# build the per-accession GBIF occurrence join for one system's occurrence counts
build_taxa_out_for_system = function(system) {
  occ_data_summary = read_csv(occ_file(system), show_col_types = FALSE) %>%
    mutate(usageKey = as.character(usageKey)) %>%
    select(usageKey,
           occ_count)

  gbif_backbone_key_occ = gbif_backbone_key %>%
    left_join(occ_data_summary, by = "usageKey") %>%
    mutate(local_occ_count = ifelse(is.na(occ_count), 0, occ_count))

  taxa_out %>%
    left_join(gbif_backbone_key_occ %>%
                filter(rank == "FAMILY") %>%
                rename(gbif_count_family_local = local_occ_count,
                       gbif_familyKey = familyKey) %>%
                select(class_q,
                       order_q,
                       family_q,
                       gbif_familyKey,
                       gbif_count_family_local),
              by = c("class" = "class_q",
                     "order" = "order_q",
                     "family" = "family_q")) %>%
    left_join(gbif_backbone_key_occ %>%
                filter(rank == "GENUS") %>%
                rename(gbif_count_genus_local = local_occ_count,
                       gbif_genusKey = genusKey) %>%
                select(class_q,
                       order_q,
                       family_q,
                       genus_q,
                       gbif_genusKey,
                       gbif_count_genus_local),
              by = c("class" = "class_q",
                     "order" = "order_q",
                     "family" = "family_q",
                     "genus" = "genus_q")) %>%
    left_join(gbif_backbone_key_occ %>%
                filter(rank %in% c("SPECIES", "SUBSPECIES"),
                       !is.na(scientific_name_q)) %>%
                rename(local_gbif_count_species = local_occ_count,
                       gbif_speciesKey = speciesKey) %>%
                select(class_q,
                       order_q,
                       family_q,
                       scientific_name_q,
                       gbif_speciesKey,
                       local_gbif_count_species) %>%
                rename(gbif_speciesKey_s = gbif_speciesKey,
                       local_gbif_count_species_s = local_gbif_count_species),
              by = c("class" = "class_q",
                     "order" = "order_q",
                     "family" = "family_q",
                     "scientificName" = "scientific_name_q")) %>%
    left_join(gbif_backbone_key_occ %>%
                filter(rank %in% c("SPECIES", "SUBSPECIES"),
                       !is.na(specific_epithet_q)) %>%
                rename(local_gbif_count_species = local_occ_count,
                       gbif_speciesKey = speciesKey) %>%
                select(class_q,
                       order_q,
                       family_q,
                       genus_q,
                       specific_epithet_q,
                       gbif_speciesKey,
                       local_gbif_count_species) %>%
                rename(gbif_speciesKey_e = gbif_speciesKey,
                       local_gbif_count_species_e = local_gbif_count_species) %>%
                distinct(),
              by = c("class" = "class_q",
                     "order" = "order_q",
                     "family" = "family_q",
                     "genus" = "genus_q",
                     "specificEpithet" = "specific_epithet_q")) %>%
    mutate(gbif_speciesKey = coalesce(gbif_speciesKey_s, gbif_speciesKey_e),
           gbif_count_species_local = coalesce(local_gbif_count_species_s, local_gbif_count_species_e),
           # correct errant parents with counts less than children
           gbif_count_genus_local = if_else((!is.na(gbif_count_species_local) & is.na(gbif_count_genus_local)) |
                                              gbif_count_species_local > gbif_count_genus_local,
                                            gbif_count_species_local,
                                            gbif_count_genus_local),
           gbif_count_family_local = if_else((!is.na(gbif_count_genus_local) & is.na(gbif_count_family_local)) |
                                               gbif_count_genus_local > gbif_count_family_local,
                                             gbif_count_genus_local,
                                             gbif_count_family_local)) %>%
    select(-gbif_speciesKey_s,
           -gbif_speciesKey_e,
           -local_gbif_count_species_s,
           -local_gbif_count_species_e) %>%
    distinct() %>%
    mutate(study_system = system)
}

# one row per (accession, study_system) -- lets downstream scripts either take a union across
# systems (for taxonomy: is this candidate plausible anywhere in the run) or look at a specific
# system (for a home-system / contamination check)
taxa_out = map_dfr(completed_systems, build_taxa_out_for_system)

cat("Done with GBIF.\n")

```

# Export results
```{r}
write.csv(taxa_out, taxa_out_file, row.names = FALSE)
```
