---
title: "Query taxonomy geography"
author: "Cob Staines"
output: html_document
---

# Description

This script queries the geographic range of taxonomies corresponding to vsearch hits, to support a more accurate, locally informed classification.

This script requires API keys for GBIF in .Renviron (see this [Tutorial](https://docs.ropensci.org/rgbif/articles/gbif_credentials.html)).

If you want to skip this step you can, just set the parameter `gbif_query` to `FALSE` in the subsequent script (`06_classify_asv.Rmd`).

# Config
```{r}
## Load in Libraries (install as needed)
librarian::shelf(tidyverse, yaml, here, rgbif)

# requires GBIF token in .Renviron

# import run config file
run_dir = here("16S_sequence_processing", "runs", "penn_2026-03-10")

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# geographic systems
#   any of: brazil, panama, california, pennsylvania
# recommend running one system at a time, for cleaner geographic distinction
study_system = "pennsylvania"

taxa_out_file = here(run_dir, "output", paste0(config$run$name, "_05_taxonomy_table_gbif_occ.csv"))

```

# Import taxa
```{r}
# load taxonomy table
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_02_taxonomy_table_filtered.csv")))) {
  taxa_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_02_taxonomy_table_filtered.csv")))
} else {
  cat("No taxonomy table found, skipping.\n")
  taxa_raw = tibble(c())
}

# load web blast table
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv")))) {
  taxa_web_blast_metazoa = read_csv(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv"))) %>%
    mutate(species = sub("^\\S+\\s+", "", species)) %>%
    select(all_of(colnames(taxa_raw))) %>%
    filter(kingdom == "Metazoa")
} else {
  cat("No web blast results found, skipping.\n")
  taxa_web_blast_metazoa = tibble(c())
}

# clean up taxa
taxa_clean = bind_rows(taxa_raw,
                       taxa_web_blast_metazoa)%>%
  mutate(
    scientificName = case_when(
      !is.na(genus) & !is.na(species) ~ paste0(genus, " ", species),
      !is.na(genus) & is.na(species) ~ genus,
      .default = NA),
    species_simple = species,
    species_simple = if_else(grepl("^sp\\.", species_simple), NA, species_simple),
    species_simple = gsub(" x .*", "", species_simple),
    species_simple = gsub("^cf\\. *", "", species_simple),
    species_simple = gsub("^aff\\. *", "", species_simple),
    specificEpithet = gsub(" .*", "", species_simple),
    intraspecificEpithet = gsub("^[^ ]+ ?", "", species_simple),
    intraspecificEpithet = if_else(intraspecificEpithet == "", NA, intraspecificEpithet))

# copy for output
taxa_out = taxa_clean

# get unique for querying
taxa_unique = taxa_clean %>%
  select(-accession) %>%
  distinct() %>%
  arrange(domain, kingdom, phylum, class, order, family, genus, species)
```


# Query GBIF
```{r}

cat("\nQuery GBIF for occurences in target countries (families, genuses & species)...\n")

## species
cat("\tLooking up GBIF species IDs\n")

# by full scientific name
gbif_query_species_sn = taxa_unique %>%
  filter(!is.na(species)) %>%
  select(class,
         order,
         family,
         scientificName) %>%
  distinct()

# query IDs by scientific name
gbif_backbone_species_sn = name_backbone_checklist(gbif_query_species_sn, bucket_size = 100)

# parse IDs
rownames(gbif_query_species_sn) = NULL
rownames(gbif_backbone_species_sn) = NULL
gbif_backbone_key_species_sn = bind_cols(gbif_query_species_sn %>%
                                           rename(class_q = class,
                                                  order_q = order,
                                                  family_q = family,
                                                  scientific_name_q = scientificName),
                                         gbif_backbone_species_sn) %>%
  mutate(local_occ_count = NA) %>%
  filter(rank %in% c("SPECIES", "SUBSPECIES"))

# by specific epithet, for those which returned NA
gbif_query_species_e = taxa_unique %>%
  filter(!is.na(specificEpithet)) %>%
  select(class,
         order,
         family,
         genus,
         specificEpithet,
         scientificName) %>%
  anti_join(gbif_backbone_key_species_sn, by = c("class" = "class_q",
                                                 "order" = "order_q",
                                                 "family" = "family_q",
                                                 "scientificName" = "scientific_name_q")) %>%
  select(-scientificName) %>%
  distinct()

# query IDs by specific epithet
gbif_backbone_species_e = name_backbone_checklist(gbif_query_species_e, bucket_size = 100)

# parse IDs
rownames(gbif_query_species_e) = NULL
rownames(gbif_backbone_species_e) = NULL
gbif_backbone_key_species_e = bind_cols(gbif_query_species_e %>%
                                          rename(class_q = class,
                                                 order_q = order,
                                                 family_q = family,
                                                 genus_q = genus,
                                                 specific_epithet_q = specificEpithet),
                                        gbif_backbone_species_e) %>%
  mutate(local_occ_count = NA) %>%
  filter(rank %in% c("SPECIES", "SUBSPECIES"))

# combine specific results
gbif_backbone_key_species = bind_rows(gbif_backbone_key_species_sn,
                                      gbif_backbone_key_species_e)

## query by genus
cat("\tLooking up genus IDs\n")

gbif_query_genus = taxa_unique %>%
  select(class,
         order,
         family,
         genus) %>%
  filter(!is.na(genus)) %>%
  distinct()

gbif_backbone_genus = name_backbone_checklist(gbif_query_genus, bucket_size = 100)

rownames(gbif_query_genus) = NULL
rownames(gbif_backbone_genus) = NULL

gbif_backbone_key_genus = bind_cols(gbif_query_genus %>%
                                      rename(class_q = class,
                                             order_q = order,
                                             family_q = family,
                                             genus_q = genus),
                                    gbif_backbone_genus) %>%
  mutate(local_occ_count = NA) %>%
  filter(rank == "GENUS")

## query by family
cat("\tLooking up family IDs\n")

gbif_query_family = taxa_unique %>%
  select(class,
         order,
         family) %>%
  filter(!is.na(family)) %>%
  distinct()

gbif_backbone_family = name_backbone_checklist(gbif_query_family, bucket_size = 100)

rownames(gbif_query_family) = NULL
rownames(gbif_backbone_family) = NULL

gbif_backbone_key_family = bind_cols(gbif_query_family %>%
                                       rename(class_q = class,
                                              order_q = order,
                                              family_q = family),
                                     gbif_backbone_family) %>%
  mutate(local_occ_count = NA) %>%
  filter(rank == "FAMILY")

# row bind all queries
gbif_backbone_key = bind_rows(gbif_backbone_key_species,
                              gbif_backbone_key_genus,
                              gbif_backbone_key_family)

cat("\tQuerying local taxa occurences\n")

gbif_country_code = recode_values(study_system,
                                  "brazil" ~ "BR",
                                  "panama" ~ "PA",
                                  "california" ~ "US",
                                  "pennsylvania" ~ "US")

gbif_state_province = recode_values(study_system,
                                    "brazil" ~ NA,
                                    "panama" ~ NA,
                                    "california" ~ "California",
                                    "pennsylvania" ~ "Pennsylvania")

# SQL query
valid_keys = na.omit(gbif_backbone_key$usageKey)
keys_sql = paste(valid_keys, collapse = ", ")

if (is.na(gbif_state_province)) {
  sql_query = paste0("
    SELECT taxonKey, speciesKey, familyKey, orderKey, 
           species, genus, family,
           COUNT(*) AS occ_count
    FROM occurrence
    WHERE countryCode = '", gbif_country_code, "'
    AND taxonKey IN (", keys_sql, ")
    GROUP BY taxonKey, speciesKey, familyKey, orderKey,
             species, genus, family
  ")
} else {
    sql_query = paste0("
    SELECT taxonKey, speciesKey, familyKey, orderKey, 
           species, genus, family,
           COUNT(*) AS occ_count
    FROM occurrence
    WHERE countryCode = '", gbif_country_code, "'
    AND stateProvince = '", gbif_state_province, "'
    AND taxonKey IN (", keys_sql, ")
    GROUP BY taxonKey, speciesKey, familyKey, orderKey,
             species, genus, family
  ")
}

# submit query
sql_submission = occ_download_sql(sql_query)

# Wait until query is complete
occ_download_wait(sql_submission)

# download data
occ_download_file = occ_download_get(sql_submission, path = here(run_dir, "output"))

# import data
occ_data = occ_download_import(occ_download_file)

# export with reusable name
write_csv(pa_data, here(run_dir, "output", paste0(config$run$name, "_05_gbif_occ_data.csv")))

# drop downloads to reduce clutter
if (file.exists(as.character(occ_download_file)) & file.exists(here(run_dir, "output", paste0(config$run$name, "_05_gbif_occ_data.csv")))) {
  file.remove(as.character(occ_download_file))
} else {
  stop("Download file error (Panama), stopping.")
}
```

# Process GBIF data
```{r}
occ_data = read_csv(here(run_dir, "output", paste0(config$run$name, "_05_gbif_occ_data.csv"))) %>%
  mutate(usageKey = as.character(taxonkey)) %>%
  select(usageKey,
         occ_count)

gbif_backbone_key_occ = gbif_backbone_key %>%
  left_join(occ_data, by = "usageKey") %>%
  mutate(local_occ_count = ifelse(is.na(occ_count), 0, occ_count))

# join with taxa_out
taxa_out = taxa_out %>%
  left_join(gbif_backbone_key_occ %>%
              filter(rank == "FAMILY") %>%
              rename(gbif_count_family_local = local_occ_count,
                     gbif_familyKey = familyKey) %>%
              select(class_q,
                     order_q,
                     family_q,
                     gbif_familyKey,
                     gbif_count_family_local),
            by = c("class" = "class_q",
                   "order" = "order_q",
                   "family" = "family_q")) %>%
  left_join(gbif_backbone_key_occ %>%
              filter(rank == "GENUS") %>%
              rename(gbif_count_genus_local = local_occ_count,
                     gbif_genusKey = genusKey) %>%
              select(class_q,
                     order_q,
                     family_q,
                     genus_q,
                     gbif_genusKey,
                     gbif_count_genus_local),
            by = c("class" = "class_q",
                   "order" = "order_q",
                   "family" = "family_q",
                   "genus" = "genus_q")) %>%
  left_join(gbif_backbone_key_occ %>%
              filter(rank %in% c("SPECIES", "SUBSPECIES"),
                     !is.na(scientific_name_q)) %>%
              rename(local_gbif_count_species = local_occ_count,
                     gbif_speciesKey = speciesKey) %>%
              select(class_q,
                     order_q,
                     family_q,
                     scientific_name_q,
                     gbif_speciesKey,
                     local_gbif_count_species) %>%
              rename(gbif_speciesKey_s = gbif_speciesKey,
                     local_gbif_count_species_s = local_gbif_count_species),
            by = c("class" = "class_q",
                   "order" = "order_q",
                   "family" = "family_q",
                   "scientificName" = "scientific_name_q")) %>%
  left_join(gbif_backbone_key_occ %>%
              filter(rank %in% c("SPECIES", "SUBSPECIES"),
                     !is.na(specific_epithet_q)) %>%
              rename(local_gbif_count_species = local_occ_count,
                     gbif_speciesKey = speciesKey) %>%
              select(class_q,
                     order_q,
                     family_q,
                     genus_q,
                     specific_epithet_q,
                     gbif_speciesKey,
                     local_gbif_count_species) %>%
              rename(gbif_speciesKey_e = gbif_speciesKey,
                     local_gbif_count_species_e = local_gbif_count_species) %>%
              distinct(),
            by = c("class" = "class_q",
                   "order" = "order_q",
                   "family" = "family_q",
                   "genus" = "genus_q",
                   "specificEpithet" = "specific_epithet_q")) %>%
  mutate(gbif_speciesKey = coalesce(gbif_speciesKey_s, gbif_speciesKey_e),
         gbif_count_species_local = coalesce(local_gbif_count_species_s, local_gbif_count_species_e),
         # correct errant parents with counts less than children
         gbif_count_genus_local = if_else((!is.na(gbif_count_species_local) & is.na(gbif_count_genus_local)) | 
                                            gbif_count_species_local > gbif_count_genus_local,
                                          gbif_count_species_local,
                                          gbif_count_genus_local),
         gbif_count_family_local = if_else((!is.na(gbif_count_genus_local) & is.na(gbif_count_family_local)) | 
                                             gbif_count_genus_local > gbif_count_family_local,
                                           gbif_count_genus_local,
                                           gbif_count_family_local)) %>%
  select(-gbif_speciesKey_s,
         -gbif_speciesKey_e,
         -local_gbif_count_species_s,
         -local_gbif_count_species_e) %>%
  distinct()

cat("Done with GBIF.\n")

```

# Export results
```{r}
write.csv(taxa_out, taxa_out_file, row.names = FALSE)
```