---
title: "Classify taxonomic hits"
author: "Cob Staines"
output: html_document
---

# Description

This script takes in hits from any search method (e.g. vsearch, blast, web blast), and optionally gbif taxonomic occurance data, and assigns a likely taxonomy to each ASV following a hierarchy `accept_method`s. All assigned taxonomy from all methods, along with all hits, are exported to `hybrid_classification_out`. The single best classifications by the most specific `accept_method` for each ASV are exported to `classification_out`.

See the algorithm below for details on each `accept_method` and to adjust parameters.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here, phylotools)

# import run config file
run_dir = here("16S_sequence_processing", "runs", "sierra_nevada_16s_2026-09-11")

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# minimum sequence overlap (0 -> no filtering) [units of Bp]
sequence_overlap_cutoff = 0

# Did you run a gbif query (script 05) for local taxa, which you want incorporated in classification?
gbif_query = FALSE

# minimum p_identical (against the local contaminant library, script 02 / s07-c) for a hit to
# flag an ASV as a likely contaminant
contaminant_identity_cutoff = 0.98

# output files
hybrid_classification_out = here(run_dir, "output", paste0(config$run$name, "_06_hybrid_classification.csv"))

classification_out = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))


```

# Prepare for classification
```{r}
# read in feature table
feature_table =
  read_tsv(here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

feature_long = feature_table %>%
  pivot_longer(cols = -asv_id,
               names_to = "sample_id",
               values_to = "asv_count") %>%
  filter(asv_count != 0) %>%
  arrange(sample_id) %>%
  select(sample_id,
         asv_id,
         asv_count)

asv_totals = feature_long %>%
  group_by(asv_id) %>%
  summarise(asv_total_count = sum(asv_count),
            sample_count = n_distinct(sample_id),
            samples = paste(sort(unique(sample_id)), collapse = ", ")) %>%
  arrange(desc(asv_total_count))

# read in sequences
sequence_table =
  read.fasta(file = here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "representative_sequences", 
                           list.files(here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "representative_sequences")), 
                           "data/dna-sequences.fasta"))%>%
  rename(asv_id = seq.name, 
         sequence = seq.text) %>%
  arrange(asv_id)

# build backbone
backbone = asv_totals %>%
  left_join(sequence_table, by = "asv_id")

# load hits
## QIIME2
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_02_qiime2_hits.csv")))) {
  qiime2_hits_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_02_qiime2_hits.csv")))
} else {
  cat("No QIIME2 hits table found, skipping.\n")
  qiime2_hits_raw = tibble(asv_id = NA)
}

## Web BLAST
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv")))) {
  web_blast_hits_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv"))) %>%
    mutate(species = sub("^\\S+\\s+", "", species))
} else {
  cat("No QIIME2 hits table found, skipping.\n")
  web_blast_hits_raw = tibble(asv_id = NA)
}

# bind with web blast
hits_raw = backbone %>%
  left_join(bind_rows(qiime2_hits_raw,
                      web_blast_hits_raw) %>%
              select(-any_of(c("asv_total_count",
                               "sample_count",
                               "samples",
                               "sequence"))), by = "asv_id")

# clean hits
hits_clean = hits_raw %>%
  filter(accession != "*") %>%
  mutate(p_identical = percent_identical / 100,
         p_overlap = seq_overlap / (s_end - s_start + 1))

# handle gbif geographic query
if (gbif_query) {
  # read in taxa + geography
  taxa_geography = read.csv(here(run_dir, "output", paste0(config$run$name, "_05_taxonomy_table_gbif_occ.csv")))
  
  hits_geography = hits_clean %>%
    left_join(taxa_geography %>%
                select(accession,
                       any_of(c("specificEpithet",
                                "intraspecificEpithet",
                                "gbif_count_family_local",
                                "gbif_count_genus_local",
                                "gbif_count_species_local"))) %>%
                distinct(), by = "accession") %>%
    mutate(specificEpithet = coalesce(specificEpithet, species), 
           local_species = case_when(
             gbif_count_species_local > 0 ~ TRUE,
             gbif_count_species_local == 0 ~ FALSE,
             gbif_count_genus_local == 0 ~ FALSE,
             gbif_count_family_local == 0 ~ FALSE,
             .default = NA))
  
} else {
  # set gbif data to NA
  hits_geography = hits_clean %>%
    mutate(specificEpithet = species,
           intraspecificEpithet = NA,
           gbif_count_family_local = NA,
           gbif_count_genus_local = NA,
           gbif_count_species_local = NA,
           local_species = NA)
}


# define classification functions

## consensus function at a specified taxonomic rank, identity threshold, and consensus threshold
find_consensus = function(hit_table, taxonomic_rank, identity_th = 0, consensus_th = 0.5, min_consensus_hits = 1, tie_handling = "error") {
  
  if (!(tie_handling %in% c("error", "pass", "drop"))) {
    stop('Invalid value for tie_handling. Must be in c("error", "pass", "drop")')
  }
  
  hit_table_filtered = hit_table %>%
    filter(p_identical >= identity_th)
  
  if (consensus_th <= 0.5) {
    warning("Using a consensus_th <= 0.5 may result in multiple consensus classification. Consensus classification(s) will be selected with the greatest consensus value, with ties handled according to tie_handling ('error', 'pass', 'drop')")
  }
  
  if (!(taxonomic_rank %in% c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"))) {
    stop('taxonnomic_rank not recognized, must be one of c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")')
  }
  
  taxon_groups_list = list(
    species = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"),
    genus = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus"),
    family = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family"),
    order = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"),
    class = c("asv_id", "method", "domain", "kingdom", "phylum", "class"),
    phylum = c("asv_id", "method", "domain", "kingdom", "phylum"),
    kingdom = c("asv_id", "method", "domain", "kingdom"),
    domain = c("asv_id", "method", "domain")
  )
  
  taxon_group = taxon_groups_list[[taxonomic_rank]]
  
  consensus_table = hit_table_filtered %>%
    group_by(asv_id, method) %>%
    mutate(asv_hits_n = n()) %>%
    ungroup() %>%
    group_by_at(taxon_group) %>%
    mutate(taxon_hits_n = n()) %>%
    ungroup() %>%
    mutate(consensus_value = taxon_hits_n / asv_hits_n) %>%
    filter(taxon_hits_n >= min_consensus_hits,
           consensus_value >= consensus_th) %>%
    arrange(desc(asv_total_count),
            asv_id,
            method,
            desc(consensus_value),
            desc(p_identical)) %>%
    group_by_at(c(taxon_group,
                  "consensus_value")) %>%
    summarise(n_hits = first(taxon_hits_n),
              p_identical_min = mean(p_identical),
              p_identical_max = mean(p_identical),
              p_identical_mean = mean(p_identical),
              seq_overlap_max = max(seq_overlap),
              .groups = "drop") %>%
    group_by(asv_id, method) %>%
    slice_max(order_by = consensus_value, n = 1, with_ties = TRUE) %>%
    mutate(n_hits_consensus = n()) %>%
    ungroup() %>%
    mutate(identity_th = identity_th,
           consensus_th = consensus_th,
           min_consensus_hits = min_consensus_hits,
           taxonomic_rank = taxonomic_rank)
  
  tie_ids = consensus_table %>%
    filter(n_hits_consensus > 1) %>%
    select(asv_id) %>%
    distinct() %>%
    pull(asv_id)
  
  if (length(tie_ids) > 0) {
    if (tie_handling == "error") {
      stop(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nAborting."))
    } else if (tie_handling == "pass") {
      warning(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nPassing to output."))
    } else if (tie_handling == "drop") {
      warning(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nDropping from output."))
      consensus_table = consensus_table %>%
        filter(consensus <= 1)
    }
  }
  
  return(consensus_table)
}

# calculate stats for taxonomic hit groups
hits_group_calc = function(hits_df, group_cols, agg_cols = c()) {
  hits_df %>%
    group_by_at(group_cols) %>%
    summarize(
      n_hits = n(),
      p_identical_mean = mean(percent_identical) / 100,
      p_identical_max = max(percent_identical) / 100,
      p_identical_min = min(percent_identical) / 100,
      seq_overlap_max = max(seq_overlap),
      across(all_of(agg_cols), \(x) max(x, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    arrange(desc(asv_total_count), asv_id, method, desc(p_identical_max)) %>%
    group_by(asv_id, method) %>%
    mutate(next_pimax = lead(p_identical_max),
           drop_to_next_pimax = p_identical_max - next_pimax,
           pimax_rank = row_number()) %>%
    ungroup() %>%
    select(all_of(group_cols),
           n_hits,
           p_identical_max,
           p_identical_mean,
           p_identical_min,
           drop_to_next_pimax,
           everything())
}

# taxonomic ranks, ordered broadest to most specific
tax_ranks = c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")

## merge two full-taxonomy hit-groups (already sorted into a fixed row order by the caller)
## into a single row, combining each rank as "value_1/value_2" wherever the pair disagrees,
## so e.g. genus_1/genus_2 always correspond to the same underlying hit-group as species_1/species_2
merge_taxon_pair = function(row1, row2) {
  merged_ranks = map(tax_ranks, function(rank) {
    v1 = row1[[rank]]
    v2 = row2[[rank]]
    case_when(
      is.na(v1) & is.na(v2) ~ NA_character_,
      !is.na(v1) & !is.na(v2) & v1 == v2 ~ v1,
      .default = paste0(coalesce(v1, "NA"), "/", coalesce(v2, "NA"))
    )
  })
  names(merged_ranks) = tax_ranks

  # taxonomic_rank = most specific rank at which both hit-groups actually have a value
  taxonomic_rank = NA_character_
  for (rank in rev(tax_ranks)) {
    if (!is.na(row1[[rank]]) && !is.na(row2[[rank]])) {
      taxonomic_rank = rank
      break
    }
  }

  bind_cols(as_tibble(merged_ranks), tibble(taxonomic_rank = taxonomic_rank))
}

## from a table of candidate hits (already filtered to p_identical == 1 & species-level
## candidates), find (asv_id, method) pairs with exactly 2 distinct full-taxonomy hit-groups,
## and merge each pair via merge_taxon_pair() into one row reporting "value_1/value_2" at
## every disagreeing rank
build_tie2_accept = function(candidate_hits, accept_method_label) {
  taxon_groups = candidate_hits %>%
    distinct(asv_id, method, asv_total_count, sample_count, samples, sequence,
             domain, kingdom, phylum, class, order, family, genus, species)

  tie2_keys = taxon_groups %>%
    count(asv_id, method, name = "n_taxa") %>%
    filter(n_taxa == 2) %>%
    select(asv_id, method)

  if (nrow(tie2_keys) == 0) {
    return(tibble())
  }

  taxon_groups %>%
    inner_join(tie2_keys, by = c("asv_id", "method")) %>%
    arrange(asv_id, method, domain, kingdom, phylum, class, order, family, genus, species) %>%
    group_by(asv_id, method, asv_total_count, sample_count, samples, sequence) %>%
    group_modify(~ merge_taxon_pair(.x[1, ], .x[2, ])) %>%
    ungroup() %>%
    left_join(candidate_hits %>%
                group_by(asv_id, method) %>%
                summarise(n_hits = n(),
                          p_identical_mean = mean(percent_identical) / 100,
                          p_identical_max = max(percent_identical) / 100,
                          p_identical_min = min(percent_identical) / 100,
                          seq_overlap_max = max(seq_overlap),
                          gbif_count_family_local = first(gbif_count_family_local),
                          gbif_count_genus_local = first(gbif_count_genus_local),
                          gbif_count_species_local = first(gbif_count_species_local),
                          local_species = any(local_species, na.rm = TRUE),
                          .groups = "drop"),
              by = c("asv_id", "method")) %>%
    mutate(accept_method = accept_method_label)
}
```

# Contaminant screening

Independently flag ASVs matching a local contaminant reference library (script 02 / s07-c, optional). This is orthogonal to the taxonomy `accept_method` hierarchy above: a contaminant match does not affect taxonomic classification, it's carried alongside it as its own flag.
```{r}
# read in local contaminant library screen, if run
if (file.exists(here(run_dir, "output", paste0(config$run$name, "_02_contaminant_hits.csv")))) {
  contaminant_hits_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_02_contaminant_hits.csv")))
} else {
  cat("No contaminant hits table found, skipping.\n")
  contaminant_hits_raw = tibble(asv_id = NA)
}

# hits meeting the contaminant identity cutoff
contaminant_hits = contaminant_hits_raw %>%
  filter(!is.na(asv_id)) %>%
  mutate(p_identical = percent_identical / 100) %>%
  filter(p_identical >= contaminant_identity_cutoff)

# best (max p_identical) contaminant match(es) per ASV
contaminant_top_hits = contaminant_hits %>%
  group_by(asv_id) %>%
  filter(p_identical == max(p_identical)) %>%
  ungroup() %>%
  distinct(asv_id, contaminant_asv_id, p_identical)

# accept method 1: unambiguous max p_identity (single best-matching contaminant reference)
accept_contaminant_max = contaminant_top_hits %>%
  group_by(asv_id) %>%
  filter(n() == 1) %>%
  ungroup() %>%
  transmute(asv_id,
            contaminant_id = contaminant_asv_id,
            contaminant_p_identity = p_identical,
            contaminant_accept_method = "max_p_identity")

# accept method 2: tie (>=2 distinct contaminant references tied at the max p_identical),
# reported as "id_1/id_2/..."
accept_contaminant_tie = contaminant_top_hits %>%
  group_by(asv_id) %>%
  filter(n() > 1) %>%
  summarise(contaminant_id = paste(sort(unique(contaminant_asv_id)), collapse = "/"),
            contaminant_p_identity = first(p_identical),
            .groups = "drop") %>%
  mutate(contaminant_accept_method = "tie")

# combine, then fill in every ASV so unflagged ASVs report contaminant_flag = FALSE
contaminant_summary = backbone %>%
  select(asv_id) %>%
  left_join(bind_rows(accept_contaminant_max, accept_contaminant_tie), by = "asv_id") %>%
  mutate(contaminant_flag = !is.na(contaminant_id))

# ASVs with zero taxonomic hits from any method (e.g. a fragment too short/degenerate to match
# any reference) never enter hits_geography, since hits_clean drops rows with no accession --
# so without this, a contaminant match on one of them would be silently lost downstream. Give
# just that subset (backbone ASVs missing from hits_geography but contaminant-flagged) a
# minimal placeholder row so they still surface in the combined outputs below.
hits_orphan_contaminant = backbone %>%
  anti_join(hits_geography, by = "asv_id") %>%
  semi_join(contaminant_summary %>% filter(contaminant_flag), by = "asv_id") %>%
  select(asv_id, asv_total_count, sample_count, samples, sequence)
```

# Group hits by taxonomy
```{r warning=FALSE}
# collect hits of common taxonomies into one candidate row, with descriptive stats

hits_species_local = hits_group_calc(hits_geography %>%
                                       filter(!(local_species %in% FALSE),
                                              !is.na(specificEpithet)),
                                     c("asv_id",
                                       "asv_total_count",
                                       "sample_count",
                                       "samples",
                                       "sequence",
                                       "method",
                                       "domain",
                                       "kingdom",
                                       "phylum",
                                       "class",
                                       "order",
                                       "family",
                                       "genus",
                                       "specificEpithet"),
                                     c("gbif_count_family_local",
                                       "gbif_count_genus_local",
                                       "gbif_count_species_local",
                                       "local_species"))

hits_species = hits_group_calc(hits_geography %>%
                                 filter(!is.na(specificEpithet)),
                               c("asv_id",
                                 "asv_total_count",
                                 "sample_count",
                                 "samples",
                                 "sequence",
                                 "method",
                                 "domain",
                                 "kingdom",
                                 "phylum",
                                 "class",
                                 "order",
                                 "family",
                                 "genus",
                                 "specificEpithet"),
                               c("gbif_count_family_local",
                                 "gbif_count_genus_local",
                                 "gbif_count_species_local",
                                 "local_species"))

hits_genus_local = hits_group_calc(hits_geography %>%
                                     filter(gbif_count_genus_local > 20,
                                            !is.na(genus)),
                                   c("asv_id",
                                     "asv_total_count",
                                     "sample_count",
                                     "samples",
                                     "sequence",
                                     "method",
                                     "domain",
                                     "kingdom",
                                     "phylum",
                                     "class",
                                     "order",
                                     "family",
                                     "genus",
                                     "gbif_count_family_local",
                                     "gbif_count_genus_local"))

hits_genus = hits_group_calc(hits_geography %>%
                               filter(!is.na(genus)),
                             c("asv_id",
                               "asv_total_count",
                               "sample_count",
                               "samples",
                               "sequence",
                               "method",
                               "domain",
                               "kingdom",
                               "phylum",
                               "class",
                               "order",
                               "family",
                               "genus",
                               "gbif_count_family_local",
                               "gbif_count_genus_local"))

hits_family_local = hits_group_calc(hits_geography %>%
                                      filter(gbif_count_family_local > 50,
                                             !is.na(family)),
                                    c("asv_id",
                                      "asv_total_count",
                                      "sample_count",
                                      "samples",
                                      "sequence",
                                      "method",
                                      "domain",
                                      "kingdom",
                                      "phylum",
                                      "class",
                                      "order",
                                      "family",
                                      "gbif_count_family_local"))

hits_family = hits_group_calc(hits_geography %>%
                                filter(!is.na(family)),
                              c("asv_id",
                                "asv_total_count",
                                "sample_count",
                                "samples",
                                "sequence",
                                "method",
                                "domain",
                                "kingdom",
                                "phylum",
                                "class",
                                "order",
                                "family",
                                "gbif_count_family_local"))

hits_order = hits_group_calc(hits_geography %>%
                               filter(!is.na(order)),
                             c("asv_id",
                               "asv_total_count",
                               "sample_count",
                               "samples",
                               "sequence",
                               "method",
                               "domain",
                               "kingdom",
                               "phylum",
                               "class",
                               "order"))
```

# Classify

Accept candidate hit groups according to a hierarchy of `accept_method`s
```{r}
# local species unanimous @ p_identical = 1
accept_species_100_unanimous_local = hits_geography %>%
  filter(p_identical == 1,
         !is.na(specificEpithet),
         local_species) %>%
  mutate(species = specificEpithet) %>%
  group_by(asv_id,
           asv_total_count,
           sample_count,
           samples,
           sequence,
           method,
           domain,
           kingdom,
           phylum,
           class,
           order,
           family,
           genus,
           species) %>%
  summarise(n_hits = n(),
            p_identical_mean = mean(percent_identical) / 100,
            p_identical_max = max(percent_identical) / 100,
            p_identical_min = min(percent_identical) / 100,
            seq_overlap_max = max(seq_overlap),
            gbif_count_family_local = first(gbif_count_family_local),
            gbif_count_genus_local = first(gbif_count_genus_local),
            gbif_count_species_local = first(gbif_count_species_local),
            local_species = any(local_species, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(accept_method = "0_species_100p_unanimous_local",
         taxonomic_rank = "species") %>%
  group_by(asv_id, method) %>%
  mutate(tie = n() > 1) %>%
  ungroup() %>%
  filter(!tie)

# species unanimous @ p_identical = 1
accept_species_100_unanimous = hits_geography %>%
  filter(p_identical == 1,
         !is.na(specificEpithet)) %>%
  mutate(species = specificEpithet) %>%
  group_by(asv_id,
           asv_total_count,
           sample_count,
           samples,
           sequence,
           method,
           domain,
           kingdom,
           phylum,
           class,
           order,
           family,
           genus,
           species) %>%
  summarise(n_hits = n(),
            p_identical_mean = mean(percent_identical) / 100,
            p_identical_max = max(percent_identical) / 100,
            p_identical_min = min(percent_identical) / 100,
            seq_overlap_max = max(seq_overlap),
            gbif_count_family_local = first(gbif_count_family_local),
            gbif_count_genus_local = first(gbif_count_genus_local),
            gbif_count_species_local = first(gbif_count_species_local),
            local_species = any(local_species, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(accept_method = "0_species_100p_unanimous",
         taxonomic_rank = "species") %>%
  group_by(asv_id, method) %>%
  mutate(tie = n() > 1) %>%
  ungroup() %>%
  filter(!tie)

# local 2-way tie @ p_identical = 1: exactly 2 equally-good candidate taxa from one search
# method; reported as "value_1/value_2" at every disagreeing rank, with taxonomic_rank set
# to the most specific rank both candidates actually resolve
accept_species_100_tie2_local = build_tie2_accept(
  hits_geography %>%
    filter(p_identical == 1,
           !is.na(specificEpithet),
           local_species) %>%
    mutate(species = specificEpithet),
  "0_species_100p_tie2_local"
)

# 2-way tie @ p_identical = 1 (see accept_species_100_tie2_local)
accept_species_100_tie2 = build_tie2_accept(
  hits_geography %>%
    filter(p_identical == 1,
           !is.na(specificEpithet)) %>%
    mutate(species = specificEpithet),
  "0_species_100p_tie2"
)

# local species consensus >=75% @ p_identity = 1
accept_species_100_consensus_75_local = find_consensus(hits_geography %>%
                                                filter(!is.na(order),
                                                       !is.na(specificEpithet),
                                                       local_species) %>%
                                                  mutate(species = specificEpithet), "species", identity_th = 1, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "1_species_100p_consensus_75_local") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order,
                     family,
                     genus,
                     species,
                     gbif_count_family_local,
                     gbif_count_genus_local,
                     gbif_count_species_local,
                     local_species) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method)

# species consensus >=75% @ p_identity = 1
accept_species_100_consensus_75 = find_consensus(hits_geography %>%
                                                filter(!is.na(order),
                                                       !is.na(specificEpithet)) %>%
                                                  mutate(species = specificEpithet), "species", identity_th = 1, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "1_species_100p_consensus_75") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order,
                     family,
                     genus,
                     species,
                     gbif_count_family_local,
                     gbif_count_genus_local,
                     gbif_count_species_local,
                     local_species) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"))

# local species unanimous @ p_identity >= .95, pmax_drop >= 0.03
accept_species_unambiguous_local = hits_species_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.95,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "2_species_95p_pdrop_03_local",
         species = specificEpithet,
         taxonomic_rank = "species")

# species unanimous @ p_identity >= .95, pmax_drop >= 0.03
accept_species_unambiguous = hits_species %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.95,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "2_species_95p_pdrop_03",
         species = specificEpithet,
         taxonomic_rank = "species")

# local 2-way tie among the top (>= 0.95) species candidates: exactly 2 equally-good species
# tied at the max p_identical_max for an (asv_id, method) -- rather than being dropped by the
# drop_to_next_pimax check above, they're merged into one row reporting "value_1/value_2" at
# every disagreeing rank (see accept_species_100_tie2_local for the analogous p_identical = 1 case)
tie2_species_unambiguous_local_keys = hits_species_local %>%
  group_by(asv_id, method) %>%
  filter(p_identical_max >= 0.95,
         p_identical_max == max(p_identical_max)) %>%
  ungroup() %>%
  mutate(species = specificEpithet) %>%
  add_count(asv_id, method, name = "n_tied") %>%
  filter(n_tied == 2)

accept_species_unambiguous_tie2_local = build_tie2_accept(
  hits_geography %>%
    filter(!(local_species %in% FALSE),
           !is.na(specificEpithet)) %>%
    mutate(species = specificEpithet) %>%
    semi_join(tie2_species_unambiguous_local_keys,
              by = c("asv_id", "method", "domain", "kingdom", "phylum", "class",
                     "order", "family", "genus", "species")),
  "2_species_95p_pdrop_03_tie2_local"
)

# local genus unanimous @ p_identity >= .90, pmax_drop >= 0.03
accept_genus_unambiguous_local = hits_genus_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.90,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "3_genus_90p_pdrop_03_local",
         taxonomic_rank = "genus")

# genus unanimous @ p_identity >= .90, pmax_drop >= 0.03
accept_genus_unambiguous = hits_genus %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.90,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "3_genus_90p_pdrop_03",
         taxonomic_rank = "genus")

# local family unanimous @ p_identity >= .80, pmax_drop >= 0.05
accept_family_unambiguous_local = hits_family_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.80,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "4_family_80p_pdrop_05_local",
         taxonomic_rank = "family")

# family unanimous @ p_identity >= .80, pmax_drop >= 0.05
accept_family_unambiguous = hits_family%>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.80,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "4_family_80p_pdrop_05",
         taxonomic_rank = "family")

# order unanimous @ p_identity >= .70, pmax_drop >= 0.05
accept_order_unambiguous = hits_order %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.70,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "5_order_70p_pdrop_05",
         taxonomic_rank = "order")

# order consensus >=75% @ p_identity >0.499
accept_order_50_consensus_75 = find_consensus(hits_geography,
                                           "order", identity_th = 0.499, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "6_order_50p_consensus_75") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order
                     ) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"))

# order consensus >=75% @ p_identity >0.499
accept_order_50_consensus_95 = find_consensus(hits_geography,
                                           "order", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "6_order_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"))

# class consensus >=95% @ p_identity >0.499
accept_class_50_consensus_95 = find_consensus(hits_geography,
                                           "class", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "7_class_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class"))

# phylum consensus >=95% @ p_identity >0.499
accept_phylum_50_consensus_95 = find_consensus(hits_geography,
                                           "phylum", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "8_phylum_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum"))

# kingdom consensus >=95% @ p_identity >0.499
accept_kingdom_50_consensus_95 = find_consensus(hits_geography,
                                           "kingdom", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "9_kingdom_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom"))

# domain consensus >=95% @ p_identity >0.499
accept_domain_50_consensus_95 = find_consensus(hits_geography,
                                           "domain", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "9_domain_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain) %>%
              distinct(), by = c("asv_id", "method", "domain"))
```

# Combine outputs

Join accepted classifications from all `accept_method`s
```{r}
accept_all = bind_rows(accept_species_100_unanimous,
                       accept_species_100_unanimous_local,
                       accept_species_100_tie2,
                       accept_species_100_tie2_local,
                       accept_species_100_consensus_75,
                       accept_species_100_consensus_75_local,
                       accept_species_unambiguous,
                       accept_species_unambiguous_local,
                       accept_species_unambiguous_tie2_local,
                       accept_genus_unambiguous,
                       accept_genus_unambiguous_local,
                       accept_family_unambiguous,
                       accept_family_unambiguous_local,
                       accept_order_unambiguous,
                       accept_order_50_consensus_95,
                       accept_class_50_consensus_95,
                       accept_phylum_50_consensus_95,
                       accept_kingdom_50_consensus_95,
                       accept_domain_50_consensus_95) %>%
  mutate(tie = if_else(is.na(tie), FALSE, tie)) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method, desc(n_hits), desc(p_identical_max)) %>%
  select(all_of(colnames(accept_species_100_unanimous)),
         everything(),
         -drop_to_next_pimax,
         -next_pimax,
         -pimax_rank)

# how many ASVs were classified?
asv_count = length(unique(accept_all$asv_id))

# which ASVs were not classified?
remainder = hits_geography %>%
  anti_join(accept_all, by = c("asv_id", "method"))

# collect hits and accepted classifications
hits_accept = hits_geography %>%
  bind_rows(accept_all) %>%
  bind_rows(hits_orphan_contaminant) %>%
  left_join(contaminant_summary, by = "asv_id") %>%
  mutate(species = if_else(!is.na(intraspecificEpithet), paste0(specificEpithet, " ", intraspecificEpithet), species),
         accept_method_priority = recode_values(accept_method,
                                                "0_species_100p_unanimous_local" ~ 1,
                                                "0_species_100p_unanimous" ~ 2,
                                                "0_species_100p_tie2_local" ~ 3,
                                                "0_species_100p_tie2" ~ 4,
                                                "1_species_100p_consensus_75_local" ~ 5,
                                                "1_species_100p_consensus_75" ~ 6,
                                                "2_species_95p_pdrop_03_local" ~ 7,
                                                "2_species_95p_pdrop_03" ~ 8,
                                                "2_species_95p_pdrop_03_tie2_local" ~ 9,
                                                "3_genus_90p_pdrop_03_local" ~ 10,
                                                "3_genus_90p_pdrop_03" ~ 11,
                                                "4_family_80p_pdrop_05_local" ~ 12,
                                                "4_family_80p_pdrop_05" ~ 13,
                                                "5_order_70p_pdrop_05" ~ 14,
                                                "6_order_50p_consensus_75" ~ 15,
                                                "7_class_50p_consensus_75" ~ 16,
                                                "8_phylum_50p_consensus_75" ~ 17,
                                                "9_kingdom_50p_consensus_75" ~ 18,
                                                "9_domain_50p_consensus_75" ~ 19,
                                                default = NA),
         search_method_priority = recode_values(method,
                                                "blast_web" ~ 1,
                                                "vsearch" ~ 2,
                                                "blast" ~ 3)) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method, desc(n_hits), desc(p_identical)) %>%
  select("asv_id", "asv_total_count", "sample_count", "samples", "sequence", 
         "method", "accession", "domain", "kingdom", "phylum", "class", 
         "order", "family", "genus", "species",
         # "specificEpithet", "intraspecificEpithet",
         "search_method_priority", "accept_method","accept_method_priority", "taxonomic_rank",
         "percent_identical", "seq_overlap",
         "seq_mismatch", "gapopen_count", "q_start", "q_end", "s_start", 
         "s_end", "p_identical", "p_overlap",  
         "gbif_count_family_local", "gbif_count_genus_local", "gbif_count_species_local", 
         "local_species", "n_hits", "p_identical_mean", "p_identical_max",
         "p_identical_min", "seq_overlap_max", "tie",
         "consensus_value", "n_hits_consensus", "identity_th", "consensus_th",
         "min_consensus_hits",
         "contaminant_flag", "contaminant_id", "contaminant_p_identity")

# write to file
write.csv(hits_accept, hybrid_classification_out, row.names = FALSE)

```

# Checks
```{r}
# list of all accept methods
all_accept_methods = hits_accept %>%
  select(accept_method) %>%
  distinct() %>%
  arrange(accept_method)

# Drop ASV classifications with overlap length below seq_overlap_cutoff
hits_accept_qc = hits_accept %>%
  filter(!(!is.na(accept_method) & seq_overlap_max < sequence_overlap_cutoff))

# species-level discrepancies
accepts_species = hits_accept_qc %>%
  filter(grepl("^0|^1|^2", accept_method)) %>%
  group_by(asv_id) %>%
  mutate(n_method = n_distinct(method),
         n_class = n_distinct(family, genus, species)) %>%
  ungroup()

## multiple classification between methods
accepts_species_conflicts = accepts_species %>%
  filter(n_class > 1)

## only classified by web blast
accepts_species_wbo = accepts_species %>%
  filter(n_method == 1,
         method == "blast_web")

## only classified by local blast
accepts_species_bo = accepts_species %>%
  filter(n_method == 1,
         method == "blast")

## only classified by vsearch
accepts_species_vo = accepts_species %>%
  filter(n_method == 1,
         method == "vsearch")

## counts by search method
accepts_species_counts = accepts_species %>%
  group_by(asv_id) %>%
  summarize(bool_wb = any(method == "blast_web"),
            bool_b = any(method == "blast"),
            bool_v = any(method == "vsearch"),
            .groups = "drop")

accepts_species_counts_summary = accepts_species_counts %>%
  summarize(spp_wb = sum(bool_wb),
            spp_b = sum(bool_b),
            spp_v = sum(bool_v))

## classifications where web blast returned nothing
accepts_species_not_wb = accepts_species_counts %>%
  filter((bool_b | bool_v) & !bool_wb) %>%
  left_join(hits_accept_qc, by = "asv_id")

# summarize by accept method
accepts_species_method = accepts_species %>%
  filter(method == "blast_web") %>%
  group_by(asv_id, method) %>%
  summarize(bool_0 = any(grepl("^0", accept_method)),
            bool_1 = any(grepl("^1", accept_method)),
            bool_2 = any(grepl("^2", accept_method)),
            .groups = "drop")

asm_summary = accepts_species_method %>%
  summarize(method_0 = sum(bool_0),
            method_1 = sum(bool_1),
            method_2 = sum(bool_2))

asm_0_not_1 = accepts_species_method %>%
  filter(bool_0 & !bool_1) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

asm_0_not_2 = accepts_species_method %>%
  filter(bool_0 & !bool_2) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

asm_1_not_2 = accepts_species_method %>%
  filter(bool_1 & !bool_2) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

# genus level
accepts_genus = hits_accept_qc %>%
  filter(grepl("^3", accept_method)) %>%
  group_by(asv_id) %>%
  mutate(n_method = n_distinct(method),
         n_class = n_distinct(family, genus)) %>%
  ungroup()

accepts_genus_conflicts = accepts_genus %>%
  filter(n_class > 1)

## cases where web blast could only classify to genus (and not species)
accepts_genus_first_wb = accepts_genus %>%
  filter(method == "blast_web") %>%
  anti_join(accepts_species, by = c("asv_id", "method")) %>%
  select(asv_id, method) %>%
  distinct() %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

```

# accepted classifications
```{r}
# filter to method of choice
# select the best classification according to hierarchy of `accept_method`s
max_search_method_priority = max(hits_accept_qc$search_method_priority, na.rm = TRUE)

accept_out = hits_accept_qc %>%
  mutate(priority = accept_method_priority + search_method_priority / max_search_method_priority) %>%
  # keep rows with an accepted taxonomy, plus taxonomically-unhit ASVs that are still
  # contaminant-flagged (see hits_orphan_contaminant above) so they aren't dropped entirely
  filter(!is.na(accept_method) | contaminant_flag) %>%
  group_by(asv_id) %>%
  slice_min(priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(asv_id,
         method,
         accept_method,
         taxonomic_rank,
         domain,
         kingdom,
         phylum,
         class,
         order,
         family,
         genus,
         species,
         n_hits,
         p_identical_max,
         seq_overlap_max,
         consensus_value,
         gbif_count_family_local,
         gbif_count_genus_local,
         gbif_count_species_local,
         contaminant_flag,
         contaminant_id,
         contaminant_p_identity)

# export classifications
write.csv(accept_out, classification_out, row.names = FALSE)
```