---
title: "Classify taxonomic hits"
author: "Cob Staines"
output: html_document
---

# Description

This script takes in hits from any search method (e.g. vsearch, blast, web blast), and optionally gbif taxonomic occurance data, and assigns a likely taxonomy to each ASV following a hierarchy `accept_method`s. All assigned taxonomy from all methods, along with all hits, are exported to `hybrid_classification_out`. The single best classifications by the most specific `accept_method` for each ASV are exported to `classification_out`.

See the algorithm below for details on each `accept_method` and to adjust parameters.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here)

# import run config file
run_dir = here("16S_sequence_processing", "runs", "methods_2026-03-13", "16S")

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# minimum sequence overlap (0 -> no filtering) [units of Bp]
sequence_overlap_cutoff = 0

# Did you run a gbif query (script 05) for local taxa, which you want incorporated in classification?
gbif_query = TRUE

# output files
hybrid_classification_out = here(run_dir, "output", paste0(config$run$name, "_06_hybrid_classification.csv"))

classification_out = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))


```

# Prepare for classification
```{r}
# read in hits from qiime2 (blast & vsearch)
qiime2_hits_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_02_qiime2_hits.csv")))

# read in hits from web blast (script 04)
web_blast_hits_raw = read_csv(here(run_dir, "output", paste0(config$run$name, "_04_ncbi_web_blast_hits.csv"))) %>%
  mutate(species = sub("^\\S+\\s+", "", species))

# pull in asv data
backbone = qiime2_hits_raw %>%
  select(asv_id,
         asv_total_count,
         sample_count,
         samples,
         sequence) %>%
  distinct()

# bind with web blast
hits_raw = bind_rows(qiime2_hits_raw,
                     backbone %>%
                     left_join(web_blast_hits_raw, by = "asv_id") %>%
                       select(all_of(colnames(qiime2_hits_raw))))

# clean hits
hits_clean = hits_raw %>%
  filter(!is.na(class),
         seq_overlap >= 50) %>%
  mutate(p_identical = percent_identical / 100,
         p_overlap = seq_overlap / (s_end - s_start + 1))

# handle gbif geographic query
if (gbif_query) {
  # read in taxa + geography
  taxa_geography = read.csv(here(run_dir, "output", paste0(config$run$name, "_05_taxonomy_table_gbif_occ.csv")))
  
  hits_geography = hits_clean %>%
    left_join(taxa_geography %>%
                select(accession,
                       any_of(c("specificEpithet",
                                "intraspecificEpithet",
                                "gbif_count_family_local",
                                "gbif_count_genus_local",
                                "gbif_count_species_local"))) %>%
                distinct(), by = "accession") %>%
    mutate(specificEpithet = coalesce(specificEpithet, species), 
           local_species = case_when(
             gbif_count_species_local > 0 ~ TRUE,
             gbif_count_species_local == 0 ~ FALSE,
             gbif_count_genus_local == 0 ~ FALSE,
             gbif_count_family_local == 0 ~ FALSE,
             .default = NA))
  
} else {
  # set gbif data to NA
  hits_geography = hits_clean %>%
    mutate(specificEpithet = species,
           intraspecificEpithet = NA,
           gbif_count_family_local = NA,
           gbif_count_genus_local = NA,
           gbif_count_species_local = NA,
           local_species = NA)
}


# define classification functions

## consensus function at a specified taxonomic rank, identity threshold, and consensus threshold
find_consensus = function(hit_table, taxonomic_rank, identity_th = 0, consensus_th = 0.5, min_consensus_hits = 1, tie_handling = "error") {
  
  if (!(tie_handling %in% c("error", "pass", "drop"))) {
    stop('Invalid value for tie_handling. Must be in c("error", "pass", "drop")')
  }
  
  hit_table_filtered = hit_table %>%
    filter(p_identical >= identity_th)
  
  if (consensus_th <= 0.5) {
    warning("Using a consensus_th <= 0.5 may result in multiple consensus classification. Consensus classification(s) will be selected with the greatest consensus value, with ties handled according to tie_handling ('error', 'pass', 'drop')")
  }
  
  if (!(taxonomic_rank %in% c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"))) {
    stop('taxonnomic_rank not recognized, must be one of c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")')
  }
  
  taxon_groups_list = list(
    species = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"),
    genus = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus"),
    family = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family"),
    order = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"),
    class = c("asv_id", "method", "domain", "kingdom", "phylum", "class"),
    phylum = c("asv_id", "method", "domain", "kingdom", "phylum"),
    kingdom = c("asv_id", "method", "domain", "kingdom"),
    domain = c("asv_id", "method", "domain")
  )
  
  taxon_group = taxon_groups_list[[taxonomic_rank]]
  
  consensus_table = hit_table_filtered %>%
    group_by(asv_id, method) %>%
    mutate(asv_hits_n = n()) %>%
    ungroup() %>%
    group_by_at(taxon_group) %>%
    mutate(taxon_hits_n = n()) %>%
    ungroup() %>%
    mutate(consensus_value = taxon_hits_n / asv_hits_n) %>%
    filter(taxon_hits_n >= min_consensus_hits,
           consensus_value >= consensus_th) %>%
    arrange(desc(asv_total_count),
            asv_id,
            method,
            desc(consensus_value),
            desc(p_identical)) %>%
    group_by_at(c(taxon_group,
                  "consensus_value")) %>%
    summarise(n_hits = first(taxon_hits_n),
              p_identical_min = mean(p_identical),
              p_identical_max = mean(p_identical),
              p_identical_mean = mean(p_identical),
              seq_overlap_max = max(seq_overlap),
              .groups = "drop") %>%
    group_by(asv_id, method) %>%
    slice_max(order_by = consensus_value, n = 1, with_ties = TRUE) %>%
    mutate(n_hits_consensus = n()) %>%
    ungroup() %>%
    mutate(identity_th = identity_th,
           consensus_th = consensus_th,
           min_consensus_hits = min_consensus_hits,
           taxonomic_rank = taxonomic_rank)
  
  tie_ids = consensus_table %>%
    filter(n_hits_consensus > 1) %>%
    select(asv_id) %>%
    distinct() %>%
    pull(asv_id)
  
  if (length(tie_ids) > 0) {
    if (tie_handling == "error") {
      stop(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nAborting."))
    } else if (tie_handling == "pass") {
      warning(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nPassing to output."))
    } else if (tie_handling == "drop") {
      warning(paste0("Ties found for the following asv_ids:\n\t", paste(tie_ids, collapse = "\n\t"), "\nDropping from output."))
      consensus_table = consensus_table %>%
        filter(consensus <= 1)
    }
  }
  
  return(consensus_table)
}

# calculate stats for taxonomic hit groups
hits_group_calc = function(hits_df, group_cols, agg_cols = c()) {
  hits_df %>%
    group_by_at(group_cols) %>%
    summarize(
      n_hits = n(),
      p_identical_mean = mean(percent_identical) / 100,
      p_identical_max = max(percent_identical) / 100,
      p_identical_min = min(percent_identical) / 100,
      seq_overlap_max = max(seq_overlap),
      across(all_of(agg_cols), \(x) max(x, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    arrange(desc(asv_total_count), asv_id, method, desc(p_identical_max)) %>%
    group_by(asv_id, method) %>%
    mutate(next_pimax = lead(p_identical_max),
           drop_to_next_pimax = p_identical_max - next_pimax,
           pimax_rank = row_number()) %>%
    ungroup() %>%
    select(all_of(group_cols),
           n_hits,
           p_identical_max,
           p_identical_mean,
           p_identical_min,
           drop_to_next_pimax,
           everything())
}
```

# Group hits by taxonomy
```{r warning=FALSE}
# collect hits of common taxonomies into one candidate row, with descriptive stats

hits_species_local = hits_group_calc(hits_geography %>%
                                       filter(!(local_species %in% FALSE),
                                              !is.na(specificEpithet)),
                                     c("asv_id",
                                       "asv_total_count",
                                       "sample_count",
                                       "samples",
                                       "sequence",
                                       "method",
                                       "domain",
                                       "kingdom",
                                       "phylum",
                                       "class",
                                       "order",
                                       "family",
                                       "genus",
                                       "specificEpithet"),
                                     c("gbif_count_family_local",
                                       "gbif_count_genus_local",
                                       "gbif_count_species_local",
                                       "local_species"))

hits_species = hits_group_calc(hits_geography %>%
                                 filter(!is.na(specificEpithet)),
                               c("asv_id",
                                 "asv_total_count",
                                 "sample_count",
                                 "samples",
                                 "sequence",
                                 "method",
                                 "domain",
                                 "kingdom",
                                 "phylum",
                                 "class",
                                 "order",
                                 "family",
                                 "genus",
                                 "specificEpithet"),
                               c("gbif_count_family_local",
                                 "gbif_count_genus_local",
                                 "gbif_count_species_local",
                                 "local_species"))

hits_genus_local = hits_group_calc(hits_geography %>%
                                     filter(gbif_count_genus_local > 20,
                                            !is.na(genus)),
                                   c("asv_id",
                                     "asv_total_count",
                                     "sample_count",
                                     "samples",
                                     "sequence",
                                     "method",
                                     "domain",
                                     "kingdom",
                                     "phylum",
                                     "class",
                                     "order",
                                     "family",
                                     "genus",
                                     "gbif_count_family_local",
                                     "gbif_count_genus_local"))

hits_genus = hits_group_calc(hits_geography %>%
                               filter(!is.na(genus)),
                             c("asv_id",
                               "asv_total_count",
                               "sample_count",
                               "samples",
                               "sequence",
                               "method",
                               "domain",
                               "kingdom",
                               "phylum",
                               "class",
                               "order",
                               "family",
                               "genus",
                               "gbif_count_family_local",
                               "gbif_count_genus_local"))

hits_family_local = hits_group_calc(hits_geography %>%
                                      filter(gbif_count_family_local > 50,
                                             !is.na(family)),
                                    c("asv_id",
                                      "asv_total_count",
                                      "sample_count",
                                      "samples",
                                      "sequence",
                                      "method",
                                      "domain",
                                      "kingdom",
                                      "phylum",
                                      "class",
                                      "order",
                                      "family",
                                      "gbif_count_family_local"))

hits_family = hits_group_calc(hits_geography %>%
                                filter(!is.na(family)),
                              c("asv_id",
                                "asv_total_count",
                                "sample_count",
                                "samples",
                                "sequence",
                                "method",
                                "domain",
                                "kingdom",
                                "phylum",
                                "class",
                                "order",
                                "family",
                                "gbif_count_family_local"))

hits_order = hits_group_calc(hits_geography %>%
                               filter(!is.na(order)),
                             c("asv_id",
                               "asv_total_count",
                               "sample_count",
                               "samples",
                               "sequence",
                               "method",
                               "domain",
                               "kingdom",
                               "phylum",
                               "class",
                               "order"))
```

# Classify

Accept candidate hit groups according to a hierarchy of `accept_method`s
```{r}
# local species unanimous @ p_identical = 1
accept_species_100_unanimous_local = hits_geography %>%
  filter(p_identical == 1,
         !is.na(specificEpithet),
         local_species) %>%
  mutate(species = specificEpithet) %>%
  group_by(asv_id,
           asv_total_count,
           sample_count,
           samples,
           sequence,
           method,
           domain,
           kingdom,
           phylum,
           class,
           order,
           family,
           genus,
           species) %>%
  summarise(n_hits = n(),
            p_identical_mean = mean(percent_identical) / 100,
            p_identical_max = max(percent_identical) / 100,
            p_identical_min = min(percent_identical) / 100,
            seq_overlap_max = max(seq_overlap),
            gbif_count_family_local = first(gbif_count_family_local),
            gbif_count_genus_local = first(gbif_count_genus_local),
            gbif_count_species_local = first(gbif_count_species_local),
            local_species = any(local_species, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(accept_method = "0_species_100p_unanimous_local",
         taxonomic_rank = "species") %>%
  group_by(asv_id) %>%
  mutate(tie = n() > 1) %>%
  ungroup() %>%
  filter(!tie)

# species unanimous @ p_identical = 1
accept_species_100_unanimous = hits_geography %>%
  filter(p_identical == 1,
         !is.na(specificEpithet)) %>%
  mutate(species = specificEpithet) %>%
  group_by(asv_id,
           asv_total_count,
           sample_count,
           samples,
           sequence,
           method,
           domain,
           kingdom,
           phylum,
           class,
           order,
           family,
           genus,
           species) %>%
  summarise(n_hits = n(),
            p_identical_mean = mean(percent_identical) / 100,
            p_identical_max = max(percent_identical) / 100,
            p_identical_min = min(percent_identical) / 100,
            seq_overlap_max = max(seq_overlap),
            gbif_count_family_local = first(gbif_count_family_local),
            gbif_count_genus_local = first(gbif_count_genus_local),
            gbif_count_species_local = first(gbif_count_species_local),
            local_species = any(local_species, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(accept_method = "0_species_100p_unanimous",
         taxonomic_rank = "species") %>%
  group_by(asv_id) %>%
  mutate(tie = n() > 1) %>%
  ungroup() %>%
  filter(!tie)

# local species consensus >=75% @ p_identity = 1
accept_species_100_consensus_75_local = find_consensus(hits_geography %>%
                                                filter(!is.na(order),
                                                       !is.na(specificEpithet),
                                                       local_species) %>%
                                                  mutate(species = specificEpithet), "species", identity_th = 1, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "1_species_100p_consensus_75_local") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order,
                     family,
                     genus,
                     species,
                     gbif_count_family_local,
                     gbif_count_genus_local,
                     gbif_count_species_local,
                     local_species) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method)

# species consensus >=75% @ p_identity = 1
accept_species_100_consensus_75 = find_consensus(hits_geography %>%
                                                filter(!is.na(order),
                                                       !is.na(specificEpithet)) %>%
                                                  mutate(species = specificEpithet), "species", identity_th = 1, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "1_species_100p_consensus_75") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order,
                     family,
                     genus,
                     species,
                     gbif_count_family_local,
                     gbif_count_genus_local,
                     gbif_count_species_local,
                     local_species) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order", "family", "genus", "species"))

# local species unanimous @ p_identity >= .95, pmax_drop >= 0.03
accept_species_unambiguous_local = hits_species_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.95,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "2_species_95p_pdrop_03_local",
         species = specificEpithet,
         taxonomic_rank = "species")

# species unanimous @ p_identity >= .95, pmax_drop >= 0.03
accept_species_unambiguous = hits_species %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.95,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "2_species_95p_pdrop_03",
         species = specificEpithet,
         taxonomic_rank = "species")

# local genus unanimous @ p_identity >= .90, pmax_drop >= 0.03
accept_genus_unambiguous_local = hits_genus_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.90,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "3_genus_90p_pdrop_03_local",
         taxonomic_rank = "genus")

# genus unanimous @ p_identity >= .90, pmax_drop >= 0.03
accept_genus_unambiguous = hits_genus %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.90,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.03) %>%
  mutate(accept_method = "3_genus_90p_pdrop_03",
         taxonomic_rank = "genus")

# local family unanimous @ p_identity >= .80, pmax_drop >= 0.05
accept_family_unambiguous_local = hits_family_local %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.80,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "4_family_80p_pdrop_05_local",
         taxonomic_rank = "family")

# family unanimous @ p_identity >= .80, pmax_drop >= 0.05
accept_family_unambiguous = hits_family%>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.80,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "4_family_80p_pdrop_05",
         taxonomic_rank = "family")

# order unanimous @ p_identity >= .70, pmax_drop >= 0.05
accept_order_unambiguous = hits_order %>%
  filter(pimax_rank == 1,
         p_identical_max >= 0.70,
         is.na(drop_to_next_pimax) | drop_to_next_pimax >= 0.05) %>%
  mutate(accept_method = "5_order_70p_pdrop_05",
         taxonomic_rank = "order")

# order consensus >=75% @ p_identity >0.499
accept_order_50_consensus_75 = find_consensus(hits_geography,
                                           "order", identity_th = 0.499, consensus_th = 0.75, tie_handling = "drop") %>%
  mutate(accept_method = "6_order_50p_consensus_75") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order
                     ) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"))

# order consensus >=75% @ p_identity >0.499
accept_order_50_consensus_95 = find_consensus(hits_geography,
                                           "order", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "6_order_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class,
                     order) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class", "order"))

# class consensus >=95% @ p_identity >0.499
accept_class_50_consensus_95 = find_consensus(hits_geography,
                                           "class", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "7_class_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum,
                     class) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum", "class"))

# phylum consensus >=95% @ p_identity >0.499
accept_phylum_50_consensus_95 = find_consensus(hits_geography,
                                           "phylum", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "8_phylum_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom,
                     phylum) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom", "phylum"))

# kingdom consensus >=95% @ p_identity >0.499
accept_kingdom_50_consensus_95 = find_consensus(hits_geography,
                                           "kingdom", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "9_kingdom_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain,
                     kingdom) %>%
              distinct(), by = c("asv_id", "method", "domain", "kingdom"))

# domain consensus >=95% @ p_identity >0.499
accept_domain_50_consensus_95 = find_consensus(hits_geography,
                                           "domain", identity_th = 0.499, consensus_th = 0.95, tie_handling = "drop") %>%
  mutate(accept_method = "9_domain_50p_consensus_95") %>%
  left_join(hits_geography %>%
              select(asv_id,
                     asv_total_count,
                     sample_count,
                     samples,
                     sequence,
                     method,
                     domain) %>%
              distinct(), by = c("asv_id", "method", "domain"))
```

# Combine outputs

Join accepted classifications from all `accept_method`s
```{r}
accept_all = bind_rows(accept_species_100_unanimous,
                       accept_species_100_unanimous_local,
                       accept_species_100_consensus_75,
                       accept_species_100_consensus_75_local,
                       accept_species_unambiguous,
                       accept_species_unambiguous_local,
                       accept_genus_unambiguous,
                       accept_genus_unambiguous_local,
                       accept_family_unambiguous,
                       accept_family_unambiguous_local,
                       accept_order_unambiguous,
                       accept_order_50_consensus_95,
                       accept_class_50_consensus_95,
                       accept_phylum_50_consensus_95,
                       accept_kingdom_50_consensus_95,
                       accept_domain_50_consensus_95) %>%
  mutate(tie = if_else(is.na(tie), FALSE, tie)) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method, desc(n_hits), desc(p_identical_max)) %>%
  select(all_of(colnames(accept_species_100_unanimous)),
         everything(),
         -drop_to_next_pimax,
         -next_pimax,
         -pimax_rank)

# how many ASVs were classified?
asv_count = length(unique(accept_all$asv_id))

# which ASVs were not classified?
remainder = hits_geography %>%
  anti_join(accept_all, by = c("asv_id", "method"))

# list of all accept methods
all_accept_methods = hits_accept %>%
  select(accept_method) %>%
  distinct() %>%
  arrange(accept_method)

# collect hits and accepted classifications
hits_accept = hits_geography %>%
  bind_rows(accept_all) %>%
  mutate(species = if_else(!is.na(intraspecificEpithet), paste0(specificEpithet, " ", intraspecificEpithet), species),
         accept_method_priority = recode_values(accept_method,
                                                "0_species_100p_unanimous_local" ~ 1,
                                                "0_species_100p_unanimous" ~ 2,
                                                "1_species_100p_consensus_75_local" ~ 3, 
                                                "1_species_100p_consensus_75" ~ 4,
                                                "2_species_95p_pdrop_03_local" ~ 5, 
                                                "2_species_95p_pdrop_03" ~ 6,
                                                "3_genus_90p_pdrop_03_local" ~ 7,
                                                "3_genus_90p_pdrop_03" ~ 8, 
                                                "4_family_80p_pdrop_05_local" ~ 9,
                                                "4_family_80p_pdrop_05" ~ 10,
                                                "5_order_70p_pdrop_05" ~ 11, 
                                                "6_order_50p_consensus_75" ~ 12,
                                                "7_class_50p_consensus_75" ~ 13,
                                                "8_phylum_50p_consensus_75" ~ 14,
                                                "9_kingdom_50p_consensus_75" ~ 15,
                                                "9_domain_50p_consensus_75" ~ 16,
                                                default = NA)) %>%
  arrange(desc(asv_total_count), asv_id, method, accept_method, desc(n_hits), desc(p_identical)) %>%
  select("asv_id", "asv_total_count", "sample_count", "samples", "sequence", 
         "method", "accession", "domain", "kingdom", "phylum", "class", 
         "order", "family", "genus", "species",
         # "specificEpithet", "intraspecificEpithet",
         "accept_method","accept_method_priority", "taxonomic_rank",
         "percent_identical", "seq_overlap",
         "seq_mismatch", "gapopen_count", "q_start", "q_end", "s_start", 
         "s_end", "p_identical", "p_overlap",  
         "gbif_count_family_local", "gbif_count_genus_local", "gbif_count_species_local", 
         "local_species", "n_hits", "p_identical_mean", "p_identical_max", 
         "p_identical_min", "seq_overlap_max", "tie", 
         "consensus_value", "n_hits_consensus", "identity_th", "consensus_th", 
         "min_consensus_hits")

# write to file
write.csv(hits_accept, hybrid_classification_out, row.names = FALSE)

```

# Checks
```{r}
# Drop ASV classifications with overlap length below seq_overlap_cutoff
hits_accept_qc = hits_accept %>%
  filter(!(!is.na(accept_method) & seq_overlap_max < seq_overlap_cutoff))

# species-level discrepancies
accepts_species = hits_accept_qc %>%
  filter(grepl("^0|^1|^2", accept_method)) %>%
  group_by(asv_id) %>%
  mutate(n_method = n_distinct(method),
         n_class = n_distinct(family, genus, species)) %>%
  ungroup()

## multiple classification between methods
accepts_species_conflicts = accepts_species %>%
  filter(n_class > 1)

## only classified by web blast
accepts_species_wbo = accepts_species %>%
  filter(n_method == 1,
         method == "blast_web")

## only classified by local blast
accepts_species_bo = accepts_species %>%
  filter(n_method == 1,
         method == "blast")

## only classified by vsearch
accepts_species_vo = accepts_species %>%
  filter(n_method == 1,
         method == "vsearch")

## counts by search method
accepts_species_counts = accepts_species %>%
  group_by(asv_id) %>%
  summarize(bool_wb = any(method == "blast_web"),
            bool_b = any(method == "blast"),
            bool_v = any(method == "vsearch"),
            .groups = "drop")

accepts_species_counts_summary = accepts_species_counts %>%
  summarize(spp_wb = sum(bool_wb),
            spp_b = sum(bool_b),
            spp_v = sum(bool_v))

## classifications where web blast returned nothing
accepts_species_not_wb = accepts_species_counts %>%
  filter((bool_b | bool_v) & !bool_wb) %>%
  left_join(hits_accept_qc, by = "asv_id")

# summarize by accept method
accepts_species_method = accepts_species %>%
  filter(method == "blast_web") %>%
  group_by(asv_id, method) %>%
  summarize(bool_0 = any(grepl("^0", accept_method)),
            bool_1 = any(grepl("^1", accept_method)),
            bool_2 = any(grepl("^2", accept_method)),
            .groups = "drop")

asm_summary = accepts_species_method %>%
  summarize(method_0 = sum(bool_0),
            method_1 = sum(bool_1),
            method_2 = sum(bool_2))

asm_0_not_1 = accepts_species_method %>%
  filter(bool_0 & !bool_1) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

asm_0_not_2 = accepts_species_method %>%
  filter(bool_0 & !bool_2) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

asm_1_not_2 = accepts_species_method %>%
  filter(bool_1 & !bool_2) %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

# genus level
accepts_genus = hits_accept_qc %>%
  filter(grepl("^3", accept_method)) %>%
  group_by(asv_id) %>%
  mutate(n_method = n_distinct(method),
         n_class = n_distinct(family, genus)) %>%
  ungroup()

accepts_genus_conflicts = accepts_genus %>%
  filter(n_class > 1)

## cases where web blast could only classify to genus (and not species)
accepts_genus_first_wb = accepts_genus %>%
  filter(method == "blast_web") %>%
  anti_join(accepts_species, by = c("asv_id", "method")) %>%
  select(asv_id, method) %>%
  distinct() %>%
  left_join(hits_accept_qc, by = c("asv_id", "method"))

```

# accepted classifications
```{r}
# filter to method of choice
# select the best classification according to hierarchy of `accept_method`s
accept_out = hits_accept_qc %>%
  filter(method == "blast_web",
         !is.na(accept_method)) %>%
  group_by(asv_id) %>%
  slice_min(accept_method_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(asv_id,
         method,
         accept_method,
         taxonomic_rank,
         domain,
         kingdom,
         phylum,
         class,
         order,
         family,
         genus,
         species,
         n_hits,
         p_identical_max,
         seq_overlap_max,
         consensus_value)

# export classifications
write.csv(accept_out, classification_out, row.names = FALSE)
```