---
title: "Classify ASV locality"
author: "Cob Staines"
output: html_document
---

# Description

This script assesses, for each study system an ASV was detected in, whether its classified taxonomy (script 06) is expected locally -- i.e. whether a detection looks like a real local occurrence or like contamination (cross-system spillover, lab, or anthropogenic DNA).

This script **only flags, it drops nothing**. Filtering reads on the basis of these flags is left to script 08. Output is one row per (asv_id, study_system) for every system whose field samples (field negatives included) contain the ASV.

Two independent lines of evidence are reported side by side:

**Locality (GBIF)** -- occurrence counts for the accepted taxon at its accepted rank, in each system (script 05). `local_status` is one of (in order of precedence):

- `anthropogenic`: taxon listed in `anthropogenic_taxa` (config), regardless of GBIF
- `local`: accepted taxon recorded in this system (or listed in `local_overrides` for this system)
- `other_system`: not recorded in this system, but recorded in another system in the run -- a likely cross-system contaminant
- `local_genus_only`: species not recorded in this system (or not resolved by GBIF), but its genus is -- a GBIF gap, split, or cryptic species?
- `not_in_run`: not recorded in any system in the run -- lab/reagent contamination, or a GBIF gap
- `unassessable`: no classification at species/genus/family rank, not Metazoa, or no GBIF data

For 2-way ties from script 06 (`A/B`), each candidate is assessed and the most local status is kept.

**Reads** -- how the ASV is distributed across samples and systems. Raw read share is biased by sequencing effort (systems have very different sample counts), so evidence is based on within-sample relative abundance (ASV reads / sample reads):

- `detection_rate`: fraction of the system's samples in which the ASV was detected (per sample, since detection varies substantially between field replicates; `n_sites_detected` is reported alongside)
- `rel_abund_max`, `rel_abund_median`: within-sample relative abundance across detected samples in the system. Cross-contamination / tag-jumping tends to appear at trace relative abundance.
- `rel_abund_mean`: mean relative abundance across *all* of the system's samples (zeros included), used to compare systems: `norm_share` (share of the ASV's summed `rel_abund_mean` across systems), `dominant_system`, and `log10_ratio_to_dominant` (0 in the dominant system).

Field negatives are pooled with field samples (their DNA is field-derived), and also reported separately. Lab controls (PCR/extraction controls and the `lab_control` pseudo-system) are excluded from the evidence and reported as descriptive columns only.

`review_case` tags the ambiguous rows, where the two lines of evidence disagree (see plots below), using the reference thresholds `locality_min_rel_abund` and `locality_log_ratio_cutoff` from the config:

- `other_system_abundant`: GBIF says the taxon belongs to another system, but reads look real here (non-trace, and not dwarfed by the dominant system)
- `local_trace`: GBIF says local, but the detection looks like spillover from another dominant system (trace, and far below the dominant system)

# Config
```{r}
librarian::shelf(tidyverse, yaml, here, plotly)

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R); this is read
# directly so the same values apply whether this script is run manually or via the driver
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

# import run config file
run_dir = here(pipeline_config$run_dir)
cat(paste0("Run_dir: ", run_dir, "\n"))

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# reference thresholds (plot lines and review_case tags only -- nothing is filtered here)
locality_min_rel_abund = pipeline_config$locality_min_rel_abund
locality_log_ratio_cutoff = pipeline_config$locality_log_ratio_cutoff

# manual status overrides ("Genus species", "Genus", or "Family")
anthropogenic_taxa = unlist(pipeline_config$anthropogenic_taxa)
local_overrides = pipeline_config$local_overrides

# synonymous species names, applied the same way as in script 06 so accepted names match GBIF lookups
species_synonyms = pipeline_config$species_synonyms

apply_species_synonyms = function(x, synonym_list) {
  for (canonical in names(synonym_list)) {
    x[x %in% synonym_list[[canonical]]] = canonical
  }
  x
}

# max ignoring NA, NA (not -Inf) when all values are NA
max_or_na = function(x) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)

possible_study_systems = c("brazil", "panama", "pennsylvania", "sierra_nevada")

tax_ranks = c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")

# precedence when combining tie candidates (first = kept)
status_levels = c("anthropogenic", "local", "other_system", "local_genus_only", "not_in_run", "unassessable")

# input files
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))
gbif_file = here(run_dir, "output", paste0(config$run$name, "_05_taxonomy_table_gbif_occ.csv"))
classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))

# output file
locality_out = here(run_dir, "output", paste0(config$run$name, "_07_asv_locality.csv"))
```

# Load data
```{r}
sample_map = read_csv(sample_map_file, show_col_types = FALSE)

asv_classification = read_csv(classification_file, show_col_types = FALSE)

feature_table =
  read_tsv(here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "sample_table/feature-table.tsv"),
           skip = 1, show_col_types = FALSE) %>%
  as.data.frame() %>%
  rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

# per-system GBIF occurrence counts (script 05), if run
if (file.exists(gbif_file)) {
  taxa_geography = read_csv(gbif_file, show_col_types = FALSE) %>%
    filter(!is.na(accession)) %>%
    mutate(specificEpithet = apply_species_synonyms(specificEpithet, species_synonyms))
} else {
  cat("No GBIF occurrence table found (script 05), all ASVs will be unassessable by GBIF.\n")
  taxa_geography = tibble(study_system = character(), family = character(), genus = character(),
                          specificEpithet = character(), gbif_count_family_local = numeric(),
                          gbif_count_genus_local = numeric(), gbif_count_species_local = numeric())
}
```

# Clean data
```{r}
# classify each sample: evidence pool (field samples + field negatives, per system) vs lab controls
sample_class = sample_map %>%
  transmute(feature_id,
            study_system,
            site,
            sample_class = case_when(
              pcr_control %in% TRUE | extraction_negative_control %in% TRUE | study_system %in% "lab_control" ~ "lab_control",
              !(study_system %in% possible_study_systems) ~ "unknown",
              field_negative_control %in% TRUE ~ "field_negative",
              .default = "field_sample"))

# within-sample relative abundance (samples with zero reads are dropped)
feature_long = feature_table %>%
  pivot_longer(cols = -asv_id,
               names_to = "feature_id",
               values_to = "reads") %>%
  group_by(feature_id) %>%
  mutate(sample_reads = sum(reads)) %>%
  ungroup() %>%
  filter(sample_reads > 0) %>%
  mutate(rel_abund = reads / sample_reads) %>%
  left_join(sample_class, by = "feature_id")

unmapped_samples = feature_long %>%
  filter(is.na(sample_class) | sample_class == "unknown") %>%
  distinct(feature_id) %>%
  pull(feature_id)

if (length(unmapped_samples) > 0) {
  cat("NOTE: excluding", length(unmapped_samples), "sample(s) with no known study system:\n\t",
      paste(unmapped_samples, collapse = "\n\t"), "\n")
}

feature_pool = feature_long %>%
  filter(sample_class %in% c("field_sample", "field_negative"))

# samples/sites per system (denominators)
system_sizes = feature_pool %>%
  distinct(study_system, feature_id, site) %>%
  group_by(study_system) %>%
  summarise(n_samples = n_distinct(feature_id),
            n_sites = n_distinct(site, na.rm = TRUE),
            .groups = "drop")

print(system_sizes)

# per (asv_id, study_system) evidence
read_evidence = feature_pool %>%
  filter(reads > 0) %>%
  group_by(asv_id, study_system) %>%
  summarise(reads = sum(reads),
            n_samples_detected = n(),
            n_sites_detected = n_distinct(site, na.rm = TRUE),
            rel_abund_max = max(rel_abund),
            rel_abund_median = median(rel_abund),
            rel_abund_sum = sum(rel_abund),
            n_field_neg_detected = sum(sample_class == "field_negative"),
            field_neg_rel_abund_max = max_or_na(rel_abund[sample_class == "field_negative"]),
            .groups = "drop") %>%
  left_join(system_sizes, by = "study_system") %>%
  mutate(detection_rate = n_samples_detected / n_samples,
         rel_abund_mean = rel_abund_sum / n_samples) %>%
  select(-rel_abund_sum) %>%
  group_by(asv_id) %>%
  mutate(n_systems_detected = n(),
         systems_detected = paste(sort(study_system), collapse = ";"),
         norm_share = rel_abund_mean / sum(rel_abund_mean),
         dominant_system = study_system[which.max(rel_abund_mean)],
         log10_ratio_to_dominant = log10(rel_abund_mean / max(rel_abund_mean))) %>%
  ungroup()

# lab controls, run-wide per ASV (descriptive only)
lab_control_evidence = feature_long %>%
  filter(sample_class == "lab_control",
         reads > 0) %>%
  group_by(asv_id) %>%
  summarise(lab_control_n_detected = n(),
            lab_control_max_reads = max(reads),
            lab_control_rel_abund_max = max(rel_abund),
            .groups = "drop")
```

# GBIF locality
```{r}
# accepted taxonomy per ASV, one row per candidate (2-way ties "A/B" from script 06 are split
# into their two candidates; non-tied ASVs have a single candidate)
split_candidate = function(x, i) {
  na_if(if_else(str_detect(x, "/"), str_split_i(x, "/", i), x), "NA")
}

classification_candidates = map_dfr(1:2, function(i) {
  asv_classification %>%
    select(asv_id, accept_method, taxonomic_rank, all_of(tax_ranks)) %>%
    mutate(across(all_of(tax_ranks), \(x) split_candidate(x, i)),
           candidate = i)
}) %>%
  distinct(across(-candidate), .keep_all = TRUE) %>%
  # accepted species may carry an intraspecific epithet; GBIF lookups are by specific epithet
  mutate(specificEpithet = word(species, 1))

# per-system GBIF counts for each rank (NA = name not resolved by GBIF)
gbif_species = taxa_geography %>%
  filter(!is.na(specificEpithet), !is.na(gbif_count_species_local)) %>%
  group_by(study_system, family, genus, specificEpithet) %>%
  summarise(gbif_count_species = max(gbif_count_species_local), .groups = "drop")

gbif_genus = taxa_geography %>%
  filter(!is.na(genus), !is.na(gbif_count_genus_local)) %>%
  group_by(study_system, family, genus) %>%
  summarise(gbif_count_genus = max(gbif_count_genus_local), .groups = "drop")

gbif_family = taxa_geography %>%
  filter(!is.na(family), !is.na(gbif_count_family_local)) %>%
  group_by(study_system, family) %>%
  summarise(gbif_count_family = max(gbif_count_family_local), .groups = "drop")

gbif_systems = sort(unique(taxa_geography$study_system))

# every candidate x every system with GBIF data, with counts in that system and in the run's
# other systems at each rank
candidate_gbif = classification_candidates %>%
  crossing(study_system = gbif_systems) %>%
  left_join(gbif_species, by = c("study_system", "family", "genus", "specificEpithet")) %>%
  left_join(gbif_genus, by = c("study_system", "family", "genus")) %>%
  left_join(gbif_family, by = c("study_system", "family")) %>%
  group_by(asv_id, candidate) %>%
  mutate(n_other_species = sum(gbif_count_species > 0, na.rm = TRUE) - (gbif_count_species > 0 & !is.na(gbif_count_species)),
         n_other_genus = sum(gbif_count_genus > 0, na.rm = TRUE) - (gbif_count_genus > 0 & !is.na(gbif_count_genus)),
         n_other_family = sum(gbif_count_family > 0, na.rm = TRUE) - (gbif_count_family > 0 & !is.na(gbif_count_family)),
         systems_species_recorded = paste(sort(study_system[gbif_count_species > 0 & !is.na(gbif_count_species)]), collapse = ";"),
         systems_genus_recorded = paste(sort(study_system[gbif_count_genus > 0 & !is.na(gbif_count_genus)]), collapse = ";"),
         systems_family_recorded = paste(sort(study_system[gbif_count_family > 0 & !is.na(gbif_count_family)]), collapse = ";")) %>%
  ungroup() %>%
  select(asv_id, candidate, study_system, gbif_count_species, gbif_count_genus, gbif_count_family,
         n_other_species, n_other_genus, n_other_family,
         systems_species_recorded, systems_genus_recorded, systems_family_recorded)

# does a candidate match a "Genus species" / "Genus" / "Family" list?
matches_taxa_list = function(family, genus, specificEpithet, taxa_list) {
  if (length(taxa_list) == 0) return(rep(FALSE, length(genus)))
  (family %in% taxa_list) | (genus %in% taxa_list) | (paste(genus, specificEpithet) %in% taxa_list)
}

# assess each (asv_id, study_system, candidate)
locality_candidates = read_evidence %>%
  select(asv_id, study_system) %>%
  left_join(classification_candidates, by = "asv_id", relationship = "many-to-many") %>%
  left_join(candidate_gbif, by = c("asv_id", "candidate", "study_system")) %>%
  rowwise() %>%
  mutate(local_override = matches_taxa_list(family, genus, specificEpithet, unlist(local_overrides[[study_system]]))) %>%
  ungroup() %>%
  mutate(anthropogenic = matches_taxa_list(family, genus, specificEpithet, anthropogenic_taxa),
         # rank the GBIF assessment is made at: the accepted rank, or genus when an accepted species
         # was not resolved by GBIF
         gbif_assessed_rank = case_when(
           !(kingdom %in% "Metazoa") ~ NA_character_,
           taxonomic_rank == "species" & !is.na(gbif_count_species) ~ "species",
           taxonomic_rank == "species" & !is.na(gbif_count_genus) ~ "genus",
           taxonomic_rank == "genus" & !is.na(gbif_count_genus) ~ "genus",
           taxonomic_rank == "family" & !is.na(gbif_count_family) ~ "family",
           .default = NA_character_),
         gbif_count_here = case_when(
           gbif_assessed_rank == "species" ~ gbif_count_species,
           gbif_assessed_rank == "genus" ~ gbif_count_genus,
           gbif_assessed_rank == "family" ~ gbif_count_family),
         n_other_systems = case_when(
           gbif_assessed_rank == "species" ~ n_other_species,
           gbif_assessed_rank == "genus" ~ n_other_genus,
           gbif_assessed_rank == "family" ~ n_other_family),
         systems_recorded = case_when(
           gbif_assessed_rank == "species" ~ systems_species_recorded,
           gbif_assessed_rank == "genus" ~ systems_genus_recorded,
           gbif_assessed_rank == "family" ~ systems_family_recorded),
         systems_recorded = na_if(systems_recorded, ""),
         # species accepted but assessed at genus (not resolved by GBIF) -> at best local_genus_only
         species_fell_back = taxonomic_rank == "species" & gbif_assessed_rank %in% "genus",
         local_status = case_when(
           anthropogenic ~ "anthropogenic",
           local_override ~ "local",
           is.na(gbif_assessed_rank) ~ "unassessable",
           gbif_count_here > 0 & !species_fell_back ~ "local",
           gbif_count_here > 0 & species_fell_back ~ "local_genus_only",
           n_other_systems > 0 ~ "other_system",
           gbif_assessed_rank == "species" & gbif_count_genus > 0 & !is.na(gbif_count_genus) ~ "local_genus_only",
           .default = "not_in_run"),
         local_status_source = case_when(
           anthropogenic ~ "anthropogenic_taxa",
           local_override ~ "local_overrides",
           local_status == "unassessable" ~ NA_character_,
           .default = "gbif"),
         local_status = factor(local_status, levels = status_levels))

# keep the most local candidate per (asv_id, study_system)
locality_gbif = locality_candidates %>%
  group_by(asv_id, study_system) %>%
  mutate(n_candidates = n_distinct(candidate)) %>%
  arrange(local_status, candidate, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(asv_id, study_system, local_status, local_status_source, gbif_assessed_rank,
         gbif_count_here, n_other_systems, systems_recorded, n_candidates)

print(count(locality_gbif, local_status))
```

# Combine ASVs
```{r}
asv_locality = read_evidence %>%
  left_join(locality_gbif, by = c("asv_id", "study_system")) %>%
  left_join(asv_classification %>%
              select(asv_id, accept_method, taxonomic_rank, all_of(tax_ranks)),
            by = "asv_id") %>%
  left_join(lab_control_evidence, by = "asv_id") %>%
  mutate(local_status = replace_na(as.character(local_status), "unassessable"),
         n_candidates = replace_na(n_candidates, 0),
         in_lab_controls = !is.na(lab_control_n_detected),
         lab_control_n_detected = replace_na(lab_control_n_detected, 0),
         is_dominant = study_system == dominant_system,
         review_case = case_when(
           local_status == "other_system" &
             rel_abund_max >= locality_min_rel_abund &
             log10_ratio_to_dominant >= locality_log_ratio_cutoff ~ "other_system_abundant",
           local_status == "local" &
             !is_dominant &
             rel_abund_max < locality_min_rel_abund &
             log10_ratio_to_dominant < locality_log_ratio_cutoff ~ "local_trace",
           .default = NA_character_),
         taxon_label = coalesce(if_else(!is.na(species), paste(genus, species), NA_character_),
                                genus, family, order, class, phylum, kingdom, domain, "unclassified")) %>%
  select(asv_id,
         study_system,
         local_status,
         local_status_source,
         review_case,
         taxon_label,
         accept_method,
         taxonomic_rank,
         all_of(tax_ranks),
         gbif_assessed_rank,
         gbif_count_here,
         n_other_systems,
         systems_recorded,
         n_candidates,
         reads,
         n_samples,
         n_samples_detected,
         detection_rate,
         n_sites,
         n_sites_detected,
         rel_abund_max,
         rel_abund_median,
         rel_abund_mean,
         norm_share,
         is_dominant,
         dominant_system,
         log10_ratio_to_dominant,
         n_systems_detected,
         systems_detected,
         n_field_neg_detected,
         field_neg_rel_abund_max,
         in_lab_controls,
         lab_control_n_detected,
         lab_control_max_reads,
         lab_control_rel_abund_max) %>%
  arrange(asv_id, study_system)

print(count(asv_locality, local_status))
print(count(asv_locality, review_case))

write_csv(asv_locality, locality_out)
```

# Visualize

Plots to help pick the thresholds script 08 will filter on. Dashed lines mark the current reference thresholds (`locality_min_rel_abund`, `locality_log_ratio_cutoff`). Hover over any mark for its ASV, taxon, and evidence.

```{r}
# status palette: the three statuses that drive threshold decisions get distinct hues; the rest
# are muted gray, distinguished by shape
status_colors = c(local = "#2a78d6",
                  other_system = "#eb6834",
                  not_in_run = "#1baf7a",
                  local_genus_only = "#8c8b86",
                  anthropogenic = "#8c8b86",
                  unassessable = "#c3c2bd")

status_shapes = c(local = 16,
                  other_system = 16,
                  not_in_run = 16,
                  local_genus_only = 1,
                  anthropogenic = 17,
                  unassessable = 4)

status_codes = c(local = "L",
                 other_system = "O",
                 not_in_run = "N",
                 local_genus_only = "G",
                 anthropogenic = "A",
                 unassessable = "U")

plot_data = asv_locality %>%
  mutate(local_status = factor(local_status, levels = status_levels),
         hover = paste0(taxon_label, " (", coalesce(taxonomic_rank, "unclassified"), ")",
                        "<br>asv: ", asv_id,
                        "<br>system: ", study_system, " | status: ", local_status,
                        "<br>GBIF recorded in: ", coalesce(systems_recorded, "none"),
                        "<br>detected ", n_samples_detected, "/", n_samples, " samples, ",
                        n_sites_detected, " sites",
                        "<br>rel. abund. max: ", signif(rel_abund_max, 2),
                        " | log10 ratio to dominant (", dominant_system, "): ", round(log10_ratio_to_dominant, 2),
                        if_else(in_lab_controls, "<br>in lab controls", "")))

theme_locality = theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "#e6e5e0", linewidth = 0.3),
        strip.text = element_text(face = "bold", hjust = 0))
```

## Shared ASVs across systems

Each row is an ASV detected in more than one system, grouped by its dominant system. Fill is the ASV's share of (sequencing-effort-normalized) relative abundance in each system; letters are `local_status` (L local, O other_system, G local_genus_only, N not_in_run, A anthropogenic, U unassessable). Look for rows with one dark cell and pale cells marked O or N elsewhere.
```{r fig.width=9, fig.height=12}
shared = plot_data %>%
  filter(n_systems_detected > 1) %>%
  mutate(row_label = paste0(taxon_label, " [", substr(asv_id, 1, 6), "]"))

row_order = shared %>%
  filter(is_dominant) %>%
  arrange(dominant_system, class, order, family, genus, species, desc(rel_abund_mean)) %>%
  pull(row_label)

shared = shared %>%
  mutate(row_label = factor(row_label, levels = rev(unique(row_order))),
         study_system = factor(study_system, levels = possible_study_systems),
         status_code = status_codes[as.character(local_status)])

p_shared = ggplot(shared, aes(x = study_system, y = row_label, text = hover)) +
  geom_tile(aes(fill = log10(norm_share)), color = "white", linewidth = 0.5) +
  geom_text(aes(label = status_code,
                color = norm_share > 0.1),
            size = 2.6) +
  scale_fill_gradient(low = "#cde2fb", high = "#0d366b", name = "log10 share") +
  scale_color_manual(values = c(`TRUE` = "#ffffff", `FALSE` = "#0b0b0b"), guide = "none") +
  scale_x_discrete(position = "top", drop = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_locality +
  theme(panel.grid.major = element_blank(),
        axis.text.y = element_text(size = 6))

ggplotly(p_shared, tooltip = "text", height = max(600, 14 * n_distinct(shared$row_label)))
```

## System exclusivity

The next three views summarize how each ASV's (sequencing-effort-normalized) relative abundance is split across systems (`norm_share`), to separate ASVs exclusive to or dominated by one system from those shared between several.
```{r}
# system identity palette (kept distinct from the status hues above)
system_colors = c(brazil = "#eda100",
                  panama = "#e87ba4",
                  pennsylvania = "#008300",
                  sierra_nevada = "#4a3aa7")

# one row per ASV: dominant system, its share, status there, and a per-system breakdown for hover
asv_summary = plot_data %>%
  arrange(asv_id, desc(norm_share)) %>%
  group_by(asv_id, taxon_label, dominant_system, n_systems_detected) %>%
  summarise(dominant_share = max(norm_share),
            total_reads = sum(reads),
            dominant_status = as.character(local_status[is_dominant]),
            share_text = paste0(study_system, ": ", signif(norm_share, 2), " (", local_status, ")",
                                collapse = "<br>"),
            .groups = "drop") %>%
  mutate(dominant_system = factor(dominant_system, levels = possible_study_systems),
         hover = paste0(taxon_label, "<br>asv: ", asv_id, "<br>", share_text))
```

### Share profile per ASV

Each thin bar is one ASV, split by its share in each system it was detected in (the dominant system's share at the bottom). ASVs are grouped by dominant system and sorted by dominant share, so ASVs exclusive to one system form the solid block on the left and shared ASVs form the mixed tail on the right.
```{r fig.width=10, fig.height=9}
structure_order = asv_summary %>%
  arrange(dominant_system, desc(dominant_share), desc(total_reads)) %>%
  group_by(dominant_system) %>%
  mutate(x = row_number(),
         panel = paste0("dominant in ", dominant_system, " (", n(), " ASVs, ",
                        sum(n_systems_detected == 1), " exclusive)")) %>%
  ungroup() %>%
  mutate(panel = fct_inorder(panel))

structure_data = plot_data %>%
  select(asv_id, study_system, norm_share) %>%
  inner_join(structure_order %>% select(asv_id, dominant_system, x, panel, hover), by = "asv_id") %>%
  group_by(asv_id) %>%
  # dominant system at the bottom, then remaining systems by decreasing share
  arrange(desc(study_system == as.character(dominant_system)), desc(norm_share), .by_group = TRUE) %>%
  mutate(ymax = cumsum(norm_share),
         ymin = ymax - norm_share) %>%
  ungroup() %>%
  mutate(study_system = factor(study_system, levels = possible_study_systems))

p_structure = ggplot(structure_data,
                     aes(xmin = x - 0.5, xmax = x + 0.5, ymin = ymin, ymax = ymax,
                         fill = study_system, text = hover)) +
  geom_rect() +
  scale_fill_manual(values = system_colors, name = "detected in", drop = FALSE) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), expand = c(0, 0)) +
  facet_wrap(~ panel, ncol = 1, scales = "free_x") +
  labs(x = "ASVs, sorted by dominant-system share", y = "share of normalized rel. abund.") +
  theme_locality +
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_blank())

ggplotly(p_structure, tooltip = "text")
```

### Shared ASVs in system space

Each system sits at a corner of the square, and each shared ASV (detected in more than one system) is placed at the share-weighted average of the corners: near a corner = dominated by that system, on an edge = split between two adjacent systems, toward the center = spread across several. Exclusive ASVs would all sit exactly on a corner, so they are counted in the corner labels rather than plotted. The center is ambiguous -- an even 4-way split and a 50/50 split between diagonal systems (brazil + pennsylvania, panama + sierra_nevada) land in the same place -- so check the hover breakdown. Color is `local_status` in the dominant system, size is total reads.
```{r fig.width=8, fig.height=8}
corners = tibble(study_system = possible_study_systems,
                 cx = c(-1, 1, 1, -1),
                 cy = c(1, 1, -1, -1))

radviz = plot_data %>%
  filter(n_systems_detected > 1) %>%
  left_join(corners, by = "study_system") %>%
  group_by(asv_id) %>%
  summarise(x = sum(norm_share * cx),
            y = sum(norm_share * cy),
            .groups = "drop") %>%
  inner_join(asv_summary, by = "asv_id") %>%
  mutate(dominant_status = factor(dominant_status, levels = status_levels))

corner_labels = corners %>%
  left_join(asv_summary %>%
              filter(n_systems_detected == 1) %>%
              count(study_system = as.character(dominant_system), name = "n_exclusive"),
            by = "study_system") %>%
  mutate(n_exclusive = replace_na(n_exclusive, 0),
         label = paste0(study_system, "\n", n_exclusive, " exclusive ASVs"),
         ly = cy * 1.18)

p_radviz = ggplot() +
  geom_polygon(data = corners, aes(x = cx, y = cy), fill = NA, color = "#c3c2bd", linewidth = 0.4) +
  geom_segment(aes(x = c(-1, -1), y = c(1, -1), xend = c(1, 1), yend = c(-1, 1)),
               color = "#e6e5e0", linetype = "dashed", linewidth = 0.3) +
  geom_text(data = corner_labels, aes(x = cx, y = ly, label = label), size = 3.2, color = "#0b0b0b") +
  geom_point(data = radviz,
             aes(x = x, y = y, color = dominant_status, shape = dominant_status,
                 size = log10(total_reads), text = hover),
             alpha = 0.75) +
  scale_color_manual(values = status_colors, drop = FALSE, name = "status in dominant system") +
  scale_shape_manual(values = status_shapes, drop = FALSE, name = "status in dominant system") +
  scale_size_continuous(range = c(1.5, 7), name = "log10 total reads") +
  coord_equal(xlim = c(-1.45, 1.45), ylim = c(-1.35, 1.35)) +
  theme_void(base_size = 11)

ggplotly(p_radviz, tooltip = "text")
```

### Directional spillover between systems

Rows are the system an ASV is dominant in; columns are the other systems it was also detected in. Off-diagonal cells give the number of ASVs (and the fraction of the row's dominant ASVs) detected in that column; the diagonal gives the number of ASVs dominant in that system. An asymmetric matrix (e.g. many pennsylvania -> panama, few panama -> pennsylvania) points to a direction of contamination, while symmetric sharing between nearby systems is more consistent with real shared taxa.
```{r fig.width=7, fig.height=6}
n_dominant = asv_summary %>%
  count(source = dominant_system, name = "n_dominant")

spillover = plot_data %>%
  filter(!is_dominant) %>%
  mutate(source = factor(dominant_system, levels = possible_study_systems),
         detected_in = factor(study_system, levels = possible_study_systems)) %>%
  group_by(source, detected_in) %>%
  summarise(n_asvs = n(),
            n_local = sum(local_status == "local"),
            n_other_system = sum(local_status == "other_system"),
            n_not_in_run = sum(local_status == "not_in_run"),
            n_other_status = n_asvs - n_local - n_other_system - n_not_in_run,
            n_review = sum(!is.na(review_case)),
            median_log10_ratio = median(log10_ratio_to_dominant),
            median_rel_abund_max = median(rel_abund_max),
            .groups = "drop") %>%
  left_join(n_dominant, by = "source") %>%
  mutate(frac_of_source = n_asvs / n_dominant)

spillover_grid = expand_grid(source = factor(possible_study_systems, levels = possible_study_systems),
                             detected_in = factor(possible_study_systems, levels = possible_study_systems)) %>%
  left_join(spillover %>% select(source, detected_in, n_asvs, frac_of_source, median_log10_ratio),
            by = c("source", "detected_in")) %>%
  left_join(n_dominant, by = "source") %>%
  mutate(diagonal = source == detected_in,
         n_asvs = replace_na(n_asvs, 0),
         frac_of_source = replace_na(frac_of_source, 0),
         label = if_else(diagonal,
                         paste0(replace_na(n_dominant, 0), "\ndominant"),
                         paste0(n_asvs, "\n", scales::percent(frac_of_source, accuracy = 0.1))),
         hover = if_else(diagonal,
                         paste0(source, ": ", replace_na(n_dominant, 0), " ASVs dominant here"),
                         paste0(n_asvs, " ASVs dominant in ", source, " also detected in ", detected_in,
                                "<br>median log10 ratio to dominant: ", round(median_log10_ratio, 2))))

p_spillover = ggplot(spillover_grid, aes(x = detected_in, y = source, text = hover)) +
  geom_tile(data = filter(spillover_grid, !diagonal), aes(fill = frac_of_source),
            color = "white", linewidth = 1) +
  geom_tile(data = filter(spillover_grid, diagonal), fill = "#e6e5e0", color = "white", linewidth = 1) +
  geom_text(aes(label = label, color = !diagonal & frac_of_source > 0.15), size = 3.2) +
  scale_fill_gradient(low = "#cde2fb", high = "#0d366b", name = "fraction of\nrow's ASVs") +
  scale_color_manual(values = c(`TRUE` = "#ffffff", `FALSE` = "#0b0b0b"), guide = "none") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev) +
  labs(x = "also detected in", y = "dominant in") +
  theme_locality +
  theme(panel.grid.major = element_blank())

ggplotly(p_spillover, tooltip = "text")

spillover %>%
  arrange(desc(n_asvs)) %>%
  mutate(across(c(median_log10_ratio, median_rel_abund_max, frac_of_source), \(x) signif(x, 2)))
```

## Detection rate vs. relative abundance, by system

One mark per (ASV, system). Likely contamination sits bottom-left (rare, trace). Confusing cases: `other_system` (orange) in the upper right -- GBIF says not local, but reads look real; `local` (blue) in the lower left of a system that isn't dominant for the ASV.
```{r fig.width=10, fig.height=8}
p_scatter = ggplot(plot_data %>%
                     filter(!(local_status %in% c("unassessable", "anthropogenic"))), aes(x = detection_rate, y = rel_abund_max,
                                  color = local_status, shape = local_status, text = hover)) +
  geom_hline(yintercept = locality_min_rel_abund, linetype = "dashed", color = "#52514e", linewidth = 0.4) +
  geom_point(size = 2, alpha = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = status_colors, drop = FALSE) +
  scale_shape_manual(values = status_shapes, drop = FALSE) +
  facet_wrap(~ study_system) +
  labs(x = "detection rate (fraction of samples)",
       y = "max within-sample relative abundance",
       color = "local_status", shape = "local_status") +
  theme_locality

ggplotly(p_scatter, tooltip = "text")
```

## Calibration: `local` vs. `other_system` in non-dominant systems

For ASVs detected in more than one system, the detections outside the ASV's dominant system, split by GBIF status. If spillover explains the `other_system` detections, they should sit left of the `local` ones on both metrics -- a threshold is defensible where the two distributions separate.
```{r fig.width=10, fig.height=6}
calibration = plot_data %>%
  filter(n_systems_detected > 1,
         !is_dominant,
         local_status %in% c("local", "other_system", "not_in_run")) %>%
  transmute(asv_id, local_status, hover,
            `log10 ratio to dominant system` = log10_ratio_to_dominant,
            `log10 max relative abundance` = log10(rel_abund_max)) %>%
  pivot_longer(cols = c(`log10 ratio to dominant system`, `log10 max relative abundance`),
               names_to = "metric", values_to = "value")

thresholds = tibble(metric = c("log10 ratio to dominant system", "log10 max relative abundance"),
                    value = c(locality_log_ratio_cutoff, log10(locality_min_rel_abund)))

p_calibration = ggplot(calibration, aes(x = value, fill = local_status)) +
  geom_histogram(bins = 40, color = "white", linewidth = 0.2) +
  geom_vline(data = thresholds, aes(xintercept = value),
             linetype = "dashed", color = "#52514e", linewidth = 0.4) +
  scale_fill_manual(values = status_colors) +
  facet_grid(local_status ~ metric, scales = "free") +
  labs(x = NULL, y = "(ASV, system) detections") +
  theme_locality +
  theme(legend.position = "none")

ggplotly(p_calibration)
```

## Cases to review

Rows where GBIF and read evidence disagree at the current reference thresholds (`review_case`).
```{r}
asv_review = asv_locality %>%
  filter(!is.na(review_case)) %>%
  arrange(review_case, study_system, desc(rel_abund_max)) %>%
  transmute(review_case,
            study_system,
            taxon = taxon_label,
            taxonomic_rank,
            asv = substr(asv_id, 1, 8),
            recorded_in = systems_recorded,
            dominant = dominant_system,
            detected = paste0(n_samples_detected, "/", n_samples),
            rel_abund_max = signif(rel_abund_max, 2),
            log10_ratio = round(log10_ratio_to_dominant, 2),
            lab_ctrl = in_lab_controls) %>%
  knitr::kable()



```

# asv locality clean systems
```{r}
asv_locality_clean = asv_locality %>%
  group_by(asv_id) %>%
  summarize(n_systems = n(),
            n_local = sum(local_status == "local"),
            .groups = "drop") %>%
  filter(n_systems == 1,
         n_local == 1) %>%
left_join(asv_locality, by = "asv_id") %>%
  filter(class == "Amphibia") %>%
  arrange(study_system,
          desc(rel_abund_max))

```