---
title: "Sample controls"
author: "Cob Staines"
output: html_document
---

# Description

This script calculates controlled ASV reads for each sample by subtracting any read counts found in controls (multiplied by a scaling factor `asv_control_th_factor`), with minimum controlled reads of 0. Lab positive and negative controls are applied to all samples globally, field negative controls are applied to corresponding field samples only.

# Config
```{r}
librarian::shelf(tidyverse, yaml, dbplyr, RPostgres, DBI, ribbitrrr, here, rio, janitor, phylotools)

# read shared pipeline config (see pipeline_config.yml / 00_run_pipeline.R); this is read
# directly so the same values apply whether this script is run manually or via the driver
pipeline_config = read_yaml(here("analysis", "general", "r", "pipeline_config.yml"))

# import run config file
run_dir = here(pipeline_config$run_dir)

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# asv control threshold factor for field contaminants
asv_control_th_factor = pipeline_config$asv_control_th_factor

# control output file
control_out = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))

```

# Load data
```{r}
asv_classification = read_csv(here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv")))

asv_contaminants = asv_classification %>%
  select(asv_id,
         contaminant_flag,
         contaminant_id,
         contaminant_p_identity,
         short_sequence_flag)

sample_map = read_csv(here(run_dir, "output", paste0(config$run$name, "_07_sample_id_map.csv")))
  # rename(field_negative_control = negative_control,
  #        field_negative_control_group_id = negative_control_group_id)

sample_map_simple = sample_map %>%
  mutate(sample_type = case_when(
    pcr_control ~ pcr_control_type,
    extraction_negative_control ~ "extraction_negative_control",
    field_negative_control ~ "field_negative_control",
    grepl("sterile", comments_edna_sample, ignore.case = TRUE) ~ "sterile",
    .default = "sample")) %>%
  select(feature_id,
         edna_sequencing_id,
         extraction_negative_control,
         pcr_control,
         pcr_control_type,
         field_negative_control,
         field_negative_control_group,
         sample_type)

feature_table =
  read_tsv(here(run_dir, "/analysis/s06_denoised_16S_eDNA/sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

sample_asv = feature_table %>%
  pivot_longer(cols = -asv_id,
               names_to = "feature_id",
               values_to = "asv_count") %>%
  # filter(asv_count != 0) %>%
  left_join(sample_map_simple, by = "feature_id")
  
```

# Identify control ASVs
```{r}
# lab positive controls
control_asv_pcr_pos = sample_asv %>%
  filter(pcr_control_type == "pcr_positive_control") %>%
  group_by(asv_id) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop") %>%
  mutate(control_th_source = "pcr_pc")

# lab negative controls
control_asv_pcr_neg = sample_asv %>%
  filter(pcr_control_type == "pcr_negative_control") %>%
  group_by(asv_id) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop") %>%
  mutate(control_th_source = "pcr_nc")

# lab controls (pooled accross run)
control_asv_extraction = sample_asv %>%
  filter(extraction_negative_control) %>%
  group_by(asv_id) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop") %>%
  mutate(control_th_source = "extraction_nc")

# field negative controls
control_asv_neg_field = sample_asv %>%
  filter(field_negative_control) %>%
  group_by(asv_id, field_negative_control_group) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop") %>%
  mutate(control_th_source = "field")
```

# Calculate controlled reads & export
```{r}
# correct asv reads by controls
control_asv = bind_rows(control_asv_pcr_pos,
                        control_asv_pcr_neg,
                        control_asv_extraction) %>%
  cross_join(bind_rows(control_asv_neg_field %>%
                         select(field_negative_control_group) %>%
                         distinct(),
                       tibble(field_negative_control_group = NA))) %>%
  bind_rows(control_asv_neg_field) %>%
  filter(max_control_asv_count > 0) %>%
  group_by(asv_id, field_negative_control_group) %>%
  slice_max(max_control_asv_count, n = 1, with_ties = FALSE) %>%
  ungroup()

sample_asv_control = sample_asv %>%
  left_join(control_asv, by = c("asv_id", "field_negative_control_group")) %>%
  left_join(asv_contaminants, by = "asv_id") %>%
  mutate(max_control_asv_count = replace_na(max_control_asv_count, 0),
         asv_control_th = asv_control_th_factor * max_control_asv_count,
         asv_count_controlled = asv_count - asv_control_th,
         asv_count_controlled = if_else(asv_count_controlled < 0, 0, asv_count_controlled),
         control_th_below = (asv_count <= asv_control_th),
         detected = (asv_count > 0),
         detected_controlled = (detected & !(control_th_below %in% c(TRUE))),
         asv_count_controlled = if_else(contaminant_flag | short_sequence_flag,
                                        NA,
                                        asv_count_controlled),
         control_th_source = case_when(
           contaminant_flag ~ "contaminant",
           short_sequence_flag ~ "short_sequence",
           .default = control_th_source
         )) %>%
  rename(asv_count_raw = asv_count) %>%
  arrange(asv_id, desc(asv_count_raw)) %>%
  select(-any_of(c("extraction_negative_control",
                   "pcr_control",
                   "pcr_control_type",
                   "field_negative_control",
                   "field_negative_control_group"))) %>%
  select(asv_id,
         feature_id,
         edna_sequencing_id,
         sample_type,
         everything())

write_csv(sample_asv_control, control_out)

```
