---
title: "08_sample_controls"
output: html_document
---

# Description

This script calculates controlled ASV reads for each sample by subtracting any read counts found in controls (multiplied by a scaling factor `asv_control_th_factor`), with minimum controlled reads of 0. Lab positive and negative controls are applied to all samples globally, field negative controls are applied to corresponding field samples only.

# Config
```{r}
librarian::shelf(tidyverse, yaml, dbplyr, RPostgres, DBI, ribbitrrr, here, rio, janitor, phylotools)

# import run config file
run_dir = here("16S_sequence_processing", "runs", "penn_2026-03-10")

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# asv control threshold factor
asv_control_th_factor = 100

# control output file
control_out = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))

```

# Load data
```{r}

sample_map = read_csv(here(run_dir, "output", paste0(config$run$name, "_07_sample_id_map.csv"))) %>%
  rename(field_negative_control = negative_control,
         field_negative_control_group_id = negative_control_group_id)

feature_table =
  read_tsv(here(run_dir, "/analysis/s06_denoised_16S_eDNA/sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

sample_asv = feature_table %>%
  pivot_longer(cols = -asv_id,
               names_to = "illumina_sample_raw",
               values_to = "asv_count") %>%
  # filter(asv_count != 0) %>%
  left_join(sample_map, by = "illumina_sample_raw") %>%
  mutate(control_type = case_when(
    grepl("^Positive", illumina_sample_raw) ~ "positive_lab",
    grepl("^Neg|NC", illumina_sample_raw) ~ "negative_lab",
    field_negative_control %in% c(TRUE) ~ "negative_field",
    .default = NA))
  
```

# Identify control ASVs
```{r}
# lab positive controls
control_asv_pos = sample_asv %>%
  filter(control_type == "positive_lab") %>%
  group_by(asv_id, control_type) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")

# lab negative controls
control_asv_neg_lab = sample_asv %>%
  filter(control_type == "negative_lab") %>%
  group_by(asv_id, control_type) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")

# field negative controls
control_asv_neg_field = sample_asv %>%
  filter(control_type == "negative_field") %>%
  group_by(asv_id, control_type, field_negative_control_group_id) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")
```

# Calculate controlled reads & export
```{r}
# correct asv reads by controls
control_asv = bind_rows(control_asv_pos,
                        control_asv_neg_lab) %>%
  cross_join(bind_rows(control_asv_neg_field %>%
                         select(field_negative_control_group_id) %>%
                         distinct(),
                       tibble(field_negative_control_group_id = NA))) %>%
  bind_rows(control_asv_neg_field) %>%
  filter(max_control_asv_count > 0) %>%
  group_by(asv_id, field_negative_control_group_id) %>%
  slice_max(max_control_asv_count, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(control_th_type = control_type)

sample_asv_control = sample_asv %>%
  left_join(control_asv, by = c("asv_id", "field_negative_control_group_id")) %>%
  mutate(max_control_asv_count = replace_na(max_control_asv_count, 0),
         asv_control_th = asv_control_th_factor * max_control_asv_count,
         asv_count_controlled = asv_count - asv_control_th,
         asv_count_controlled = if_else(asv_count_controlled < 0, 0, asv_count_controlled),
         control_th_below = (asv_count <= asv_control_th),
         detected = (asv_count > 0),
         detected_controlled = (detected & !(control_th_below %in% c(TRUE)))) %>%
  rename(asv_count_raw = asv_count)

write_csv(sample_asv_control, control_out)

```
