---
title: "10_results_overview"
output: html_document
---

# Description

This script provides a first look at the processed and analyzed results, including some checks and preliminary summaries to identify any anomalies or curiosities.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here, phylotools, dbplyr, RPostgres, DBI, ribbitrrr, vegan, janitor)

# import run config file
run_dir = here("16S_sequence_processing", "runs", "panama_16s_2026-09-16")

# read in config file
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

```

# load results
```{r include=FALSE}
# hybrid classification
res_06_class = read_csv(here(run_dir, "output", paste0(config$run$name, "_06_hybrid_classification.csv")))

# # controls
# res_08_controlled = read_csv(here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv")))

# sample_metadata
res_07_sample_map = read_csv(here(run_dir, "output", paste0(config$run$name, "_07_sample_id_map.csv")))

# exported file
res_09_export = read_csv(here(run_dir, "output", paste0(config$run$name, "_09_prelim_results.csv")))

peace = res_09_export %>%
  select(-edna_sequencing_id,
         -feature_id,
         -sample_type,
         ) %>%
  distinct()

# metadata_in = read_csv(here(run_dir, "output", paste0(config$run$name, "_09_prelim_results_metadata.csv")))

# res_09_positive = res_09_export %>%
#   filter(asv_reads_controlled > 0) %>%
#   arrange(class, desc(asv_reads_controlled))

```

# Read counts
```{r}
# summary stats for samples (including controls)
sample_read_count = res_09_export %>%
  # drop all controls
  # filter(sample_type == "sample") %>%
  group_by(edna_sequencing_id) %>%
  mutate(detected_raw = asv_reads_raw > 0,
         detected_controlled = asv_reads_controlled > 0) %>%
  summarize(
    # number of reads
    n_reads = sum(asv_reads_raw),
    # number of reads classified
    n_reads_class = sum(if_else(!is.na(taxonomic_rank), asv_reads_raw, 0L), na.rm = TRUE),
    # number of reads classified chordata
    n_reads_class_chord = sum(if_else(phylum == "Chordata", asv_reads_raw, 0L), na.rm = TRUE),
    # number of reads classified as amphibians
    n_reads_class_chord_amph = sum(if_else(class == "Amphibia", asv_reads_raw, 0L), na.rm = TRUE),
    # controlled number of reads classified as amphibians
    n_reads_class_chord_amph_ctrl = sum(if_else(class == "Amphibia", asv_reads_controlled, 0L), na.rm = TRUE),
    # number of reads classified as amphibians to the species level
    n_reads_class_chord_amph_spp = sum(if_else((class == "Amphibia") & (taxonomic_rank == "species"), asv_reads_raw, 0L), na.rm = TRUE),
    # controlled number of reads classified as amphibians to the species level
    n_reads_class_chord_amph_spp_ctrl = sum(if_else((class == "Amphibia") & (taxonomic_rank == "species"), asv_reads_controlled, 0L), na.rm = TRUE),
    # number of reads classified as human
    n_reads_class_chord_sapien = sum(if_else((species == "sapiens"), asv_reads_raw, 0L), na.rm = TRUE),
    # number of controlled reads classified as human
    n_reads_class_chord_sapien_ctrl = sum(if_else((species == "sapiens"), asv_reads_controlled, 0L), na.rm = TRUE),
    ## ASVs
    # number of asvs
    n_asvs = length(unique(asv_id[asv_reads_raw > 0])),
    # number of asvs classified
    n_asvs_class = length(unique(asv_id[asv_reads_raw > 0 & !is.na(taxonomic_rank)])),
    # number of asvs classified chordata
    n_asvs_class_chord = length(unique(asv_id[asv_reads_raw > 0 & phylum %in% "Chordata"])),
    # number of asvs classified as amphibians
    n_asvs_class_chord_amph = length(unique(asv_id[asv_reads_raw > 0 & class %in% "Amphibia"])),
    # controlled number of asvs classified as amphibians
    n_asvs_class_chord_amph_ctrl = length(unique(asv_id[asv_reads_controlled > 0 & class %in% "Amphibia"])),
    # number of asvs classified as amphibians to the species level
    n_asvs_class_chord_amph_spp = length(unique(asv_id[asv_reads_raw > 0 & class %in% "Amphibia" & taxonomic_rank == "species"])),
    # controlled number of asvs classified as amphibians to the species level
    n_asvs_class_chord_amph_spp_ctrl = length(unique(asv_id[asv_reads_controlled > 0 & class %in% "Amphibia" & taxonomic_rank == "species"])),
    # number of asvs classified as human
    n_asvs_class_chord_sapien = length(unique(asv_id[asv_reads_raw > 0 & species %in% "sapiens"])),
    # number of controlled asvs classified as human
    n_asvs_class_chord_sapien_ctrl = length(unique(asv_id[asv_reads_controlled > 0 & species %in% "sapiens"])),
    .groups = "drop") %>%
  mutate(ratio_amph = n_reads_class_chord_amph_ctrl / n_reads,
         ratio_sapiens = n_reads_class_chord_sapien / n_reads)

# pivot longer for plotting
sample_read_count_long = sample_read_count %>%
  pivot_longer(cols = c(n_reads,
                        n_reads_class,
                        n_reads_class_chord,
                        n_reads_class_chord_amph,
                        n_reads_class_chord_amph_ctrl,
                        n_reads_class_chord_amph_spp,
                        n_reads_class_chord_amph_spp_ctrl,
                        n_reads_class_chord_sapien,
                        n_reads_class_chord_sapien_ctrl),
               names_to = "count_type",
               values_to = "n") %>%
  mutate(log10_n = log10(n + 1))

sample_asv_count_long = sample_read_count %>%
  pivot_longer(cols = c(n_asvs,
                        n_asvs_class,
                        n_asvs_class_chord,
                        n_asvs_class_chord_amph,
                        n_asvs_class_chord_amph_ctrl,
                        n_asvs_class_chord_amph_spp,
                        n_asvs_class_chord_amph_spp_ctrl,
                        n_asvs_class_chord_sapien,
                        n_asvs_class_chord_sapien_ctrl),
               names_to = "count_type",
               values_to = "n")

# plot
ggplot(sample_read_count_long, aes(y = log10_n, color = count_type)) +
  geom_boxplot()

ggplot(sample_asv_count_long, aes(y = n, color = count_type)) +
  geom_boxplot()

```

# Taxa groups
```{r}
sample_taxa_group = res_09_export %>%
  mutate(taxa_group = case_when(
    domain == "Bacteria" ~ "Bacteria",
    species == "sapiens" ~ "Homo sapiens",
    class == "Amphibia" ~ "Amphibia",
    class == "Mammalia" ~ "Mammalia Other",
    class == "Aves" ~ "Aves",
    class == "Actinopteri" ~ "Actinopteri",
    class == "Reptilia" ~ "Reptilia",
    #class == "Branchiopoda" ~ "Branchiopoda",
    #class == "Insecta" ~ "Insecta",
    #class == "Lepidosauria" ~ "Lepidosauria",
    #class == "Ostracoda" ~ "Ostracoda",
    kingdom == "Metazoa" ~ "Metazoa Other",
    # kingdom == "Viridiplantae" ~ "Viridiplantae",
    !is.na(domain) ~ "Other",
    .default = "Unassigned"))

# 1. proportions per replicate
replicate_props_co = sample_taxa_group |>
  group_by(edna_sequencing_id, taxa_group) |>
  summarize(
    taxa_reads      = sum(asv_reads_raw),
    taxa_reads_ctrl = sum(asv_reads_controlled, na.rm = TRUE),
    .groups = "drop_last"  # drops taxa_group, keeping replicate groups for total
  ) |>
  mutate(
    replicate_reads      = sum(taxa_reads),
    replicate_reads_ctrl = sum(taxa_reads_ctrl, na.rm = TRUE),
    prop      = taxa_reads      / replicate_reads,
    prop_ctrl = taxa_reads_ctrl / replicate_reads_ctrl
  ) |>
  ungroup()

# 2. average proportions across replicates within each sample
sample_props_co = replicate_props_co |>
  group_by(edna_sequencing_id, taxa_group) |>
  summarize(
    prop      = mean(prop, na.rm = TRUE),
    prop_ctrl = mean(prop_ctrl, na.rm = TRUE),
    .groups = "drop"
  )

# 3. average proportions across samples within each gene
gene_props_co = sample_props_co |>
  group_by(taxa_group) |>
  summarize(
    group_reads      = mean(prop, na.rm = TRUE),
    group_reads_ctrl = mean(prop_ctrl, na.rm = TRUE),
    .groups = "drop"
  )

sample_taxa_group_agg_long_co = gene_props_co %>%
  pivot_longer(cols = c("group_reads",
                        "group_reads_ctrl"),
               names_to = "ctrl",
               values_to = "group_reads") %>%
  mutate(ctrl = recode_values(ctrl,
                              "group_reads" ~ "Uncontrolled",
                              "group_reads_ctrl" ~ "Controlled")) %>%
  group_by(ctrl) %>%
  mutate(group_read_proportion = group_reads / sum(group_reads),
         ctrl = factor(ctrl, levels = c("Uncontrolled", "Controlled")))

# Build a named color vector
taxa_colors <- setNames(
  c(RColorBrewer::brewer.pal(8, "Dark2"), "grey80"),
  c("Actinopteri", "Amphibia", "Aves", "Bacteria",
    "Homo sapiens", "Mammalia Other", "Metazoa Other", "Other",
    "Unassigned")
)

p_taxa_prop_co = ggplot(sample_taxa_group_agg_long_co,
                       aes(x = ctrl, y = group_reads, fill = taxa_group)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  geom_text(
    aes(label = ifelse(group_read_proportion > 0.05, taxa_group, "")),
    stat     = "identity",
    position = position_fill(vjust = 0.5),
    size     = 3,
    color    = "white",
    fontface = "bold"
  ) +
  # facet_grid(country ~ gene) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = taxa_colors) +        # <-- replaces scale_fill_brewer
  labs(
    # title = "Taxonomic group proportions before and after control, by gene",
    x     = "Study System",
    y     = "Mean proportion of reads per sample",
    fill  = "Taxonomic group"
  ) +
  theme_bw() +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    axis.text.x     = element_text(size = 10),
    legend.position = "bottom"
  )

p_taxa_prop_co

```

# Classification
## aggregate by ASV
```{r}
asv_classification_vars = res_06_class %>%
  filter(is.na(percent_identical),
         !is.na(accept_method_priority)) %>%
  group_by(asv_id) %>%
  slice_min(accept_method_priority, n = 1) %>%
  ungroup() %>%
  select(asv_id,
         accept_method,
         accept_method_priority,
         p_identical_mean,
         p_identical_max,
         p_identical_min,
         seq_overlap_max) %>%
  distinct()

asv_agg = res_09_export %>%
  select("asv_id", 
         "asv_sequence", "asv_reads_raw", "asv_reads_controlled", "control_th_source", 
         "taxonomic_rank", "domain", "kingdom", "phylum", "class", "order", 
         "family", "genus", "species", "n_hits", "gbif_count_family_local", 
         "gbif_count_genus_local", "gbif_count_species_local") %>%
  left_join(asv_classification_vars, by = "asv_id") %>%
  group_by(across(-c(asv_reads_raw,
                     asv_reads_controlled,
                     control_th_source))) %>%
  summarize(total_asv_reads_raw = sum(asv_reads_raw),
            total_asv_reads_controlled = sum(asv_reads_controlled),
            .groups = "drop")

peace = get_dupes(asv_classification_vars, asv_id)

```

## Amphibian species ASVs
```{r}
amphibian_asvs_raw = asv_agg %>%
  filter(total_asv_reads_raw > 0,
         class %in% "Amphibia") %>%
  arrange(taxonomic_rank,
          desc(total_asv_reads_raw))

amphibian_asvs_controlled = asv_agg %>%
  filter(total_asv_reads_controlled > 0,
         class %in% "Amphibia") %>%
  arrange(taxonomic_rank,
          desc(total_asv_reads_controlled))

amphibian_asvs_controlled_hybrid = res_06_class %>%
  filter(asv_id %in% amphibian_asvs_controlled$asv_id)
```

## pick out specific samples
```{r}
samples_filtered = res_09_export %>%
  filter(asv_id == "74ce083ac47674eb5740d03e521385c2",
         asv_reads_raw > 0) %>%
  left_join(res_07_sample_map, by = c("edna_sequencing_id",
                                      "feature_id"))

```

## GBIF checks
```{r}
check_gbif = asv_agg %>%
  filter(total_asv_reads_controlled > 0,
         (taxonomic_rank == "species" & gbif_count_species_local == 0) |
           (taxonomic_rank == "genus" & gbif_count_genus_local == 0) |
           (taxonomic_rank == "family" & gbif_count_family_local == 0)) %>%
  arrange(taxonomic_rank,
          desc(total_asv_reads_controlled))
```

## accept methods discrepancies
```{r}
tax_cols = c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")

check_accept_discr = res_06_class %>%
  filter(is.na(percent_identical)) %>%
  group_by(asv_id) %>%
  filter(if_any(all_of(tax_cols), ~ n_distinct(.x, na.rm = TRUE) > 1)) %>%
  ungroup()

n_accept_discr = check_accept_discr %>%
  pull(n_distinct(asv_id)) %>%
  length()

# double check that this works...
check_unclassified = res_06_class %>%
  filter(is.na(accept_method) & is.na(percent_identical))
```

# Sample-level variation

This section looks at how different samples are from one another, using
`res_07_sample_map` (site, region, coordinates, environmental readings) to
give the read-based comparisons spatial and environmental context, and to
check how well replicates from the same site/date agree.

## Community ordination (Bray-Curtis PCoA)
```{r}

res_09_prefilter = res_09_export %>%
  filter(class %in% "Amphibia")

sample_type_lookup = res_09_prefilter  %>%
  distinct(edna_sequencing_id, sample_type)

sample_sciname_reads = res_09_prefilter %>%
  filter(sample_type == "sample") %>%
  mutate(sciname = if_else(is.na(species),
                           genus,
                           paste(genus, species)),
         # genus_grp = if_else(is.na(genus), "Unclassified", genus)
         ) %>%
  group_by(edna_sequencing_id, sciname) %>%
  summarize(reads = sum(asv_reads_controlled, na.rm = TRUE), .groups = "drop")

sample_totals = sample_sciname_reads %>%
  group_by(edna_sequencing_id) %>%
  summarize(total_reads = sum(reads), .groups = "drop") %>%
  filter(total_reads > 0)

# community matrix
comm_matrix = sample_sciname_reads %>%
  inner_join(sample_totals, by = "edna_sequencing_id") %>%
  mutate(prop = reads / total_reads) %>%
  select(edna_sequencing_id, sciname, prop) %>%
  pivot_wider(names_from = sciname, values_from = prop, values_fill = 0) %>%
  column_to_rownames("edna_sequencing_id") %>%
  as.matrix()

bray_dist = vegan::vegdist(comm_matrix, method = "bray")

pcoa = cmdscale(bray_dist, k = 2, eig = TRUE)
colnames(pcoa$points) = c("pcoa1", "pcoa2")
pct_var = round(100 * pcoa$eig[1:2] / sum(pcoa$eig[pcoa$eig > 0]), 1)

sample_meta = res_07_sample_map %>%
  distinct(edna_sequencing_id, site, region, date, transect)

pcoa_scores = as_tibble(pcoa$points, rownames = "edna_sequencing_id") %>%
  left_join(sample_meta, by = "edna_sequencing_id")

ggplot(pcoa_scores, aes(x = pcoa1, y = pcoa2, color = as.factor(site))) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = paste0("PCoA1 (", pct_var[1], "%)"),
       y = paste0("PCoA2 (", pct_var[2], "%)"),
       color = "Site",
       title = "Sample community composition (Bray-Curtis PCA)") +
  theme_bw()

ggplot(pcoa_scores, aes(x = pcoa1, y = pcoa2, color = region)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = paste0("PCA1 (", pct_var[1], "%)"),
       y = paste0("PCA2 (", pct_var[2], "%)"),
       color = "Region",
       title = "Sample community composition (Bray-Curtis PCA)") +
  theme_bw()

ggplot(pcoa_scores, aes(x = pcoa1, y = pcoa2, color = date)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(x = paste0("PCA1 (", pct_var[1], "%)"),
       y = paste0("PCA2 (", pct_var[2], "%)"),
       color = "Date",
       title = "Sample community composition (Bray-Curtis PCA)") +
  theme_bw()

```

## Replicate concordance
```{r}
# pairwise Bray-Curtis distance between every pair of samples, long format
bray_mat = as.matrix(bray_dist)

bray_pairs = as.data.frame(as.table(bray_mat)) %>%
  rename(seq_a = Var1, seq_b = Var2, bray_dist = Freq) %>%
  mutate(seq_a = as.character(seq_a), seq_b = as.character(seq_b)) %>%
  filter(seq_a < seq_b)

pair_meta = sample_meta %>%
  select(edna_sequencing_id, site, date)

bray_pairs = bray_pairs %>%
  left_join(pair_meta, by = c("seq_a" = "edna_sequencing_id")) %>%
  rename(site_a = site, date_a = date) %>%
  left_join(pair_meta, by = c("seq_b" = "edna_sequencing_id")) %>%
  rename(site_b = site, date_b = date) %>%
  mutate(pair_type = if_else(site_a == site_b & date_a == date_b,
                              "Site / date replicates",
                              "Distinct site / date"))

# overall: are same site/date pairs more similar than everything else?
ggplot(bray_pairs, aes(x = pair_type, y = bray_dist)) +
  geom_boxplot() +
  labs(x = NULL,
       y = "Bray-Curtis dissimilarity",
       title = "Replicate concordance: within site/date vs. across") +
  theme_bw()

# per site/date group: which replicate groups agree poorly?
replicate_concordance = bray_pairs %>%
  filter(pair_type == "Same site & date (replicate)") %>%
  group_by(site_a, date_a) %>%
  summarize(n_pairs = n(),
            mean_bray = mean(bray_dist),
            max_bray = max(bray_dist),
            .groups = "drop") %>%
  arrange(desc(mean_bray))

replicate_concordance

```

## Environmental variable patterns
```{r}
env_vars = res_07_sample_map %>%
  distinct(edna_sequencing_id, water_temp_c, p_h, conductivity_us_cm)

# how complete is env data for this run?
env_vars %>%
  summarize(across(c(water_temp_c, p_h, conductivity_us_cm), ~ sum(!is.na(.))))

env_join = pcoa_scores %>%
  left_join(env_vars, by = "edna_sequencing_id") %>%
  left_join(sample_read_count %>% select(edna_sequencing_id, ratio_amph),
            by = "edna_sequencing_id")

env_long = env_join %>%
  pivot_longer(cols = c(water_temp_c, p_h, conductivity_us_cm),
               names_to = "env_var",
               values_to = "env_value") %>%
  filter(!is.na(env_value))

# amphibian detection vs environmental variables
ggplot(env_long, aes(x = env_value, y = ratio_amph)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ env_var, scales = "free_x") +
  labs(x = "Environmental value",
       y = "Amphibian read proportion (controlled)",
       title = "Amphibian detection vs. environmental variables") +
  theme_bw()

# community composition (PCoA1) vs environmental variables
ggplot(env_long, aes(x = env_value, y = pcoa1)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ env_var, scales = "free_x") +
  labs(x = "Environmental value",
       y = "PCoA1 (community composition)",
       title = "Community composition (PCoA1) vs. environmental variables") +
  theme_bw()

```

## Spatial detection map
```{r}
site_amph = sample_read_count %>%
  left_join(sample_type_lookup, by = "edna_sequencing_id") %>%
  filter(sample_type == "sample") %>%
  left_join(res_07_sample_map %>%
              distinct(edna_sequencing_id, site, site_latitude, site_longitude, region),
            by = "edna_sequencing_id") %>%
  filter(!is.na(site_latitude), !is.na(site_longitude)) %>%
  group_by(site, site_latitude, site_longitude, region) %>%
  summarize(
    n_samples = n(),
    detection_rate = mean(n_reads_class_chord_amph_ctrl > 0),
    mean_ratio_amph = mean(ratio_amph, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(site_amph, aes(x = site_longitude, y = site_latitude,
                       size = n_samples, color = detection_rate)) +
  geom_point(alpha = 0.85) +
  scale_color_viridis_c(labels = scales::percent, limits = c(0, 1)) +
  coord_quickmap() +
  labs(x = "Longitude",
       y = "Latitude",
       color = "Amphibian\ndetection rate",
       size = "n samples",
       title = "Amphibian detection rate by site") +
  theme_bw()

```