---
title: "Network analysis of ASV detections"
author: "Cob Staines"
output: html_document
---

# Description

This script builds the sample--ASV network and uses its structure to separate signal from contamination, alongside the GBIF / read evidence in script 07. It **only flags, it drops nothing**, and nothing downstream reads its outputs: it is for exploration (including in Gephi / Cytoscape) and for checking the contamination filters of script 08.

`network_input` (config) picks which detections the network is built from:

- `raw`: the denoised feature table, before any filtering (needs scripts 03, 06, 07)
- `filtered`: script 08's controlled reads (needs script 08 too)

Running it on both and comparing is a before/after check of script 08: if the filters remove contamination rather than ecology, cross-system similarity edges should drop, ubiquitous contaminant communities should disappear, and association with site should rise relative to association with sequencing run. All outputs carry the input in their name (`_11_raw_...`, `_11_filtered_...`).

## Network export

Node and edge tables for Gephi (Data Laboratory > Import Spreadsheet: the nodes table first, then an edges table, appended to the same workspace). Three graphs share the one nodes table (`node_type` is `sample` or `asv`):

- **detection** edges (sample -- ASV): every non-zero ASV in every sample, lab controls included. `Weight` is the within-sample relative abundance, so trace detections barely pull on a force layout; filter on `rel_abund` / `log10_rel_abund` to show only the non-trace skeleton. `run_ratio_to_max` is the detection's relative abundance over the ASV's maximum in any sample of the same sequencing run (1 = the run's source; ~1e-3 and below = candidate cross-talk).
- **co-occurrence** edges (ASV -- ASV): pairs of ASVs detected in the same samples (all samples, controls included), `Weight` = Jaccard similarity, thresholded by `network_min_shared_samples` and `network_min_jaccard`. Sample nodes have no co-occurrence edges; filter `node_type == asv` in Gephi.
- **similarity** edges (sample -- sample): each sample with at least `network_knn_min_asvs` non-anthropogenic ASVs linked to its `network_knn_k` most similar samples (cosine similarity of sqrt relative abundance on non-anthropogenic reads, each ASV weighted by inverse prevalence so near-ubiquitous ASVs count little), `Weight` = similarity. `cross_system` and `same_run` flag edges between systems and within a sequencing run; ASV nodes have no similarity edges.

Field-sample nodes carry `anthropogenic_read_share` and `nonlocal_read_share`: the share of the sample's reads from ASVs flagged `anthropogenic`, or `other_system` / `not_in_run`, in its system (script 07). They are kept apart because human DNA dominates many samples and would mask cross-system contamination. ASV nodes carry script 07's `local_status` in each system (`status_<system>`), limited to the systems the ASV is detected in with this input.

## Network analysis

The guiding question is whether an ASV's presence is better explained by **ecological** structure (system, site, field replicates) or by **lab** structure (sequencing run, extraction batch, the sample's own negative controls). Three kinds of evidence are produced:

**Detection features** -- one row per (sample, ASV) detection in field samples and field negatives:

- `run_ratio_to_max`: relative abundance over the ASV's maximum in any sample of the same sequencing run. ~1e-3 and below is typical of cross-talk from the run's source sample.
- `own_negative_rel_abund_max`, `log10_ratio_to_own_negatives`: the ASV's maximum relative abundance in the sample's *own* linked negative controls (same field negative, extraction negative, or PCR control group), and the sample's relative abundance against it. A detection at or below its own negatives is not distinguishable from contamination.
- `n_replicate_libraries`, `n_replicates_detected`: for field samples sequenced as several libraries, in how many the ASV was detected. Real eDNA tends to replicate; trace leaks tend not to.

**ASV co-occurrence communities** -- groups of ASVs that tend to be detected in the same samples (Leiden communities on the co-occurrence graph). Each community is tested for association with lab batches vs. ecology (see below). A community tied to a sequencing run or extraction batch, or concentrated in controls, is a candidate *contaminant cohort*.

**Sample similarity communities** -- groups of samples with similar composition (Leiden communities on the k-nearest-neighbour similarity graph), tested the same way, plus per-sample neighbourhood shares (how many of a sample's nearest neighbours share its system, run, and site).

**Batch vs. ecology association.** Runs and systems are confounded (some systems were sequenced on their own runs), so each association is tested against a stratified permutation null:

- *run beyond system*: mutual information with `sequencing_date`, shuffling run labels **within** each system
- *extraction beyond system*: the same for `extraction_date`
- *site beyond run*: mutual information with `site`, shuffling site labels **within** each run

`*_excess` is the observed mutual information minus its permutation mean (a chance-corrected effect size, comparable across labels with different numbers of categories), and `*_p` the permutation p-value.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here, igraph, plotly)

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R)
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

run_dir = here(pipeline_config$run_dir)
cat(paste0("Run_dir: ", run_dir, "\n"))

config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# which detections to build the network from: "raw" (feature table) or "filtered" (script 08)
# network_input = pipeline_config$network_input
network_input = "raw"
if (!(network_input %in% c("raw", "filtered"))) {
  stop("network_input must be 'raw' or 'filtered', not '", network_input, "'")
}

# network export
network_min_shared_samples = pipeline_config$network_min_shared_samples
network_min_jaccard = pipeline_config$network_min_jaccard
network_knn_k = pipeline_config$network_knn_k
network_knn_min_asvs = pipeline_config$network_knn_min_asvs

# network analysis
network_seed = pipeline_config$network_seed
network_leiden_resolution = pipeline_config$network_leiden_resolution
network_n_permutations = pipeline_config$network_n_permutations
network_min_community_size = pipeline_config$network_min_community_size

possible_study_systems = c("brazil", "panama", "pennsylvania", "sierra_nevada")

tax_ranks = c("domain", "kingdom", "phylum", "class", "order", "family", "genus", "species")

# samples whose DNA is field-derived (the evidence pool, as in script 07)
field_classes = c("field_sample", "field_negative")

# input files
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))
classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))
locality_file = here(run_dir, "output", paste0(config$run$name, "_07_asv_locality.csv"))
feature_table_file = here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "sample_table/feature-table.tsv")
controlled_file = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))

# output files, named by input
out_prefix = paste0(config$run$name, "_11_", network_input, "_")
network_nodes_out = here(run_dir, "output", paste0(out_prefix, "network_nodes.csv"))
network_detection_edges_out = here(run_dir, "output", paste0(out_prefix, "network_edges_detection.csv"))
network_cooccurrence_edges_out = here(run_dir, "output", paste0(out_prefix, "network_edges_cooccurrence.csv"))
network_similarity_edges_out = here(run_dir, "output", paste0(out_prefix, "network_edges_similarity.csv"))
detection_features_out = here(run_dir, "output", paste0(out_prefix, "detection_network_features.csv"))
asv_metrics_out = here(run_dir, "output", paste0(out_prefix, "asv_network_metrics.csv"))
sample_metrics_out = here(run_dir, "output", paste0(out_prefix, "sample_network_metrics.csv"))
community_association_out = here(run_dir, "output", paste0(out_prefix, "community_association.csv"))
cytoscape_session_out = here(run_dir, "output", paste0(out_prefix, "networks.cys"))
```

# Load data
```{r}
input_files = c(sample_map_file, classification_file, locality_file,
                if (network_input == "raw") feature_table_file else controlled_file)
missing_inputs = input_files[!file.exists(input_files)]
if (length(missing_inputs) > 0) {
  stop("Missing input(s) for network_input = '", network_input, "':\n\t", paste(missing_inputs, collapse = "\n\t"))
}

# guess_max = Inf: some sample map columns are NA for most rows, and would otherwise be guessed
# as logical and their later values dropped
sample_map = read_csv(sample_map_file, show_col_types = FALSE, guess_max = Inf)
asv_classification = read_csv(classification_file, show_col_types = FALSE)
asv_locality = read_csv(locality_file, show_col_types = FALSE)

# (asv_id, feature_id, reads) for every sample x ASV
if (network_input == "raw") {
  asv_reads = read_tsv(feature_table_file, skip = 1, show_col_types = FALSE) %>%
    rename(asv_id = "#OTU ID") %>%
    pivot_longer(cols = -asv_id, names_to = "feature_id", values_to = "reads")
} else {
  # script 08 clean reads (0 for removed detections)
  asv_reads = read_csv(controlled_file, show_col_types = FALSE) %>%
    transmute(asv_id, feature_id, reads = asv_count_controlled)
}

# fixed row order, so edge orientation and floating-point sums don't depend on the input's order
asv_reads = asv_reads %>%
  arrange(asv_id)
```

# Network export
```{r}
# sample class and labels (same classes as script 07)
sample_classes = sample_map %>%
  transmute(feature_id,
            study_system,
            site,
            sequencing_date,
            sample_class = case_when(
              pcr_control %in% TRUE | extraction_negative_control %in% TRUE | study_system %in% "lab_control" ~ "lab_control",
              !(study_system %in% possible_study_systems) ~ "unknown",
              field_negative_control %in% TRUE ~ "field_negative",
              .default = "field_sample"))

# within-sample relative abundance (samples with zero reads are dropped)
feature_long = asv_reads %>%
  group_by(feature_id) %>%
  mutate(sample_reads = sum(reads)) %>%
  ungroup() %>%
  filter(sample_reads > 0) %>%
  mutate(rel_abund = reads / sample_reads) %>%
  left_join(sample_classes, by = "feature_id")

# every detection, with its run-level ratio and script 07's (asv, system) status
network_detections = feature_long %>%
  filter(reads > 0) %>%
  group_by(asv_id, sequencing_date) %>%
  mutate(run_ratio_to_max = rel_abund / max(rel_abund)) %>%
  ungroup() %>%
  left_join(asv_locality %>% select(asv_id, study_system, local_status),
            by = c("asv_id", "study_system"))

# systems each ASV is detected in with this input, and its dominant system (highest mean relative
# abundance over the system's field-pool samples, as in script 07)
system_n_samples = feature_long %>%
  filter(sample_class %in% field_classes) %>%
  distinct(study_system, feature_id) %>%
  count(study_system, name = "n_system_samples")

asv_systems = network_detections %>%
  filter(sample_class %in% field_classes) %>%
  group_by(asv_id, study_system) %>%
  summarise(rel_abund_sum = sum(rel_abund), .groups = "drop") %>%
  left_join(system_n_samples, by = "study_system") %>%
  mutate(rel_abund_mean = rel_abund_sum / n_system_samples)

asv_system_summary = asv_systems %>%
  group_by(asv_id) %>%
  summarise(n_systems_detected = n(),
            systems_detected = paste(sort(study_system), collapse = ";"),
            dominant_system = study_system[which.max(rel_abund_mean)],
            .groups = "drop") %>%
  left_join(asv_locality %>% select(asv_id, dominant_system = study_system, dominant_status = local_status),
            by = c("asv_id", "dominant_system"))

# sample nodes
sample_load = network_detections %>%
  filter(sample_class %in% field_classes) %>%
  group_by(feature_id) %>%
  summarise(anthropogenic_read_share = sum(reads[local_status %in% "anthropogenic"]) / sum(reads),
            nonlocal_read_share = sum(reads[local_status %in% c("other_system", "not_in_run")]) / sum(reads),
            .groups = "drop")

sample_nodes = feature_long %>%
  distinct(feature_id, study_system, site, sample_class, sample_reads) %>%
  left_join(network_detections %>% count(feature_id, name = "n_asvs"), by = "feature_id") %>%
  left_join(sample_load, by = "feature_id") %>%
  left_join(sample_map %>%
              select(feature_id, region, sequencing_date, extraction_date, edna_sample_id,
                     pcr_control_type, field_negative_control_group,
                     extraction_negative_control_group, pcr_control_group),
            by = "feature_id") %>%
  transmute(Id = feature_id,
            Label = feature_id,
            node_type = "sample",
            sample_class = coalesce(sample_class, "unknown"),
            study_system,
            region,
            site,
            sequencing_date,
            extraction_date,
            edna_sample_id,
            pcr_control_type,
            field_negative_control_group,
            extraction_negative_control_group,
            pcr_control_group,
            sample_reads,
            n_asvs,
            anthropogenic_read_share,
            nonlocal_read_share)

# ASV nodes (status in each detected system as separate columns, NA elsewhere)
asv_status_wide = asv_locality %>%
  semi_join(asv_systems, by = c("asv_id", "study_system")) %>%
  select(asv_id, study_system, local_status) %>%
  pivot_wider(names_from = study_system, values_from = local_status, names_prefix = "status_")

asv_nodes = network_detections %>%
  group_by(asv_id) %>%
  summarise(total_reads = sum(reads),
            n_samples_detected = n(),
            n_runs_detected = n_distinct(sequencing_date, na.rm = TRUE),
            n_field_negatives_detected = sum(sample_class %in% "field_negative"),
            n_lab_controls_detected = sum(sample_class %in% "lab_control"),
            rel_abund_max = max(rel_abund),
            .groups = "drop") %>%
  left_join(asv_classification %>% select(asv_id, taxonomic_rank, all_of(tax_ranks)), by = "asv_id") %>%
  left_join(asv_system_summary, by = "asv_id") %>%
  left_join(asv_status_wide, by = "asv_id") %>%
  mutate(taxon_label = coalesce(if_else(!is.na(species), paste(genus, species), NA_character_),
                                genus, family, order, class, phylum, kingdom, domain, "unclassified"),
         Label = paste0(taxon_label, " [", substr(asv_id, 1, 6), "]"),
         node_type = "asv") %>%
  select(Id = asv_id,
         Label,
         node_type,
         taxon_label,
         taxonomic_rank,
         phylum, class, order, family, genus, species,
         dominant_system,
         dominant_status,
         starts_with("status_"),
         n_systems_detected,
         systems_detected,
         n_samples_detected,
         n_runs_detected,
         n_field_negatives_detected,
         n_lab_controls_detected,
         total_reads,
         rel_abund_max)

nodes = bind_rows(sample_nodes, asv_nodes)

# detection edges (sample -- ASV)
detection_edges = network_detections %>%
  transmute(Source = feature_id,
            Target = asv_id,
            Type = "Undirected",
            Weight = rel_abund,
            reads,
            rel_abund,
            log10_rel_abund = log10(rel_abund),
            run_ratio_to_max,
            sample_class = coalesce(sample_class, "unknown"),
            study_system,
            sequencing_date,
            local_status)

# co-occurrence edges (ASV -- ASV), from shared-sample counts
presence = network_detections %>%
  distinct(asv_id, feature_id) %>%
  mutate(present = 1) %>%
  pivot_wider(names_from = feature_id, values_from = present, values_fill = 0) %>%
  column_to_rownames("asv_id") %>%
  as.matrix()

shared_samples = tcrossprod(presence)
n_detected = diag(shared_samples)

pairs = which(upper.tri(shared_samples) & shared_samples >= network_min_shared_samples, arr.ind = TRUE)

cooccurrence_edges = tibble(Source = rownames(shared_samples)[pairs[, 1]],
                            Target = colnames(shared_samples)[pairs[, 2]],
                            n_shared = shared_samples[pairs]) %>%
  mutate(jaccard = n_shared / (n_detected[Source] + n_detected[Target] - n_shared)) %>%
  filter(jaccard >= network_min_jaccard) %>%
  transmute(Source, Target, Type = "Undirected", Weight = jaccard, jaccard, n_shared)

# similarity edges (sample -- sample): each sample to its k nearest samples by cosine similarity of
# sqrt(relative abundance), on non-anthropogenic reads so samples don't cluster by how much human
# DNA they carry. Each ASV is weighted by inverse prevalence (log(n samples / n samples detected)),
# so near-ubiquitous ASVs (e.g. positive-control leak, unclassified background) don't dominate
anthropogenic_asvs = asv_locality %>%
  filter(local_status == "anthropogenic") %>%
  distinct(asv_id) %>%
  pull(asv_id)

composition = network_detections %>%
  filter(!(asv_id %in% anthropogenic_asvs)) %>%
  group_by(feature_id) %>%
  mutate(p = reads / sum(reads)) %>%
  ungroup() %>%
  select(feature_id, asv_id, p) %>%
  pivot_wider(names_from = asv_id, values_from = p, values_fill = 0) %>%
  column_to_rownames("feature_id") %>%
  as.matrix()

asv_idf = log(n_distinct(network_detections$feature_id) / colSums(composition > 0))
weighted = sweep(sqrt(composition), 2, asv_idf, `*`)
weighted = weighted[rowSums(composition > 0) >= network_knn_min_asvs & rowSums(weighted) > 0, , drop = FALSE]
weighted = weighted / sqrt(rowSums(weighted^2))

sample_similarity = tcrossprod(weighted)
diag(sample_similarity) = NA

sample_knn = map_dfr(rownames(sample_similarity), function(s) {
  sims = sample_similarity[s, ]
  nn = head(order(sims, decreasing = TRUE), network_knn_k)
  tibble(from = s, to = colnames(sample_similarity)[nn], similarity = sims[nn])
}) %>%
  filter(similarity > 0)

sample_attrs = sample_map %>%
  select(feature_id, study_system, sequencing_date)

# undirected: a pair found from both ends is kept once and marked mutual
similarity_edges = sample_knn %>%
  mutate(Source = pmin(from, to),
         Target = pmax(from, to)) %>%
  group_by(Source, Target) %>%
  summarise(similarity = first(similarity),
            mutual = n() == 2,
            .groups = "drop") %>%
  left_join(sample_attrs %>% rename(Source = feature_id, source_system = study_system, source_run = sequencing_date),
            by = "Source") %>%
  left_join(sample_attrs %>% rename(Target = feature_id, target_system = study_system, target_run = sequencing_date),
            by = "Target") %>%
  transmute(Source, Target, Type = "Undirected", Weight = similarity, similarity, mutual,
            cross_system = source_system != target_system,
            same_run = source_run == target_run)

cat("Network input:", network_input, "\n")
cat("Network nodes:", nrow(sample_nodes), "samples,", nrow(asv_nodes), "ASVs\n")
cat("Detection edges:", nrow(detection_edges), "\n")
cat("Co-occurrence edges:", nrow(cooccurrence_edges), "\n")
cat("Similarity edges:", nrow(similarity_edges), "\n")

# empty strings (not "NA") so Gephi reads missing values as missing
write_csv(nodes, network_nodes_out, na = "")
write_csv(detection_edges, network_detection_edges_out, na = "")
write_csv(cooccurrence_edges, network_cooccurrence_edges_out, na = "")
write_csv(similarity_edges, network_similarity_edges_out, na = "")
```

# Build graphs
```{r}
# sample -- ASV detections (bipartite: type TRUE = ASV)
g_detection = graph_from_data_frame(detection_edges, vertices = nodes, directed = FALSE)
V(g_detection)$type = V(g_detection)$node_type == "asv"

# ASV -- ASV co-occurrence, and sample -- sample similarity, each on its own node subset
g_cooccurrence = graph_from_data_frame(cooccurrence_edges, vertices = asv_nodes, directed = FALSE)
g_similarity = graph_from_data_frame(similarity_edges, vertices = sample_nodes, directed = FALSE)

cat("detection graph:", vcount(g_detection), "nodes,", ecount(g_detection), "edges\n")
cat("co-occurrence graph:", sum(degree(g_cooccurrence) > 0), "connected ASVs,", ecount(g_cooccurrence), "edges\n")
cat("similarity graph:", sum(degree(g_similarity) > 0), "connected samples,", ecount(g_similarity), "edges\n")
```

# Helpers
```{r}
# mutual information (nats) between two categorical vectors, ignoring NA pairs
mutual_information = function(x, y) {
  keep = !is.na(x) & !is.na(y)
  p = table(x[keep], y[keep]) / sum(keep)
  expected = outer(rowSums(p), colSums(p))
  nonzero = p > 0
  sum(p[nonzero] * log(p[nonzero] / expected[nonzero]))
}

# shuffle labels within strata
permute_within = function(label, strata) {
  strata = coalesce(as.character(strata), "NA")
  out = label
  for (s in unique(strata)) {
    i = which(strata == s)
    out[i] = label[i][sample.int(length(i))]
  }
  out
}

# association of x with label beyond strata: observed MI vs. a stratified permutation null
association_test = function(x, label, strata, n_perm = network_n_permutations) {
  observed = mutual_information(x, label)
  null = replicate(n_perm, mutual_information(x, permute_within(label, strata)))
  tibble(mi = observed,
         excess = observed - mean(null),
         p = (sum(null >= observed) + 1) / (n_perm + 1))
}

# the three batch vs. ecology tests for a categorical grouping x over samples (rows of `samples`)
batch_vs_ecology = function(x, samples) {
  bind_cols(association_test(x, samples$sequencing_date, samples$study_system) %>% rename_with(~ paste0("run_", .x)),
            association_test(x, samples$extraction_date, samples$study_system) %>% rename_with(~ paste0("extraction_", .x)),
            association_test(x, samples$site, samples$sequencing_date) %>% rename_with(~ paste0("site_", .x)))
}
```

# Detection features
```{r}
# each sample's own negative controls, via the control-group ids shared with them
link_controls = function(group_col, is_control, control_type) {
  groups = sample_map %>%
    select(feature_id, group = all_of(group_col)) %>%
    filter(!is.na(group))
  groups %>%
    inner_join(groups %>%
                 filter(feature_id %in% sample_map$feature_id[is_control]) %>%
                 rename(control_id = feature_id),
               by = "group", relationship = "many-to-many") %>%
    filter(feature_id != control_id) %>%
    transmute(feature_id, control_id, control_type = control_type)
}

control_links = bind_rows(
  link_controls("field_negative_control_group", sample_map$field_negative_control %in% TRUE, "field_negative"),
  link_controls("extraction_negative_control_group", sample_map$extraction_negative_control %in% TRUE, "extraction_negative"),
  link_controls("pcr_control_group", sample_map$pcr_control_type %in% "pcr_negative_control", "pcr_negative"))

n_own_negatives = control_links %>%
  group_by(feature_id) %>%
  summarise(n_own_negatives = n_distinct(control_id),
            own_negative_types_linked = paste(sort(unique(control_type)), collapse = ";"),
            .groups = "drop")

own_negative_evidence = control_links %>%
  inner_join(detection_edges %>% select(control_id = Source, asv_id = Target, control_rel_abund = rel_abund),
             by = "control_id", relationship = "many-to-many") %>%
  group_by(feature_id, asv_id) %>%
  summarise(own_negative_rel_abund_max = max(control_rel_abund),
            own_negative_types_detected = paste(sort(unique(control_type)), collapse = ";"),
            .groups = "drop")

# replicate libraries of the same field sample
replicate_detection = detection_edges %>%
  select(feature_id = Source, asv_id = Target) %>%
  left_join(sample_map %>% select(feature_id, edna_sample_id), by = "feature_id") %>%
  filter(!is.na(edna_sample_id)) %>%
  count(edna_sample_id, asv_id, name = "n_replicates_detected")

n_replicate_libraries = sample_map %>%
  filter(!is.na(edna_sample_id)) %>%
  count(edna_sample_id, name = "n_replicate_libraries")

detection_features = detection_edges %>%
  filter(sample_class %in% field_classes) %>%
  rename(feature_id = Source, asv_id = Target) %>%
  left_join(sample_map %>% select(feature_id, site, extraction_date, edna_sample_id), by = "feature_id") %>%
  left_join(n_own_negatives, by = "feature_id") %>%
  left_join(own_negative_evidence, by = c("feature_id", "asv_id")) %>%
  left_join(n_replicate_libraries, by = "edna_sample_id") %>%
  left_join(replicate_detection, by = c("edna_sample_id", "asv_id")) %>%
  mutate(n_own_negatives = replace_na(n_own_negatives, 0),
         in_own_negatives = !is.na(own_negative_rel_abund_max),
         log10_ratio_to_own_negatives = log10(rel_abund / own_negative_rel_abund_max),
         log10_run_ratio_to_max = log10(run_ratio_to_max)) %>%
  select(feature_id, asv_id, study_system, site, sequencing_date, extraction_date, sample_class,
         reads, rel_abund, local_status,
         run_ratio_to_max, log10_run_ratio_to_max,
         n_own_negatives, own_negative_types_linked, in_own_negatives, own_negative_types_detected,
         own_negative_rel_abund_max, log10_ratio_to_own_negatives,
         edna_sample_id, n_replicate_libraries, n_replicates_detected)

cat("field detections:", nrow(detection_features), "\n")
cat("  with linked negatives:", sum(detection_features$n_own_negatives > 0), "\n")
cat("  also in own negatives:", sum(detection_features$in_own_negatives), "\n")
cat("  in replicated field samples:", sum(detection_features$n_replicate_libraries > 1, na.rm = TRUE), "\n")
```

# ASV co-occurrence communities
```{r}
set.seed(network_seed)

asv_leiden = cluster_leiden(g_cooccurrence,
                            objective_function = "modularity",
                            weights = E(g_cooccurrence)$Weight,
                            resolution = network_leiden_resolution,
                            n_iterations = -1)

asv_community = tibble(asv_id = V(g_cooccurrence)$name,
                       community = membership(asv_leiden),
                       cooccurrence_degree = degree(g_cooccurrence),
                       cooccurrence_strength = strength(g_cooccurrence, weights = E(g_cooccurrence)$Weight)) %>%
  add_count(community, name = "community_size") %>%
  mutate(asv_community = if_else(cooccurrence_degree > 0 & community_size >= network_min_community_size,
                                 paste0("asv_c", community), NA_character_)) %>%
  select(asv_id, asv_community, cooccurrence_degree, cooccurrence_strength)

cat("ASV communities (>=", network_min_community_size, "ASVs):", n_distinct(na.omit(asv_community$asv_community)), "\n")

# field-pool samples, with labels for the association tests
pool_samples = sample_nodes %>%
  filter(sample_class %in% field_classes) %>%
  select(feature_id = Id, study_system, site, sequencing_date, extraction_date)

# per community: which samples contain any member ASV, its control share, and batch vs. ecology
asv_community_detections = detection_edges %>%
  inner_join(asv_community %>% filter(!is.na(asv_community)), by = c("Target" = "asv_id"))

asv_community_association = asv_community_detections %>%
  group_by(asv_community) %>%
  summarise(n_asvs = n_distinct(Target),
            n_samples = n_distinct(Source),
            n_runs = n_distinct(sequencing_date, na.rm = TRUE),
            n_systems = n_distinct(study_system[sample_class %in% field_classes], na.rm = TRUE),
            control_detection_share = mean(sample_class %in% c("field_negative", "lab_control")),
            lab_control_detection_share = mean(sample_class %in% "lab_control"),
            median_rel_abund = median(rel_abund),
            samples = list(unique(Source)),
            .groups = "drop") %>%
  mutate(tests = map(samples, \(s) batch_vs_ecology(pool_samples$feature_id %in% s, pool_samples))) %>%
  unnest(tests) %>%
  select(-samples) %>%
  left_join(asv_community %>%
              filter(!is.na(asv_community)) %>%
              left_join(asv_nodes %>% select(asv_id = Id, Label, total_reads), by = "asv_id") %>%
              group_by(asv_community) %>%
              slice_max(total_reads, n = 5, with_ties = FALSE) %>%
              summarise(top_members = paste(Label, collapse = "; "), .groups = "drop"),
            by = "asv_community") %>%
  mutate(graph = "asv_cooccurrence", .before = 1) %>%
  rename(community = asv_community) %>%
  arrange(desc(run_excess))

asv_community_association
```

# Sample similarity communities
```{r}
set.seed(network_seed)

sample_leiden = cluster_leiden(g_similarity,
                               objective_function = "modularity",
                               weights = E(g_similarity)$Weight,
                               resolution = network_leiden_resolution,
                               n_iterations = -1)

sample_community = tibble(feature_id = V(g_similarity)$name,
                          community = membership(sample_leiden),
                          similarity_degree = degree(g_similarity)) %>%
  add_count(community, name = "community_size") %>%
  mutate(sample_community = if_else(similarity_degree > 0 & community_size >= network_min_community_size,
                                    paste0("sample_c", community), NA_character_)) %>%
  select(feature_id, sample_community, similarity_degree)

# share of each sample's nearest neighbours that share its system / run / site
neighbor_shares = similarity_edges %>%
  select(a = Source, b = Target) %>%
  { bind_rows(., rename(., a = b, b = a)) } %>%
  left_join(sample_map %>% select(a = feature_id, a_system = study_system, a_run = sequencing_date, a_site = site), by = "a") %>%
  left_join(sample_map %>% select(b = feature_id, b_system = study_system, b_run = sequencing_date, b_site = site), by = "b") %>%
  group_by(feature_id = a) %>%
  summarise(n_neighbors = n(),
            neighbor_same_system_share = mean(a_system == b_system, na.rm = TRUE),
            neighbor_same_run_share = mean(a_run == b_run, na.rm = TRUE),
            neighbor_same_site_share = mean(a_site == b_site, na.rm = TRUE),
            .groups = "drop")

# one test over the whole partition (samples in a community, field pool only)
community_samples = pool_samples %>%
  inner_join(sample_community %>% filter(!is.na(sample_community)), by = "feature_id")

sample_partition_association = batch_vs_ecology(community_samples$sample_community, community_samples) %>%
  mutate(graph = "sample_similarity", community = "all (partition)",
         n_samples = nrow(community_samples), .before = 1)

# per community: membership indicator vs. labels
sample_community_association = community_samples %>%
  group_by(community = sample_community) %>%
  summarise(n_samples = n(),
            n_runs = n_distinct(sequencing_date, na.rm = TRUE),
            n_systems = n_distinct(study_system, na.rm = TRUE),
            systems = paste(sort(unique(study_system)), collapse = ";"),
            samples = list(feature_id),
            .groups = "drop") %>%
  mutate(tests = map(samples, \(s) batch_vs_ecology(community_samples$feature_id %in% s, community_samples))) %>%
  unnest(tests) %>%
  select(-samples) %>%
  mutate(graph = "sample_similarity", .before = 1) %>%
  arrange(desc(run_excess))

sample_partition_association
sample_community_association
```

# ASV and sample metrics
```{r}
asv_metrics = asv_nodes %>%
  select(asv_id = Id, Label, taxon_label, taxonomic_rank, dominant_system, dominant_status,
         starts_with("status_"), n_systems_detected, n_samples_detected, n_runs_detected,
         n_field_negatives_detected, n_lab_controls_detected, total_reads, rel_abund_max) %>%
  left_join(asv_classification %>% select(asv_id, short_sequence_flag, contaminant_flag), by = "asv_id") %>%
  left_join(asv_community, by = "asv_id") %>%
  left_join(detection_features %>%
              group_by(asv_id) %>%
              summarise(n_field_detections = n(),
                        share_run_trace = mean(run_ratio_to_max < 1e-3),
                        share_in_own_negatives = mean(in_own_negatives),
                        median_log10_ratio_to_own_negatives = median(log10_ratio_to_own_negatives, na.rm = TRUE),
                        replicate_detection_rate = sum(pmax(n_replicates_detected - 1, 0), na.rm = TRUE) /
                          sum(pmax(n_replicate_libraries - 1, 0), na.rm = TRUE),
                        .groups = "drop"),
            by = "asv_id") %>%
  mutate(lab_control_share = n_lab_controls_detected / n_samples_detected,
         replicate_detection_rate = if_else(is.nan(replicate_detection_rate), NA_real_, replicate_detection_rate))

sample_metrics = sample_nodes %>%
  select(feature_id = Id, sample_class, study_system, site, sequencing_date, extraction_date,
         sample_reads, n_asvs, anthropogenic_read_share, nonlocal_read_share) %>%
  left_join(sample_community, by = "feature_id") %>%
  left_join(neighbor_shares, by = "feature_id") %>%
  left_join(n_own_negatives, by = "feature_id")

detection_features = detection_features %>%
  left_join(asv_community %>% select(asv_id, asv_community), by = "asv_id") %>%
  left_join(sample_community %>% select(feature_id, sample_community), by = "feature_id")

community_association = bind_rows(asv_community_association,
                                  sample_partition_association,
                                  sample_community_association)

write_csv(detection_features, detection_features_out)
write_csv(asv_metrics, asv_metrics_out)
write_csv(sample_metrics, sample_metrics_out)
write_csv(community_association, community_association_out)
```

# Visualize

## ASV communities: batch vs. ecology

One point per ASV co-occurrence community. x is the chance-corrected association with sequencing run (beyond system), y with site (beyond run); size is the number of ASVs, fill the share of the community's detections that are in negatives or lab controls. Contaminant cohorts should sit to the right, low, and dark; ecological communities high and pale.
```{r fig.width=8, fig.height=6}
p_communities = asv_community_association %>%
  mutate(hover = paste0(community, ": ", n_asvs, " ASVs, ", n_samples, " samples, ", n_runs, " runs",
                        "<br>run excess MI: ", signif(run_excess, 2), " (p = ", signif(run_p, 2), ")",
                        "<br>site excess MI: ", signif(site_excess, 2), " (p = ", signif(site_p, 2), ")",
                        "<br>control share: ", round(control_detection_share, 2),
                        "<br>", top_members)) %>%
  ggplot(aes(x = run_excess, y = site_excess, size = n_asvs, fill = control_detection_share, text = hover)) +
  geom_hline(yintercept = 0, color = "#c3c2bd", linewidth = 0.3) +
  geom_vline(xintercept = 0, color = "#c3c2bd", linewidth = 0.3) +
  geom_point(shape = 21, color = "white", stroke = 0.4, alpha = 0.9) +
  scale_fill_gradient(low = "#cde2fb", high = "#0d366b", limits = c(0, 1), name = "control share") +
  scale_size_continuous(range = c(2, 10), name = "ASVs") +
  labs(x = "association with sequencing run (excess MI, beyond system)",
       y = "association with site (excess MI, beyond run)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_communities, tooltip = "text")
```

## Detection features by GBIF status

Distribution of the cross-talk and own-negative features across script 07's `local_status`. If the features carry signal, `other_system` / `not_in_run` detections should sit lower than `local` ones.
```{r fig.width=10, fig.height=5}
detection_features %>%
  filter(local_status %in% c("local", "other_system", "not_in_run", "local_genus_only")) %>%
  transmute(local_status,
            `log10 run ratio to max` = log10_run_ratio_to_max,
            `log10 ratio to own negatives` = log10_ratio_to_own_negatives) %>%
  pivot_longer(-local_status, names_to = "feature", values_to = "value") %>%
  filter(is.finite(value)) %>%
  ggplot(aes(x = value, y = local_status)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.4, fill = "#cde2fb", color = "#52514e", linewidth = 0.3) +
  facet_wrap(~ feature, scales = "free_x") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

# Cytoscape (optional)

Sends the co-occurrence and similarity graphs, with community memberships, to a running Cytoscape (via RCy3) and saves a session. Skipped when RCy3 is not installed or Cytoscape is not open. Nothing here feeds the outputs above.
```{r}
cytoscape_available = requireNamespace("RCy3", quietly = TRUE) &&
  tryCatch({ RCy3::cytoscapePing(); TRUE }, error = function(e) FALSE)

if (cytoscape_available) {
  system_colors = c(brazil = "#eda100", panama = "#e87ba4", pennsylvania = "#008300", sierra_nevada = "#4a3aa7")

  # ASV co-occurrence, colored by community
  V(g_cooccurrence)$asv_community = asv_community$asv_community[match(V(g_cooccurrence)$name, asv_community$asv_id)]
  g_cooc_connected = induced_subgraph(g_cooccurrence, degree(g_cooccurrence) > 0)
  RCy3::createNetworkFromIgraph(g_cooc_connected, title = "ASV co-occurrence", collection = config$run$name)
  RCy3::createVisualStyle("asv_cooccurrence", defaults = list(NODE_SHAPE = "ellipse", EDGE_TRANSPARENCY = 120))
  RCy3::setNodeColorMapping("asv_community", mapping.type = "d", style.name = "asv_cooccurrence")
  RCy3::setNodeSizeMapping("total_reads", mapping.type = "c", style.name = "asv_cooccurrence")
  RCy3::setNodeLabelMapping("Label", style.name = "asv_cooccurrence")
  RCy3::setVisualStyle("asv_cooccurrence")
  RCy3::layoutNetwork("force-directed")

  # sample similarity, colored by system
  V(g_similarity)$sample_community = sample_community$sample_community[match(V(g_similarity)$name, sample_community$feature_id)]
  g_sim_connected = induced_subgraph(g_similarity, degree(g_similarity) > 0)
  RCy3::createNetworkFromIgraph(g_sim_connected, title = "sample similarity", collection = config$run$name)
  RCy3::createVisualStyle("sample_similarity", defaults = list(NODE_SHAPE = "ellipse", EDGE_TRANSPARENCY = 120))
  RCy3::setNodeColorMapping("study_system", names(system_colors), unname(system_colors),
                            mapping.type = "d", style.name = "sample_similarity")
  RCy3::setEdgeLineWidthMapping("Weight", c(0, 1), c(0.5, 4), style.name = "sample_similarity")
  RCy3::setVisualStyle("sample_similarity")
  RCy3::layoutNetwork("force-directed")

  RCy3::saveSession(cytoscape_session_out)
  cat("Cytoscape session saved:", cytoscape_session_out, "\n")
} else {
  cat("Cytoscape not available (RCy3 not installed or Cytoscape not running) -- skipped.\n")
}
```
