---
title: "ASV fingerprints"
author: "Cob Staines"
format:
  html:
    embed-resources: true
    page-layout: full
    code-fold: true
    code-summary: "Code"
execute:
  warning: false
  message: false
---

# Description

An interactive key for spotting contamination mechanisms by eye. Each ASV gets one row of panels, in a fixed column order: field samples and field negatives for each study system, then the lab controls (extraction negatives, PCR negatives, PCR positives). In each panel:

- **points** are the libraries the ASV was detected in, at their within-sample relative abundance (log10 scale, shared by every panel and every ASV). Relative abundance is over all of the sample's raw reads, as in scripts 07 and 11.
- **the bar under the panel** is the share of the group's libraries the ASV was detected in (`k/n` below it), so non-detections can be read too.

Hovering a point shows the library, and fades every point in the row that is not from the same sequencing run: cross-talk shows up as a trace smear sharing a run with a high-abundance source.

Reads are **raw** (the denoised feature table, via script 08's `asv_count_raw`), so the key shows every detection before any filtering. It only displays, it writes nothing. It needs scripts 03, 06, 07 and 08.

**Signal** is DNA that was in the environment at the sampled site when the sample was taken. **Contamination** is DNA introduced by the sampling and lab process -- by the sampler or gear in the field, during extraction or PCR, or by sequencing cross-talk. DNA can be physically in the sample and still be contamination: what counts is whether it came from the environment or from the process. Human DNA is the clearest case: from the environment (e.g. sewage, people upstream) it is signal; from the sampler's hands it is contamination. Field negatives receive the same handling as samples, so DNA at levels comparable to the field negatives is consistent with handling. Use `uncertain` when the pattern cannot tell the sources apart.

Patterns to look for:

| pattern | likely mechanism |
|---|---|
| dense, high cloud in one system; sparse or absent in its field negatives and the lab columns | real local taxon |
| high in one system, plus a faint smear (~1e-4 and below) in other systems / controls, sharing that run | cross-talk / tag-jumping |
| thin presence across every column, extraction negatives detected as often as field samples or more | reagent (kitome) contamination |
| high in PCR positives, trace elsewhere | positive-control leak |
| concentrated in one system's field negatives and samples, absent in lab controls | field / gear carryover |

# Fingerprints

```{=html}
<style>
#fp-app { font-size: 13px; }
#fp-controls, #fp-sortbar { display: flex; flex-wrap: wrap; gap: 6px 14px; align-items: center; margin-bottom: 6px; }
#fp-controls label, #fp-sortbar label { display: flex; gap: 4px; align-items: center; margin: 0; }
#fp-controls input[type=text] { width: 16em; }
#fp-controls input[type=number] { width: 4.5em; }
#fp-controls select, #fp-controls input, #fp-sortbar select, #fp-sortbar input { font-size: 12px; padding: 1px 4px; }
#fp-body { display: grid; grid-template-columns: minmax(0, 1fr) 165px; gap: 12px; align-items: start; }
#fp-legend { position: sticky; top: 8px; display: flex; flex-direction: column; gap: 3px; font-size: 12px; color: #444;
  border-left: 1px solid #ddd; padding: 2px 0 2px 10px; }
#fp-legend .head { font-weight: 600; margin-top: 6px; }
#fp-legend .head:first-child { margin-top: 0; }
#fp-legend .lk { cursor: default; border-radius: 3px; padding: 0 3px; margin-left: -3px; }
#fp-legend .lk:hover { background: #eef1f5; }
#fp-legend .sw { display: inline-block; width: 10px; height: 10px; border-radius: 50%; margin-right: 4px; vertical-align: -1px; }
.fp-pager { display: flex; gap: 8px; align-items: center; margin: 6px 0; }
.fp-pager button { font-size: 12px; padding: 0 8px; }
#fp-header-wrap { position: sticky; top: 0; z-index: 5; background: #fff; border-bottom: 1px solid #ddd; }
#fp-header-wrap svg, .fp-row svg { display: block; width: 100%; height: auto; }
.fp-row { border-bottom: 1px solid #eee; padding: 4px 0 2px; }
.fp-meta { display: flex; flex-wrap: wrap; gap: 3px 8px; align-items: baseline; line-height: 1.4; }
.fp-taxon { font-weight: 600; font-size: 14px; }
.fp-sub { color: #666; font-size: 12px; }
.fp-id, .fp-seq { cursor: pointer; font-size: 11px; color: #555; background: #f2f2f2; padding: 0 4px; border-radius: 3px; }
.fp-id:hover, .fp-seq:hover { background: #e3e3e3; }
.fp-badge { font-size: 11px; padding: 0 5px; border-radius: 8px; border: 1px solid; white-space: nowrap; }
.fp-row.hl circle.pt:not(.on) { opacity: 0.08; }
.fp-row.hl circle.pt.on { stroke: #000; stroke-width: 0.8; }
#fp-tooltip { position: fixed; pointer-events: none; z-index: 10; background: rgba(255,255,255,0.97); border: 1px solid #bbb;
  border-radius: 4px; padding: 5px 8px; font-size: 12px; line-height: 1.35; box-shadow: 0 2px 6px rgba(0,0,0,0.15); display: none; max-width: 340px; }
#fp-tooltip b { font-weight: 600; }
#fp-empty { padding: 20px; color: #666; }
#fp-labelbar { margin-bottom: 6px; }
#fp-labelbar button { font-size: 12px; padding: 1px 10px; }
#fp-labelpanel { display: none; border: 1px solid #c9d1dc; border-radius: 6px; padding: 6px 10px; margin-bottom: 8px; background: #f8fafc; }
#fp-app.labelling #fp-labelpanel { display: block; }
#fp-app.labelling #fp-controls, #fp-app.labelling #fp-sortbar, #fp-app.labelling .fp-pager, #fp-app.labelling #fp-labelbar { display: none; }
#fp-labelpanel .lrow { display: flex; flex-wrap: wrap; gap: 6px 12px; align-items: center; margin: 4px 0; }
#fp-labelpanel button { font-size: 12px; padding: 1px 8px; }
#fp-labelpanel button.lab { font-size: 13px; padding: 2px 12px; border: 1.5px solid; border-radius: 4px; background: #fff; }
#fp-labelpanel .lab-signal { border-color: #4e79a7; color: #4e79a7; }
#fp-labelpanel .lab-contamination { border-color: #e15759; color: #e15759; }
#fp-labelpanel .lab-uncertain { border-color: #8a8a8a; color: #6a6a6a; }
#fp-labelpanel button.lab.on { color: #fff; }
#fp-labelpanel .lab-signal.on { background: #4e79a7; }
#fp-labelpanel .lab-contamination.on { background: #e15759; }
#fp-labelpanel .lab-uncertain.on { background: #8a8a8a; }
#fp-labelpanel .warn { color: #b3261e; font-weight: 600; }
#fp-labelpanel input[type=text] { font-size: 12px; padding: 1px 4px; }
</style>

<div id="fp-app">
  <div id="fp-labelbar"></div>
  <div id="fp-labelpanel"></div>
  <div id="fp-controls">
    <label>Search <input type="text" id="fp-q" placeholder="taxon, family, ASV id…"></label>
    <label>Class <select id="fp-class"></select></label>
    <label>Order <select id="fp-order"></select></label>
    <label>Family <select id="fp-family"></select></label>
    <label>Status (any system) <select id="fp-status"><option value="">any</option></select></label>
    <label>Detected in <select id="fp-detin"><option value="">any group</option></select></label>
    <label>Min detections <input type="number" id="fp-mindet" min="0" value="1"></label>
    <label><input type="checkbox" id="fp-known"> known contaminants only</label>
  </div>
  <div id="fp-sortbar">
    <label>Sort <select id="fp-sort">
      <option value="reads">total reads</option>
      <option value="detections">detections</option>
      <option value="systems">systems detected</option>
      <option value="ext_neg">extraction neg detection rate</option>
      <option value="field_neg">field neg detection rate</option>
      <option value="lab_max">max rel. abund. in lab controls</option>
      <option value="rel_max">max rel. abund.</option>
      <option value="taxon">taxon</option>
      <option value="length">sequence length</option>
    </select></label>
    <label>Direction <select id="fp-desc">
      <option value="desc">descending</option>
      <option value="asc">ascending</option>
    </select></label>
    <label>Per page <select id="fp-pagesize">
      <option>10</option><option selected>25</option><option>50</option><option>100</option>
    </select></label>
  </div>
  <div id="fp-body">
    <div id="fp-main">
      <div class="fp-pager" id="fp-pager-top"></div>
      <div id="fp-header-wrap"><svg id="fp-header"></svg></div>
      <div id="fp-rows"></div>
      <div class="fp-pager" id="fp-pager-bottom"></div>
    </div>
    <div id="fp-legend"></div>
  </div>
  <div id="fp-tooltip"></div>
</div>

<script>
// runs after the page (and the embedded data, further down) has loaded
document.addEventListener("DOMContentLoaded", function () {
  const D = window.FP_DATA;
  const A = D.asvs, S = D.samples, det = D.detections;
  const nA = A.asv_id.length, nD = det.asv.length;

  // ---- groups / geometry ----------------------------------------------------------------------
  const groups = D.groups.group.map((g, i) => ({
    id: g, label: D.groups.label[i], sublabel: D.groups.sublabel[i],
    kind: D.groups.kind[i], system: D.groups.study_system[i], n: 0 }));
  for (let s = 0; s < S.feature_id.length; s++) groups[S.group[s]].n++;
  const groupsOfKind = k => groups.map((g, i) => g.kind === k ? i : -1).filter(i => i >= 0);
  const fieldIdx = groupsOfKind("field_sample"), fieldNegIdx = groupsOfKind("field_negative");
  const extIdx = groups.findIndex(g => g.id === "extraction_negative");
  const nFieldNeg = fieldNegIdx.reduce((t, g) => t + groups[g].n, 0);

  const G = { left: 40, right: 6, panelW: 92, gap: 5, groupGap: 14, top: 3, plotH: 104, barGap: 4, barH: 7, countH: 13 };
  const px = [];
  { let x = G.left;
    groups.forEach((g, i) => {
      if (i > 0) { x += G.gap; if (groups[i - 1].system !== g.system) x += G.groupGap; }
      px.push(x); x += G.panelW;
    });
    G.W = x + G.right; }
  G.plotTop = G.top; G.plotBottom = G.top + G.plotH;
  G.barY = G.plotBottom + G.barGap;
  G.H = G.barY + G.barH + G.countH;

  let minRel = 1;
  for (let d = 0; d < nD; d++) if (det.rel_abund[d] > 0 && det.rel_abund[d] < minRel) minRel = det.rel_abund[d];
  const yMinLog = Math.min(-3, Math.floor(Math.log10(minRel)));
  const y = r => G.plotTop + 3 + (-Math.log10(r) / -yMinLog) * (G.plotH - 6);

  // deterministic jitter per library, so points don't move between redraws
  const jitter = S.feature_id.map((f, i) => { let h = 2166136261; for (const c of f) { h ^= c.charCodeAt(0); h = Math.imul(h, 16777619); } return ((h >>> 0) / 4294967295) - 0.5; });

  // ---- colours --------------------------------------------------------------------------------
  const runPalette = ["#4e79a7", "#f28e2b", "#e15759", "#76b7b2", "#59a14f", "#edc948", "#b07aa1", "#ff9da7",
                      "#9c755f", "#17becf", "#7f7f7f", "#bcbd22"];
  const runColor = D.runs.map((r, i) => runPalette[i % runPalette.length]);
  const statusColors = { local: "#2e7d32", local_genus_only: "#7cb342", other_system: "#ef6c00", not_in_run: "#c62828",
                         anthropogenic: "#7b1fa2", unassessable: "#757575" };

  // ---- per-ASV detections and stats -----------------------------------------------------------
  const asvDet = Array.from({ length: nA }, () => []);
  for (let d = 0; d < nD; d++) asvDet[det.asv[d]].push(d);
  // within each ASV, draw low-abundance points last so trace detections stay visible
  asvDet.forEach(ds => ds.sort((a, b) => det.rel_abund[b] - det.rel_abund[a]));

  const asvStats = asvDet.map(ds => {
    const per = new Array(groups.length).fill(0);
    let n = 0, maxRel = 0, maxLab = 0;
    for (const d of ds) {
      const g = S.group[det.sample[d]], r = det.rel_abund[d];
      per[g]++; n++;
      if (r > maxRel) maxRel = r;
      if (groups[g].kind === "lab_control" && r > maxLab) maxLab = r;
    }
    return { per, n, maxRel, maxLab,
             nSystems: fieldIdx.filter(g => per[g] > 0).length,
             extRate: extIdx >= 0 && groups[extIdx].n ? per[extIdx] / groups[extIdx].n : 0,
             fieldNegRate: nFieldNeg ? fieldNegIdx.reduce((t, g) => t + per[g], 0) / nFieldNeg : 0 };
  });
  const stats = () => asvStats;

  const statusCols = D.study_systems.map(s => A["status_" + s] || new Array(nA).fill(null));
  const searchText = A.asv_id.map((id, a) =>
    [A.label[a], id, A.class[a], A.order[a], A.family[a], A.known_contaminant[a]]
      .filter(Boolean).join(" ").toLowerCase());

  // ---- controls -------------------------------------------------------------------------------
  const state = { q: "", class: "", order: "", family: "", sort: "reads", desc: true, status: "", detIn: "", minDet: 1, knownOnly: false,
                  page: 0, pageSize: 25 };
  const $ = id => document.getElementById(id);
  const esc = s => String(s).replace(/[&<>"]/g, c => ({ "&": "&amp;", "<": "&lt;", ">": "&gt;", '"': "&quot;" }[c]));

  const statuses = [...new Set(statusCols.flat().filter(Boolean))].sort();
  statuses.forEach(s => $("fp-status").insertAdjacentHTML("beforeend", `<option value="${s}">${s}</option>`));
  groups.forEach((g, i) => $("fp-detin").insertAdjacentHTML("beforeend",
    `<option value="${i}">${g.label === "Lab" ? "" : g.label + " "}${g.sublabel}</option>`));

  const bind = (id, key, ev, parse) => $(id).addEventListener(ev, e => {
    state[key] = parse(e.target); state.page = 0; render(); });
  bind("fp-q", "q", "input", t => t.value);

  // taxonomy filters: each list only offers values under the ranks chosen above it, with ASV counts
  const taxonRanks = ["class", "order", "family"];
  const NONE = "(unclassified)";
  const taxonValue = (rank, a) => A[rank][a] || NONE;
  const taxonMatch = (a, upTo) => taxonRanks.slice(0, upTo).every(r => !state[r] || taxonValue(r, a) === state[r]);
  function fillTaxonSelects(from) {
    taxonRanks.forEach((rank, k) => {
      if (k < from) return;
      const counts = new Map();
      for (let a = 0; a < nA; a++) if (taxonMatch(a, k)) counts.set(taxonValue(rank, a), (counts.get(taxonValue(rank, a)) || 0) + 1);
      const values = [...counts.keys()].sort((x, y) => (x === NONE) - (y === NONE) || x.localeCompare(y));
      if (!counts.has(state[rank])) state[rank] = "";
      $("fp-" + rank).innerHTML = `<option value="">any</option>` + values.map(v =>
        `<option value="${esc(v)}"${v === state[rank] ? " selected" : ""}>${esc(v)} (${counts.get(v)})</option>`).join("");
    });
  }
  taxonRanks.forEach((rank, k) => $("fp-" + rank).addEventListener("change", e => {
    state[rank] = e.target.value;
    taxonRanks.slice(k + 1).forEach(r => { state[r] = ""; });
    fillTaxonSelects(k + 1);
    state.page = 0;
    render();
  }));
  fillTaxonSelects(0);
  // a new sort field starts in its natural direction (taxon A-Z, everything else largest first)
  $("fp-sort").addEventListener("change", e => {
    state.sort = e.target.value;
    state.desc = !ascendingByDefault.has(state.sort);
    $("fp-desc").value = state.desc ? "desc" : "asc";
    state.page = 0;
    render();
  });
  bind("fp-desc", "desc", "change", t => t.value === "desc");
  bind("fp-status", "status", "change", t => t.value);
  bind("fp-detin", "detIn", "change", t => t.value);
  bind("fp-mindet", "minDet", "input", t => +t.value || 0);
  bind("fp-known", "knownOnly", "change", t => t.checked);
  bind("fp-pagesize", "pageSize", "change", t => +t.value);

  // ---- filter / sort --------------------------------------------------------------------------
  // sort keys; the direction comes from the order drop-down, unknown values go last, ties by total reads
  const sortKeys = {
    reads: (a, st) => A.total_reads[a],
    detections: (a, st) => st[a].n,
    systems: (a, st) => st[a].nSystems,
    ext_neg: (a, st) => st[a].extRate,
    field_neg: (a, st) => st[a].fieldNegRate,
    lab_max: (a, st) => st[a].maxLab,
    rel_max: (a, st) => st[a].maxRel,
    taxon: (a, st) => A.label[a],
    length: (a, st) => A.sequence_length[a],
  };
  const ascendingByDefault = new Set(["taxon"]);

  function selection() {
    const st = stats(), q = state.q.trim().toLowerCase(), out = [];
    for (let a = 0; a < nA; a++) {
      const s = st[a];
      if (s.n < state.minDet) continue;
      if (q && !searchText[a].includes(q)) continue;
      if (!taxonMatch(a, taxonRanks.length)) continue;
      if (state.status && !statusCols.some(col => col[a] === state.status)) continue;
      if (state.knownOnly && !A.known_contaminant[a]) continue;
      if (state.detIn !== "" && !(s.per[+state.detIn] > 0)) continue;
      out.push(a);
    }
    const key = sortKeys[state.sort], dir = state.desc ? -1 : 1;
    return out.sort((a, b) => {
      const ka = key(a, st), kb = key(b, st);
      if (ka == null || kb == null) return (ka == null) - (kb == null);
      const c = typeof ka === "string" ? ka.localeCompare(kb) : ka - kb;
      return dir * c || A.total_reads[b] - A.total_reads[a];
    });
  }

  // ---- drawing --------------------------------------------------------------------------------
  const fmtInt = n => n.toLocaleString();
  const fmtRel = r => r >= 0.01 ? (100 * r).toFixed(1) + "%" : r.toExponential(1);
  const decLabel = k => k === 0 ? "1" : "1e" + k;

  function panelsBackground() {
    let s = "";
    groups.forEach((g, i) => {
      s += `<rect x="${px[i]}" y="${G.plotTop}" width="${G.panelW}" height="${G.plotH}" fill="${g.kind === "lab_control" ? "#f7f4ef" : g.kind === "field_negative" ? "#f3f5f8" : "#f8f9fa"}"/>`;
    });
    for (let k = 0; k >= yMinLog; k--) {
      const yy = y(Math.pow(10, k));
      groups.forEach((g, i) => { s += `<line x1="${px[i]}" x2="${px[i] + G.panelW}" y1="${yy}" y2="${yy}" stroke="#e2e4e8" stroke-width="0.6"/>`; });
      if (k % 2 === 0) s += `<text x="${G.left - 4}" y="${yy + 3}" text-anchor="end" font-size="9" fill="#777">${decLabel(k)}</text>`;
    }
    s += `<text x="${G.left - 4}" y="${G.barY + G.barH - 1}" text-anchor="end" font-size="9" fill="#777">det.</text>`;
    return s;
  }
  let background = panelsBackground();

  function drawHeader() {
    const h = 40;
    let s = "";
    groups.forEach((g, i) => {
      const cx = px[i] + G.panelW / 2;
      const showLabel = i === 0 || groups[i - 1].system !== g.system;
      if (showLabel) s += `<text x="${px[i]}" y="12" font-size="11" font-weight="600" fill="#333">${esc(g.label)}</text>`;
      s += `<text x="${cx}" y="25" font-size="10.5" text-anchor="middle" fill="#444">${esc(g.sublabel)}</text>`;
      s += `<text x="${cx}" y="37" font-size="9.5" text-anchor="middle" fill="#888">n = ${g.n}</text>`;
    });
    s += `<text x="${G.left - 4}" y="37" font-size="9" text-anchor="end" fill="#777">rel. ab.</text>`;
    const svg = $("fp-header");
    svg.setAttribute("viewBox", `0 0 ${G.W} ${h}`);
    svg.innerHTML = s;
  }

  function badge(text, color, title) {
    return `<span class="fp-badge" style="color:${color};border-color:${color}"${title ? ` title="${esc(title)}"` : ""}>${esc(text)}</span>`;
  }

  // target: a detection index to highlight, for blind labelling (hides model information)
  function rowHtml(a, st, target) {
    const s = st[a], blind = target !== undefined;
    let meta = `<span class="fp-taxon">${esc(A.label[a])}</span>`;
    const lineage = [A.class[a], A.order[a], A.family[a]].filter(Boolean).join(" / ");
    const identity = A.p_identical_max[a] != null ? ` · ${(100 * A.p_identical_max[a]).toFixed(1)}% identity` : "";
    meta += `<span class="fp-sub" title="max percent identity of the taxonomic classification (script 06)">${esc(A.taxonomic_rank[a] || "")}${lineage ? " · " + esc(lineage) : ""}${identity}</span>`;
    meta += `<code class="fp-id" data-id="${A.asv_id[a]}" title="click to copy full ASV id">${A.asv_id[a].slice(0, 8)}</code>`;
    if (A.sequence[a]) meta += `<code class="fp-seq" data-a="${a}" title="click to copy the sequence">${A.sequence_length[a]} bp</code>`;
    meta += `<span class="fp-sub">${A.sequence[a] ? "" : "? bp · "}${fmtInt(A.total_reads[a])} reads · ${asvDet[a].length} detections</span>`;
    // status tags (GBIF status per system, then flags), on a line of their own
    let status = "";
    D.study_systems.forEach((sys, k) => {
      const v = statusCols[k][a];
      if (v) status += badge(`${D.system_codes[sys]}: ${v}`, statusColors[v] || "#555", D.system_labels[sys]);
    });
    if (A.short_sequence_flag[a]) status += badge("short sequence", "#9c755f");
    if (A.contaminant_flag[a]) status += badge("contaminant library" + (A.contaminant_id[a] ? ": " + A.contaminant_id[a] : ""), "#b07aa1");
    if (A.positive_control_component[a]) status += badge("PCR+ component", "#f28e2b");
    if (A.known_contaminant[a]) status += badge("known contaminant", "#000", A.known_contaminant[a]);

    let svg = background;
    groups.forEach((g, i) => {
      const k = s.per[i], rate = g.n ? k / g.n : 0;
      const bw = k > 0 ? Math.max(1.5, rate * G.panelW) : 0;
      svg += `<rect x="${px[i]}" y="${G.barY}" width="${G.panelW}" height="${G.barH}" fill="#e6e6e6"/>`;
      if (bw) svg += `<rect x="${px[i]}" y="${G.barY}" width="${bw}" height="${G.barH}" fill="#37474f"/>`;
      svg += `<text x="${px[i] + G.panelW / 2}" y="${G.barY + G.barH + 10}" font-size="9.5" text-anchor="middle" fill="${k ? "#333" : "#aaa"}">${k}/${g.n}</text>`;
    });
    for (const d of asvDet[a]) {
      const smp = det.sample[d], g = S.group[smp];
      const cx = px[g] + G.panelW / 2 + jitter[smp] * G.panelW * 0.8;
      svg += `<circle class="pt" data-d="${d}" cx="${cx.toFixed(1)}" cy="${y(det.rel_abund[d]).toFixed(1)}" r="2.6" fill="${runColor[S.run[smp]]}" fill-opacity="0.8"/>`;
    }
    if (blind) {
      const smp = det.sample[target], g = S.group[smp];
      const cx = px[g] + G.panelW / 2 + jitter[smp] * G.panelW * 0.8, cy = y(det.rel_abund[target]);
      svg += `<rect x="${px[g]}" y="${G.plotTop}" width="${G.panelW}" height="${G.plotH}" fill="none" stroke="#000" stroke-width="1"/>`;
      svg += `<circle cx="${cx.toFixed(1)}" cy="${cy.toFixed(1)}" r="7" fill="none" stroke="#000" stroke-width="1.8" pointer-events="none"/>`;
    }
    return `<div class="fp-row"><div class="fp-meta">${meta}</div>${status ? `<div class="fp-meta">${status}</div>` : ""}
      <svg viewBox="0 0 ${G.W} ${G.H}">${svg}</svg></div>`;
  }

  function drawLegend() {
    let s = "";
    if (state.labelMode) {
      s += `<span class="head">Sequencing run</span>`;
      D.runs.forEach((r, i) => { s += `<span class="lk" data-k="run:${i}"><span class="sw" style="background:${runColor[i]}"></span>${esc(r)}</span>`; });
      s += `<span class="head">Symbols</span><span><svg width="16" height="16"><circle cx="8" cy="8" r="6" fill="none" stroke="#000" stroke-width="1.8"/></svg> edge to label</span>`;
      $("fp-legend").innerHTML = s;
      return;
    }
    const nRun = D.runs.map(() => 0);
    S.run.forEach(r => nRun[r]++);
    s += `<span class="head">Sequencing run</span>`;
    D.runs.forEach((r, i) => { s += `<span class="lk" data-k="run:${i}"><span class="sw" style="background:${runColor[i]}"></span>${esc(r)} (${nRun[i]})</span>`; });
    $("fp-legend").innerHTML = s;
  }

  function drawPager(total) {
    const pages = Math.max(1, Math.ceil(total / state.pageSize));
    state.page = Math.min(state.page, pages - 1);
    const html = `<button data-p="prev" ${state.page === 0 ? "disabled" : ""}>‹ prev</button>
      <span>page ${state.page + 1} of ${pages} · ${fmtInt(total)} ASVs</span>
      <button data-p="next" ${state.page >= pages - 1 ? "disabled" : ""}>next ›</button>`;
    $("fp-pager-top").innerHTML = html;
    $("fp-pager-bottom").innerHTML = html;
  }
  document.querySelectorAll(".fp-pager").forEach(el => el.addEventListener("click", e => {
    const p = e.target.dataset && e.target.dataset.p;
    if (!p) return;
    state.page += p === "next" ? 1 : -1;
    render();
    if (e.currentTarget.id === "fp-pager-bottom") $("fp-pager-top").scrollIntoView();
  }));

  function render() {
    if (state.labelMode) return renderLabelling();
    const sel = selection(), st = stats();
    drawPager(sel.length);
    drawLegend();
    const page = sel.slice(state.page * state.pageSize, (state.page + 1) * state.pageSize);
    $("fp-rows").innerHTML = page.length ? page.map(a => rowHtml(a, st)).join("") : `<div id="fp-empty">No ASVs match.</div>`;
  }

  // ---- interaction ----------------------------------------------------------------------------
  const tip = $("fp-tooltip");
  $("fp-rows").addEventListener("mouseover", e => {
    const c = e.target.closest("circle.pt");
    if (!c) return;
    const d = +c.dataset.d, smp = det.sample[d], g = groups[S.group[smp]];
    const row = c.closest(".fp-row");
    row.classList.add("hl");
    row.querySelectorAll("circle.pt").forEach(p => p.classList.toggle("on", S.run[det.sample[+p.dataset.d]] === S.run[smp]));
    tip.innerHTML = `<b>${esc(S.feature_id[smp])}</b><br>
      ${esc(g.label === "Lab" ? g.sublabel : g.label + " " + g.sublabel)}${S.site[smp] ? " · " + esc(S.site[smp]) : ""}<br>
      run ${esc(S.run[smp] != null ? D.runs[S.run[smp]] : "?")} · extracted ${esc(S.extraction_date[smp] || "?")}<br>
      ${fmtInt(det.reads[d])} / ${fmtInt(S.sample_reads[smp])} reads (${fmtRel(det.rel_abund[d])})`;
    tip.style.display = "block";
  });
  $("fp-rows").addEventListener("mousemove", e => {
    if (tip.style.display !== "block") return;
    const pad = 14, w = tip.offsetWidth, h = tip.offsetHeight;
    tip.style.left = Math.min(e.clientX + pad, window.innerWidth - w - 4) + "px";
    tip.style.top = (e.clientY + pad + h > window.innerHeight ? e.clientY - h - pad : e.clientY + pad) + "px";
  });
  $("fp-rows").addEventListener("mouseout", e => {
    const c = e.target.closest("circle.pt");
    if (!c) return;
    c.closest(".fp-row").classList.remove("hl");
    tip.style.display = "none";
  });
  // hovering a key entry highlights its points in every row on the page
  const legendMatch = {
    run: k => d => S.run[det.sample[d]] === k,
  };
  $("fp-legend").addEventListener("mouseover", e => {
    const item = e.target.closest(".lk");
    if (!item) return;
    const [type, k] = item.dataset.k.split(":");
    const match = legendMatch[type](+k);
    document.querySelectorAll("#fp-rows .fp-row").forEach(row => {
      row.classList.add("hl");
      row.querySelectorAll("circle.pt").forEach(p => p.classList.toggle("on", match(+p.dataset.d)));
    });
  });
  $("fp-legend").addEventListener("mouseout", e => {
    if (!e.target.closest(".lk")) return;
    document.querySelectorAll("#fp-rows .fp-row.hl").forEach(row => row.classList.remove("hl"));
  });
  // click the ASV id or the sequence length to copy the full id or the sequence
  $("fp-rows").addEventListener("click", e => {
    const el = e.target.closest(".fp-id, .fp-seq");
    if (!el) return;
    const text = el.classList.contains("fp-seq") ? A.sequence[+el.dataset.a] : el.dataset.id;
    navigator.clipboard?.writeText(text).then(() => {
      const t = el.textContent; el.textContent = "copied"; setTimeout(() => { el.textContent = t; }, 800);
    });
  });

  // ---- blind labelling of the validation sample (script 14) -----------------------------------
  // Shows one sampled edge at a time, with model information hidden. Labels autosave in
  // this browser and are exported as a CSV to the validation folder; a re-render reads them back.
  const V = D.validation && D.validation.validation_id ? D.validation : null;
  const labelNames = ["signal", "contamination", "uncertain"];
  const storeKey = "fp-validation-labels:" + D.run_name;
  const exportKey = "fp-validation-exported:" + D.run_name;
  const fileLabels = {};
  let labels = {};
  let vi = 0;
  const store = {
    get: k => { try { return localStorage.getItem(k); } catch (e) { return null; } },
    set: (k, v) => { try { localStorage.setItem(k, v); } catch (e) {} },
  };
  // keep the most recent label for each edge
  function mergeLabel(id, rec) {
    if (!rec || !labelNames.includes(rec.label)) return;
    const cur = labels[id];
    if (!cur || (rec.labelled_at || "") > (cur.labelled_at || "")) labels[id] = { label: rec.label, note: rec.note || "", labelled_at: rec.labelled_at || "" };
  }
  if (V) {
    const L = D.validation_labels;
    if (L && L.validation_id) L.validation_id.forEach((id, i) => {
      fileLabels[id] = L.labelled_at[i] || "";
      mergeLabel(id, { label: L.label[i], note: L.note[i], labelled_at: L.labelled_at[i] });
    });
    try { Object.entries(JSON.parse(store.get(storeKey) || "{}")).forEach(([id, rec]) => mergeLabel(id, rec)); } catch (e) {}
    $("fp-labelbar").innerHTML = `<button id="fp-label-start">Label validation sample (blind)</button>`;
    $("fp-label-start").addEventListener("click", () => {
      state.labelMode = true;
      const open = V.validation_id.findIndex(id => !labels[id]);
      vi = open >= 0 ? open : 0;
      $("fp-app").classList.add("labelling");
      render();
    });
  }
  const saveLabels = () => store.set(storeKey, JSON.stringify(labels));
  function unexportedCount() {
    const exported = store.get(exportKey) || "";
    return Object.entries(labels).filter(([id, r]) => r.labelled_at > exported && r.labelled_at !== fileLabels[id]).length;
  }

  function renderLabelling() {
    const d = V.detection[vi];
    $("fp-rows").innerHTML = rowHtml(det.asv[d], stats(), d);
    drawLegend();
    drawLabelPanel();
  }

  function drawLabelPanel() {
    const n = V.validation_id.length, id = V.validation_id[vi], d = V.detection[vi];
    const smp = det.sample[d], g = groups[S.group[smp]], rec = labels[id];
    const nDone = V.validation_id.filter(x => labels[x]).length, nOpen = unexportedCount();
    $("fp-labelpanel").innerHTML = `
      <div class="lrow"><b>${esc(id)}</b><span>${vi + 1} of ${n}</span><span>${nDone} labelled</span>
        ${nOpen ? `<span class="warn">${nOpen} not exported</span>` : ""}
        <span style="flex:1"></span>
        <button data-v="prev">‹ prev [←]</button><button data-v="next">next [→] ›</button><button data-v="open">next unlabelled »</button></div>
      <div class="lrow"><span>Edge: <b>${esc(A.label[det.asv[d]])}</b> in <b>${esc(S.feature_id[smp])}</b></span>
        <span>${esc(g.label + " " + g.sublabel)}${S.site[smp] ? " · " + esc(S.site[smp]) : ""}</span>
        <span>run ${esc(D.runs[S.run[smp]] ?? "?")} · extracted ${esc(S.extraction_date[smp] || "?")}</span>
        <span>${fmtInt(det.reads[d])} / ${fmtInt(S.sample_reads[smp])} reads (${fmtRel(det.rel_abund[d])})</span></div>
      <div class="lrow">${labelNames.map(l => `<button class="lab lab-${l}${rec && rec.label === l ? " on" : ""}" data-label="${l}">${l} [${l[0]}]</button>`).join("")}
        <input type="text" id="fp-note" size="44" placeholder="note (optional, saved with the label)" value="${esc(rec ? rec.note : "")}">
        ${rec ? `<span class="fp-sub">saved ${esc(rec.labelled_at.slice(0, 16).replace("T", " "))}</span>` : ""}</div>
      <div class="lrow"><button data-v="export">export labels CSV</button>
        <label class="fp-sub">import labels CSV <input type="file" id="fp-import" accept=".csv,text/csv"></label>
        <span style="flex:1"></span><button data-v="exit">exit labelling</button></div>`;
  }

  function setLabel(label) {
    const id = V.validation_id[vi];
    mergeLabel(id, { label, note: ($("fp-note") || {}).value || "", labelled_at: new Date().toISOString() });
    saveLabels();
    if (vi < V.validation_id.length - 1) vi++;
    render();
  }

  const csvField = v => /[",\n\r]/.test(v) ? `"${String(v).replace(/"/g, '""')}"` : String(v);
  function exportLabels() {
    const head = ["validation_id", "asv_id", "feature_id", "label", "note", "labelled_at"];
    const lines = [head.join(",")];
    V.validation_id.forEach((id, i) => {
      const r = labels[id];
      if (!r) return;
      const d = V.detection[i];
      lines.push([id, A.asv_id[det.asv[d]], S.feature_id[det.sample[d]], r.label, r.note, r.labelled_at].map(csvField).join(","));
    });
    const url = URL.createObjectURL(new Blob([lines.join("\n") + "\n"], { type: "text/csv" }));
    const link = document.createElement("a");
    link.href = url;
    link.download = `${D.run_name}_validation_labels.csv`;
    document.body.appendChild(link);
    link.click();
    link.remove();
    setTimeout(() => URL.revokeObjectURL(url), 1000);
    store.set(exportKey, new Date().toISOString());
    render();
  }

  function parseCsv(text) {
    const rows = [];
    let row = [], field = "", quoted = false;
    for (let i = 0; i < text.length; i++) {
      const c = text[i];
      if (quoted) {
        if (c === '"') { if (text[i + 1] === '"') { field += '"'; i++; } else quoted = false; }
        else field += c;
      } else if (c === '"') quoted = true;
      else if (c === ",") { row.push(field); field = ""; }
      else if (c === "\n" || c === "\r") {
        if (c === "\r" && text[i + 1] === "\n") i++;
        row.push(field); rows.push(row); row = []; field = "";
      } else field += c;
    }
    if (field !== "" || row.length) { row.push(field); rows.push(row); }
    const head = rows.shift() || [];
    return rows.filter(r => r.length > 1).map(r => Object.fromEntries(head.map((h, i) => [h, r[i] ?? ""])));
  }

  if (V) {
    $("fp-labelpanel").addEventListener("click", e => {
      const b = e.target.closest("button");
      if (!b) return;
      if (b.dataset.label) return setLabel(b.dataset.label);
      const v = b.dataset.v, n = V.validation_id.length;
      if (v === "prev") vi = Math.max(0, vi - 1);
      else if (v === "next") vi = Math.min(n - 1, vi + 1);
      else if (v === "open") { const k = V.validation_id.findIndex((id, i) => i > vi && !labels[id]); if (k >= 0) vi = k; }
      else if (v === "export") return exportLabels();
      else if (v === "exit") {
        state.labelMode = false;
        $("fp-app").classList.remove("labelling");
      }
      render();
    });
    $("fp-labelpanel").addEventListener("change", e => {
      if (e.target.id === "fp-note") {
        const id = V.validation_id[vi];
        if (labels[id]) { labels[id].note = e.target.value; labels[id].labelled_at = new Date().toISOString(); saveLabels(); }
      } else if (e.target.id === "fp-import" && e.target.files[0]) {
        const reader = new FileReader();
        reader.onload = () => { parseCsv(reader.result).forEach(r => mergeLabel(r.validation_id, r)); saveLabels(); render(); };
        reader.readAsText(e.target.files[0]);
      }
    });
    document.addEventListener("keydown", e => {
      if (!state.labelMode || e.target.tagName === "INPUT" || e.metaKey || e.ctrlKey || e.altKey) return;
      const n = V.validation_id.length;
      if (e.key === "s") setLabel("signal");
      else if (e.key === "c") setLabel("contamination");
      else if (e.key === "u") setLabel("uncertain");
      else if (e.key === "ArrowLeft") { vi = Math.max(0, vi - 1); render(); }
      else if (e.key === "ArrowRight") { vi = Math.min(n - 1, vi + 1); render(); }
    });
  }

  drawHeader();
  render();
});
</script>
```

```{r}
#| include: false
# ---- config ----
librarian::shelf(tidyverse, yaml, here, jsonlite)

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R)
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

run_dir = here(pipeline_config$run_dir)
cat(paste0("Run_dir: ", run_dir, "\n"))

config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# hand-reviewed known contaminants (scripts 13 and 14), shown as a badge with their reason
known_contaminants = pipeline_config$known_contaminants

possible_study_systems = c("brazil", "panama", "pennsylvania", "sierra_nevada")
system_labels = c(brazil = "Brazil", panama = "Panama", pennsylvania = "Pennsylvania", sierra_nevada = "Sierra Nevada")
system_codes = c(brazil = "BR", panama = "PAN", pennsylvania = "PENN", sierra_nevada = "SN")

# input files
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))
classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))
locality_file = here(run_dir, "output", paste0(config$run$name, "_07_asv_locality.csv"))
controlled_file = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))
sequence_file = Sys.glob(here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"),
                              "representative_sequences", "*", "data", "dna-sequences.fasta"))[1]
# optional (script 14): adds the blind labelling mode for the validation sample, starting from any
# labels already exported
validation_dir = here(run_dir, pipeline_config$validation$dir)
validation_sample_file = file.path(validation_dir, paste0(config$run$name, "_validation_sample.csv"))
validation_labels_file = file.path(validation_dir, paste0(config$run$name, "_validation_labels.csv"))
```

```{r}
#| include: false
# ---- load data ----
input_files = c(sample_map_file, classification_file, locality_file, controlled_file, sequence_file)
missing_inputs = input_files[is.na(input_files) | !file.exists(input_files)]
if (length(missing_inputs) > 0) {
  stop("Missing input(s):\n\t", paste(missing_inputs, collapse = "\n\t"))
}

# guess_max = Inf: some sample map columns are NA for most rows, and would otherwise be guessed
# as logical and their later values dropped
sample_map = read_csv(sample_map_file, show_col_types = FALSE, guess_max = Inf)
asv_classification = read_csv(classification_file, show_col_types = FALSE)
asv_locality = read_csv(locality_file, show_col_types = FALSE)
asv_controlled = read_csv(controlled_file, show_col_types = FALSE, guess_max = Inf) %>%
  select(asv_id, feature_id, asv_count_raw, positive_control_component)

stopifnot(!anyDuplicated(sample_map$feature_id))

# ASV sequences (multi-line records allowed)
fasta_lines = read_lines(sequence_file)
is_header = str_starts(fasta_lines, ">")
record = cumsum(is_header)
sequences = tibble(asv_id = word(str_remove(fasta_lines[is_header], "^>"), 1),
                   sequence = unname(tapply(fasta_lines[!is_header], record[!is_header], paste, collapse = ""))) %>%
  transmute(asv_id, sequence, sequence_length = nchar(sequence))
```

```{r}
#| include: false
# ---- groups and samples ----
study_systems = intersect(possible_study_systems, sample_map$study_system)

# panel columns, in display order
groups = bind_rows(
  map_dfr(study_systems, \(s) tibble(group = paste0(s, c("_field", "_field_negative")),
                                     label = system_labels[[s]],
                                     sublabel = c("field", "field neg"),
                                     kind = c("field_sample", "field_negative"),
                                     study_system = s)),
  tibble(group = c("extraction_negative", "pcr_negative", "pcr_positive"),
         label = "Lab",
         sublabel = c("extraction neg", "PCR neg", "PCR pos"),
         kind = "lab_control",
         study_system = NA_character_))

sample_reads = asv_controlled %>%
  group_by(feature_id) %>%
  summarise(sample_reads = sum(asv_count_raw), .groups = "drop")

samples_all = sample_map %>%
  transmute(feature_id,
            group = case_when(
              pcr_control_type %in% "pcr_positive_control" ~ "pcr_positive",
              pcr_control_type %in% "pcr_negative_control" ~ "pcr_negative",
              extraction_negative_control %in% TRUE ~ "extraction_negative",
              study_system %in% study_systems & field_negative_control %in% TRUE ~ paste0(study_system, "_field_negative"),
              study_system %in% study_systems ~ paste0(study_system, "_field"),
              .default = NA_character_),
            site,
            sequencing_date = as.character(sequencing_date),
            extraction_date = as.character(extraction_date)) %>%
  inner_join(sample_reads, by = "feature_id")

cat("libraries without a panel group (not shown):", sum(is.na(samples_all$group)), "\n")
cat("libraries with zero reads (not shown):", sum(samples_all$sample_reads == 0), "\n")

samples = samples_all %>%
  filter(!is.na(group), sample_reads > 0) %>%
  arrange(match(group, groups$group), sequencing_date, feature_id)

runs = sort(unique(samples$sequencing_date))

samples %>%
  count(group) %>%
  right_join(groups %>% select(group, label, sublabel), by = "group") %>%
  arrange(match(group, groups$group)) %>%
  transmute(label, sublabel, n_libraries = replace_na(n, 0)) %>%
  knitr::kable()
```

```{r}
#| include: false
# ---- detections and asvs ----
# every non-zero raw count
detections = asv_controlled %>%
  filter(asv_count_raw > 0) %>%
  inner_join(samples %>% select(feature_id, sample_reads), by = "feature_id") %>%
  transmute(asv_id, feature_id,
            reads = as.integer(asv_count_raw),
            rel_abund = asv_count_raw / sample_reads)

taxonomy = asv_classification %>%
  transmute(asv_id, taxonomic_rank, class, order, family, p_identical_max,
            taxon_06 = coalesce(if_else(!is.na(species), paste(genus, species), NA_character_),
                                genus, family, order, class, phylum),
            short_sequence_flag, contaminant_flag, contaminant_id)

# script 07 status per system, one column per system (status_<system>)
locality_status = asv_locality %>%
  select(asv_id, study_system, local_status) %>%
  pivot_wider(names_from = study_system, values_from = local_status, names_prefix = "status_")
for (s in study_systems) {
  if (!(paste0("status_", s) %in% names(locality_status))) locality_status[[paste0("status_", s)]] = NA_character_
}

known_contaminant_notes = map_dfr(known_contaminants, \(x) tibble(asv_id = x$asv_id, known_contaminant = x$note %||% "known contaminant"))

positive_control_components = asv_controlled %>%
  group_by(asv_id) %>%
  summarise(positive_control_component = any(positive_control_component %in% TRUE), .groups = "drop")

asvs = detections %>%
  group_by(asv_id) %>%
  summarise(total_reads = as.integer(sum(reads)), .groups = "drop") %>%
  left_join(asv_locality %>% distinct(asv_id, taxon_label), by = "asv_id") %>%
  left_join(taxonomy, by = "asv_id") %>%
  left_join(sequences, by = "asv_id") %>%
  left_join(locality_status, by = "asv_id") %>%
  left_join(positive_control_components, by = "asv_id") %>%
  left_join(known_contaminant_notes, by = "asv_id") %>%
  mutate(label = coalesce(taxon_label, taxon_06, "unclassified"),
         sequence_length = as.integer(sequence_length)) %>%
  select(asv_id, label, taxonomic_rank, class, order, family, p_identical_max, sequence, sequence_length, total_reads,
         short_sequence_flag, contaminant_flag, contaminant_id, positive_control_component,
         known_contaminant, all_of(paste0("status_", study_systems))) %>%
  arrange(desc(total_reads))

# validation sample, as 0-based detection indices in labelling order (split and strata are left out:
# labelling is blind to them)
validation = NULL
validation_labels = NULL
if (file.exists(validation_sample_file)) {
  validation_sample = read_csv(validation_sample_file, show_col_types = FALSE) %>%
    arrange(labelling_order) %>%
    mutate(detection = match(paste(asv_id, feature_id), paste(detections$asv_id, detections$feature_id)) - 1L)
  if (anyNA(validation_sample$detection)) {
    warning(sum(is.na(validation_sample$detection)), " validation edges are not among the detections and are left out")
  }
  validation = validation_sample %>%
    filter(!is.na(detection)) %>%
    select(validation_id, detection)
  if (file.exists(validation_labels_file)) {
    validation_labels = read_csv(validation_labels_file, col_types = cols(.default = col_character())) %>%
      select(validation_id, label, note, labelled_at)
  }
  cat("validation sample:", nrow(validation), "edges,", if (is.null(validation_labels)) 0 else nrow(validation_labels), "labelled\n")
}

cat("ASVs:", nrow(asvs), "\n")
cat("detections:", nrow(detections), "\n")
```

```{r}
#| echo: false
#| output: asis
# ---- embed data (read by the fingerprints viewer above) ----
fp_data = list(
  run_name = unbox(config$run$name),
  study_systems = study_systems,
  system_labels = map(as.list(system_labels[study_systems]), unbox),
  system_codes = map(as.list(system_codes[study_systems]), unbox),
  groups = groups,
  runs = runs,
  validation = validation,
  validation_labels = validation_labels,
  samples = samples %>%
    transmute(feature_id,
              group = match(group, groups$group) - 1L,
              site,
              run = match(sequencing_date, runs) - 1L,
              extraction_date,
              sample_reads = as.integer(sample_reads)),
  asvs = asvs,
  # indices are 0-based, into asvs / samples
  detections = detections %>%
    transmute(asv = match(asv_id, asvs$asv_id) - 1L,
              sample = match(feature_id, samples$feature_id) - 1L,
              reads,
              rel_abund))

fp_json = toJSON(fp_data, dataframe = "columns", na = "null", digits = I(4)) %>%
  str_replace_all(fixed("</"), "<\\/")

cat("```{=html}\n<script>window.FP_DATA = ", fp_json, ";</script>\n```\n", sep = "")
```
