---
title: "Edge classification"
author: "Cob Staines"
output: html_document
---

# Description

This script classifies every **edge** of the sample--ASV network -- a detection of an ASV in a field library -- as `signal` or `contamination`. A library is one sequenced unit (`feature_id`); an eDNA sample may have several. It **only classifies, it drops nothing**, and nothing downstream reads its outputs: it is a test bed for the classification scheme, to be compared with script 08's filters.

**Evidence.** Each edge gets a log-odds score L, the sum of terms each scaled to [-1, 1] and multiplied by a weight (`edge_classification$weights`, the term's maximum contribution). Positive is evidence for signal. Psig = 1 / (1 + exp(-L)), and confidence = |2 Psig - 1|.

Fixed terms, from the data alone:

- **U1 own negatives**: log10((x + eps) / (m + eps)), where m is the ASV's maximum x in the library's *own* linked negatives (same field negative, extraction negative or PCR control group), clipped to -3 decades below and `own_negatives_max_decades_above` above, then divided by 3 -- absence from a few negatives is weak evidence for signal, while sitting at or below them is strong evidence against. 0 when the library has no linked negatives -- missing controls are not evidence.
- **U2 cross-talk** (one-sided, towards contamination only): k = log10(x / s), where s is the ASV's maximum x in any *other* library of the same sequencing run, negatives excluded (negatives are sinks, never sources; positive controls are sources). No evidence while k is above log10(`crosstalk_cutoff`), full evidence at or below log10(`crosstalk_full`).
- **U3 replicates** (mostly one-sided, towards signal): the share of the eDNA sample's other field libraries that also detect the ASV; `replicate_miss_penalty` when none do; 0 without other libraries.
- **U4 control share**, per ASV x system: logit((mean x in the system's field libraries + eps) / (that + mean x in lab negatives and the system's field negatives + 2 eps)), clipped to -4.6 below and `control_share_max_logit` above, then divided by 4.6 -- like U1, absence from negatives is weak evidence. Means are per library, zeros included; positive controls are excluded.

Relational terms, from the labels of neighbouring edges (S signal votes, C contamination votes, the edge itself excluded; `kappa` damps small counts):

- **R1 same ASV, same system**: (S - C) / (S + C + kappa) -- real taxa are consistent within a system.
- **R2 same ASV, other systems** (towards contamination only): -C / (C + kappa) -- contamination elsewhere suggests a common contaminant; signal elsewhere is a potential source, not evidence of signal here.
- **R3 same library, other ASVs** (towards contamination only): -C / (S + C + kappa) -- the library's sink score. A library's signal edges do not vouch for its other edges.

**Relative abundance x** is the edge's reads over the library's reads, excluding hard-flagged ASVs and reads already labelled contamination (the *dynamic* denominator), never below `denominator_floor` of the library's reads, plus `denominator_pseudo_reads` -- so a high x resting on few remaining reads (e.g. a library that is mostly clamped human DNA) carries less evidence. The edge's own reads always stay in its own denominator, so labelling an edge cannot inflate its own x. Control libraries keep a fixed denominator.

**Clamps** (fixed contamination, never re-scored):

- hard flags from script 06 (`short_sequence`, `contaminant_library`): left out of everything, denominators included
- every edge in a negative control (PCR, extraction, field) and in a PCR positive control: they inform U1, U2 (positive controls as sources) and U4, but do not vote in R1-R3
- ASVs in `known_contaminants` (a hand-reviewed list shared with script 14: lab stocks and positive-control constructs), and PCR positive-control component ASVs (script 08) when `clamp_positive_control_components`: every detection, local or not; removed from denominators from the start, do not vote

Anthropogenic ASVs are **not** clamped: they are classified like any other ASV, and vote in R3.

**Iteration.** Three variants are run with the same scoring:

- `hard` (primary): edges start unassigned. At each threshold in `tau_schedule`, every unassigned edge with confidence >= tau is assigned by the direction of Psig, scores are recomputed, and this repeats until nothing changes; then the next (lower) threshold. An assigned edge whose confidence drops below `tau_drop`, or whose direction flips, is unassigned again; after `max_flips` it stays unassigned (`unstable`). Edges still below the last threshold stay `unassigned`, for manual review.
- `soft`: all edges are updated together, each vote weighted by its confidence (2 Psig - 1), damped, until Psig stops changing; labelled at the last threshold.
- `hard_fixed`: as `hard`, but x keeps a fixed denominator (sensitivity check on the dynamic denominator).

**Evaluation.** Model versions are scored against the **validation sample** (script 14), labelled blind in script 12, following the validation protocol in `analysis/README.md`: weighted by stratum, on the development set only unless `validation$evaluate_test_set` is set, with error costs from `validation$costs`. The **reference cases** (`labels_file`: hand-picked ASV x system labels, with an optional `feature_id` override and `min_x` / `max_x` limits) are a sanity check only and are never scored. GBIF locality (script 07) is not used to classify, and is shown as a secondary check.

## Model formulation

**Notation.** An edge $e = (a, \ell)$ is a detection of ASV $a$ in field library $\ell$, with $n_e$ reads. Library $\ell$ belongs to study system $\sigma(\ell)$ and sequencing run $r(\ell)$, and has $N_\ell$ reads excluding hard-flagged ASVs. $F$ is the set of free (unclamped) edges. Each free edge carries a label $y_e \in \{+1, 0, -1\}$ (signal, unassigned, contamination) and a signal / contamination vote

$$
s_e = \mathbb{1}[y_e = +1], \quad c_e = \mathbb{1}[y_e = -1] \qquad \text{(hard)}, \qquad\qquad
s_e = \max(0,\, 2p_e - 1), \quad c_e = \max(0,\, 1 - 2p_e) \qquad \text{(soft)},
$$

where $p_e$ is the edge's current $P_{sig}$.

**Relative abundance.** With $K_\ell$ the reads of clamped ASVs in $\ell$, $\phi$ = `denominator_floor` and $n_0$ = `denominator_pseudo_reads`:

$$
x_e = \frac{n_e}{\max\!\left(N_\ell - K_\ell - \sum_{e' \in F_\ell \setminus \{e\}} n_{e'}\, c_{e'},\;\; \phi N_\ell\right) + n_0}
$$

where $F_\ell$ are the free edges of library $\ell$. The sum is dropped in the `hard_fixed` variant. Control libraries use $x = n_e / N_\ell$.

**Fixed terms.** With $\epsilon$ = `epsilon`, $\mathrm{clip}(z; lo, hi) = \min(\max(z, lo), hi)$, $d$ = `own_negatives_max_decades_above`, $\lambda$ = `control_share_max_logit`, $\gamma_1$ = `crosstalk_cutoff`, $\gamma_2$ = `crosstalk_full`, and $\pi$ = `replicate_miss_penalty`:

$$
U_1(e) = \mathbb{1}[\ell \text{ has linked negatives}] \cdot \frac{1}{3}\,\mathrm{clip}\!\left(\log_{10}\frac{x_e + \epsilon}{m_e + \epsilon};\; -3,\; d\right),
\qquad m_e = \max_{\ell' \in \mathrm{neg}(\ell)} x_{a\ell'}
$$

$$
U_2(e) = -\,\mathrm{clip}\!\left(\frac{\log_{10}\gamma_1 - \log_{10}(x_e / q_e)}{\log_{10}\gamma_1 - \log_{10}\gamma_2};\; 0,\; 1\right),
\qquad q_e = \max_{\substack{\ell' \neq \ell,\; r(\ell') = r(\ell) \\ \ell' \notin \text{negatives}}} x_{a\ell'}
\quad (U_2 = 0 \text{ if no such } \ell')
$$

$$
U_3(e) = \begin{cases} 0 & R_e = 0 \\ k_e / R_e & k_e > 0 \\ \pi & \text{otherwise} \end{cases}
\qquad R_e, k_e = \text{other field libraries of the eDNA sample, and how many detect } a
$$

$$
U_4(e) = \frac{1}{4.6}\,\mathrm{clip}\!\left(\mathrm{logit}\,\frac{\mu^{f}_{a\sigma} + \epsilon}{\mu^{f}_{a\sigma} + \mu^{n}_{a\sigma} + 2\epsilon};\; -4.6,\; \lambda\right)
$$

where $\mu^f_{a\sigma}$ is the mean $x$ of $a$ over all field libraries of system $\sigma$ (zeros included, fixed denominator), and $\mu^n_{a\sigma}$ the same over lab negatives and the field negatives of $\sigma$.

**Relational terms.** With $\kappa$ = `kappa`, and the edge itself always excluded:

$$
S_1, C_1 = \sum_{e' \in F_{a\sigma} \setminus \{e\}} s_{e'},\; c_{e'}
\qquad
C_2 = \sum_{e' \in F_{a} \setminus F_{a\sigma}} c_{e'}
\qquad
S_3, C_3 = \sum_{e' \in F_{\ell} \setminus \{e\}} s_{e'},\; c_{e'}
$$

$$
R_1(e) = \frac{S_1 - C_1}{S_1 + C_1 + \kappa},
\qquad
R_2(e) = -\frac{C_2}{C_2 + \kappa},
\qquad
R_3(e) = -\frac{C_3}{S_3 + C_3 + \kappa}
$$

where $F_{a\sigma}$ are the free edges of ASV $a$ in system $\sigma$, $F_a$ all free edges of $a$, and $F_\ell$ the free edges of library $\ell$.

**Score.** Every term lies in $[-1, 1]$ and is scaled by its weight $w$ (`weights`, the term's maximum contribution in log-odds):

$$
L_e = \sum_{k=1}^{4} w_{U_k}\, U_k(e) + \sum_{j=1}^{3} w_{R_j}\, R_j(e),
\qquad
P_{sig}(e) = \frac{1}{1 + e^{-L_e}},
\qquad
\mathrm{conf}(e) = \left|2P_{sig}(e) - 1\right|
$$

**Hard iteration.** For thresholds $\tau_1 > \tau_2 > \dots > \tau_K$ (`tau_schedule`), repeat at each $\tau_k$ until nothing changes (at most `max_steps_per_tau` steps): recompute $L$ for all free edges; unassign every assigned edge with $\mathrm{conf}(e) < \tau_{drop}$ or $\mathrm{sign}(L_e) \neq y_e$ (counting a flip); then set $y_e = \mathrm{sign}(L_e)$ for every unassigned edge with $\mathrm{conf}(e) \ge \tau_k$ and fewer than `max_flips` flips. Edges still unassigned after $\tau_K$ are left for review.

**Soft iteration.** Starting from $p_e = 0.5$, with damping $\eta$ = `soft_damping`, until $\max_e |\Delta p_e|$ < `soft_tolerance`:

$$
p_e \leftarrow p_e + \eta\left(\frac{1}{1 + e^{-L_e(p)}} - p_e\right)
$$

then $y_e = \mathrm{sign}(2p_e - 1)$ where $|2p_e - 1| \ge \tau_K$, otherwise unassigned.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here)

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R)
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

run_dir = here(pipeline_config$run_dir)
cat(paste0("Run_dir: ", run_dir, "\n"))

config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

ec = pipeline_config$edge_classification
model_version = ec$model_version
edge_labels_file = here(ec$labels_file)
known_contaminants = map_chr(pipeline_config$known_contaminants, "asv_id")
clamp_positive_control_components = isTRUE(ec$clamp_positive_control_components)
epsilon = as.numeric(ec$epsilon)
own_negatives_max_decades_above = as.numeric(ec$own_negatives_max_decades_above)
kappa = as.numeric(ec$kappa)
denominator_floor = as.numeric(ec$denominator_floor)
denominator_pseudo_reads = as.numeric(ec$denominator_pseudo_reads)
control_share_max_logit = as.numeric(ec$control_share_max_logit)
crosstalk_cutoff = as.numeric(ec$crosstalk_cutoff)
crosstalk_full = as.numeric(ec$crosstalk_full)
replicate_miss_penalty = as.numeric(ec$replicate_miss_penalty)
weights = map_dbl(ec$weights, as.numeric)
tau_schedule = sort(as.numeric(unlist(ec$tau_schedule)), decreasing = TRUE)
tau_drop = as.numeric(ec$tau_drop)
max_flips = as.integer(ec$max_flips)
max_steps_per_tau = as.integer(ec$max_steps_per_tau)
soft_damping = as.numeric(ec$soft_damping)
soft_tolerance = as.numeric(ec$soft_tolerance)
soft_max_iterations = as.integer(ec$soft_max_iterations)

term_names = c("own_negatives", "crosstalk", "replicates", "control_share", "asv_system", "asv_other_systems", "library")
stopifnot(all(term_names %in% names(weights)))
weights = weights[term_names]

# validation (see the validation protocol in analysis/README.md)
vc = pipeline_config$validation
validation_dir = here(vc$dir)
validation_costs = map_dbl(vc$costs, as.numeric)
evaluate_test_set = isTRUE(vc$evaluate_test_set)
bootstrap_reps = as.integer(vc$bootstrap_reps)
validation_sample_file = file.path(validation_dir, paste0(config$run$name, "_validation_sample.csv"))
validation_labels_file = file.path(validation_dir, paste0(config$run$name, "_validation_labels.csv"))

possible_study_systems = c("brazil", "panama", "pennsylvania", "sierra_nevada")
negative_classes = c("field_negative", "extraction_negative", "pcr_negative")

# input files
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))
classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))
locality_file = here(run_dir, "output", paste0(config$run$name, "_07_asv_locality.csv"))
controlled_file = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))

# output files
edges_out = here(run_dir, "output", paste0(config$run$name, "_13_edge_classification.csv"))
libraries_out = here(run_dir, "output", paste0(config$run$name, "_13_library_sink_scores.csv"))
asv_system_out = here(run_dir, "output", paste0(config$run$name, "_13_asv_system_summary.csv"))
iteration_log_out = here(run_dir, "output", paste0(config$run$name, "_13_iteration_log.csv"))
validation_metrics_out = here(run_dir, "output", paste0(config$run$name, "_13_validation_metrics.csv"))
```

# Load data
```{r}
input_files = c(sample_map_file, classification_file, locality_file, controlled_file)
missing_inputs = input_files[!file.exists(input_files)]
if (length(missing_inputs) > 0) {
  stop("Missing input(s):\n\t", paste(missing_inputs, collapse = "\n\t"))
}

# guess_max = Inf: some sample map columns are NA for most rows, and would otherwise be guessed
# as logical and their later values dropped
sample_map = read_csv(sample_map_file, show_col_types = FALSE, guess_max = Inf)
asv_classification = read_csv(classification_file, show_col_types = FALSE)
asv_locality = read_csv(locality_file, show_col_types = FALSE)
asv_controlled = read_csv(controlled_file, show_col_types = FALSE, guess_max = Inf) %>%
  select(asv_id, feature_id, asv_count_raw, asv_count_controlled, positive_control_component)

stopifnot(!anyDuplicated(sample_map$feature_id))

edge_labels = if (file.exists(edge_labels_file)) {
  read_csv(edge_labels_file, col_types = cols(.default = col_character())) %>%
    mutate(min_x = as.numeric(min_x), max_x = as.numeric(max_x),
           feature_id = na_if(feature_id, ""))
} else {
  warning("No labels file at ", edge_labels_file, ": evaluation against manual labels is skipped")
  tibble(asv_id = character(), study_system = character(), feature_id = character(), label = character(),
         min_x = numeric(), max_x = numeric(), note = character())
}
stopifnot(all(edge_labels$label %in% c("signal", "contamination")))
```

# Libraries and edges
```{r}
study_systems = intersect(possible_study_systems, sample_map$study_system)

libraries = sample_map %>%
  transmute(feature_id,
            library_class = case_when(
              pcr_control_type %in% "pcr_positive_control" ~ "pcr_positive",
              pcr_control_type %in% "pcr_negative_control" ~ "pcr_negative",
              extraction_negative_control %in% TRUE ~ "extraction_negative",
              study_system %in% study_systems & field_negative_control %in% TRUE ~ "field_negative",
              study_system %in% study_systems ~ "field_sample",
              .default = NA_character_),
            study_system = if_else(library_class %in% c("field_sample", "field_negative"), study_system, NA_character_),
            sequencing_date = as.character(sequencing_date),
            edna_sample_id)

cat("libraries without a class (left out):", sum(is.na(libraries$library_class)), "\n")
libraries = libraries %>% filter(!is.na(library_class))

# hard flags (script 06): left out of everything
hard_flags = asv_classification %>%
  filter(short_sequence_flag %in% TRUE | contaminant_flag %in% TRUE) %>%
  transmute(asv_id, hard_flag = if_else(short_sequence_flag %in% TRUE, "short_sequence", "contaminant_library"))

detections = asv_controlled %>%
  filter(asv_count_raw > 0) %>%
  inner_join(libraries, by = "feature_id") %>%
  left_join(hard_flags, by = "asv_id") %>%
  rename(reads = asv_count_raw)

positive_control_component_asvs = if (clamp_positive_control_components) {
  unique(asv_controlled$asv_id[asv_controlled$positive_control_component %in% TRUE])
} else {
  character()
}
cat("positive-control component ASVs clamped:", length(positive_control_component_asvs), "\n")

edges = detections %>%
  filter(is.na(hard_flag)) %>%
  mutate(role = case_when(library_class %in% negative_classes ~ "negative_control",
                          library_class == "pcr_positive" ~ "positive_control",
                          asv_id %in% known_contaminants ~ "known_contaminant",
                          asv_id %in% positive_control_component_asvs ~ "positive_control_component",
                          .default = "free"))

# library reads excluding hard flags (N), and reads of clamped ASVs (removed from field libraries'
# denominators from the start)
library_reads = edges %>%
  group_by(feature_id) %>%
  summarise(N = sum(reads),
            clamp_reads = sum(reads[role %in% c("known_contaminant", "positive_control_component")]),
            .groups = "drop")

libraries = libraries %>%
  left_join(library_reads, by = "feature_id") %>%
  mutate(N = replace_na(N, 0), clamp_reads = replace_na(clamp_reads, 0))

# x with a fixed denominator: control libraries over N; field libraries over N less clamped reads,
# plus the pseudo reads
edges = edges %>%
  left_join(libraries %>% select(feature_id, N, clamp_reads), by = "feature_id") %>%
  mutate(x_fixed = if_else(library_class == "field_sample",
                           reads / (pmax(N - clamp_reads, denominator_floor * N) + denominator_pseudo_reads),
                           reads / N))

edges %>% count(role, library_class) %>% knitr::kable()
cat("hard-flagged detections (left out):", sum(!is.na(detections$hard_flag)), "\n")
```

# Fixed terms
```{r}
# free edges: field-library detections to classify
fe = edges %>%
  filter(role == "free") %>%
  arrange(asv_id, feature_id)

# U1: each library's own linked negatives, via the control-group ids shared with them
link_controls = function(group_col, control_class) {
  groups = sample_map %>%
    select(feature_id, group = all_of(group_col)) %>%
    filter(!is.na(group))
  controls = libraries$feature_id[libraries$library_class == control_class]
  groups %>%
    inner_join(groups %>% filter(feature_id %in% controls) %>% rename(control_id = feature_id),
               by = "group", relationship = "many-to-many") %>%
    filter(feature_id != control_id) %>%
    distinct(feature_id, control_id)
}

control_links = bind_rows(link_controls("field_negative_control_group", "field_negative"),
                          link_controls("extraction_negative_control_group", "extraction_negative"),
                          link_controls("pcr_control_group", "pcr_negative")) %>%
  distinct()

own_negative_max = control_links %>%
  inner_join(edges %>% filter(role == "negative_control") %>% select(control_id = feature_id, asv_id, x_fixed),
             by = "control_id", relationship = "many-to-many") %>%
  group_by(feature_id, asv_id) %>%
  summarise(own_negative_max = max(x_fixed), .groups = "drop")

fe = fe %>%
  mutate(has_negatives = feature_id %in% control_links$feature_id) %>%
  left_join(own_negative_max, by = c("feature_id", "asv_id")) %>%
  mutate(own_negative_max = replace_na(own_negative_max, 0))

# U3: other field libraries of the same eDNA sample
replicate_libraries = libraries %>%
  filter(library_class == "field_sample", !is.na(edna_sample_id), N > 0) %>%
  select(feature_id, edna_sample_id)

replicate_counts = replicate_libraries %>% count(edna_sample_id, name = "n_libraries")
replicate_detections = edges %>%
  filter(library_class == "field_sample") %>%
  inner_join(replicate_libraries %>% select(feature_id), by = "feature_id") %>%
  count(edna_sample_id, asv_id, name = "n_detected")

fe = fe %>%
  left_join(replicate_counts, by = "edna_sample_id") %>%
  left_join(replicate_detections, by = c("edna_sample_id", "asv_id")) %>%
  mutate(n_other_replicates = replace_na(n_libraries - 1L, 0L),
         n_other_replicates_detected = replace_na(n_detected - 1L, 0L),
         u_replicates = case_when(n_other_replicates == 0 ~ 0,
                                  n_other_replicates_detected > 0 ~ n_other_replicates_detected / n_other_replicates,
                                  .default = replicate_miss_penalty)) %>%
  select(-n_libraries, -n_detected)

# U4: per ASV x system, mean x over the system's field libraries vs. over lab negatives and the
# system's field negatives (per library, zeros included)
n_field_libraries = libraries %>%
  filter(library_class == "field_sample", N > 0) %>%
  count(study_system, name = "n_field")

lab_negatives = libraries %>% filter(library_class %in% c("extraction_negative", "pcr_negative"), N > 0)
n_negative_libraries = libraries %>%
  filter(library_class == "field_negative", N > 0) %>%
  count(study_system, name = "n_field_negative") %>%
  mutate(n_negative = n_field_negative + nrow(lab_negatives))

field_mean = fe %>%
  group_by(asv_id, study_system) %>%
  summarise(x_sum = sum(x_fixed), .groups = "drop") %>%
  left_join(n_field_libraries, by = "study_system") %>%
  transmute(asv_id, study_system, field_mean = x_sum / n_field)

lab_negative_sum = edges %>%
  filter(feature_id %in% lab_negatives$feature_id) %>%
  group_by(asv_id) %>%
  summarise(lab_sum = sum(x_fixed), .groups = "drop")

field_negative_sum = edges %>%
  filter(library_class == "field_negative") %>%
  group_by(asv_id, study_system) %>%
  summarise(field_negative_sum = sum(x_fixed), .groups = "drop")

control_share = field_mean %>%
  left_join(lab_negative_sum, by = "asv_id") %>%
  left_join(field_negative_sum, by = c("asv_id", "study_system")) %>%
  left_join(n_negative_libraries, by = "study_system") %>%
  mutate(negative_mean = (replace_na(lab_sum, 0) + replace_na(field_negative_sum, 0)) / n_negative,
         control_share_a = (field_mean + epsilon) / (field_mean + negative_mean + 2 * epsilon),
         u_control_share = pmin(pmax(qlogis(control_share_a), -4.6), control_share_max_logit) / 4.6) %>%
  select(asv_id, study_system, field_mean, negative_mean, control_share_a, u_control_share)

fe = fe %>% left_join(control_share, by = c("asv_id", "study_system"))

cat("free edges:", nrow(fe), "\n")
cat("  in libraries with linked negatives:", sum(fe$has_negatives), "\n")
cat("  also in own negatives:", sum(fe$own_negative_max > 0), "\n")
cat("  with other replicate libraries:", sum(fe$n_other_replicates > 0), "\n")
```

# Scoring
```{r}
# group codes for the free edges
n_fe = nrow(fe)
lib_g = as.integer(factor(fe$feature_id))
asv_g = as.integer(factor(fe$asv_id))
asv_system_g = as.integer(factor(paste(fe$asv_id, fe$study_system)))

# U2 source pool: free edges (dynamic x) and positive-control edges (fixed x), by ASV x run
positive_edges = edges %>% filter(role == "positive_control")
asv_run_levels = unique(c(paste(fe$asv_id, fe$sequencing_date), paste(positive_edges$asv_id, positive_edges$sequencing_date)))
pool_g = c(match(paste(fe$asv_id, fe$sequencing_date), asv_run_levels),
           match(paste(positive_edges$asv_id, positive_edges$sequencing_date), asv_run_levels))

# per-element sum of v over its group, with groups coded 1..n
group_sum = function(v, g) {
  r = rowsum(v, g)
  out = numeric(max(g))
  out[as.integer(rownames(r))] = r[, 1]
  out[g]
}

# per-element maximum of x over the *other* elements of its group (0 if none)
leave_one_out_max = function(g, x) {
  o = order(g, -x)
  gs = g[o]
  xs = x[o]
  first = which(!duplicated(gs))
  second = first + 1L
  has_second = second <= length(gs) & gs[pmin(second, length(gs))] == gs[first]
  top1 = xs[first]
  top2 = if_else(has_second, xs[pmin(second, length(gs))], 0)
  gi = match(g, gs[first])
  is_top = seq_along(x) == o[first][gi]
  if_else(is_top, top2[gi], top1[gi])
}

# term contributions and log odds for every free edge, given signal / contamination votes
# (hard: 0/1 labels; soft: max(0, 2 Psig - 1) and max(0, 1 - 2 Psig))
score_edges = function(v_signal, v_contamination, dynamic) {
  # dynamic denominator: reads labelled contamination drop out, except the edge's own
  removed = fe$clamp_reads
  if (dynamic) {
    contamination_reads = fe$reads * v_contamination
    removed = removed + group_sum(contamination_reads, lib_g) - contamination_reads
  }
  x = fe$reads / (pmax(fe$N - removed, denominator_floor * fe$N) + denominator_pseudo_reads)

  u_own_negatives = if_else(fe$has_negatives,
                            pmin(pmax(log10((x + epsilon) / (fe$own_negative_max + epsilon)), -3), own_negatives_max_decades_above) / 3,
                            0)

  source_x = leave_one_out_max(pool_g, c(x, positive_edges$x_fixed))[seq_len(n_fe)]
  k = if_else(source_x > 0, log10(x / source_x), 0)
  u_crosstalk = -pmin(pmax((log10(crosstalk_cutoff) - k) / (log10(crosstalk_cutoff) - log10(crosstalk_full)), 0), 1)

  s_same = group_sum(v_signal, asv_system_g) - v_signal
  c_same = group_sum(v_contamination, asv_system_g) - v_contamination
  u_asv_system = (s_same - c_same) / (s_same + c_same + kappa)

  c_other = group_sum(v_contamination, asv_g) - group_sum(v_contamination, asv_system_g)
  u_asv_other_systems = -c_other / (c_other + kappa)

  s_lib = group_sum(v_signal, lib_g) - v_signal
  c_lib = group_sum(v_contamination, lib_g) - v_contamination
  u_library = -c_lib / (s_lib + c_lib + kappa)

  contributions = cbind(own_negatives = u_own_negatives,
                        crosstalk = u_crosstalk,
                        replicates = fe$u_replicates,
                        control_share = fe$u_control_share,
                        asv_system = u_asv_system,
                        asv_other_systems = u_asv_other_systems,
                        library = u_library) %*% diag(weights)
  colnames(contributions) = term_names
  list(x = x, source_x = source_x, contributions = contributions, log_odds = rowSums(contributions))
}
```

# Classify
```{r}
label_names = c("contamination", "unassigned", "signal")  # label codes -1, 0, 1

classify_hard = function(dynamic) {
  label = integer(n_fe)
  flips = integer(n_fe)
  assigned_step = rep(NA_integer_, n_fe)
  assigned_tau = rep(NA_real_, n_fe)
  psig_at_assignment = rep(NA_real_, n_fe)
  x_at_assignment = rep(NA_real_, n_fe)
  log = list()
  step = 0L
  for (tau in tau_schedule) {
    for (i in seq_len(max_steps_per_tau)) {
      step = step + 1L
      sc = score_edges(as.numeric(label == 1L), as.numeric(label == -1L), dynamic)
      psig = plogis(sc$log_odds)
      confidence = abs(2 * psig - 1)
      direction = if_else(psig >= 0.5, 1L, -1L)

      dropped = label != 0L & (confidence < tau_drop | direction != label)
      flips[dropped] = flips[dropped] + 1L
      label[dropped] = 0L
      assigned_step[dropped] = NA_integer_
      assigned_tau[dropped] = NA_real_
      psig_at_assignment[dropped] = NA_real_
      x_at_assignment[dropped] = NA_real_

      added = label == 0L & flips < max_flips & confidence >= tau
      label[added] = direction[added]
      assigned_step[added] = step
      assigned_tau[added] = tau
      psig_at_assignment[added] = psig[added]
      x_at_assignment[added] = sc$x[added]

      log[[step]] = tibble(step = step, tau = tau, n_added = sum(added), n_dropped = sum(dropped),
                           signal = sum(label == 1L), contamination = sum(label == -1L), unassigned = sum(label == 0L))
      if (!any(added) && !any(dropped)) break
    }
  }
  sc = score_edges(as.numeric(label == 1L), as.numeric(label == -1L), dynamic)
  label_text = label_names[label + 2L]
  label_text[label == 0L & flips >= max_flips] = "unstable"
  list(label = label_text, psig = plogis(sc$log_odds), score = sc, flips = flips,
       assigned_step = assigned_step, assigned_tau = assigned_tau,
       psig_at_assignment = psig_at_assignment, x_at_assignment = x_at_assignment,
       log = bind_rows(log))
}

classify_soft = function(dynamic) {
  psig = rep(0.5, n_fe)
  for (iteration in seq_len(soft_max_iterations)) {
    sc = score_edges(pmax(0, 2 * psig - 1), pmax(0, 1 - 2 * psig), dynamic)
    psig_new = plogis(sc$log_odds)
    change = max(abs(psig_new - psig))
    psig = psig + soft_damping * (psig_new - psig)
    if (change < soft_tolerance) break
  }
  cat("soft:", iteration, "iterations, last change", signif(change, 2), "\n")
  confidence = abs(2 * psig - 1)
  label = if_else(confidence >= min(tau_schedule), if_else(psig >= 0.5, "signal", "contamination"), "unassigned")
  list(label = label, psig = psig, iterations = iteration)
}

hard = classify_hard(dynamic = TRUE)
soft = classify_soft(dynamic = TRUE)
hard_fixed = classify_hard(dynamic = FALSE)

iteration_log = bind_rows(hard$log %>% mutate(variant = "hard"),
                          hard_fixed$log %>% mutate(variant = "hard_fixed"))

tibble(variant = c("hard", "soft", "hard_fixed"),
       result = list(hard$label, soft$label, hard_fixed$label)) %>%
  mutate(result = map(result, \(l) as_tibble(as.list(table(factor(l, levels = c("signal", "contamination", "unassigned", "unstable"))))))) %>%
  unnest(result) %>%
  knitr::kable()
```

# Outputs
```{r}
contribution_cols = as_tibble(hard$score$contributions) %>% rename_with(~ paste0("c_", .x))

fe_result = fe %>%
  select(asv_id, feature_id) %>%
  mutate(label = hard$label,
         psig = hard$psig,
         confidence = abs(2 * psig - 1),
         log_odds = hard$score$log_odds,
         x = hard$score$x,
         run_source_x = hard$score$source_x,
         assigned_step = hard$assigned_step,
         assigned_tau = hard$assigned_tau,
         psig_at_assignment = hard$psig_at_assignment,
         x_at_assignment = hard$x_at_assignment,
         n_flips = hard$flips,
         label_soft = soft$label,
         psig_soft = soft$psig,
         label_fixed_denominator = hard_fixed$label,
         psig_fixed_denominator = hard_fixed$psig) %>%
  bind_cols(contribution_cols)

# every detection, clamped ones included (labelled contamination, with the clamp as the reason)
edge_classification = detections %>%
  left_join(edges %>% select(asv_id, feature_id, role, x_fixed), by = c("asv_id", "feature_id")) %>%
  left_join(fe %>% select(asv_id, feature_id, own_negative_max, n_other_replicates, n_other_replicates_detected,
                          control_share_a),
            by = c("asv_id", "feature_id")) %>%
  left_join(fe_result, by = c("asv_id", "feature_id")) %>%
  mutate(role = coalesce(role, "hard_flag"),
         clamp_reason = case_when(role == "hard_flag" ~ hard_flag,
                                  role == "free" ~ NA_character_,
                                  .default = role),
         across(c(label, label_soft, label_fixed_denominator), ~ if_else(role == "free", .x, "contamination")),
         removed_by_08 = asv_count_controlled == 0) %>%
  left_join(asv_locality %>% select(asv_id, study_system, local_status), by = c("asv_id", "study_system")) %>%
  mutate(model_version = model_version) %>%
  select(model_version, asv_id, feature_id, library_class, study_system, sequencing_date, edna_sample_id, reads,
         role, clamp_reason, label, psig, confidence, log_odds, x_fixed, x, run_source_x,
         starts_with("c_"), own_negative_max, n_other_replicates, n_other_replicates_detected, control_share_a,
         assigned_step, assigned_tau, psig_at_assignment, x_at_assignment, n_flips,
         label_soft, psig_soft, label_fixed_denominator, psig_fixed_denominator,
         removed_by_08, local_status)

library_sink_scores = edge_classification %>%
  filter(role == "free") %>%
  group_by(feature_id, study_system, sequencing_date) %>%
  summarise(n_edges = n(),
            n_signal = sum(label == "signal"),
            n_contamination = sum(label == "contamination"),
            n_unassigned = sum(label %in% c("unassigned", "unstable")),
            sink_score = n_contamination / pmax(n_signal + n_contamination, 1),
            contamination_read_share = sum(reads[label == "contamination"]) / sum(reads),
            .groups = "drop")

asv_system_summary = edge_classification %>%
  filter(role == "free") %>%
  group_by(asv_id, study_system) %>%
  summarise(n_edges = n(),
            n_signal = sum(label == "signal"),
            n_contamination = sum(label == "contamination"),
            n_unassigned = sum(label %in% c("unassigned", "unstable")),
            n_disagree_soft = sum(label != label_soft),
            n_disagree_fixed_denominator = sum(label != label_fixed_denominator),
            max_x_signal = suppressWarnings(max(x_fixed[label == "signal"])),
            max_x_contamination = suppressWarnings(max(x_fixed[label == "contamination"])),
            local_status = first(local_status),
            .groups = "drop") %>%
  mutate(across(starts_with("max_x"), ~ if_else(is.finite(.x), .x, NA_real_))) %>%
  left_join(asv_locality %>% distinct(asv_id, taxon_label), by = "asv_id")

write_csv(edge_classification, edges_out)
write_csv(library_sink_scores, libraries_out)
write_csv(asv_system_summary, asv_system_out)
write_csv(iteration_log, iteration_log_out)
```

# Iteration
```{r}
iteration_log %>%
  pivot_longer(c(signal, contamination, unassigned), names_to = "label", values_to = "n_edges") %>%
  ggplot(aes(step, n_edges, colour = label)) +
  geom_step() +
  facet_wrap(~ variant) +
  scale_colour_manual(values = c(signal = "#4e79a7", contamination = "#e15759", unassigned = "#999999")) +
  labs(x = "step", y = "free edges", title = "Hard assignment by step") +
  theme_minimal()

iteration_log %>%
  group_by(variant, tau) %>%
  summarise(steps = n(), added = sum(n_added), dropped = sum(n_dropped), .groups = "drop") %>%
  arrange(variant, desc(tau)) %>%
  knitr::kable()
```

# Agreement between variants
```{r}
free_edges = edge_classification %>% filter(role == "free")

free_edges %>% count(hard = label, soft = label_soft) %>%
  pivot_wider(names_from = soft, values_from = n, values_fill = 0, names_prefix = "soft_") %>%
  knitr::kable(caption = "hard (rows) vs soft (columns)")

free_edges %>% count(hard = label, fixed = label_fixed_denominator) %>%
  pivot_wider(names_from = fixed, values_from = n, values_fill = 0, names_prefix = "fixed_") %>%
  knitr::kable(caption = "hard, dynamic denominator (rows) vs fixed denominator (columns)")
```

# Reference cases

Hand-picked cases, a sanity check only -- never scored, since they are not a random sample. Each label is expanded to the free edges it covers (`feature_id` rows override their ASV x system row). x limits use x with the fixed denominator.
```{r}
asv_system_labels = edge_labels %>% filter(is.na(feature_id)) %>% mutate(label_row = row_number())
edge_level_labels = edge_labels %>% filter(!is.na(feature_id))

labelled_edges = bind_rows(
  free_edges %>%
    inner_join(asv_system_labels %>% select(asv_id, study_system, expected = label, min_x, max_x, note, label_row),
               by = c("asv_id", "study_system"), relationship = "many-to-many") %>%
    filter(is.na(min_x) | x_fixed >= min_x, is.na(max_x) | x_fixed <= max_x) %>%
    anti_join(edge_level_labels, by = c("asv_id", "feature_id")),
  free_edges %>%
    inner_join(edge_level_labels %>% select(asv_id, feature_id, expected = label, note), by = c("asv_id", "feature_id")))

label_evaluation = labelled_edges %>%
  pivot_longer(c(label, label_soft, label_fixed_denominator), names_to = "variant", values_to = "result") %>%
  mutate(variant = recode(variant, label = "hard", label_soft = "soft", label_fixed_denominator = "hard_fixed"),
         outcome = case_when(result %in% c("unassigned", "unstable") ~ "unassigned",
                             result == expected ~ "agree",
                             .default = "disagree"))

label_evaluation %>%
  count(variant, expected, outcome) %>%
  pivot_wider(names_from = outcome, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "edges under manual labels")

label_evaluation %>%
  filter(variant == "hard") %>%
  group_by(asv_id, study_system, expected, note) %>%
  summarise(n_edges = n(), agree = sum(outcome == "agree"), disagree = sum(outcome == "disagree"),
            unassigned = sum(outcome == "unassigned"), .groups = "drop") %>%
  knitr::kable(caption = "hard, per label")
```

# Validation

Scored against the blind-labelled validation sample (script 14), following the validation protocol in `analysis/README.md`. Each labelled edge is weighted by its stratum's population over the number of labelled edges of that stratum in the evaluated split, so the estimates hold for the whole population (all eDNA) or for its amphibian strata. Edges labelled `uncertain` are not scored. Cost per edge: `false_signal` for a contamination edge called signal, `false_contamination` for a signal edge called contamination, `unassigned` for an edge left unassigned. Clamped edges count as contamination; `script_08` (removed = contamination, kept = signal) and `all_signal` are baselines. Confidence intervals are from a stratified bootstrap, resampling edges within strata; differences are paired (the same resamples for every model).
```{r}
evaluated_split = if (evaluate_test_set) "test" else "development"
if (evaluate_test_set) {
  cat("NOTE: scoring the sealed TEST set -- only for a final comparison of candidate versions\n")
}

model_names = c("hard", "soft", "hard_fixed", "script_08", "all_signal")
domains = c("amphibian", "all eDNA")

validation_metrics = tibble()
if (!file.exists(validation_sample_file) || !file.exists(validation_labels_file)) {
  cat("No validation labels yet (", validation_labels_file, "): label the sample in script 12, export, and re-run.\n")
} else {
  validation_sample = read_csv(validation_sample_file, show_col_types = FALSE)
  validation_labels = read_csv(validation_labels_file, col_types = cols(.default = col_character())) %>%
    select(validation_id, truth = label)
  stopifnot(all(validation_labels$truth %in% c("signal", "contamination", "uncertain")))

  predictions = edge_classification %>%
    transmute(asv_id, feature_id,
              hard = label, soft = label_soft, hard_fixed = label_fixed_denominator,
              script_08 = if_else(removed_by_08, "contamination", "signal"),
              all_signal = "signal")

  scored = validation_sample %>%
    filter(split == evaluated_split) %>%
    inner_join(validation_labels, by = "validation_id") %>%
    left_join(predictions, by = c("asv_id", "feature_id")) %>%
    group_by(stratum) %>%
    mutate(weight = N_h / n()) %>%
    ungroup()

  # every sampled edge must have a prediction: an NA prediction would otherwise score as cost 0
  unmatched = scored %>% filter(is.na(hard))
  if (nrow(unmatched) > 0) {
    stop(nrow(unmatched), " labelled validation edge(s) have no prediction (population mismatch with script 14?): ",
         paste(head(unmatched$validation_id, 10), collapse = ", "))
  }

  cat(nrow(scored), "labelled", evaluated_split, "edges (of", sum(validation_sample$split == evaluated_split), "in the split),",
      sum(scored$truth == "uncertain"), "uncertain;", n_distinct(scored$stratum), "of", n_distinct(validation_sample$stratum), "strata covered\n")

  # per-edge cost for each model (NA for edges labelled uncertain)
  edge_cost = function(truth, prediction) {
    prediction = if_else(prediction %in% c("unassigned", "unstable"), "unassigned", prediction)
    case_when(truth == "uncertain" ~ NA_real_,
              prediction == "unassigned" ~ validation_costs[["unassigned"]],
              truth == "contamination" & prediction == "signal" ~ validation_costs[["false_signal"]],
              truth == "signal" & prediction == "contamination" ~ validation_costs[["false_contamination"]],
              .default = 0)
  }
  cost_matrix = sapply(model_names, \(m) edge_cost(scored$truth, scored[[m]]))

  weighted_share = function(w, hit, among) if (sum(w[among]) > 0) sum(w[among & hit]) / sum(w[among]) else NA_real_

  metrics = expand_grid(domain = domains, model = model_names) %>%
    mutate(m = pmap(list(domain, model), \(dom, m) {
      in_domain = if (dom == "amphibian") scored$group == "amphibian" else rep(TRUE, nrow(scored))
      truth = scored$truth[in_domain]
      pred = if_else(scored[[m]][in_domain] %in% c("unassigned", "unstable"), "unassigned", scored[[m]][in_domain])
      w = scored$weight[in_domain]
      decided = truth != "uncertain"
      assigned = pred != "unassigned"
      tibble(n_labelled = sum(in_domain), n_uncertain = sum(!decided),
             cost = sum(w[decided] * cost_matrix[in_domain, m][decided]) / sum(w[decided]),
             coverage = weighted_share(w, assigned, decided),
             false_signal_rate = weighted_share(w, pred == "signal", decided & truth == "contamination"),
             false_contamination_rate = weighted_share(w, pred == "contamination", decided & truth == "signal"),
             accuracy_assigned = weighted_share(w, pred == truth, decided & assigned))
    })) %>%
    unnest(m)

  # stratified bootstrap: resample edges within strata, the same resamples for every model
  set.seed(1)
  strata_rows = split(seq_len(nrow(scored)), scored$stratum)
  boot_costs = map(domains, \(dom) {
    in_domain = if (dom == "amphibian") scored$group == "amphibian" else rep(TRUE, nrow(scored))
    rows_by_stratum = keep(map(strata_rows, \(r) r[in_domain[r]]), \(r) length(r) > 0)
    t(replicate(bootstrap_reps, {
      rows = unlist(map(rows_by_stratum, \(r) r[sample.int(length(r), length(r), replace = TRUE)]))
      w = scored$weight[rows]
      decided = scored$truth[rows] != "uncertain"
      colSums(w[decided] * cost_matrix[rows, , drop = FALSE][decided, , drop = FALSE]) / sum(w[decided])
    }))
  }) %>% set_names(domains)

  cost_ci = imap_dfr(boot_costs, \(b, dom) tibble(domain = dom, model = colnames(b),
                                                  cost_lo = apply(b, 2, quantile, 0.025, na.rm = TRUE),
                                                  cost_hi = apply(b, 2, quantile, 0.975, na.rm = TRUE),
                                                  diff_vs_08 = colMeans(b - b[, "script_08"], na.rm = TRUE),
                                                  diff_vs_08_lo = apply(b - b[, "script_08"], 2, quantile, 0.025, na.rm = TRUE),
                                                  diff_vs_08_hi = apply(b - b[, "script_08"], 2, quantile, 0.975, na.rm = TRUE)))

  validation_metrics = metrics %>%
    left_join(cost_ci, by = c("domain", "model")) %>%
    mutate(model_version = model_version, split = evaluated_split, scored_on = as.character(Sys.Date()), .before = 1)

  validation_metrics %>%
    transmute(domain, model, n_labelled, n_uncertain,
              cost = sprintf("%.3f [%.3f, %.3f]", cost, cost_lo, cost_hi),
              `cost - script 08` = sprintf("%+.3f [%+.3f, %+.3f]", diff_vs_08, diff_vs_08_lo, diff_vs_08_hi),
              coverage = round(coverage, 3), false_signal_rate = round(false_signal_rate, 3),
              false_contamination_rate = round(false_contamination_rate, 3), accuracy_assigned = round(accuracy_assigned, 3)) %>%
    knitr::kable(caption = paste0("model ", model_version, ", ", evaluated_split, " set (weighted; 95% bootstrap intervals)")) %>%
    print()

  scored %>%
    mutate(prediction = if_else(hard %in% c("unassigned", "unstable"), "unassigned", hard)) %>%
    count(group, truth, prediction) %>%
    pivot_wider(names_from = prediction, values_from = n, values_fill = 0) %>%
    knitr::kable(caption = "hard: labelled edges, truth (rows) vs prediction (columns), unweighted counts") %>%
    print()

  write_csv(validation_metrics, validation_metrics_out)
}
```

# Comparison with script 08
```{r}
free_edges %>%
  count(script_08 = if_else(removed_by_08, "removed", "kept"), label) %>%
  pivot_wider(names_from = label, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "free edges: script 08 (rows) vs hard label (columns)")
```

# Validation against GBIF locality

Script 07's `local_status` was not used to classify. `other_system` and `not_in_run` edges should mostly be contamination (barring GBIF gaps); `local` edges mostly, but not all, signal (within-system cross-talk is invisible to GBIF).
```{r}
free_edges %>%
  mutate(local_status = replace_na(local_status, "(none)")) %>%
  count(local_status, label) %>%
  group_by(local_status) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(share, local_status, fill = label)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(values = c(signal = "#4e79a7", contamination = "#e15759", unassigned = "#999999", unstable = "#f28e2b")) +
  labs(x = "share of free edges", y = "script 07 local_status", title = "Hard labels by GBIF locality") +
  theme_minimal()
```

# Term contributions
```{r}
free_edges %>%
  select(label, starts_with("c_")) %>%
  pivot_longer(-label, names_to = "term", values_to = "contribution") %>%
  mutate(term = factor(str_remove(term, "^c_"), levels = term_names)) %>%
  ggplot(aes(contribution, term, fill = label)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_fill_manual(values = c(signal = "#4e79a7", contamination = "#e15759", unassigned = "#999999", unstable = "#f28e2b")) +
  labs(x = "log-odds contribution (final state)", y = NULL, title = "Term contributions by hard label") +
  theme_minimal()
```
