---
title: "Edge classification"
author: "Cob Staines"
output: html_document
---

# Description

This script classifies every **edge** of the sample--ASV network -- a detection of an ASV in a field library -- as `signal` or `contamination`, by the **model ladder** and **validation protocol** in `analysis/README.md`. It **only classifies, it drops nothing**, and nothing downstream reads its outputs.

**Edges.** Hard-flagged ASVs (script 06) are left out. ASVs in `known_contaminants` are contamination in every model (role `known_contaminant`). Every other field detection is a **free** edge, scored by each model.

**Features** (one row per free edge; data only, never labels):

- `log10_x`: x = the edge's reads over its library's reads, excluding hard-flagged and known-contaminant reads.
- `home_share` = mu(a, this system) / (mu(a, this system) + max over other systems of mu(a, other system)), where mu(a, g) is the mean x of ASV a over the libraries of group g, zeros included.
- `pc_component`: the ASV is a PCR positive-control component (script 08).
- `gbif_local` (local, local genus only) and `gbif_not_local` (other system, not in run), from script 07; anthropogenic and unassessable are neither.
- `kit_share` = mu(a, this system) / (mu(a, this system) + mu(a, lab negatives)); lab negatives are extraction and PCR negatives.
- `human_share`: the share of the rest of the library's reads (the edge's own reads left out) from ASVs of genus *Homo* -- an indicator of low target biomass.
- `dirty_asv`: the mean `human_share` of the ASV's other edges in the system, minus the mean over all free edges of the system (0 when the ASV has no other edge there).
- `gbif_local_amphibian`, `gbif_not_local_amphibian`: the GBIF features for amphibian ASVs (script 06 class Amphibia) only, 0 for other ASVs (refinement M2a).
- `neg_present`: the ASV is in any of the library's linked negatives (field, extraction, PCR); `neg_ratio`: log10 of the edge's x over the ASV's largest x in those negatives, capped at +/- 3, 0 when absent (refinement M2n).

**Models.** Baselines are fixed rules: `all_signal`; `script_08` (removed by step 8 = contamination); `neg_presence` (contamination if the ASV is in any of the library's linked negatives); `neg_subtract` (contamination if the edge's reads minus the ASV's largest read count in a linked negative is <= 0); `legacy` (the v2.1 labels from `13_legacy_edge_classification.Rmd`, when its output exists). The **ladder** rungs B1, M1, M1b, M2, M3, M4, M5 are logistic regressions, each adding one mechanism's features to the rung below (refinement rungs, e.g. M2a and M2n, branch off a ladder rung: see the change log), with each coefficient's sign fixed in advance and a Normal(0, `prior_sd`) prior on the coefficients of the standardised features. They are fitted on the **development** labels only (`signal` vs `contamination`; `uncertain` left out), unweighted.

**Decisions.** A model's probability of signal p becomes the decision of least expected cost under `validation$costs`: with costs 2 / 1 / 0.25, contamination if p < 0.25, signal if p > 0.875, unassigned in between.

**Performance** is not shown unless `validation$report_performance` is true: then models are compared by repeated cross-validation within the development set, with folds grouped by ASV x study system (weighted cost per edge, amphibian and all eDNA), the simplest rung within one standard error of the best is marked, and a drop-one analysis removes each mechanism from the top rung. The sealed **test** set is read only when `validation$evaluate_test_set` is true, for the models in `validation$final_candidates`.

# Config
```{r}
librarian::shelf(tidyverse, yaml, here)

# read shared pipeline config (see 00_pipeline_config.yml / 00_run_pipeline.R)
pipeline_config = read_yaml(here("analysis", "general", "r", "00_pipeline_config.yml"))

run_dir = here(pipeline_config$run_dir)
cat(paste0("Run_dir: ", run_dir, "\n"))

config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

known_contaminants = map_chr(pipeline_config$known_contaminants, "asv_id")

ec = pipeline_config$edge_classification
model_version = ec$model_version
prior_sd = as.numeric(ec$prior_sd)
cv_folds = as.integer(ec$cv_folds)
cv_repeats = as.integer(ec$cv_repeats)
cv_seed = as.integer(ec$cv_seed)

# validation (see the validation protocol in analysis/README.md)
vc = pipeline_config$validation
validation_dir = here(run_dir, vc$dir)
costs = map_dbl(vc$costs, as.numeric)
report_performance = isTRUE(vc$report_performance)
evaluate_test_set = isTRUE(vc$evaluate_test_set)
final_candidates = as.character(unlist(vc$final_candidates))
bootstrap_reps = as.integer(vc$bootstrap_reps)
validation_sample_file = file.path(validation_dir, paste0(config$run$name, "_validation_sample.csv"))
validation_labels_file = file.path(validation_dir, paste0(config$run$name, "_validation_labels.csv"))

possible_study_systems = c("brazil", "panama", "pennsylvania", "sierra_nevada")
negative_classes = c("field_negative", "extraction_negative", "pcr_negative")

# the ladder: each rung adds one mechanism's features to the rung below; signs fixed in advance
# (+1: evidence for signal, -1: evidence for contamination)
feature_signs = c(log10_x = 1, home_share = 1, pc_component = -1, gbif_local = 1, gbif_not_local = -1,
                  kit_share = 1, human_share = -1, dirty_asv = -1,
                  gbif_local_amphibian = 1, gbif_not_local_amphibian = -1, neg_present = -1, neg_ratio = 1)
mechanisms = list(abundance = "log10_x", home = "home_share", positive_control = "pc_component",
                  gbif = c("gbif_local", "gbif_not_local"), kitome = "kit_share", library_human = "human_share",
                  asv_dirty = "dirty_asv", gbif_amphibian = c("gbif_local_amphibian", "gbif_not_local_amphibian"),
                  own_negatives = c("neg_present", "neg_ratio"))
ladder = list(B1 = "abundance", M1 = "home", M1b = "positive_control", M2 = "gbif", M3 = "kitome",
              M4 = "library_human", M5 = "asv_dirty") %>%
  accumulate(c)
# refinement rungs (see the change log in analysis/README.md): branch off a ladder rung
refinements = list(M2a = c(ladder$M1b, "gbif_amphibian"),
                   M2n = c(ladder$M1b, "gbif_amphibian", "own_negatives"))
# rungs in order of complexity, refinements after the ladder rung they parallel
rungs = c(ladder[c("B1", "M1", "M1b", "M2")], refinements[c("M2a", "M2n")], ladder[c("M3", "M4", "M5")]) %>%
  map(\(m) unname(unlist(mechanisms[m])))
# drop-one analysis starts from the top of the ladder
top_rung = "M5"
baseline_names = c("all_signal", "script_08", "neg_presence", "neg_subtract", "legacy")

# input files
sample_map_file = here(run_dir, "output", paste0(config$run$name, "_03_sample_id_map.csv"))
classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))
locality_file = here(run_dir, "output", paste0(config$run$name, "_07_asv_locality.csv"))
controlled_file = here(run_dir, "output", paste0(config$run$name, "_08_asv_controlled.csv"))
# optional: the legacy model's labels (13_legacy_edge_classification.Rmd)
legacy_file = here(run_dir, "output", paste0(config$run$name, "_13_legacy_edge_classification.csv"))

# output files
edges_out = here(run_dir, "output", paste0(config$run$name, "_13_edge_classification.csv"))
coefficients_out = here(run_dir, "output", paste0(config$run$name, "_13_ladder_coefficients.csv"))
network_edges_out = here(run_dir, "output", paste0(config$run$name, "_13_network_edges_detection.csv"))
cv_metrics_out = here(run_dir, "output", paste0(config$run$name, "_13_cv_metrics.csv"))
drop_one_out = here(run_dir, "output", paste0(config$run$name, "_13_drop_one.csv"))
test_metrics_out = here(run_dir, "output", paste0(config$run$name, "_13_test_metrics.csv"))
```

# Load data
```{r}
input_files = c(sample_map_file, classification_file, locality_file, controlled_file)
missing_inputs = input_files[!file.exists(input_files)]
if (length(missing_inputs) > 0) {
  stop("Missing input(s):\n\t", paste(missing_inputs, collapse = "\n\t"))
}

# guess_max = Inf: some sample map columns are NA for most rows, and would otherwise be guessed
# as logical and their later values dropped
sample_map = read_csv(sample_map_file, show_col_types = FALSE, guess_max = Inf)
asv_classification = read_csv(classification_file, show_col_types = FALSE)
asv_locality = read_csv(locality_file, show_col_types = FALSE)
asv_controlled = read_csv(controlled_file, show_col_types = FALSE, guess_max = Inf) %>%
  select(asv_id, feature_id, asv_count_raw, asv_count_controlled, positive_control_component)

stopifnot(!anyDuplicated(sample_map$feature_id))
```

# Libraries and edges
```{r}
study_systems = intersect(possible_study_systems, sample_map$study_system)

libraries = sample_map %>%
  transmute(feature_id,
            library_class = case_when(
              pcr_control_type %in% "pcr_positive_control" ~ "pcr_positive",
              pcr_control_type %in% "pcr_negative_control" ~ "pcr_negative",
              extraction_negative_control %in% TRUE ~ "extraction_negative",
              study_system %in% study_systems & field_negative_control %in% TRUE ~ "field_negative",
              study_system %in% study_systems ~ "field_sample",
              .default = NA_character_),
            study_system = if_else(library_class %in% c("field_sample", "field_negative"), study_system, NA_character_),
            sequencing_date = as.character(sequencing_date))

cat("libraries without a class (left out):", sum(is.na(libraries$library_class)), "\n")
libraries = libraries %>% filter(!is.na(library_class))

hard_flagged = asv_classification %>%
  filter(short_sequence_flag %in% TRUE | contaminant_flag %in% TRUE) %>%
  pull(asv_id)
human_asvs = asv_classification %>% filter(genus %in% "Homo") %>% pull(asv_id)
amphibian_asvs = asv_classification %>% filter(class %in% "Amphibia") %>% pull(asv_id)

# every detection in a classified library; hard-flagged ASVs are left out of everything
detections = asv_controlled %>%
  filter(asv_count_raw > 0) %>%
  inner_join(libraries, by = "feature_id") %>%
  rename(reads = asv_count_raw) %>%
  mutate(role = case_when(asv_id %in% hard_flagged ~ "hard_flag",
                          asv_id %in% known_contaminants ~ "known_contaminant",
                          library_class == "field_sample" ~ "free",
                          .default = "control"))

# library reads excluding hard-flagged and known-contaminant reads (the denominator of x), and the
# human reads among them
library_reads = detections %>%
  filter(role %in% c("free", "control")) %>%
  group_by(feature_id) %>%
  summarise(N = sum(reads), human_reads = sum(reads[asv_id %in% human_asvs]), .groups = "drop")

libraries = libraries %>%
  left_join(library_reads, by = "feature_id") %>%
  mutate(N = replace_na(N, 0), human_reads = replace_na(human_reads, 0))

counted = detections %>%
  filter(role %in% c("free", "control")) %>%
  left_join(libraries %>% select(feature_id, N, human_reads), by = "feature_id") %>%
  mutate(x = reads / N)

detections %>% count(role, library_class) %>% knitr::kable()
```

# Features
```{r}
# mu(a, g): mean x of ASV a over the libraries of group g, zeros included
n_field = libraries %>% filter(library_class == "field_sample", N > 0) %>% count(study_system, name = "n")
mu_system = counted %>%
  filter(library_class == "field_sample") %>%
  group_by(asv_id, study_system) %>%
  summarise(x_sum = sum(x), .groups = "drop") %>%
  left_join(n_field, by = "study_system") %>%
  transmute(asv_id, study_system, mu = x_sum / n)

n_lab = sum(libraries$library_class %in% c("extraction_negative", "pcr_negative") & libraries$N > 0)
mu_lab = counted %>%
  filter(library_class %in% c("extraction_negative", "pcr_negative")) %>%
  group_by(asv_id) %>%
  summarise(mu_lab = sum(x) / n_lab, .groups = "drop")

# home share: this system against the strongest other system
asv_system = mu_system %>%
  group_by(asv_id) %>%
  mutate(mu_other_max = map_dbl(seq_along(mu), \(i) max(c(0, mu[-i])))) %>%
  ungroup() %>%
  left_join(mu_lab, by = "asv_id") %>%
  mutate(mu_lab = replace_na(mu_lab, 0),
         home_share = mu / (mu + mu_other_max),
         kit_share = mu / (mu + mu_lab))

gbif = asv_locality %>%
  transmute(asv_id, study_system, local_status,
            gbif_local = as.numeric(local_status %in% c("local", "local_genus_only")),
            gbif_not_local = as.numeric(local_status %in% c("other_system", "not_in_run")))

pc_components = asv_controlled %>%
  group_by(asv_id) %>%
  summarise(pc_component = as.numeric(any(positive_control_component %in% TRUE)), .groups = "drop")

fe = counted %>%
  filter(role == "free") %>%
  left_join(asv_system %>% select(asv_id, study_system, mu, home_share, kit_share), by = c("asv_id", "study_system")) %>%
  left_join(gbif, by = c("asv_id", "study_system")) %>%
  left_join(pc_components, by = "asv_id") %>%
  mutate(across(c(gbif_local, gbif_not_local, pc_component), \(v) replace_na(v, 0)),
         # GBIF for amphibians only (refinement M2a)
         gbif_local_amphibian = gbif_local * (asv_id %in% amphibian_asvs),
         gbif_not_local_amphibian = gbif_not_local * (asv_id %in% amphibian_asvs),
         log10_x = log10(x),
         # human share of the rest of the library: the edge's own reads left out, so a human edge
         # does not count itself
         own_human = if_else(asv_id %in% human_asvs, reads, 0),
         human_share = if_else(N - reads > 0, (human_reads - own_human) / (N - reads), 0))

# dirty ASV: mean human share of the ASV's other edges in the system, against the system mean
fe = fe %>%
  group_by(study_system) %>%
  mutate(system_mean_human = mean(human_share)) %>%
  group_by(asv_id, study_system) %>%
  mutate(n_other = n() - 1L,
         dirty_asv = if_else(n_other > 0, (sum(human_share) - human_share) / n_other - system_mean_human, 0)) %>%
  ungroup() %>%
  select(-own_human, -n_other, -system_mean_human)

# each library's linked negatives, via the control-group ids shared with them
link_controls = function(group_col, control_class) {
  groups = sample_map %>%
    select(feature_id, group = all_of(group_col)) %>%
    filter(!is.na(group))
  controls = libraries$feature_id[libraries$library_class == control_class]
  groups %>%
    inner_join(groups %>% filter(feature_id %in% controls) %>% rename(control_id = feature_id),
               by = "group", relationship = "many-to-many") %>%
    filter(feature_id != control_id) %>%
    distinct(feature_id, control_id)
}

control_links = bind_rows(link_controls("field_negative_control_group", "field_negative"),
                          link_controls("extraction_negative_control_group", "extraction_negative"),
                          link_controls("pcr_control_group", "pcr_negative")) %>%
  distinct()

own_negative_reads = control_links %>%
  inner_join(counted %>% filter(library_class %in% negative_classes) %>% select(control_id = feature_id, asv_id, reads, x),
             by = "control_id", relationship = "many-to-many") %>%
  group_by(feature_id, asv_id) %>%
  summarise(own_negative_max_reads = max(reads), own_negative_max_x = max(x), .groups = "drop")

# own negatives (refinement M2n): is the ASV in the library's linked negatives, and how far above them
fe = fe %>%
  left_join(own_negative_reads, by = c("feature_id", "asv_id")) %>%
  mutate(own_negative_max_reads = replace_na(own_negative_max_reads, 0),
         own_negative_max_x = replace_na(own_negative_max_x, 0),
         neg_present = as.numeric(own_negative_max_x > 0),
         neg_ratio = if_else(own_negative_max_x > 0, pmin(pmax(log10(x / own_negative_max_x), -3), 3), 0))

stopifnot(!anyNA(fe[names(feature_signs)]))

fe %>%
  select(all_of(names(feature_signs))) %>%
  pivot_longer(everything(), names_to = "feature") %>%
  group_by(feature = factor(feature, names(feature_signs))) %>%
  summarise(mean = mean(value), sd = sd(value), min = min(value), max = max(value)) %>%
  knitr::kable(digits = 3, caption = paste(nrow(fe), "free edges"))
```

# Baselines
```{r}
legacy = if (file.exists(legacy_file)) {
  read_csv(legacy_file, show_col_types = FALSE, guess_max = Inf) %>%
    transmute(asv_id, feature_id, legacy = if_else(label == "unstable", "unassigned", label))
} else {
  cat("No legacy output (", legacy_file, "): the legacy baseline is left out\n")
  NULL
}

fe = fe %>%
  mutate(all_signal = "signal",
         script_08 = if_else(asv_count_controlled == 0, "contamination", "signal"),
         neg_presence = if_else(own_negative_max_reads > 0, "contamination", "signal"),
         neg_subtract = if_else(reads - own_negative_max_reads <= 0, "contamination", "signal"))
if (!is.null(legacy)) fe = fe %>% left_join(legacy, by = c("asv_id", "feature_id"))
baselines = intersect(baseline_names, names(fe))

fe %>%
  select(all_of(baselines)) %>%
  pivot_longer(everything(), names_to = "model", values_to = "label") %>%
  count(model = factor(model, baselines), label) %>%
  pivot_wider(names_from = label, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "free edges by baseline label")
```

# Fitting
```{r}
# the decision of least expected cost for probability of signal p
decide = function(p) {
  expected = cbind(signal = costs[["false_signal"]] * (1 - p),
                   contamination = costs[["false_contamination"]] * p,
                   unassigned = rep(costs[["unassigned"]], length(p)))
  colnames(expected)[max.col(-expected, ties.method = "last")]
}

# sign-constrained logistic regression with a Normal(0, prior_sd) prior on the coefficients of the
# z-scored features (intercept unpenalised). A feature constant in the training rows gets coefficient 0.
fit_logistic = function(X, y) {
  center = colMeans(X)
  scale = apply(X, 2, sd)
  usable = scale > 0
  Z = sweep(sweep(X, 2, center), 2, ifelse(usable, scale, 1), "/")
  Z1 = cbind(1, Z)
  signs = feature_signs[colnames(X)]
  objective = function(b) {
    eta = drop(Z1 %*% b)
    sum(log1p(exp(-abs(eta))) + pmax(eta, 0) - y * eta) + sum(b[-1]^2) / (2 * prior_sd^2)
  }
  gradient = function(b) {
    p = plogis(drop(Z1 %*% b))
    drop(crossprod(Z1, p - y)) + c(0, b[-1] / prior_sd^2)
  }
  lower = c(-Inf, ifelse(usable & signs > 0, 0, ifelse(usable, -Inf, 0)))
  upper = c(Inf, ifelse(usable & signs < 0, 0, ifelse(usable, Inf, 0)))
  start = c(qlogis(min(max(mean(y), 0.01), 0.99)), rep(0, ncol(X)))
  opt = optim(start, objective, gradient, method = "L-BFGS-B", lower = lower, upper = upper)
  if (opt$convergence != 0) warning("fit did not converge: ", opt$message)
  list(intercept = opt$par[1], beta = setNames(opt$par[-1], colnames(X)), center = center,
       scale = ifelse(usable, scale, 1))
}

predict_logistic = function(fit, X) {
  Z = sweep(sweep(X[, names(fit$beta), drop = FALSE], 2, fit$center), 2, fit$scale, "/")
  plogis(fit$intercept + drop(Z %*% fit$beta))
}

# training data: labelled development edges only. The test split is dropped before any label is
# joined, so its labels are never read here.
if (!file.exists(validation_sample_file) || !file.exists(validation_labels_file)) {
  stop("No validation sample / labels in ", validation_dir, ": draw (script 14) and label (script 12) first")
}
validation_sample = read_csv(validation_sample_file, show_col_types = FALSE)
development = validation_sample %>% filter(split == "development")
development_labels = read_csv(validation_labels_file, col_types = cols(.default = col_character())) %>%
  filter(validation_id %in% development$validation_id) %>%
  select(validation_id, truth = label)
stopifnot(!anyDuplicated(development_labels$validation_id),
          all(development_labels$truth %in% c("signal", "contamination", "uncertain")))

labelled = development %>%
  inner_join(development_labels, by = "validation_id") %>%
  # weight: the stratum's population over its labelled development edges (uncertain included)
  group_by(stratum) %>%
  mutate(weight = N_h / n()) %>%
  ungroup() %>%
  left_join(fe %>% select(asv_id, feature_id, all_of(names(feature_signs)), all_of(baselines)),
            by = c("asv_id", "feature_id"))

# every sampled edge must be a free edge here (the same population as script 14)
unmatched = labelled %>% filter(is.na(log10_x))
if (nrow(unmatched) > 0) {
  stop(nrow(unmatched), " labelled development edge(s) are not free edges here (population mismatch with script 14?): ",
       paste(head(unmatched$validation_id, 10), collapse = ", "))
}

training = labelled %>% filter(truth != "uncertain") %>% mutate(y = as.numeric(truth == "signal"))
cat(nrow(development), "development edges,", nrow(labelled), "labelled,", nrow(training), "used for fitting (",
    sum(training$y), "signal,", sum(1 - training$y), "contamination ),", n_distinct(paste(training$asv_id, training$study_system)),
    "ASV x system groups\n")

fits = map(rungs, \(features) fit_logistic(as.matrix(training[features]), training$y))

coefficients = imap_dfr(fits, \(fit, rung) tibble(
  model_version = model_version,
  rung = rung,
  term = c("(intercept)", names(fit$beta)),
  sign = c(NA, feature_signs[names(fit$beta)]),
  estimate_standardised = c(fit$intercept, fit$beta),
  # per unit of the raw feature (intercept at the raw features' zero)
  estimate = c(fit$intercept - sum(fit$beta * fit$center / fit$scale), fit$beta / fit$scale),
  feature_mean = c(NA, fit$center),
  feature_sd = c(NA, fit$scale),
  n_train = nrow(training),
  fitted_on = as.character(Sys.Date())))

coefficients %>%
  select(rung, term, estimate_standardised) %>%
  mutate(rung = factor(rung, names(rungs)), term = factor(term, c("(intercept)", names(feature_signs)))) %>%
  pivot_wider(names_from = rung, values_from = estimate_standardised) %>%
  arrange(term) %>%
  knitr::kable(digits = 2, caption = "coefficients per rung, on the standardised features (log-odds per SD); 0 = held at its sign bound")

write_csv(coefficients, coefficients_out)
```

# Classify
```{r}
X_all = as.matrix(fe[names(feature_signs)])
for (rung in names(rungs)) {
  p = predict_logistic(fits[[rung]], X_all)
  fe[[paste0("p_", rung)]] = p
  fe[[rung]] = decide(p)
}
models = c(baselines, names(rungs))

# label counts over all free edges (no labels involved)
fe %>%
  select(all_of(models)) %>%
  pivot_longer(everything(), names_to = "model", values_to = "label") %>%
  count(model = factor(model, models), label) %>%
  pivot_wider(names_from = label, values_from = n, values_fill = 0) %>%
  knitr::kable(caption = "free edges by model decision")

# every detection, clamped ones included (contamination in every model, with the reason)
edge_classification = detections %>%
  select(asv_id, feature_id, library_class, study_system, sequencing_date, reads, role) %>%
  filter(library_class == "field_sample") %>%
  left_join(fe %>% select(asv_id, feature_id, x, all_of(names(feature_signs)), local_status,
                          own_negative_max_reads, starts_with("p_"), all_of(models)),
            by = c("asv_id", "feature_id")) %>%
  mutate(across(all_of(models), \(v) if_else(role == "free", v, "contamination")),
         model_version = model_version, .before = 1)

write_csv(edge_classification, edges_out)
```

# Network export

A detection edge table (sample -- ASV) for Gephi / Cytoscape with every model's decision, to colour or filter the network by model. It covers every detection in every classified library, with the same `Source` / `Target` ids as script 11: import script 11's nodes table (`<run>_11_raw_network_nodes.csv`) first, then this edges table. Each model column holds the model's decision for free edges, `contamination` for known contaminants and hard-flagged ASVs, and `control` for detections in control libraries; `p_<rung>` is a rung's probability of signal. `Weight` is the within-library relative abundance over all raw reads, as in script 11.
```{r}
raw_library_reads = detections %>% group_by(feature_id) %>% summarise(raw_reads = sum(reads), .groups = "drop")

network_edges = detections %>%
  left_join(raw_library_reads, by = "feature_id") %>%
  left_join(fe %>% select(asv_id, feature_id, starts_with("p_"), all_of(models)), by = c("asv_id", "feature_id")) %>%
  transmute(Source = feature_id, Target = asv_id, Type = "Undirected", Weight = reads / raw_reads,
            reads, library_class, study_system, sequencing_date, role,
            across(all_of(models), \(v) case_when(role == "free" ~ v,
                                                  role %in% c("known_contaminant", "hard_flag") ~ "contamination",
                                                  .default = "control")),
            across(starts_with("p_")))

# empty strings (not "NA") so Gephi reads missing values as missing
write_csv(network_edges, network_edges_out, na = "")
cat(nrow(network_edges), "detection edges ->", network_edges_out, "\n")
```

# Performance (development set, cross-validated)

Shown only when `validation$report_performance` is true. Each rung is refitted on the training folds and scored on the held-out fold; folds are groups of ASV x study system, so an ASV's detections in a system are never split between fitting and scoring. Baselines need no fitting and are scored on the same folds, so differences are paired. Costs are weighted per edge (stratum population over its labelled development edges); `uncertain` edges are not scored.
```{r}
# per-edge cost of a decision (NA for uncertain truth)
edge_cost = function(truth, decision) {
  case_when(truth == "uncertain" ~ NA_real_,
            decision %in% c("unassigned", "unstable") ~ costs[["unassigned"]],
            truth == "contamination" & decision == "signal" ~ costs[["false_signal"]],
            truth == "signal" & decision == "contamination" ~ costs[["false_contamination"]],
            .default = 0)
}

# repeated grouped cross-validation: decisions for every scored edge, model and repeat
cross_validate = function(data, model_features, seed) {
  set.seed(seed)
  groups = paste(data$asv_id, data$study_system)
  unique_groups = unique(groups)
  map_dfr(seq_len(cv_repeats), \(r) {
    fold_of_group = setNames(sample(rep(seq_len(cv_folds), length.out = length(unique_groups))), unique_groups)
    fold = unname(fold_of_group[groups])
    map_dfr(seq_len(cv_folds), \(k) {
      train = data[fold != k, ]
      test = data[fold == k, ]
      fitted = map(model_features, \(features) {
        fit = fit_logistic(as.matrix(train[features]), train$y)
        decide(predict_logistic(fit, as.matrix(test[features])))
      })
      as_tibble(fitted) %>% mutate(row = which(fold == k), repeat_id = r, fold = k)
    })
  })
}

# weighted cost per model: the mean over repeats of the pooled cost, and the standard error from
# the spread over folds (mean over repeats of sd(fold costs) / sqrt(folds))
summarise_cv = function(cv_decisions, data, model_names, in_domain) {
  long = cv_decisions %>%
    pivot_longer(all_of(model_names), names_to = "model", values_to = "decision") %>%
    mutate(truth = data$truth[row], weight = data$weight[row], cost = edge_cost(truth, decision)) %>%
    filter(in_domain[row])
  folds = long %>%
    group_by(model, repeat_id, fold) %>%
    summarise(fold_cost = sum(weight * cost) / sum(weight), .groups = "drop")
  pooled = long %>%
    group_by(model, repeat_id) %>%
    summarise(cost = sum(weight * cost) / sum(weight),
              coverage = sum(weight[decision != "unassigned"]) / sum(weight),
              false_signal_rate = sum(weight[truth == "contamination" & decision == "signal"]) / sum(weight[truth == "contamination"]),
              false_contamination_rate = sum(weight[truth == "signal" & decision == "contamination"]) / sum(weight[truth == "signal"]),
              .groups = "drop") %>%
    group_by(model) %>%
    summarise(across(c(cost, coverage, false_signal_rate, false_contamination_rate), mean), .groups = "drop")
  se = folds %>%
    group_by(model, repeat_id) %>%
    summarise(se = sd(fold_cost) / sqrt(n()), .groups = "drop") %>%
    group_by(model) %>%
    summarise(se = mean(se), .groups = "drop")
  pooled %>% left_join(se, by = "model")
}

if (!report_performance) {
  cat("Performance reporting is off (validation$report_performance: false): models are trained, not scored.\n")
} else {
  scored = training
  # baselines as fixed decisions, repeated on the same rows and folds as the rungs
  cv_decisions = cross_validate(scored, rungs, cv_seed)
  for (b in baselines) cv_decisions[[b]] = scored[[b]][cv_decisions$row]

  domains = list(amphibian = scored$group == "amphibian", `all eDNA` = rep(TRUE, nrow(scored)))
  cv_metrics = imap_dfr(domains, \(in_domain, domain) {
    summarise_cv(cv_decisions, scored, models, in_domain) %>% mutate(domain = domain, .before = 1)
  }) %>%
    mutate(model = factor(model, models)) %>%
    arrange(domain, model)

  # the simplest rung within one standard error of the best rung (per domain)
  cv_metrics = cv_metrics %>%
    group_by(domain) %>%
    mutate(is_rung = model %in% names(rungs),
           best_rung_cost = min(cost[is_rung]),
           best_rung_se = se[is_rung][which.min(cost[is_rung])],
           within_1se = is_rung & cost <= best_rung_cost + best_rung_se,
           selected = within_1se & as.integer(model) == min(as.integer(model)[within_1se])) %>%
    ungroup() %>%
    select(-is_rung, -best_rung_cost, -best_rung_se) %>%
    mutate(model_version = model_version, scored_on = as.character(Sys.Date()), .before = 1)

  cv_metrics %>%
    select(domain, model, cost, se, coverage, false_signal_rate, false_contamination_rate, within_1se, selected) %>%
    knitr::kable(digits = 3, caption = paste0("development set, ", cv_repeats, " x ", cv_folds, "-fold grouped cross-validation (weighted)")) %>%
    print()

  # drop-one: remove each mechanism from the top rung, on the same folds
  top = top_rung
  in_top = keep(names(mechanisms), \(m) any(mechanisms[[m]] %in% rungs[[top]]))
  drop_features = map(set_names(in_top), \(m) setdiff(rungs[[top]], mechanisms[[m]]))
  drop_features = c(set_names(list(rungs[[top]]), top), drop_features[lengths(drop_features) > 0])
  drop_decisions = cross_validate(scored, drop_features, cv_seed)
  drop_one = imap_dfr(domains, \(in_domain, domain) {
    summarise_cv(drop_decisions, scored, names(drop_features), in_domain) %>% mutate(domain = domain, .before = 1)
  }) %>%
    group_by(domain) %>%
    mutate(change_vs_top = cost - cost[model == top]) %>%
    ungroup() %>%
    mutate(dropped = if_else(model == top, "(none)", model), .after = domain) %>%
    select(-model)

  drop_one %>%
    select(domain, dropped, cost, se, change_vs_top) %>%
    knitr::kable(digits = 3, caption = paste("drop-one from", top, "(a positive change means the mechanism helps)")) %>%
    print()

  write_csv(cv_metrics, cv_metrics_out)
  write_csv(drop_one, drop_one_out)
}
```

# Test set

Read only when `validation$evaluate_test_set` is true: the models in `validation$final_candidates` are fitted on all labelled development edges (as above) and scored once on the sealed test set, with 95% stratified bootstrap intervals for each model's cost and for paired differences from the standard-practice baselines. Record the scoring in the change log.
```{r}
if (!evaluate_test_set) {
  cat("The test set is sealed (validation$evaluate_test_set: false).\n")
} else {
  stopifnot(length(final_candidates) > 0, all(final_candidates %in% models))
  cat("NOTE: scoring the sealed TEST set -- only for the final comparison of candidate models\n")
  test = validation_sample %>% filter(split == "test")
  test_labels = read_csv(validation_labels_file, col_types = cols(.default = col_character())) %>%
    filter(validation_id %in% test$validation_id) %>%
    select(validation_id, truth = label)
  test_scored = test %>%
    inner_join(test_labels, by = "validation_id") %>%
    group_by(stratum) %>%
    mutate(weight = N_h / n()) %>%
    ungroup() %>%
    left_join(fe %>% select(asv_id, feature_id, all_of(models)), by = c("asv_id", "feature_id"))
  stopifnot(!anyNA(test_scored[final_candidates]))

  comparison = unique(c(final_candidates, intersect(c("script_08", "neg_presence", "neg_subtract"), models)))
  cost_matrix = sapply(comparison, \(m) edge_cost(test_scored$truth, test_scored[[m]]))
  set.seed(1)
  domains = list(amphibian = test_scored$group == "amphibian", `all eDNA` = rep(TRUE, nrow(test_scored)))
  test_metrics = imap_dfr(domains, \(in_domain, domain) {
    rows_by_stratum = keep(map(split(which(in_domain), test_scored$stratum[in_domain]), identity), \(r) length(r) > 0)
    decided = !is.na(cost_matrix[, 1])
    point = colSums(test_scored$weight[in_domain & decided] * cost_matrix[in_domain & decided, , drop = FALSE]) /
      sum(test_scored$weight[in_domain & decided])
    boot = t(replicate(bootstrap_reps, {
      rows = unlist(map(rows_by_stratum, \(r) r[sample.int(length(r), length(r), replace = TRUE)]))
      rows = rows[decided[rows]]
      colSums(test_scored$weight[rows] * cost_matrix[rows, , drop = FALSE]) / sum(test_scored$weight[rows])
    }))
    tibble(domain = domain, model = comparison, n_scored = sum(in_domain & decided), cost = point,
           cost_lo = apply(boot, 2, quantile, 0.025), cost_hi = apply(boot, 2, quantile, 0.975),
           diff_vs_08 = point - point[["script_08"]],
           diff_vs_08_lo = apply(boot - boot[, "script_08"], 2, quantile, 0.025),
           diff_vs_08_hi = apply(boot - boot[, "script_08"], 2, quantile, 0.975),
           diff_vs_neg_presence = point - point[["neg_presence"]],
           diff_vs_neg_presence_lo = apply(boot - boot[, "neg_presence"], 2, quantile, 0.025),
           diff_vs_neg_presence_hi = apply(boot - boot[, "neg_presence"], 2, quantile, 0.975))
  }) %>%
    mutate(model_version = model_version, scored_on = as.character(Sys.Date()), .before = 1)

  test_metrics %>%
    select(domain, model, n_scored, cost, cost_lo, cost_hi, diff_vs_08, diff_vs_neg_presence) %>%
    knitr::kable(digits = 3, caption = "TEST set (weighted; 95% stratified bootstrap intervals in the csv)") %>%
    print()
  write_csv(test_metrics, test_metrics_out)
}
```
