---
title: "Sample controls (methods)"
author: "Cob Staines"
output: html_document
---

# Description

This script calculates controlled ASV reads for each sample by subtracting any read counts found in controls (multiplied by a scaling factor `asv_control_th_factor`), with minimum controlled reads of 0. Lab positive and negative controls are applied to all samples globally, field negative controls are applied to corresponding field samples only.

# Config
```{r}
librarian::shelf(tidyverse, yaml, dbplyr, RPostgres, DBI, ribbitrrr, here, rio, janitor, phylotools)

# data directory
ddir = Sys.getenv("data_dir")

run_dir_methods = here("16S_sequence_processing", "runs", "methods_2026-03-13")

# read in config files
config_12s = read_yaml(here(run_dir_methods, "12S", "output", "metadata", "config.yml"))
config_16s = read_yaml(here(run_dir_methods, "16S", "output", "metadata", "config.yml"))

# asv control threshold factor
asv_control_th_factor = 100

# control output file
control_out = here(run_dir_methods, "12S", "output", paste0(config_12s$run$name, "_08_asv_controlled_joined.csv"))
```

# Load data
## metadata
```{r}
# sample metadata file
sample_lookup_md = import(here(Sys.getenv("data_dir"), "edna_results", "Illumina_runs_for_paper_2026-03-05.xlsx")) %>%
  clean_names() %>%
  filter(illumina_run %in% c("February_2023",
                             "December_2023")) %>%
  rename(passive_active_rz = passive_active_rz_lab) %>%
  mutate(passive_active_rz = if_else(passive_active_rz == "NA", NA, passive_active_rz),
         filter_size_um_rz = if_else(filter_size_um_rz == "NA", NA, filter_size_um_rz)) %>%
  select(-date,
         -region,
         -site,
         -x22)


sample_lookup_rib = read_csv(here(run_dir_methods, "12S", "output", paste0(config_12s$run$name, "_07_sample_id_map.csv")))

# augment control types
field_controls = tibble(sample_id = c("PAN_14-12S",
                                      "PAN_14-12S-CRISPR",
                                      "PAN_14-16Sblock",
                                      "PAN_14-16Sblock-CRISPR",
                                      "PAN_14-16Snoblock",
                                      "PAN_14-16Snoblock-CRISPR",
                                      "PENN_08-12S",
                                      "PENN_08-12S-CRISPR",
                                      "PENN_08-16Sblock",
                                      "PENN_08-16Sblock-CRISPR",
                                      "PENN_08-16Snoblock",
                                      "PENN_08-16Snoblock-CRISPR",
                                      "Tissue_63"),
                       control_type = "negative_field")

sample_lookup = sample_lookup_rib %>%
  rename(ribbitr_sample_name = edna_sample_name) %>%
  full_join(sample_lookup_md, by = "id_e_dna_sample") %>%
  mutate(sample_id = sub("-", "_", gsub("_", "-", sequencing_sample_name))) %>%
  left_join(field_controls, by = "sample_id") %>%
  select(sample_id,
         control_type,
         everything()) %>%
  mutate(crispr = grepl("CRISPR", sample_id),
         block = grepl("block", sample_id) & !grepl("noblock", sample_id),
         gene = case_when(
           grepl("12S", sample_id) ~ "12S",
           grepl("16S|Tissue|PCI|Neg_Control", sample_id) ~ "16S",
           .default = NA),
         control_type = case_when(
           grepl("^Positive", sample_id) ~ "positive",
           grepl("^Neg", sample_id) ~ "negative_lab",
           .default = control_type),
         country = if_else(is.na(country), "usa", country),
         extraction_type = if_else(grepl("PCI", sample_id), "PCI", extraction_type))

sample_simple = sample_lookup %>%
  select(sample_id,
         ribbitr_sample_name,
         gene,
         filter_size_um_rz,
         passive_active_rz,
         crispr,
         block,
         extraction_type,
         control_type)

# summary stats
sample_stats = sample_simple %>%
  group_by(gene,
           filter_size_um_rz,
           passive_active_rz,
           crispr,
           block,
           extraction_type,
           control_type) %>%
  count() %>%
  ungroup() %>%
  complete(gene,
           filter_size_um_rz,
           passive_active_rz,
           crispr,
           block,
           extraction_type,
           control_type,
           fill = list(n = 0)) %>%
  arrange(gene,
          filter_size_um_rz,
          passive_active_rz,
          crispr,
          block,
          extraction_type,
          control_type)

```

## Feature tables
```{r}
## 12s
feature_table_12s =
  read_tsv(here("16S_sequence_processing", config_12s$run$runDir, "analysis/s06_denoised_12S_eDNA/sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  dplyr::rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

feature_long_12s = feature_table_12s %>%
  pivot_longer(cols = -asv_id,
               names_to = "sample_id",
               values_to = "asv_count") %>%
  # filter(asv_count != 0) %>%
  select(sample_id,
         everything()) %>%
  arrange(sample_id,
          desc(asv_count))

samples_12s = feature_long_12s %>%
  select(sample_id) %>%
  distinct() %>%
  mutate(gene = "12S")

## 16s
feature_table_16s =
  read_tsv(here("16S_sequence_processing", config_16s$run$runDir, "analysis/s06_denoised_16S_eDNA/sample_table/feature-table.tsv"), skip = 1) %>%
  as.data.frame() %>%
  dplyr::rename("asv_id" = "#OTU ID") %>%
  arrange(asv_id)

feature_long_16s = feature_table_16s %>%
  pivot_longer(cols = -asv_id,
               names_to = "sample_id",
               values_to = "asv_count") %>%
  # filter(asv_count != 0) %>%
  select(sample_id,
         everything()) %>%
  arrange(sample_id,
          desc(asv_count))

samples_16s = feature_long_16s %>%
  select(sample_id) %>%
  distinct() %>%
  mutate(gene = "16S")

# # check alignment
# sample_align = sample_simple %>%
#   left_join(bind_rows(samples_12s,
#                       samples_16s), by = "sample_id")
# 
# sample_align$gene.x == sample_align$gene.y
# 
# # check no common asv_ids
# peace = get_dupes(bind_rows(bind_rows(feature_long_12s %>%
#                                         select(asv_id) %>%
#                                         distinct(),
#                                       feature_long_16s %>%
#                                         select(asv_id) %>%
#                                         distinct())), asv_id)
```


## Sequence tables
```{r}
## 12s
asv_seq_12s =
  read.fasta(here("16S_sequence_processing", paste0(config_12s$run$runDir, "/analysis/s06_denoised_", config_12s$taxonomy$gene, "_eDNA/representative_sequences/", 
                           list.files(here("16S_sequence_processing", paste0(config_12s$run$runDir, "/analysis/s06_denoised_", config_12s$taxonomy$gene, "_eDNA/representative_sequences"))), 
                           "/data/dna-sequences.fasta"))) %>%
  dplyr::rename(asv_id = seq.name) %>%
  arrange(asv_id)

## 16s
asv_seq_16s =
  read.fasta(here("16S_sequence_processing", paste0(config_16s$run$runDir, "/analysis/s06_denoised_", config_16s$taxonomy$gene, "_eDNA/representative_sequences/", 
                           list.files(here("16S_sequence_processing", paste0(config_16s$run$runDir, "/analysis/s06_denoised_", config_16s$taxonomy$gene, "_eDNA/representative_sequences"))), 
                           "/data/dna-sequences.fasta"))) %>%
  dplyr::rename(asv_id = seq.name) %>%
  arrange(asv_id)
```

## Classification tables
```{r}
classified_12s = read.csv(here(run_dir_methods, "12S", "output", "methods_2026-03-13_12s_06_asv_classification.csv"))
classified_16s = read.csv(here(run_dir_methods, "16S", "output", "methods_2026-03-13_16s_06_asv_classification.csv"))

```

# Controls
```{r}
sample_asv = sample_simple %>%
  left_join(bind_rows(feature_long_12s,
                      feature_long_16s), by = "sample_id")

control_asv_pos = sample_asv %>%
  filter(control_type == "positive") %>%
  group_by(asv_id, control_type) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")

control_asv_neg_lab = sample_asv %>%
  filter(control_type == "negative_lab") %>%
  group_by(asv_id, control_type) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")

control_asv_neg_field = sample_asv %>%
  filter(control_type == "negative_field") %>%
  group_by(asv_id, control_type) %>%
  summarise(max_control_asv_count = max(asv_count),
            .groups = "drop")

control_asv = bind_rows(control_asv_pos,
                        control_asv_neg_lab,
                        control_asv_neg_field) %>%
  filter(max_control_asv_count > 0) %>%
  group_by(asv_id) %>%
  slice_max(max_control_asv_count, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(asv_control_th = asv_control_th_factor * max_control_asv_count) %>%
  rename(control_th_type = control_type)

sample_asv_control = sample_asv %>%
  left_join(control_asv, by = "asv_id") %>%
  mutate(asv_count_controlled = asv_count - asv_control_th,
         asv_count_controlled = if_else(asv_count_controlled < 0, 0, asv_count_controlled),
         control_th_below = (asv_count <= asv_control_th),
         detected = (asv_count > 0),
         detected_controlled = (detected & !(control_th_below %in% c(TRUE))))


sample_asv_control_seq_class = sample_asv_control %>%
  left_join(bind_rows(asv_seq_12s,
                      asv_seq_16s), by = "asv_id") %>%
  left_join(bind_rows(classified_12s,
                      classified_16s), by = "asv_id") %>%
  left_join(sample_lookup %>%
              select(sample_id,
                     edna_id,
                     collection_time,
                     filter_volume_ml,
                     air_temp_c,
                     wind_speed_m_s,
                     cloud_cover_percent,
                     water_temp_c,
                     p_h,
                     tds_ppm,
                     conductivity_us_cm,
                     stream_side_rz_lab,
                     date,
                     site,
                     region,
                     country), by = "sample_id") %>%
  rename(seq = seq.text)



write_csv(sample_asv_control_seq_class, control_out)
```
