---
title: "Methods modeling & analysis (frequentist)"
author: "Cob Staines"
output: html_document
---

# load data
```{r include=FALSE}
librarian::shelf(tidyverse, here, yaml, lme4, lmerTest, emmeans, MASS, glmmTMB, DHARMa, ggrepel)

run_dir_methods = here("16S_sequence_processing", "runs", "methods_2026-03-13")

model_dir = here("analysis", "methods", "output", "models")

# read in config files
config_12s = read_yaml(here(run_dir_methods, "12S", "output", "metadata", "config.yml"))
config_16s = read_yaml(here(run_dir_methods, "16S", "output", "metadata", "config.yml"))

data_combined = read.csv(here(run_dir_methods, "12S", "output", paste0(config_12s$run$name, "_08_asv_controlled_joined.csv")))
```

# aggregate ASV reads
```{r}
# summary stats for samples (including controls)
sample_read_count = data_combined %>%
  # drop all controls
  filter(is.na(control_type)) %>%
  group_by(sample_id,
           gene,
           filter_size_um_rz,
           passive_active_rz,
           crispr,
           block,
           extraction_type,
           control_type,
           site,
           region,
           country) %>%
  summarize(
    # number of reads
    n_reads = sum(asv_count),
    # number of reads classified
    n_reads_class = sum(if_else(!is.na(taxonomic_rank), asv_count, 0L), na.rm = TRUE),
    # number of reads classified chordata
    n_reads_class_chord = sum(if_else(phylum == "Chordata", asv_count, 0L), na.rm = TRUE),
    # number of reads classified as amphibians
    n_reads_class_chord_amph = sum(if_else(class == "Amphibia", asv_count, 0L), na.rm = TRUE),
    # controlled number of reads classified as amphibians
    n_reads_class_chord_amph_ctrl = sum(if_else(class == "Amphibia", asv_count_controlled, 0L), na.rm = TRUE),
    # number of reads classified as amphibians to the species level
    n_reads_class_chord_amph_spp = sum(if_else((class == "Amphibia") & (taxonomic_rank == "species"), asv_count, 0L), na.rm = TRUE),
    # controlled number of reads classified as amphibians to the species level
    n_reads_class_chord_amph_spp_ctrl = sum(if_else((class == "Amphibia") & (taxonomic_rank == "species"), asv_count_controlled, 0L), na.rm = TRUE),
    # number of reads classified as human
    n_reads_class_chord_sapien = sum(if_else((species == "sapiens"), asv_count, 0L), na.rm = TRUE),
    ## ASVs
    # number of ASVs (before controls)
    n_asv = sum(detected),
    # number of ASVs (after controls)
    n_asv_controlled = sum(detected_controlled, na.rm = TRUE),
    # number of ASVs classified
    n_asv_controlled_class = sum(detected_controlled & !is.na(taxonomic_rank), na.rm = TRUE),
    # number of ASVs classified chordata
    n_asv_controlled_class_chord = sum(detected_controlled & (phylum == "Chordata"), na.rm = TRUE),
    # number of ASVs classified as amphibians
    n_asv_controlled_class_xamph = sum(detected_controlled & (class == "Amphibia"), na.rm = TRUE),
    # number of ASVs classified as amphibians to the species level
    n_asv_controlled_class_xamph_spp = sum(detected_controlled & (class == "Amphibia") & (taxonomic_rank == "species"), na.rm = TRUE),
    .groups = "drop") %>%
  mutate(ratio_amph = n_reads_class_chord_amph_ctrl / n_reads,
         ratio_sapiens = n_reads_class_chord_sapien / n_reads)

# pivot longer for plotting
sample_read_count_long = sample_read_count %>%
  pivot_longer(cols = c(n_reads,
                        n_reads_class,
                        n_reads_class_chord,
                        n_reads_class_chord_amph,
                        n_reads_class_chord_amph_ctrl,
                        n_reads_class_chord_amph_spp,
                        n_reads_class_chord_amph_spp_ctrl,
                        n_reads_class_chord_sapien),
               names_to = "count_type",
               values_to = "n") %>%
  mutate(log10_n = log10(n + 1))

# samples
sample_read_count_summary = sample_read_count %>%
  group_by(gene, crispr, block) %>%
  count()

# gene
ggplot(sample_read_count_long, aes(x = gene, y = log10_n, color = count_type)) +
  geom_boxplot()
```
# examine data

```{r}
ggplot(sample_read_count, aes(x = n_reads)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  labs(title = "All read counts") +
  facet_grid(gene ~ crispr)

# ggplot(sample_read_count %>%
#   filter(n_reads_class_chord_amph_ctrl > 0), aes(x = n_reads_class_chord_amph_ctrl)) +
#   geom_histogram() +
#   scale_x_log10() +
#   labs(title = "Amphibian read counts (controlled)")
# 
# ggplot(sample_read_count %>%
#   filter(n_reads_class_chord_sapien > 0), aes(x = n_reads_class_chord_sapien)) +
#   geom_histogram() +
#   scale_x_log10() +
#   labs(title = "Homo sapiens read counts")

ggplot(sample_read_count, aes(x = n_reads_class_chord_sapien / n_reads)) +
  geom_histogram() +
  labs(title = "Proportion of human reads") +
  facet_grid(gene ~ crispr)

ggplot(sample_read_count, aes(x = n_reads_class_chord_amph_ctrl / n_reads)) +
  geom_histogram() +
  labs(title = "Proportion of amphibian reads") +
  facet_grid(gene ~ crispr)

ggplot(sample_read_count, aes(x = n_asv_controlled_class_xamph)) +
  geom_histogram() +
  labs(title = "Amphibian ASV counts") +
  facet_grid(gene ~ crispr)


```


# read proportion


## 16S block & crispr
### model fitting
```{r}
sample_read_count_bk = sample_read_count %>%
  filter(gene == "16S",
         extraction_type != "PCI",
         passive_active_rz == "Active",
         # !crispr,
         !grepl("^Tissue_", sample_id))
# %>%
#   filter(ratio_sapiens < 0.999)

ggplot(sample_read_count_bk, aes(x = n_reads_class_chord_sapien / n_reads, fill = country)) +
  geom_histogram() +
  labs(title = "Ratio human to total reads") +
  facet_grid(block ~ .)

ggplot(sample_read_count_bk, aes(x = n_reads_class_chord_amph_ctrl / n_reads, fill = country)) +
  geom_histogram() +
  labs(title = "Ratio amphibian to total reads") +
  facet_grid(block ~ .)

# no zero inflation
m0_amph_bk = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ block + (1 | site),
  ziformula  = ~0,          # no zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)

m0_sapiens_bk = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ block * crispr + (1 | site),
  ziformula  = ~0,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)


# xero inflated
m1_amph_bk = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ block * crispr + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)

m1_sapiens_bk = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ block + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)



# You can also model zero-inflation as a function of covariates,
# e.g. sequencing depth driving structural non-detection:
m2_amph_bk = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ block + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)

m2_sapiens_bk = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ block + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_bk
)

m3_sapiens_bk <- glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ block + (1 | site),
  dispformula = ~block,      # phi varies by group
  ziformula   = ~0,
  family      = betabinomial(link = "logit"),
  data        = sample_read_count_bk
)

```

### model comparison
```{r}
# AIC-based model comparison (analogous to LOO-IC comparison in brms)
# Lower AIC = better out-of-sample predictive fit

(aic_comp_amph = AIC(m0_amph_bk,
                     m1_amph_bk,
                     m2_amph_bk))

(aic_comp_sapiens = AIC(m0_sapiens_bk,
                        m1_sapiens_bk,
                        m2_sapiens_bk,
                        m3_sapiens_bk))

```

### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m0_sapiens_bk, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_read_count_bk$block)
plotResiduals(sim_res, form = sample_read_count_bk$n_reads)   # sequencing depth

```
### hypothesis testing
```{r}
em_amph_bk = emmeans(m1_amph_bk, ~ block * crispr)
em_sapiens_bk = emmeans(m0_sapiens_bk, ~ block * crispr)

# response scale
em_amph_bk_resp = emmeans(m1_amph_bk, ~ block * crispr, type = "response")
em_sapiens_bk_resp = emmeans(m0_sapiens_bk, ~ block * crispr, type = "response")
```

#### amphibians
```{r}
# effect fo block within gene group
contrast(em_amph_bk, "pairwise")

# plot contrasts
p_block_amph = as.data.frame(em_amph_bk_resp) %>%
  ggplot(aes(x = block, y = prob, color = crispr,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Block",
       title = "Proportion of 16S amphibian reads vs. block & CRISPR") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_16s_block_crispr_amph.png"),
       p_block_amph,
       width = 6, height = 8)

p_block_amph

```

#### sapiens
```{r}
# effect fo block within gene group
contrast(em_sapiens_bk, "pairwise")

# Or with ggplot for more control
p_block_sapiens = as.data.frame(em_sapiens_bk_resp) %>%
  ggplot(aes(x = block, y = prob, color = crispr,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of human reads",
       x = "Block",
       title = "Proportion of 16S human reads vs. block & CRISPR") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_16s_block_crispr_sapiens.png"),
       p_block_sapiens,
       width = 6, height = 8)

p_block_sapiens

```


## 12S crispr
### model fitting
```{r}
sample_read_count_nb = sample_read_count %>%
  filter(gene == "12S",
         extraction_type != "PCI",
         passive_active_rz == "Active",
         !grepl("^Tissue_", sample_id))
# %>%
  # filter(ratio_sapiens < 0.999)

ggplot(sample_read_count_nb, aes(x = n_reads_class_chord_sapien / n_reads, color = country)) +
  geom_histogram() +
  labs(title = "Ratio human to total reads") +
  facet_wrap("crispr")

ggplot(sample_read_count_nb, aes(x = n_reads_class_chord_amph_ctrl / n_reads, color = country)) +
  geom_histogram() +
  labs(title = "Ratio amphibian to total reads") +
  facet_wrap("crispr")


# no zero inflation
m0_amph = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ crispr + (1 | site),
  ziformula  = ~0,          # no zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)

m0_sapiens = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ crispr + (1 | site),
  ziformula  = ~0,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)

# xero inflated
m1_amph = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ crispr + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)

m1_sapiens = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ crispr + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)



# You can also model zero-inflation as a function of covariates,
# e.g. sequencing depth driving structural non-detection:
m2_amph = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ crispr + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)

m2_sapiens = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ crispr + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_nb
)

m3_sapiens <- glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~
    crispr + (1 | site),
  dispformula = ~ crispr,      # phi varies by group
  ziformula   = ~0,
  family      = betabinomial(link = "logit"),
  data        = sample_read_count_nb
)

```

### model comparison
```{r}
# AIC-based model comparison (analogous to LOO-IC comparison in brms)
# Lower AIC = better out-of-sample predictive fit

(aic_comp_amph = AIC(m0_amph,
                     m1_amph,
                     m2_amph))

(aic_comp_sapiens = AIC(m0_sapiens,
                        m1_sapiens,
                        m2_sapiens,
                        m3_sapiens))
```

### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m0_amph, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_read_count_nb$gene)
plotResiduals(sim_res, form = sample_read_count_nb$crispr)
plotResiduals(sim_res, form = sample_read_count_nb$n_reads)   # sequencing depth

```
### hypothesis testing
```{r}
em_amph = emmeans(m1_amph, ~ crispr)
em_sapiens = emmeans(m0_sapiens, ~ crispr)

# response scale
em_amph_resp = emmeans(m1_amph, ~ crispr, type = "response")
em_sapiens_resp = emmeans(m0_sapiens, ~ crispr, type = "response")
```

#### amphibians
```{r}
# effect fo crispr within gene group
contrast(em_amph, "pairwise")


# plot contrasts
p_gene_crispr_amph = as.data.frame(em_amph_resp) %>%
  ggplot(aes(x = crispr, y = prob, color = crispr,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Gene", color = "CRISPR",
       title = "Proportion of 12S amphibian reads vs. CRISPR") +
  theme_bw()


ggsave(here("analysis", "methods", "output", "figures", "contrast_12s_crispr_amph.png"),
       p_gene_crispr_amph,
       width = 6, height = 8)

p_gene_crispr_amph

```

#### sapiens
```{r}
# effect fo crispr within gene group
contrast(em_sapiens, "pairwise")

# Or with ggplot for more control
p_gene_crispr_sapiens = as.data.frame(em_sapiens_resp) %>%
  ggplot(aes(x = crispr, y = prob, color = crispr,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of human reads",
       x = "Gene", color = "CRISPR",
       title = "Proportion of 12S human reads vs. CRISPR") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_12s_crispr_sapiens.png"),
       p_gene_crispr_sapiens,
       width = 6, height = 8)

p_gene_crispr_sapiens
```

## extraction kit
### model fitting
```{r}
sample_read_count_ex = sample_read_count %>%
  filter(grepl("^Tissue_|^PCI_", sample_id))
# %>%
#   filter(ratio_sapiens < 0.999)

ggplot(sample_read_count_ex, aes(x = n_reads_class_chord_sapien / n_reads)) +
  geom_histogram() +
  labs(title = "Ratio human to total reads") +
  facet_grid(extraction_type ~ passive_active_rz)

ggplot(sample_read_count_ex, aes(x = n_reads_class_chord_amph_ctrl / n_reads)) +
  geom_histogram() +
  labs(title = "Ratio amphibian to total reads") +
  facet_grid(extraction_type ~ passive_active_rz)


# no zero inflation
m0_amph_ex = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~0,          # no zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)

m0_sapiens_ex = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~0,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)


# xero inflated
m1_amph_ex = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)

m1_sapiens_ex = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)



# You can also model zero-inflation as a function of covariates,
# e.g. sequencing depth driving structural non-detection:
m2_amph_ex = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)

m2_sapiens_ex = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ extraction_type * passive_active_rz + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_ex
)

m3_sapiens_ex <- glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ extraction_type * passive_active_rz + (1 | site),
  dispformula = ~extraction_type,      # phi varies by group
  ziformula   = ~0,
  family      = betabinomial(link = "logit"),
  data        = sample_read_count_ex
)

```

### model comparison
```{r}
# AIC-based model comparison (analogous to LOO-IC comparison in brms)
# Lower AIC = better out-of-sample predictive fit

(aic_comp_amph = AIC(m0_amph_ex,
                     m1_amph_ex,
                     m2_amph_ex))

(aic_comp_sapiens = AIC(m0_sapiens_ex,
                        m1_sapiens_ex,
                        m2_sapiens_ex,
                        m3_sapiens_ex))
```
### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m1_amph_ex, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_read_count_ex$extraction_type)
plotResiduals(sim_res, form = sample_read_count_ex$passive_active_rz)
plotResiduals(sim_res, form = sample_read_count_ex$n_reads)   # sequencing depth

```

### hypothesis testing
```{r}
em_amph_ex = emmeans(m1_amph_ex, ~ extraction_type * passive_active_rz)
em_sapiens_ex = emmeans(m0_sapiens_ex, ~ extraction_type * passive_active_rz)

# response scale
em_amph_ex_resp = emmeans(m1_amph_ex, ~ extraction_type * passive_active_rz, type = "response")
em_sapiens_ex_resp = emmeans(m0_sapiens_ex, ~ extraction_type * passive_active_rz, type = "response")
```

#### amphibians
```{r}
# effect of extraction type within filtration group

contrast(em_amph_ex, "pairwise", by = "passive_active_rz")

contrast(em_amph_ex, "pairwise", by = "extraction_type")


# plot contrasts
p_ex_filt_amph = as.data.frame(em_amph_ex_resp) %>%
  ggplot(aes(x = extraction_type, y = prob, color = passive_active_rz,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Extraction type", color = "Filtration type",
       title = "Proportion of 16S amphibian reads vs. Extraction & Filtration") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_16s_ex_filt_amph.png"),
       p_ex_filt_amph,
       width = 6, height = 8)

p_ex_filt_amph
```

#### sapiens
```{r}
# effect of extraction type within filtration group
contrast(em_sapiens_ex, "pairwise", by = "passive_active_rz")

# effect of filtration type within extraction group
contrast(em_sapiens_ex, "pairwise", by = "extraction_type")

# Or with ggplot for more control
p_ex_filt_sapiens = as.data.frame(em_sapiens_ex_resp) %>%
  ggplot(aes(x = extraction_type, y = prob, color = passive_active_rz,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Extraction type", color = "Filtration type",
       title = "Proportion of 16S human reads vs. Extraction & Filtration") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_16s_ex_filt_sapiens.png"),
       p_ex_filt_sapiens,
       width = 6, height = 8)

p_ex_filt_sapiens
```

## gene
## 16S block & crispr
### model fitting
```{r}
sample_read_count_g = sample_read_count %>%
  filter(extraction_type != "PCI",
         passive_active_rz == "Active",
         # !crispr,
         !grepl("^Tissue_", sample_id))
# %>%
#   filter(ratio_sapiens < 0.999)

ggplot(sample_read_count_g, aes(x = n_reads_class_chord_sapien / n_reads, fill = country)) +
  geom_histogram() +
  labs(title = "Ratio human to total reads") +
  facet_grid(gene ~ .)

ggplot(sample_read_count_g, aes(x = n_reads_class_chord_amph_ctrl / n_reads, fill = country)) +
  geom_histogram() +
  labs(title = "Ratio amphibian to total reads") +
  facet_grid(gene ~ .)

# no zero inflation
m0_amph_g = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ gene + (1 | site),
  ziformula  = ~0,          # no zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)

m0_sapiens_g = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ gene + (1 | site),
  ziformula  = ~0,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)


# xero inflated
m1_amph_g = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ gene + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)

m1_sapiens_g = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ gene + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)



# You can also model zero-inflation as a function of covariates,
# e.g. sequencing depth driving structural non-detection:
m2_amph_g = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ gene + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)

m2_sapiens_g = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ gene + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_g
)

m3_sapiens_g <- glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~ gene + (1 | site),
  dispformula = ~gene,      # phi varies by group
  ziformula   = ~0,
  family      = betabinomial(link = "logit"),
  data        = sample_read_count_g
)

```

### model comparison
```{r}
# AIC-based model comparison (analogous to LOO-IC comparison in brms)
# Lower AIC = better out-of-sample predictive fit

(aic_comp_amph = AIC(m0_amph_g,
                     m1_amph_g,
                     m2_amph_g))

(aic_comp_sapiens = AIC(m0_sapiens_g,
                        m1_sapiens_g,
                        m2_sapiens_g,
                        m3_sapiens_g))

```

### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m0_sapiens_g, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_read_count_g$gene)
plotResiduals(sim_res, form = sample_read_count_g$n_reads)   # sequencing depth

```
### hypothesis testing
```{r}
em_amph_g = emmeans(m1_amph_g, ~ gene)
em_sapiens_g = emmeans(m0_sapiens_g, ~ gene)

# response scale
em_amph_g_resp = emmeans(m1_amph_g, ~ gene, type = "response")
em_sapiens_g_resp = emmeans(m0_sapiens_g, ~ gene, type = "response")
```

#### amphibians
```{r}
# effect fo block within gene group
contrast(em_amph_g, "pairwise")

# plot contrasts
p_gene_amph = as.data.frame(em_amph_g_resp) %>%
  ggplot(aes(x = gene, y = prob,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Gene",
       title = "Proportion of amphibian reads vs. gene") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_gene_amph.png"),
       p_gene_amph,
       width = 6, height = 8)

p_gene_amph

```

#### sapiens
```{r}
# effect fo block within gene group
contrast(em_sapiens_g, "pairwise")

# Or with ggplot for more control
p_gene_sapiens = as.data.frame(em_sapiens_g_resp) %>%
  ggplot(aes(x = gene, y = prob,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of human reads",
       x = "Gene",
       title = "Proportion of human reads vs. gene") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_gene_sapiens.png"),
       p_gene_sapiens,
       width = 6, height = 8)

p_gene_sapiens

```



## filter size
### model fitting
```{r}
sample_read_count_fii = sample_read_count %>%
  filter(gene == "16S",
         extraction_type != "PCI",
         !crispr,
         !block,
         !grepl("^Tissue_", sample_id))

sample_read_count_gcf = sample_read_count %>%
  filter(!block,
         extraction_type != "PCI",
         passive_active_rz == "Active",
         !grepl("^Tissue_", sample_id)) %>%
  filter(ratio_sapiens < 0.999)

ggplot(sample_read_count_gcf, aes(x = n_reads_class_chord_sapien / n_reads, color = crispr)) +
  geom_histogram() +
  labs(title = "Ratio human to total reads") +
  facet_grid(gene ~ filter_size_um_rz)

ggplot(sample_read_count_gcf, aes(x = n_reads_class_chord_amph_ctrl / n_reads)) +
  geom_histogram() +
  labs(title = "Ratio amphibian to total reads") +
  facet_grid(gene ~ crispr)


# no zero inflation
m0_amph_gcf = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~ gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~0,          # no zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)

m0_sapiens_gcf = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~  gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~0,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)


# xero inflated
m1_amph_gcf = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~  gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)

m1_sapiens_gcf = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~  gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~1,          # constant zero-inflation probability
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)



# You can also model zero-inflation as a function of covariates,
# e.g. sequencing depth driving structural non-detection:
m2_amph_gcf = glmmTMB(
  cbind(n_reads_class_chord_amph_ctrl, n_reads - n_reads_class_chord_amph_ctrl) ~  gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)

m2_sapiens_gcf = glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~  gene * crispr * filter_size_um_rz + (1 | site),
  ziformula  = ~log(n_reads),     # lower-depth samples more likely structural zero
  family     = betabinomial(link = "logit"),
  data       = sample_read_count_gcf
)

m3_sapiens_gcf <- glmmTMB(
  cbind(n_reads_class_chord_sapien, n_reads - n_reads_class_chord_sapien) ~  gene * crispr * filter_size_um_rz + (1 | site),
  dispformula = ~ filter_size_um_rz,      # phi varies by group
  ziformula   = ~0,
  family      = betabinomial(link = "logit"),
  data        = sample_read_count_gcf
)

```

### model comparison
```{r}
# AIC-based model comparison (analogous to LOO-IC comparison in brms)
# Lower AIC = better out-of-sample predictive fit

(aic_comp_amph = AIC(m0_amph_gcf,
                     m1_amph_gcf,
                     m2_amph_gcf))

(aic_comp_sapiens = AIC(m0_sapiens_gcf,
                        m1_sapiens_gcf,
                        m2_sapiens_gcf,
                        m3_sapiens_gcf))
```

### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m0_sapiens_gcf, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_read_count_gcf$gene)
plotResiduals(sim_res, form = sample_read_count_gcf$crispr)
plotResiduals(sim_res, form = sample_read_count_gcf$filter_size_um_rz)
plotResiduals(sim_res, form = sample_read_count_gcf$n_reads)   # sequencing depth

```

### hypothesis testing
```{r}
em_amph_fs = emmeans(m0_amph_gcf, ~ gene * filter_size_um_rz)
em_sapiens_fs = emmeans(m3_sapiens_fs, ~ gene * filter_size_um_rz)

# response scale
em_amph_fs_resp = emmeans(m0_amph_gcf, ~ gene * filter_size_um_rz, type = "response")
em_sapiens_fs_resp = emmeans(m3_sapiens_fs, ~ gene * filter_size_um_rz, type = "response")
```

#### amphibians
```{r}
# effect fo block within gene group
contrast(em_amph_fs, "pairwise", by = "gene")

# plot contrasts
p_block_amph = as.data.frame(em_amph_bk_resp) %>%
  ggplot(aes(x = block, y = prob, color = block,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of amphibian reads",
       x = "Block",
       title = "Proportion of amphibian reads vs. block") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_block_amph.png"),
       p_block_amph,
       width = 6, height = 8)

```

#### sapiens
```{r}
# effect fo block within gene group
contrast(em_sapiens_bk, "pairwise")

# Or with ggplot for more control
p_block_sapiens = as.data.frame(em_sapiens_bk_resp) %>%
  ggplot(aes(x = block, y = prob, color = block,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Estimated proportion of human reads",
       x = "Block",
       title = "Proportion of human reads vs. block") +
  theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "contrast_block_sapiens.png"),
       p_block_sapiens,
       width = 6, height = 8)

```

# ASV counts

## filtration
### model fitting
```{r}
sample_asv_count_nb = sample_read_count %>%
  mutate(filter_size_um_rz = as.character(filter_size_um_rz),
         gene_filt = interaction(gene, passive_active_rz, drop = TRUE))

peace = sample_asv_count_nb %>%
  filter(gene_filt == "12S.Active")

m_asv_nb = glmmTMB(
  n_asv_controlled_class_xamph ~ filter_size_um_rz * gene_filt + (1 | site),
  family = nbinom2(link = "log"),    # nbinom2 = NB with quadratic variance
  data   = sample_asv_count_nb
)

# Zero-inflated version if you have many zero-species samples
m_asv_zinb = glmmTMB(
  n_asv_controlled_class_xamph ~ filter_size_um_rz * gene_filt + (1 | site),
  ziformula = ~1,
  family = nbinom2(link = "log"),    # nbinom2 = NB with quadratic variance
  data   = sample_asv_count_nb
)

```

### model comparison
```{r}
(aic_asv_amph = AIC(m_asv_nb,
                    m_asv_zinb))
```

### model checks
```{r}
# Simulate residuals -- use more simulations for stable results
sim_res <- simulateResiduals(m_asv_nb, n = 1000, seed = 42)

# Omnibus plot: QQ plot + residuals vs fitted
# Look for: S-shaped QQ = distributional misfit
#           Fanning/trends in right panel = heteroscedasticity
plot(sim_res)

# Test for specific problems
testUniformity(sim_res)     # overall distributional fit
testDispersion(sim_res)     # over/underdispersion relative to model
testZeroInflation(sim_res)  # are there more zeros than model predicts?
testOutliers(sim_res)       # outlier observations

# Residuals against your predictors -- should show no pattern
plotResiduals(sim_res, form = sample_asv_count_nb$gene_filt)
plotResiduals(sim_res, form = sample_asv_count_nb$filter_size_um_rz)

```

### hypothesis testing
```{r}
sample_asv_count_nb %>%
  count(gene_filt, filter_size_um_rz) %>%
  print(n = Inf)

em_amph_asv = emmeans(m_asv_nb, ~ filter_size_um_rz * gene_filt)

# response scale
em_amph_asv_resp = emmeans(m_asv_nb, ~ filter_size_um_rz * gene_filt, type = "response")
```

```{r}
joint_tests(m_asv_nb)
joint_tests(m_asv_nb, by = "gene_filt")   # filter size effect within each group
joint_tests(m_asv_nb, by = "filter_size_um_rz")  # gene_filt effect within each filter
```

#### asvs
```{r}
summary(em_amph_asv_resp)

# Update all-zero upper bound (which fails for Wald CI) with simple Bayesian argument: Gamma(1, n) posterior under Jeffreys prior
n_obs = 16
gamma_upper = qgamma(0.95, shape = 1, rate = n_obs)

plot_data = as.data.frame(em_amph_asv_resp) %>%
  mutate(asymp.UCL = if_else(gene_filt == "12S.Active" & filter_size_um_rz == 5 & asymp.UCL == Inf, gamma_upper, asymp.UCL))

p_gene_filt_pore = plot_data %>%
  ggplot(aes(x = gene_filt, y = response, color = filter_size_um_rz,
             ymin = asymp.LCL, ymax = asymp.UCL)) +
  geom_pointrange(position = position_dodge(width = 0.3)) +
  labs(title = "Filtration Method and Pore Size by Gene",
       x = "Gene & filtration method",
       y = "# of amphibian ASVs per sample (controlled)",
       color = "Filter pore\nsize [um]") +
  theme_bw()

p_gene_filt_pore

ggsave(here("analysis", "methods", "output", "figures", "contrast_gene_filtration_pore.png"),
       p_gene_filt_pore,
       width = 6, height = 8)

```

# Descriptive read proportions
```{r}
peace = data_combined %>%
  filter(domain == "Eukaryota") %>%
  group_by(kingdom, phylum, class) %>%
  count()

sample_taxa_group = data_combined %>%
  mutate(taxa_group = case_when(
    domain == "Bacteria" ~ "Bacteria",
    species == "sapiens" ~ "Homo sapiens",
    class == "Amphibia" ~ "Amphibia",
    class == "Mammalia" ~ "Mammalia Other",
    class == "Aves" ~ "Aves",
    class == "Actinopteri" ~ "Actinopteri",
    class == "Reptilia" ~ "Reptilia",
    #class == "Branchiopoda" ~ "Branchiopoda",
    #class == "Insecta" ~ "Insecta",
    #class == "Lepidosauria" ~ "Lepidosauria",
    #class == "Ostracoda" ~ "Ostracoda",
    kingdom == "Metazoa" ~ "Metazoa Other",
    # kingdom == "Viridiplantae" ~ "Viridiplantae",
    !is.na(domain) ~ "Other",
    .default = "Unassigned"))

sample_taxa_group_agg = sample_taxa_group %>%
  group_by(gene, taxa_group) %>%
  summarize(group_reads = sum(asv_count),
            group_reads_ctrl = sum(asv_count_controlled),
            .groups = "drop") 
# %>%
  # group_by(gene) %>%
  # mutate(group_read_proportion = group_reads / sum(group_reads),
  #        group_read_ctrl_proportion = group_reads_ctrl / sum(group_reads_ctrl)) %>%
  # ungroup()

sample_taxa_group_agg_long = sample_taxa_group_agg %>%
  pivot_longer(cols = c("group_reads",
                        "group_reads_ctrl"),
               names_to = "ctrl",
               values_to = "group_reads") %>%
  mutate(ctrl = recode_values(ctrl,
                              "group_reads" ~ "Uncontrolled",
                              "group_reads_ctrl" ~ "Controlled")) %>%
  group_by(gene, ctrl) %>%
  mutate(group_read_proportion = group_reads / sum(group_reads),
         ctrl = factor(ctrl, levels = c("Uncontrolled", "Controlled")))

# Build a named color vector
taxa_colors <- setNames(
  c(RColorBrewer::brewer.pal(8, "Dark2"), "grey80"),
  c("Actinopteri", "Amphibia", "Aves", "Bacteria",
    "Homo sapiens", "Mammalia Other", "Metazoa Other", "Other",
    "Unassigned")
)

p_taxa_gene_ctrl_prop = ggplot(sample_taxa_group_agg_long,
                       aes(x = ctrl, y = group_reads, fill = taxa_group)) +
  geom_bar(stat = "identity", position = "fill", color = "black") +
  geom_text(
    aes(label = ifelse(group_read_proportion > 0.05, taxa_group, "")),
    stat     = "identity",
    position = position_fill(vjust = 0.5),
    size     = 3,
    color    = "white",
    fontface = "bold"
  ) +
  facet_wrap(~ gene, nrow = 1) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = taxa_colors) +        # <-- replaces scale_fill_brewer
  labs(
    # title = "Taxonomic group proportions before and after control, by gene",
    x     = NULL,
    y     = "Proportion of reads",
    fill  = "Taxonomic group"
  ) +
  theme_bw() +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    axis.text.x     = element_text(size = 10),
    legend.position = "bottom"
  )

p_taxa_gene_ctrl_count = ggplot(sample_taxa_group_agg_long,
       aes(x = ctrl, y = group_reads, fill = taxa_group)) +
  geom_bar(stat = "identity", position = "stack", color = "black") +
  geom_text(
    aes(label = ifelse(group_reads > 120000, taxa_group, "")),
    stat     = "identity",
    position = position_stack(vjust = 0.5),
    size     = 3,
    color    = "white",
    fontface = "bold"
  ) +
  facet_wrap(~ gene, nrow = 1) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = taxa_colors) +
  labs(
    x    = NULL,
    y    = "Reads",
    fill = "Taxonomic group"
  ) +
  theme_bw() +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    axis.text.x     = element_text(size = 10),
    legend.position = "bottom"
  )

# p_taxa_gene_ctrl_count = ggplot(sample_taxa_group_agg_long, aes(x = ctrl, y = group_reads, fill = taxa_group)) +
#   geom_bar(stat = "identity", color = "black") + 
#   facet_wrap("gene") +
#   labs(title = "Taxanomic group proportions by Gene",
#        x = "Gene",
#        y = "Proportion of taxanomic group reads",
#        fill = "Taxanomic group") +
#   theme_bw()

# p_taxa_gene_ctrl = ggplot(sample_taxa_group_agg, aes(x = gene, y = group_reads_ctrl, fill = taxa_group)) +
#   geom_bar(stat = "identity", position = "fill") + 
#   labs(title = "Taxonomic group proportions by Gene",
#        x = "Gene",
#        y = "Proportion of taxanomic group reads",
#        fill = "Taxonomic group") +
#   theme_bw()

ggsave(here("analysis", "methods", "output", "figures", "reads_taxa_gene_ctrl_proportion.png"),
       p_taxa_gene_ctrl_prop,
       width = 8, height = 5)

ggsave(here("analysis", "methods", "output", "figures", "reads_taxa_gene_ctrl_count.png"),
       p_taxa_gene_ctrl_count,
       width = 8, height = 5)

# ggsave(here("analysis", "methods", "output", "figures", "proportion_taxa_gene_controlled.png"),
#        p_taxa_gene_ctrl,
#        width = 6, height = 8)

p_taxa_gene_ctrl_prop
p_taxa_gene_ctrl_count

```

# lolipop
```{r}
p_slope <- ggplot(taxa_long,
                  aes(x = control_status, y = proportion,
                      color = taxa_group, group = taxa_group)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ gene, nrow = 1) +
  scale_y_continuous(labels = scales::percent) +
  # scale_color_brewer(palette = "Paired") +
  labs(
    title = "Shift in taxonomic group proportions after control, by gene",
    x     = NULL,
    y     = "Proportion of reads",
    color = "Taxonomic group"
  ) +
  theme_bw() +
  theme(
    strip.text      = element_text(face = "bold", size = 11),
    axis.text.x     = element_text(size = 10),
    legend.position = "bottom"
  )

# ggsave(
#   here("analysis", "methods", "output", "figures", "taxa_slope.png"),
#   p_slope,
#   width = 10, height = 6
# )

p_slope

```

```{r}
# Reshape to long format with a control_status column
taxa_long <- sample_taxa_group_agg %>%
  dplyr::select(gene, taxa_group,
         Before = group_read_proportion,
         After  = group_read_ctrl_proportion) %>%
  pivot_longer(
    cols      = c(Before, After),
    names_to  = "control_status",
    values_to = "proportion"
  ) %>%
  mutate(
    control_status = factor(control_status, levels = c("Before", "After")),
    taxa_group     = reorder(taxa_group, proportion, FUN = sum)
  )

p_facet_bars <- ggplot(taxa_long,
                       aes(x = control_status, y = proportion, fill = taxa_group)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ gene, nrow = 1) +          # one panel per gene, side by side
  scale_y_continuous(labels = scales::percent) +
  # scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Taxonomic group proportions before and after control, by gene",
    x     = NULL,
    y     = "Proportion of reads",
    fill  = "Taxonomic group"
  ) +
  theme_bw() +
  theme(
    strip.text   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(size = 10),
    legend.position = "bottom"
  )

# ggsave(
#   here("analysis", "methods", "output", "figures", "taxa_facet_bars.png"),
#   p_facet_bars,
#   width = 10, height = 7
# )

p_facet_bars


```

#sankey
```{r}
library(ggalluvial)
library(tidyr)

# --- Reshape data for 3-axis alluvial ---
# Gene → Taxa Group → Control Status (Retained / Filtered)
sankey_data <- sample_taxa_group_agg %>%
  mutate(reads_filtered = group_reads - group_reads_ctrl) %>%
  dplyr::select(gene, taxa_group,
         Retained = group_reads_ctrl,
         Filtered = reads_filtered) %>%
  pivot_longer(
    cols      = c(Retained, Filtered),
    names_to  = "control_status",
    values_to = "reads"
  ) %>%
  filter(reads > 0) %>%
  # Optional: order taxa so the largest groups sit at the top of each stratum
  mutate(taxa_group = reorder(taxa_group, reads, FUN = sum))

# --- Plot ---
p_sankey <- ggplot(
  sankey_data,
  aes(
    # axis1 = gene,
      axis1 = taxa_group,
      axis2 = control_status,
      y     = reads)
) +
  geom_alluvium(
    aes(fill = taxa_group),
    alpha = 0.7,
    width = 1/4,
    knot.pos = 0.4       # controls curve tension between axes
  ) +
  geom_stratum(width = 1/4, fill = "grey90", color = "grey40") +
  geom_text(
    stat  = "stratum",
    aes(label = after_stat(stratum)),
    size  = 3,
    fontface = "bold"
  ) +
  scale_x_discrete(
    limits = c("Gene", "Taxonomic Group", "Control Status"),
    expand = c(0.12, 0.05)
  ) +
  # scale_fill_brewer(palette = "Paired") +   # swap for your preferred palette
  labs(
    title   = "Read flow by Gene, Taxonomic Group, and Control Status",
    y       = "Reads",
    fill    = "Taxonomic Group"
  ) +
  theme_bw() +
  theme(
    axis.text.x  = element_text(size = 11, face = "bold"),
    axis.text.y  = element_text(size = 9),
    panel.grid   = element_blank()
  ) +
  facet_wrap("gene")

# ggsave(
#   here("analysis", "methods", "output", "figures", "sankey_taxa_gene.png"),
#   p_sankey,
#   width = 10, height = 8
# )

p_sankey

```