---
title: "x8_genbank_json_export"
output: html_document
---
# Genbank BLAST submission

Steps:

1. Register with NCBI and create an Entrez API key
  - obtain a key following this guidance: https://support.nlm.nih.gov/kbArticle/?pn=KA-05317
  - save this key to your local .Renviron file as: ENTREZ_KEY = "your-key-here". This will allow you to look up NCBI taxa through taxize without additional steps.
2. Locate the ASV representative sequences .fasta file at: `[run-directory]/analysis/s06_denoised_16S_eDNA/representative sequences/.../data/dna-sequences.fasta`
3. Upload the ASV representative sequences .fasta file to NCBI's Web BLAST: Nucleotide BLAST service
  - Visit https://blast.ncbi.nlm.nih.gov/Blast.cgi, click on Nucleotide BLAST
  - Click `Browse` and navigate to the ASV representatitve sequences .fasta file
  - Under `Algorithm parameters`
    - adjust the desired max number of hits as desired (10 is ideal)
    - adjust the `Expect threshold` (0.03 is ideal)
  - Click `BLAST` and wait for the query to finish
4. In the main results panel, to the right of `RID`, click `Download All` and select `Single-file JSON`. Save the JSON report file to `[run-directory]/outout/`
5. Adjust the parameters in the Config section below to match your needs.

# Config
```{r}
librarian::shelf(tidyverse, jsonlite, here, taxize)

# import json file
json_blast_results = fromJSON(here("16S_sequence_processing", "runs", "penn_2026-03-10", "output", "V3SY4J9U014-Alignment.json"))

output_file = here("16S_sequence_processing", "runs", "penn_2026-03-10", "output", "ncbi_web_blast_hits.csv")
```

# Parse BLAST results
```{r}
# strip hit taxa
search_results = json_blast_results$BlastOutput2$report$results$search

# preallocate hits table
all_hits = tibble(asv_id = c())

cat("Reading hits...\n")
# extract hits from JSON
pb = txtProgressBar(min = 1, max = length(search_results$hits), style = 3)
for (ii in 1:length(search_results$hits)) {
  # print(ii)
  
  ii_hits = search_results$hits[[ii]]$description %>%
    map_dfr(as_tibble, .id = "hit_id") %>%
    mutate(asv_id = search_results$query_title[ii])
  
  ii_identity = search_results$hits[[ii]]$hsps %>%
    map_dfr(as_tibble, .id = "hit_id")
  
  if (nrow(ii_hits) > 0) {
    ii_identity = ii_identity %>%
      rename(seq_overlap = align_len,
             gapopen_count = gaps,
             q_start = query_from,
             q_end = query_to,
             s_start = hit_from,
             s_end = hit_to) %>%
      mutate(percent_identical = 100 * identity / seq_overlap,
             seq_mismatch = seq_overlap - identity,) %>%
      select(hit_id,
             percent_identical,
             seq_overlap,
             seq_mismatch,
             gapopen_count,
             q_start,
             q_end,
             s_start,
             s_end) %>%
      group_by(hit_id) %>%
      slice_max(percent_identical, with_ties = FALSE) %>%
      ungroup()
    
    ii_out = ii_hits %>%
      left_join(ii_identity, by = "hit_id")
  } else {
    ii_out = tibble()
  }
  
  all_hits = bind_rows(all_hits, ii_out)
  setTxtProgressBar(pb, ii)
}
close(pb)

all_hits = all_hits %>%
  select(asv_id,
         everything())
```


# query NCBI taxonomy
```{r}
# pull taxa IDs
all_taxids = unique(all_hits$taxid)

# retreive taxa classifications
cat("Querying NCBI taxa...\n")
ncbi_taxa = classification(all_taxids, db = "ncbi")

# extract taxa
all_taxa = tibble(taxid = c())

cat("Extracting taxa...\n")
pb = txtProgressBar(min = 1, max = length(all_taxids), style = 3)
for (ii in 1:length(all_taxids)) {
  # print(ii)
  ii_taxa = ncbi_taxa[[ii]] %>%
    select(-id) %>%
    filter(rank %in% c("domain",
                       "kingdom",
                       "phylum",
                       "class",
                       "order",
                       "family",
                       "genus",
                       "species",
                       "subspecies")) %>%
    pivot_wider(names_from = rank,
                values_from = name) %>%
    mutate(taxid = all_taxids[ii])
  
  all_taxa = bind_rows(all_taxa, ii_taxa)
  setTxtProgressBar(pb, ii)
}
close(pb)
```

# Merge and export
```{r}
all_hits_out = all_hits %>%
  left_join(all_taxa, by = "taxid") %>%
  mutate(method = "blast_web") %>%
  select(asv_id,
         hit_id,
         method,
         accession,
         taxid,
         title,
         percent_identical,
         domain,
         kingdom,
         phylum,
         class,
         order,
         family,
         genus,
         species,
         subspecies)

write_csv(all_hits_out, output_file)

```