---
title: "build_local_classifier"
author: "Cob Staines"
output: html_document
---

# Description

This is a utility script which converts the ASV representative sequences and accepted taxonomic classifications from a completed sequence processing + analysis run into the same file/directory structure used by the local reference classifiers (e.g. `amphibia16s_classifier`, `vertebrata16s_classifier`, `vertebrata12s_classifier`), so the run's ASVs can be referenced as a local BLAST/vsearch classifier in future runs (`taxonomy.classifierDir` in `config.yml`).

All ASVs present in the run's representative sequences are included, whether or not they received an accepted classification in `06_classify_asv.Rmd` (unclassified ASVs are carried through as `Unassigned`, matching QIIME2 convention).

*Note* - this script builds real QIIME2 `.qza` artifacts (`FeatureData[Sequence]`, `FeatureData[Taxonomy]`), extracted the same way `fetch-ncbi-seq-classifier-*.sh` does. It therefore requires the `qiime` CLI to be on `PATH` -- run this from inside the pipeline's Podman/Docker container (see `16S_sequence_processing/README.md`), e.g.:

```bash
podman run -it --rm -v $(pwd):/data:z edna-pipeline bash
```

then work through the chunks below from an R session started in that shell (or `Rscript -e "rmarkdown::render('utilities/build_local_classifier.Rmd')"`, noting `rmarkdown` is not installed in the container image by default).

# Config
```{r}
library(tidyverse)
library(yaml)
library(here)
library(phylotools)

# run to build the classifier from
run_dir = here("16S_sequence_processing", "runs", "lab_controls_2026-09-03")

# read in run config file (for gene & run name)
config = read_yaml(here(run_dir, "output", "metadata", "config.yml"))

# name of the new classifier directory, created as a subfolder of 16S_sequence_processing/
# (this is the value you would later set as `taxonomy.classifierDir` in config.yml)
classifier_name = "lab_contaminants_classifier"

# taxon string assigned to ASVs with no accepted classification from 06_classify_asv.Rmd
unassigned_taxon = "Unassigned"
```

# Function: build_local_classifier

Converts a run's representative ASV sequences + accepted taxonomy into a local classifier directory matching the structure of e.g. `vertebrata16s_classifier`:

- `<classifier_dir>/derep1_seqs_extracted.qza`
- `<classifier_dir>/derep1_seqs_extracted/.../data/dna-sequences.fasta`
- `<classifier_dir>/derep1_taxa_extracted.qza`
- `<classifier_dir>/derep1_taxa_extracted/.../data/taxonomy.tsv`
- `<classifier_dir>/classifier_metadata.yml`

```{r}
build_local_classifier = function(run_dir, config, classifier_name, unassigned_taxon = "Unassigned") {

  if (Sys.which("qiime") == "") {
    stop("qiime CLI not found on PATH. Run this script inside the pipeline's QIIME2 container -- see 16S_sequence_processing/README.md for setup.")
  }

  classifier_dir = here("16S_sequence_processing", classifier_name)

  if (dir.exists(classifier_dir)) {
    stop(paste0("Classifier directory already exists: ", classifier_dir, "\n\tRemove or rename it before rebuilding."))
  }

  dir.create(classifier_dir, recursive = TRUE)
  temp_dir = file.path(classifier_dir, "temp")
  dir.create(temp_dir)

  # representative sequences (all ASVs from denoising -- unfiltered)
  sequence_table =
    read.fasta(file = here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "representative_sequences",
                           list.files(here(run_dir, "analysis", paste0("s06_denoised_", config$taxonomy$gene, "_eDNA"), "representative_sequences")),
                           "data", "dna-sequences.fasta")) %>%
    rename(asv_id = seq.name,
           sequence = seq.text) %>%
    arrange(asv_id)

  # accepted classification per ASV, if present (06_classify_asv.Rmd output)
  classification_file = here(run_dir, "output", paste0(config$run$name, "_06_asv_classification.csv"))

  if (file.exists(classification_file)) {
    asv_classification = read_csv(classification_file, show_col_types = FALSE) %>%
      select(asv_id, domain, kingdom, phylum, class, order, family, genus, species)
  } else {
    cat("No accepted classification table found, all ASVs will be written as", unassigned_taxon, "\n")
    asv_classification = tibble(asv_id = character())
  }

  # build taxon strings; every ASV in sequence_table is retained, classified or not
  taxonomy_table = sequence_table %>%
    select(asv_id) %>%
    left_join(asv_classification, by = "asv_id") %>%
    mutate(taxon = if_else(
      is.na(kingdom),
      unassigned_taxon,
      paste0("k__", replace_na(kingdom, ""),
             ";p__", replace_na(phylum, ""),
             ";c__", replace_na(class, ""),
             ";o__", replace_na(order, ""),
             ";f__", replace_na(family, ""),
             ";g__", replace_na(genus, ""),
             ";s__", replace_na(species, ""))
    )) %>%
    select(`Feature ID` = asv_id, Taxon = taxon)

  # write intermediate fasta + taxonomy tsv
  fasta_path = file.path(temp_dir, "sequences.fasta")
  taxonomy_path = file.path(temp_dir, "taxonomy.tsv")

  writeLines(paste0(">", sequence_table$asv_id, "\n", sequence_table$sequence), fasta_path)
  write_tsv(taxonomy_table, taxonomy_path)

  # import as QIIME2 artifacts
  seqs_qza = file.path(classifier_dir, "derep1_seqs_extracted.qza")
  taxa_qza = file.path(classifier_dir, "derep1_taxa_extracted.qza")

  cat("Importing sequences as FeatureData[Sequence]\n")
  system2("qiime", c("tools", "import",
                     "--type", "FeatureData[Sequence]",
                     "--input-path", fasta_path,
                     "--output-path", seqs_qza))

  cat("Importing taxonomy as FeatureData[Taxonomy]\n")
  system2("qiime", c("tools", "import",
                     "--type", "FeatureData[Taxonomy]",
                     "--input-path", taxonomy_path,
                     "--output-path", taxa_qza))

  # extract, matching the layout referenced by qiime-seq-process_*.sh
  cat("Extracting artifacts\n")
  system2("qiime", c("tools", "extract", "--input-path", seqs_qza, "--output-path", file.path(classifier_dir, "derep1_seqs_extracted")))
  system2("qiime", c("tools", "extract", "--input-path", taxa_qza, "--output-path", file.path(classifier_dir, "derep1_taxa_extracted")))

  unlink(temp_dir, recursive = TRUE)

  # write classifier metadata
  metadata = list(
    classifier = list(
      name = classifier_name,
      buildScript = "utilities/build_local_classifier.Rmd",
      sourceRunName = config$run$name,
      sourceRunDir = config$run$runDir,
      buildEndTimestamp = format(Sys.time(), '%Y-%m-%d %H:%M:%S %z')
    )
  )
  write_yaml(metadata, file.path(classifier_dir, "classifier_metadata.yml"))

  cat("Done. Classifier written to", classifier_dir, "\n")
  invisible(classifier_dir)
}
```

# Run
```{r}
build_local_classifier(run_dir, config, classifier_name, unassigned_taxon)
```
