---
title: "copy_seqs_by_system"
author: "Cob Staines"
output: html_document
---

# Copy Sequence Files by System

This is a utility script which pulls sequencing files from the AWS S3 eDNA server, identifies the sequence_id, and references a lookup table to determine which system they belong to, and copies the desired system(s) to a specified folder. Contrast with `copy_seqs_by_system_local`, which sources sequence files from a locally accessible directory.

*Note* - To access the AWS S3 eDNA server, you must have variables `aws_edna_access_key` and `aws_edna_secret_access_key` saved in your .Renviron file (`usethis::edit_r_environ()`). If you need these, contact your lab PI or the RIBBiTR data manager.

# Configuration

Fill out the parameters below to point to the corresponding local directories and files.
```{r}
# load required packages
librarian::shelf(tidyverse, RPostgres, DBI, RIBBiTR-BII/ribbitrrr, here, janitor, aws.s3)

# directory to copy files to
destination_dir = here("16S_sequence_processing", "runs", "lab_negatives_2026-09-03", "sequences")

# system(s) of interest, any of: brazil, panama, pennsylvania, sierra_nevada, lab_controls (case sensitive)
soi = c("lab_controls")

# Boolean: if true, drops duplicate sequences with later sequencing dates
most_recent_seqs_only = TRUE

# hard coded extractions to include for each study system
include_extractions_list = list(
  brazil = c("RP_330",
             "RP_343",
             "RP_345",
             "RP_346"),
  pennsylvania = c("Tissue_63"),
  panama = c(),
  sierra_nevada = c(),
  lab_negatives = c()
)

```

# database connection
```{r}
# connection
dbcon = hopToDB("ribbitr")

# pointers=
db_edna_samp = tbl(dbcon, Id("survey_data", "edna_sample"))
db_edna_ext = tbl(dbcon, Id("survey_data", "edna_extraction"))
db_edna_seq = tbl(dbcon, Id("survey_data", "edna_sequencing"))

db_survey = tbl(dbcon, Id("survey_data", "survey"))
db_visit = tbl(dbcon, Id("survey_data", "visit"))
db_site = tbl(dbcon, Id("survey_data", "site"))
db_region = tbl(dbcon, Id("survey_data", "region"))
db_country = tbl(dbcon, Id("survey_data", "country"))

```

# metadata pull
```{r}
# check soi
if (any(!(soi %in% c("brazil", "panama", "pennsylvania", "sierra_nevada", "lab_controls")))) {
  cat(paste0("Unrecognized study systems in config variable `soi`\n\tPermitted values: brazil, panama, pennsylvania, sierra_nevada, lab_controls\n\tUnrecognized value(s): ", paste(soi[!(soi %in% c("brazil", "panama", "pennsylvania", "sierra_nevada", "lab_controls"))], collapse = ", ")))
}

# build study system filter expression
filter_criteria = list(
  brazil = quo(country == "brazil"),
  panama = quo(country == "panama"),
  pennsylvania = quo(region == "pennsylvania"),
  sierra_nevada = quo(region == "california"),
  lab_controls = quo(FALSE)
)

selected_criteria = filter_criteria[soi]
filter_expression = reduce(selected_criteria, ~ expr(!!.x | !!.y))

surveys_of_interest = db_survey %>%
  left_join(db_visit, by = "visit_id") %>%
  left_join(db_site, by = "site_id") %>%
  left_join(db_region, by = "region_id") %>%
  left_join(db_country, by = "country_id") %>%
  filter(detection_type == "edna",
         !!filter_expression)

sids = surveys_of_interest %>%
  pull(survey_id)

# --- samples ---
samples_nc_groups = db_edna_samp %>%
  filter(survey_id %in% sids,
         !is.na(field_negative_control_group)) %>%
  pull(field_negative_control_group)

samples_nc_ids = db_edna_samp %>%
  filter(field_negative_control,
         field_negative_control_group %in% samples_nc_groups) %>%
  pull(edna_sample_id)

samples_ids = db_edna_samp %>%
  filter(survey_id %in% sids | edna_sample_id %in% samples_nc_ids) %>%
  pull(edna_sample_id)

# --- extractions ---
extractions_nc_groups = db_edna_ext %>%
  filter(edna_sample_id %in% samples_ids,
         !is.na(extraction_negative_control_group)) %>%
  pull(extraction_negative_control_group)

extractions_nc_ids = db_edna_ext %>%
  filter(extraction_negative_control,
         extraction_negative_control_group %in% extractions_nc_groups) %>%
  pull(edna_extraction_id)

extractions_ids = db_edna_ext %>%
  filter(edna_sample_id %in% samples_ids | edna_extraction_id %in% extractions_nc_ids) %>%
  pull(edna_extraction_id)

# --- sequences ---
sequences_c_groups = db_edna_seq %>%
  filter(edna_extraction_id %in% extractions_ids,
         !is.na(pcr_control_group)) %>%
  pull(pcr_control_group)

sequences_c_ids = db_edna_seq %>%
  filter(pcr_control,
         pcr_control_group %in% sequences_c_groups) %>%
  pull(edna_sequencing_id)

sequences_ids = db_edna_seq %>%
  filter(edna_extraction_id %in% extractions_ids | edna_sequencing_id %in% sequences_c_ids) %>%
  pull(edna_sequencing_id)

include_extractions = unlist(include_extractions_list[soi])

if ("lab_controls" %in% soi) {
  lab_controls_query = quo(extraction_negative_control | pcr_control)
} else {
  lab_controls_query = quo(FALSE)
}

all_seqs_data = db_edna_seq %>%
  left_join(db_edna_ext, by = "edna_extraction_id") %>%
  left_join(db_edna_samp, by = "edna_sample_id") %>%
  left_join(surveys_of_interest, by = "survey_id") %>%
  filter(edna_sequencing_id %in% sequences_ids |
           edna_extraction_name %in% include_extractions |
           lab_controls_query) %>%
  select(all_of(colnames(db_edna_seq)),
         all_of(colnames(db_edna_ext)),
         all_of(colnames(db_edna_samp)),
         time_of_day,
         date,
         transect,
         site,
         region,
         country) %>%
  collect() %>%
  group_by(sequencing_s3_filepath) %>%
  slice_head(n = 1) %>%
  ungroup()

if (most_recent_seqs_only) {
  all_seqs_data = all_seqs_data %>%
    arrange(edna_sequencing_name, read_direction, sequencing_date) %>%
    group_by(edna_sequencing_name, read_direction) %>%
    slice_tail(n = 1) %>%
    ungroup()
}

anti_seq = db_edna_seq %>%
  left_join(db_edna_ext, by = "edna_extraction_id") %>%
  left_join(db_edna_samp, by = "edna_sample_id") %>%
  left_join(surveys_of_interest, by = "survey_id") %>%
  select(all_of(colnames(db_edna_seq)),
         all_of(colnames(db_edna_ext)),
         all_of(colnames(db_edna_samp)),
         time_of_day,
         date,
         transect,
         site,
         region,
         country) %>%
  filter(!(edna_sequencing_id %in% all_seqs_data$edna_sequencing_id)) %>%
  collect()

check_sequence_read_run_dupes = get_dupes(all_seqs_data, edna_sequencing_name, read_direction)

```

# Filter to system of interest
```{r}
all_seqs_data

# inventory s3
inventory = get_bucket_df(bucket = "ribbitr-edna-sequences",
                          prefix = "edna_sequences",
                          max = Inf,
                          key = Sys.getenv("aws_edna_access_key"),
                          secret = Sys.getenv("aws_edna_secret_access_key"),
                          region = "us-west-1") %>%
  clean_names() %>%
  filter(key %in% all_seqs_data$sequencing_s3_filepath)

all_seqs_inventory = all_seqs_data %>%
  left_join(inventory, by = c("sequencing_s3_filepath" = "key"), relationship = "one-to-one", unmatched = "error")

total_filesize_gb = sum(as.integer(all_seqs_inventory$size), na.rm = TRUE) / 1073741824

cat(paste0("---Pre-Copying Summary---\n\t",
           "File count: ", nrow(all_seqs_inventory), "\n\t",
           "Sequencing ids: ", length(unique(all_seqs_inventory$edna_sequencing_name)), "\n\t",
           "Study system(s): ", paste0(soi, collapse = ", "), "\n\t",
           "Total file size: ", total_filesize_gb, " Gb\n\t",
           "From AWS S3 Bucket: ribbitr-edna-sequences\n\t",
           "To directory: ", destination_dir, "\n\n"))

copy_input = readline(prompt = "Proceed with copy? [y/n]: ");

if (tolower(copy_input == "n")) {
  stop("User Aborted.\n")
} else if (tolower(copy_input != "y")) {
  stop("Unexpected input. Copy aborted.\n")
}
```

# Copy files to copy directory
```{r}

# create destination dir if it does not exist
if (!dir.exists(destination_dir)) {
  dir.create(destination_dir, recursive = TRUE)
}

# delete all existing files in directory first
if (length(list.files(destination_dir)) > 0) {
  drop_input = readline(prompt = "Files exist in the copy directory. Do you want to delete these files before copying? [y/n]"); 
  
  if (tolower(drop_input == "y")) {
    cat("Deleting existing files in destination directory.\n")
    unlink(here(destination_dir, "*"))
  } else if (tolower(drop_input == "n")) {
    cat("Leaving existing files in destination directory. Proceeding to copy new files.\n")
  } else {
    stop("Input not recognized. Copy aborted.")
  }
}

cat(paste0("Copying ", as.character(nrow(all_seqs_inventory)), " sequence files to destination directory from s3 server... "))

pb = txtProgressBar(min = 1, max = nrow(all_seqs_inventory), style = 3)

for (ii in 1:length(all_seqs_inventory$sequencing_s3_filepath)) {
  kk = all_seqs_inventory$sequencing_s3_filepath[ii]
  save_object(
    object = kk,
    bucket = "ribbitr-edna-sequences",
    key = Sys.getenv("aws_edna_access_key"),
    secret = Sys.getenv("aws_edna_secret_access_key"),
    region = "us-west-1",
    file = file.path(destination_dir, basename(kk))
  )
  setTxtProgressBar(pb, ii)
}

cat("\ndone.")
```

# copied files check
```{r}
dest_dir_files_raw = tibble(absolute_filepath = c(list.files(destination_dir, full.names = FALSE, recursive = TRUE))) %>%
  mutate(filename = basename(absolute_filepath))

check_missing_files = all_seqs_inventory %>%
  mutate(filename = basename(sequencing_s3_filepath)) %>%
  anti_join(dest_dir_files_raw, by = "filename")

if (nrow(check_missing_files) > 0) {
  cat(paste0(nrow(check_missing_files), " expected files missing from destination directory. See variable `check_missing_files` for details.\n"))
}

# write_csv(missing_files, here(destination_dir, "missing_copied_files.csv"))
# 
# # attempt to recopy
# source_dir_filtered_files = missing_files

```