---
title: "copy_seqs_by_system"
author: "Cob Staines"
output: html_document
---

# Copy Sequence Files by System

This is a utility script which pulls sequencing files from the AWS S3 eDNA server, identifies the sequence_id, and references a lookup table to determine which system they belong to, and copies the desired system(s) to a specified folder. Contrast with `copy_seqs_by_system_local`, which sources sequence files from a locally accessible directory.

*Note* - To access the AWS S3 eDNA server, you must have variables `aws_edna_access_key` and `aws_edna_secret_access_key` saved in your .Renviron file. If you need these, contact your lab PI or the RIBBiTR data manager.

# Configuration

Fill out the parameters below to point to the corresponding local directories and files.
```{r}
# load required packages
librarian::shelf(tidyverse, fuzzyjoin, here, janitor, rio, readr, aws.s3)

# Point to RIBBiTR eDNA Processing 22-25 lookup table for extraction_ids & study systems
sample_lookup_file = here(Sys.getenv("data_dir"), "edna_results", "RIBBiTR eDNA Processing 22-25_2026-08-13.xlsx")

# directory to copy files to
destination_dir = here("16S_sequence_processing", "runs", "brazil_copy_test", "sequences")

# system(s) of interest: any of: Brazil, Panama, Pennsylvania, Sierra Nevada
soi = c("Brazil")

```

# import lookup table
```{r}
lookup_samples = import(sample_lookup_file, which = "SEQUENCING", na.strings = c("NA", ""), guess_max = 2000) %>%
  clean_names() %>%
  filter(!is.na(sequencing_id),
         sequencing_date != "Not sent")
```

# Parse search directory
```{r}
inventory = get_bucket_df(bucket = "ribbitr-edna-sequences",
                          prefix = "edna_sequences",
                          max = Inf,
                          key = Sys.getenv("aws_edna_access_key"),
                          secret = Sys.getenv("aws_edna_secret_access_key"),
                          region = "us-west-1")

inventory_parsed = inventory %>%
  clean_names() %>%
  filter(str_detect(key, "\\.fastq(\\.gz)?$")) %>%
  mutate(filename = basename(key),
         filebase = gsub(".fastq", "", gsub(".gz", "", filename)),
         parent_dir = dirname(key),
         processing_id = str_match(filebase, "(_S[0-9]+_L[0-9]+_R[12]_[0-9]+)$")[,2],
         sequencing_base = gsub("-", "_", str_remove(filebase, paste0(processing_id, "$"))),
         read_dir = case_when(
           grepl("_R1_", processing_id) ~ "forward",
           grepl("_R2_", processing_id) ~ "reverse",
           .default = NA)) %>%
  select(key,
         e_tag,
         parent_dir,
         filename,
         filebase,
         sequencing_base,
         read_dir,
         size)

inventory_samples = inventory_parsed %>%
  pivot_wider(id_cols = c("parent_dir", "sequencing_base"),
              names_from = read_dir,
              values_from = key)

# check for sequencing_base duplicates
check_etag_dupes = get_dupes(inventory_parsed, e_tag)

if (nrow(check_etag_dupes) > 0) {
  stop(paste0("Duplicate files found:\n\t", paste0(unique(check_etag_dupes$sequencing_base), collapse = "\n\t")))
}

# check for sequencing_base duplicates
check_sample_dupes = get_dupes(inventory_samples, sequencing_base)

if (nrow(check_sample_dupes) > 0) {
  stop(paste0("Duplicate sequence files found. See variable `check_sample_dupes` to identify the prefered `relative_parent_dir` and add it to `priority_dir` (in configuration chunk). Offending samples:\n\t", paste0(unique(check_sample_dupes$sequencing_base), collapse = "\n\t")))
}

# check for any without a forward or reverse read
check_sample_missing_direction = inventory_samples %>%
  filter(is.na(forward) |is.na(reverse))

if (nrow(check_sample_missing_direction) > 0) {
  stop(paste0("Samples found with unpaired sequencing reads (only forward or only reverse), see variable `check_sample_missing_direction` to investigate. Offening samples:\n\t", paste0(check_sample_missing_dir$relative_filepath, collapse = "\n\t")))
}
```

# align with lookup table
```{r}
match_fun = function(a, b) {
  pattern = paste0("(?<![0-9A-Za-z])", stringr::str_escape(b), "$")
  stringr::str_detect(a, pattern)
}

lookup_file_matched = fuzzy_full_join(
  inventory_samples, lookup_samples,
  by = c("sequencing_base" = "sequencing_id"),
  match_fun = match_fun)

# checks
## one sequence file to many samples (generally not good)
check_m2o = lookup_file_matched %>%
  count(sequencing_base) %>%
  filter(n > 1,
         !is.na(sequencing_base))

if (nrow(check_m2o) > 0) {
  stop(paste0("Multiple samples associated with a given sequencing file. See variable `check_m2o` and resolve before proceeding. Offending sequences:\n\t", paste0(check_m20$sequencing, collapse = "\n\t")))
}

## one sample to many sequence files (generally ok)
check_o2m = lookup_file_matched %>%
  count(sequencing_id) %>%
  filter(n > 1,
         !is.na(sequencing_id))

## no sequence
check_no_seq = lookup_file_matched %>%
  filter(is.na(sequencing_base))

## no extraction_id
check_no_ext = lookup_file_matched %>%
  filter(is.na(sequencing_id))

check_misaligned = lookup_file_matched %>%
  filter(is.na(sequencing_base) | is.na(sequencing_id)) %>%
  arrange(sequencing_base,
         sequencing_id) %>%
  select(sequencing_base,
         sequencing_id,
         parent_dir,
         sequencing_date,
         everything())

if (nrow(check_no_seq) > 0) {
  cat(paste0(nrow(check_no_seq), " samples with no associated sequencing files. See variable `check_misaligned` for details.\n"))
}

if (nrow(check_no_ext) > 0) {
  cat(paste0(nrow(check_no_ext), " sequencing file pairs with no associated sample metadata. See variable `check_misaligned` for details.\n"))
}
```

# Filter to system of interest
```{r}
# files of interest
lookup_file_aligned = lookup_file_matched %>%
  filter(!is.na(sequencing_base),
         !is.na(sequencing_id))

lookup_file_filtered = lookup_file_aligned %>%
  filter(study_system %in% soi)

lookup_file_filtered_files = lookup_file_filtered %>%
  pivot_longer(cols = c("forward", "reverse"),
               names_to = "read_dir",
               values_to = "key") %>%
  left_join(inventory_parsed %>%
              select(key,
                     size), by = "key")

total_filesize_gb = sum(as.integer(lookup_file_filtered_files$size), na.rm = TRUE) / 1073741824

cat(paste0("---Pre-Copying Summary---\n\t",
           "File count: ", nrow(lookup_file_filtered_files), "\n\t",
           "Sample count: ", nrow(lookup_file_filtered), "\n\t",
           "Study system(s): ", paste0(soi, collapse = ", "), "\n\t",
           "Total file size: ", total_filesize_gb, " Gb\n\t",
           "From AWS S3 Bucket: ribbitr-edna-sequences\n\t",
           "To directory: ", destination_dir, "\n\n"))

copy_input = readline(prompt = "Proceed with copy? [y/n]: ");

if (tolower(copy_input == "n")) {
  stop("User Aborted.\n")
} else if (tolower(copy_input != "y")) {
  stop("Unexpected input. Copy aborted.\n")
}
```

# Copy files to copy directory
```{r}

# create destination dir if it does not exist
if (!dir.exists(destination_dir)) {
  dir.create(destination_dir, recursive = TRUE)
}

# delete all existing files in directory first
if (length(list.files(destination_dir)) > 0) {
  drop_input = readline(prompt = "Files exist in the copy directory. Do you want to delete these files before copying? [y/n]"); 
  
  if (tolower(drop_input == "y")) {
    cat("Deleting existing files in destination directory.\n")
    unlink(here(destination_dir, "*"))
  } else if (tolower(drop_input == "n")) {
    cat("Leaving existing files in destination directory. Proceeding to copy new files.\n")
  } else {
    stop("Input not recognized. Copy aborted.")
  }
}

cat(paste0("Copying ", as.character(nrow(lookup_file_filtered_files)), " sequence files to destination directory... "))

pb = txtProgressBar(min = 1, max = nrow(lookup_file_filtered_files), style = 3)

for (ii in 1:length(lookup_file_filtered_files$key)) {
  kk = lookup_file_filtered_files$key[ii]
  save_object(
    object = kk,
    bucket = "ribbitr-edna-sequences",
    key = Sys.getenv("aws_edna_access_key"),
    secret = Sys.getenv("aws_edna_secret_access_key"),
    region = "us-west-1",
    file = file.path(destination_dir, basename(kk))
  )
  setTxtProgressBar(pb, ii)
}

cat("done.")
```

# copied files check
```{r}
dest_dir_files_raw = tibble(absolute_filepath = c(list.files(destination_dir, full.names = FALSE, recursive = TRUE))) %>%
  mutate(filename = basename(absolute_filepath))

missing_files = lookup_file_filtered_files %>%
  mutate(filename = basename(key)) %>%
  anti_join(dest_dir_files_raw, by = "filename")

if (nrow(check_missing_files) > 0) {
  cat(paste0(nrow(missing_files), " expected files missing from destination directory. See variable `check_missing_files` for details.\n"))
}

# write_csv(missing_files, here(destination_dir, "missing_copied_files.csv"))
# 
# # attempt to recopy
# source_dir_filtered_files = missing_files

```