---
title: "copy_seqs_by_system"
author: "Cob Staines"
output: html_document
---

# Copy Sequence Files by System

This is a utility script which parses through nested folders for .fastq sequencing files, identifies the sequence_id, and references a lookup table to determine which system they belong to, and copies the desired system(s) to a specified folder.

# Configuration

Fill out the parameters below to point to the corresponding local directories and files.
```{r}
# load required packages
librarian::shelf(tidyverse, fuzzyjoin, here, janitor, rio)

# directory to search recursively
source_dir = here(Sys.getenv("data_dir"), "edna_results")

# directories (or start of directories) to exclude from search and copy
excluded_dirs = c(
  here(source_dir, "July_2024_poorquality"),  # re-sequenced in August 2024/sierra_2022
  here(source_dir, "December 2023/All_Sequences/MINI824"),  # re-sequenced in Mini816
  here(source_dir, "December 2023/All_Sequences/MINI825"),  # re-sequenced in Mini816
  here(source_dir, "April 2024", "Sequencing_April2024", "Full_Run", "Miseq_835_Brandon_47942-417083877"),  # duplicated in April 2024
  here(source_dir, "July_2_2026/Project_Fluidigm_PITT-509339831/BaseSpace_CLI_07_02_2026_1_53_11-937308385")  # single duplicated sample
)
# 
# excluded_dirs = c(here(source_dir, "July_2024_poorquality"),
#                   here(source_dir, "December 2023", "All_Sequences", "Mini816"),
#                   here(source_dir, "December 2023", "All_Sequences", "MINI824"),
#                   here(source_dir, "December 2023", "All_Sequences", "Miseq_816"))

# directory to copy files to
destination_dir = here("16S_sequence_processing", "runs", "brazil_2026-08-12", "sequences")

# Point to RIBBiTR eDNA Processing 22-25 lookup table for extraction_ids & study systems
sample_lookup_file = here(source_dir, "RIBBiTR eDNA Processing 22-25_2026-08-12.xlsx")

# system(s) of interest: any of: Brazil, Panama, Pennsylvania, Sierra Nevada
soi = c("Brazil")

```

# import lookup table
```{r}
lookup_samples = import(sample_lookup_file, which = "SEQUENCING") %>%
  clean_names() %>%
  filter(!is.na(sequencing_id),
         sequencing_date != "Not sent")
```

# Parse search directory
```{r}
# search sequencing files exhaustively, excluding those in excluded dirs, and extract identifiers from file names
source_dir_files = tibble(relative_filepath = c(list.files(source_dir, recursive = TRUE))) %>%
  filter(str_detect(relative_filepath, "\\.fastq(\\.gz)?$")) %>%
  mutate(absolute_filepath = here(source_dir, relative_filepath),
         relative_parent_dir = dirname(relative_filepath),
         filename = basename(relative_filepath),
         filebase = gsub(".fastq", "", gsub(".gz", "", filename)),
         batch_id = str_match(filebase, "^([A-Za-z0-9]+-)")[,2],
         processing_id = str_match(filebase, "(_S[0-9]+_L[0-9]+_R[12]_[0-9]+)$")[,2],
         sequencing_base = gsub("-", "_", str_remove(filebase, paste0(processing_id, "$"))),
         read_dir = case_when(
           grepl("_R1_", processing_id) ~ "forward",
           grepl("_R2_", processing_id) ~ "reverse",
           .default = NA)) %>%
  filter(!purrr::map_lgl(absolute_filepath, function(f) {
    any(str_starts(f, fixed(excluded_dirs)))
  }))

# pivot wider to have 1 row per sample, with columns for forward and reverse reads
source_dir_samples = source_dir_files %>%
  pivot_wider(id_cols = c("relative_parent_dir", "sequencing_base"),
              names_from = read_dir,
              values_from = absolute_filepath)

# check for sequencing_base duplicates
check_sample_dupes = get_dupes(source_dir_samples, sequencing_base)

# check for any without a forward or reverse read
check_sample_missing_dir = source_dir_samples %>%
  filter(is.na(forward) |is.na(reverse))

if (nrow(check_sample_missing_dir) > 0) {
  stop(paste0("Unpaired reads found for the following samples:\n\t", paste0(check_sample_missing_dir$relative_filepath, collapse = "\n\t")))
}
```

# align with lookup table
```{r}
match_fun = function(a, b) {
  pattern = paste0("(?<![0-9])", stringr::str_escape(b), "(?![0-9])")
  stringr::str_detect(a, pattern)
}

source_dir_matched = fuzzy_full_join(
  source_dir_samples, lookup_samples,
  by = c("sequencing_base" = "sequencing_id"),
  match_fun = match_fun)

# checks
## many to one
check_m2o = source_dir_matched %>%
  count(sequencing_base) %>%
  filter(n > 1,
         !is.na(sequencing_base))  # rows in A matched by more than one b

## one to many
check_o2m = source_dir_matched %>%
  count(sequencing_id) %>%
  filter(n > 1,
         !is.na(sequencing_id))  # rows in B matched by more than one a

## no sequence
check_no_seq = source_dir_matched %>%
  filter(is.na(sequencing_base))

## no extraction_id
check_no_ext = source_dir_matched %>%
  filter(is.na(sequencing_id))

check_misaligned = source_dir_matched %>%
  filter(is.na(sequencing_base) | is.na(sequencing_id)) %>%
  arrange(sequencing_base,
         sequencing_id) %>%
  select(sequencing_base,
         sequencing_id,
         relative_parent_dir,
         sequencing_date,
         everything())

```

# Filter to system of interest
```{r}
# files of interest
source_dir_filtered = source_dir_matched %>%
  filter(!is.na(sequencing_base),
         !is.na(sequencing_id),
         study_system %in% soi)

source_dir_filtered_files = source_dir_filtered %>%
  pivot_longer(cols = c("forward", "reverse"),
               names_to = "read_dir",
               values_to = "absolute_filepath")

## for testing
# source_dir_filtered_files = head(source_dir_filtered_files, n = 10)

```

# Copy files to copy directory
```{r}

# create destination dir if it does not exist
if (!dir.exists(destination_dir)) {
  dir.create(destination_dir, recursive = TRUE)
}

# delete all existing files in directory first
if (length(list.files(destination_dir)) > 0) {
  drop_input = readline(prompt = "Files exist in the copy directory. Do you want to delete these files before copying? [y/n]"); 
  
  if (tolower(drop_input == "y")) {
    cat("Deleting existing files in destination directory.\n")
    unlink(here(destination_dir, "*"))
  } else if (tolower(drop_input == "n")) {
    cat("Leaving existing files in destination directory. Proceeding to copy new files.\n")
  } else {
    stop("Input not recognized. Copy aborted.")
  }
}

cat(paste0("Copying ", as.character(nrow(source_dir_filtered_files)), " sequence files to destination directory... "))

file.copy(source_dir_filtered_files$absolute_filepath, destination_dir)

cat("done.")
```