---
title: "copy_seqs_by_system"
author: "Cob Staines"
output: html_document
---

# Copy Sequence Files by System (remote)

This is a utility script which parses through nested folders for .fastq sequencing files, identifies the sequence_id, and references a lookup table to determine which system they belong to, and copies the desired system(s) to a specified folder. Contrast with `copy_seqs_by_system_remote`, which sources sequence files from the AWS S3 eDNA server.

# Configuration

Fill out the parameters below to point to the corresponding local directories and files.
```{r}
# load required packages
librarian::shelf(tidyverse, fuzzyjoin, here, janitor, rio, readr)

# Point to RIBBiTR eDNA Processing 22-25 lookup table for extraction_ids & study systems
sample_lookup_file = here(Sys.getenv("data_dir"), "edna_results", "RIBBiTR eDNA Processing 22-25_2026-08-13.xlsx")

# directory to copy files to
destination_dir = here("16S_sequence_processing", "runs", "brazil_copy_test", "sequences")

# system(s) of interest: any of: Brazil, Panama, Pennsylvania, Sierra Nevada
soi = c("Brazil")

```

# import lookup table
```{r}
lookup_samples = import(sample_lookup_file, which = "SEQUENCING", na.strings = c("NA", ""), guess_max = 2000) %>%
  clean_names() %>%
  filter(!is.na(sequencing_id),
         sequencing_date != "Not sent")
```

# Parse search directory
```{r}
# defin function
resolve_priority_dupes = function(df, priority_dirs,
                                    id_col = "sequencing_base",
                                    dir_col = "relative_parent_dir") {

  # for a single directory value, find the rank of the first (highest-priority)
  # priority_dirs entry that is an ancestor of (or equal to) it
  match_rank = function(dir_val) {
    is_match = (dir_val == priority_dirs) |
      startsWith(dir_val, paste0(priority_dirs, "/"))
    if (any(is_match)) min(which(is_match)) else NA_integer_
  }

  df %>%
    mutate(.priority_rank = vapply(.data[[dir_col]], match_rank, integer(1))) %>%
    group_by(.data[[id_col]]) %>%
    mutate(
      .has_priority_match = any(!is.na(.priority_rank)),
      .min_rank = suppressWarnings(min(.priority_rank, na.rm = TRUE))
    ) %>%
    filter(!.has_priority_match | .priority_rank == .min_rank) %>%
    ungroup() %>%
    select(-.priority_rank, -.has_priority_match, -.min_rank)
}


source_dir_files_raw = tibble(relative_filepath = c(list.files(source_dir, recursive = TRUE)))

# search sequencing files exhaustively, excluding those in excluded dirs, and extract identifiers from file names
source_dir_files = source_dir_files_raw %>%
  filter(str_detect(relative_filepath, "\\.fastq(\\.gz)?$")) %>%
  mutate(absolute_filepath = here(source_dir, relative_filepath),
         relative_parent_dir = dirname(relative_filepath),
         filename = basename(relative_filepath),
         filebase = gsub(".fastq", "", gsub(".gz", "", filename)),
         batch_id = str_match(filebase, "^([A-Za-z0-9]+-)")[,2],
         processing_id = str_match(filebase, "(_S[0-9]+_L[0-9]+_R[12]_[0-9]+)$")[,2],
         sequencing_base = gsub("-", "_", str_remove(filebase, paste0(processing_id, "$"))),
         read_dir = case_when(
           grepl("_R1_", processing_id) ~ "forward",
           grepl("_R2_", processing_id) ~ "reverse",
           .default = NA))

# %>%
#   filter(!purrr::map_lgl(absolute_filepath, function(f) {
#     any(str_starts(f, fixed(excluded_dirs)))
#   }))


# pivot wider to have 1 row per sample, with columns for forward and reverse reads
source_dir_samples = source_dir_files %>%
  pivot_wider(id_cols = c("relative_parent_dir", "sequencing_base"),
              names_from = read_dir,
              values_from = absolute_filepath) %>%
  resolve_priority_dupes(priority_dirs = priority_dirs)

# check for sequencing_base duplicates
check_sample_dupes = get_dupes(source_dir_samples, sequencing_base)

if (nrow(check_sample_dupes) > 0) {
  stop(paste0("Duplicate sequence files found. See variable `check_sample_dupes` to identify the prefered `relative_parent_dir` and add it to `priority_dir` (in configuration chunk). Offending samples:\n\t", paste0(unique(check_sample_dupes$sequencing_base), collapse = "\n\t")))
}

# check for any without a forward or reverse read
check_sample_missing_direction = source_dir_samples %>%
  filter(is.na(forward) |is.na(reverse))

if (nrow(check_sample_missing_direction) > 0) {
  stop(paste0("Samples found with unpaired sequencing reads (only forward or only reverse), see variable `check_sample_missing_direction` to investigate. Offening samples:\n\t", paste0(check_sample_missing_dir$relative_filepath, collapse = "\n\t")))
}
```

# align with lookup table
```{r}
match_fun = function(a, b) {
  pattern = paste0("(?<![0-9A-Za-z])", stringr::str_escape(b), "$")
  stringr::str_detect(a, pattern)
}

source_dir_matched = fuzzy_full_join(
  source_dir_samples, lookup_samples,
  by = c("sequencing_base" = "sequencing_id"),
  match_fun = match_fun)

# checks
## one sequence file to many samples (generally not good)
check_m2o = source_dir_matched %>%
  count(sequencing_base) %>%
  filter(n > 1,
         !is.na(sequencing_base))

if (nrow(check_m2o) > 0) {
  stop(paste0("Multiple samples associated with a given sequencing file. See variable `check_m2o` and resolve before proceeding. Offending sequences:\n\t", paste0(check_m20$sequencing, collapse = "\n\t")))
}

## one sample to many sequence files (generally ok)
check_o2m = source_dir_matched %>%
  count(sequencing_id) %>%
  filter(n > 1,
         !is.na(sequencing_id))

## no sequence
check_no_seq = source_dir_matched %>%
  filter(is.na(sequencing_base))

## no extraction_id
check_no_ext = source_dir_matched %>%
  filter(is.na(sequencing_id))

check_misaligned = source_dir_matched %>%
  filter(is.na(sequencing_base) | is.na(sequencing_id)) %>%
  arrange(sequencing_base,
         sequencing_id) %>%
  select(sequencing_base,
         sequencing_id,
         relative_parent_dir,
         sequencing_date,
         everything())

if (nrow(check_no_seq) > 0) {
  cat(paste0(nrow(check_no_seq), " samples with no associated sequencing files. See variable `check_misaligned` for details.\n"))
}

if (nrow(check_no_ext) > 0) {
  cat(paste0(nrow(check_no_ext), " sequencing file pairs with no associated sample metadata. See variable `check_misaligned` for details.\n"))
}

```

# Filter to system of interest
```{r}
# files of interest
source_dir_aligned = source_dir_matched %>%
  filter(!is.na(sequencing_base),
         !is.na(sequencing_id))

source_dir_filtered = source_dir_aligned %>%
  filter(study_system %in% soi)

source_dir_filtered_files = source_dir_filtered %>%
  pivot_longer(cols = c("forward", "reverse"),
               names_to = "read_dir",
               values_to = "absolute_filepath")

total_filesize_gb = sum(file.size(source_dir_filtered_files$absolute_filepath), na.rm = TRUE) / 1073741824

cat(paste0("---Pre-Copying Summary---\n\t",
           "File count: ", nrow(source_dir_filtered_files), "\n\t",
           "Sample count: ", nrow(source_dir_filtered), "\n\t",
           "Study system(s): ", paste0(soi, collapse = ", "), "\n\t",
           "Total file size: ", total_filesize_gb, " Gb\n\t",
           "From directory: ", source_dir, "\n\t",
           "To directory: ", destination_dir, "\n\n"))

copy_input = readline(prompt = "Proceed with copy? [y/n]: ");

if (tolower(copy_input == "n")) {
  stop("User Aborted.\n")
} else if (tolower(copy_input != "y")) {
  stop("Unexpected input. Copy aborted.\n")
}
```

# Copy files to copy directory
```{r}

# create destination dir if it does not exist
if (!dir.exists(destination_dir)) {
  dir.create(destination_dir, recursive = TRUE)
}

# delete all existing files in directory first
if (length(list.files(destination_dir)) > 0) {
  drop_input = readline(prompt = "Files exist in the copy directory. Do you want to delete these files before copying? [y/n]"); 
  
  if (tolower(drop_input == "y")) {
    cat("Deleting existing files in destination directory.\n")
    unlink(here(destination_dir, "*"))
  } else if (tolower(drop_input == "n")) {
    cat("Leaving existing files in destination directory. Proceeding to copy new files.\n")
  } else {
    stop("Input not recognized. Copy aborted.")
  }
}

cat(paste0("Copying ", as.character(nrow(source_dir_filtered_files)), " sequence files to destination directory... "))

file.copy(source_dir_filtered_files$absolute_filepath, destination_dir)

cat("done.")
```

# copied files check
```{r}
dest_dir_files_raw = tibble(absolute_filepath = c(list.files(destination_dir, full.names = FALSE, recursive = TRUE))) %>%
  mutate(filename = basename(absolute_filepath))

missing_files = source_dir_filtered_files %>%
  mutate(filename = basename(absolute_filepath)) %>%
  anti_join(dest_dir_files_raw, by = "filename")

write_csv(missing_files, here(destination_dir, "missing_copied_files.csv"))

# attempt to recopy
source_dir_filtered_files = missing_files

```